### 2023-11-16 how_to_select_causal_models.tex
intial commit

%\documentclass[MYBIB,STIX1COL]{WileyNJD-v2}
\documentclass[10pt]{article}

% declared functions and additional packages
\usepackage{caption}
\usepackage{subcaption}
\usepackage{threeparttable}
\usepackage{times}
\usepackage{hyperref}
\def\mycitecolor{green!50!black}
\def\mylinkcolor{red!60!black}
\hypersetup{ colorlinks=true, linkcolor=\mylinkcolor, citecolor=\mycitecolor }
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb} %mathbb packages
\usepackage{siunitx}
\usepackage{algorithm, algorithmicx, algpseudocode}
\usepackage{multirow}
\usepackage[sorting=nyt,style=numeric, natbib=true,doi=false,isbn=false,url=false,eprint=false, maxnames=999, maxcitenames=2]{biblatex} 
\DeclareNameAlias{default}{last-first}
\usepackage{placeins}
% associated arxiv class
\usepackage{arxiv}
\usepackage{multicol}
\setlength{\columnsep}{1cm}
\usepackage{authblk}% for author affiliations
\usepackage[dvipsnames]{xcolor}

\usepackage{comment}
%\usepackage{snaptodo}

\setlength{\marginparwidth}{1.5cm}
\setlength\textwidth{177.8mm}

% \snaptodoset{block rise=1em}
% \snaptodoset{margin block/.style={font=\tiny}} \newcommand{\gv}[1]{%
%     \snaptodo[margin block/.append style=green!50!black]{%
% 	\sloppy\textbf{Gael}: #1}}

% \newcommand{\md}[1]{\snaptodo[margin block/.append style=blue]{\textbf{Matt}:
% #1}} \newcommand{\idea}[1]{\snaptodo[margin block/.append
% style=red]{\textbf{Idea}: #1}}

% maths commands
\newcommand{\indep}{\perp \!\!\! \perp}
\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
\usepackage{mathtools}
\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}

\newtheorem{definition}{Definition}
\newtheorem{condition}{Condition}
\newtheorem{property}{Property}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}
\newtheorem{proposition*}{Proposition}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{assumption}{Assumption}
\newtheorem{proof}{Proof}

\title{How to select predictive models for causal inference?}
\addbibresource{biblio.bib}

\author[1,2,*]{Matthieu Doutreligne}
\author[1]{Gaël Varoquaux}

\affil[1]{Inria, Soda, Saclay, France}
\affil[2]{Mission Data, Haute Autorité de Santé, Saint-Denis, France}
\affil[*]{Corresponding author: matthieu.doutreligne@inria.fr}

%\raggedbottom

\begin{document}

%\corres{*\email{matthieu.doutreligne@inria.fr}}

\maketitle

\begin{abstract}
    As predictive models --\emph{eg} from machine learning-- give likely
    outcomes, they may be used to reason on the effect of an intervention,
    a causal-inference task. The increasing complexity of
    health data has opened the door to a plethora of models, but also the
    Pandora box of model selection: which of these models yield the most
    valid causal estimates?
    Here we highlight that classic machine-learning model selection
    does not select the best outcome models for causal inference.
    Indeed, causal model selection should control both outcome errors
    for each individual, treated or not
    treated, whereas only one outcome is observed.
    Theoretically, simple risks used in machine learning do not control
    causal effects when treated and non-treated population differ too much.
    More elaborate risks build proxies of the causal error using
    ``nuisance'' re-weighting to compute it on the observed data. But
    does computing these nuisance adds noise to model selection? Drawing
    from an extensive empirical study, we outline a good causal
    model-selection procedure: using the so-called
    $R\text{-risk}$; using flexible estimators to
    compute the nuisance models on the train set; and splitting out 10\% of
    the data to compute risks.
\end{abstract}
\keywords{Model Selection, Treatment Effect, G-formula, Observational Study, Machine Learning, }
\clearpage

\begin{multicols}{2}

    \section{Background}\label{sec:intro}

    \subsection{Extending prediction to prescription requires causal model selection}

    % Framing:
    % - Increasing rich observational, ``real-life'' data.
    %   - routinely collected health care
    %   - convenient to build prognostic model (eg using claims or
    %   administrative data
    %   \cite{desai2020comparison} ) \cite{yurkovich2015systematic},
    %   electronic health records \cite{fontana2019can}, encompassing notes and questionnaires
    %   \cite{horng2017creating,simon2018predicting}
    % - Motivates building increasingly complex predictive model, such as
    %   prognostic models
    % - Plethora of predictive models
    % - From prediction to prescription There is often a desire to reason on intervention => causal inference is necessary

    Increasingly rich data drives new predictive models. In health, new risks
    or prognostic models leverage routinely-collected data, sometimes with
    machine learning \cite{mooney2018bigdata}: predicting morbidity-related
    outcomes from administrative data \cite{yurkovich2015systematic}, heart
    failure from claims \cite{desai2020comparison}, sepsis from clinical
    records \cite{horng2017creating}, suicide attempts from patient records
    and questionnaires \cite{simon2018predicting}... Data may be difficult to
    control and model, but claims of accurate prediction can be established
    on left-out data
    \cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
    Given a model predicting of an outcome of interest, it is tempting to use
    it to guide decisions: will an individual benefit or not from an
    intervention such as surgery \cite{fontana2019can}? This is a
    causal-inference task and principled approaches can build
    on contrasting the
    prediction of the outcome with and without the treatment
    \cite{snowden_implementation_2011,blakely2020reflection}.

    % - Causal inference is difficult
    % - Beyond the dream of assessing effectiveness and safety in real-word
    %   practice \cite{black1996we} include off-label drug usage 
    %   \cite{radley2006off} (which may lead to drug repurposing)
    % - Plug in that outcome modeling is complementary to RCTs: observational
    %   data gives weaker evidence than trials, but on both data outcome
    %   modeling opens the door to individual.
    % - Epidemiology has focused on propensity-score methods
    % - Principle causal inference with outcome models is possible
    % - It brings the promises of \emph{individualized treatment effects} --a
    %   goal related to capturing heterogeneity and thus CATE--
    % - It also performs well on causal-inference competition, although they
    %   focused on the ATE

    % The plethora of approaches approaches give different causal-inference
    % estimates => need for guideline on causal model selection.

    Outcome modeling is an integral part of the causal modeling toolkit,
    under the names of
    G-computation, G-formula \cite{robins_role_1986}, Q-model
    \cite{snowden_implementation_2011}, or conditional mean
    regression \cite{wendling_comparing_2018}. On observational data, causal
    inference of treatment effects is brittle to un-accounted for
    confounding, but it can assess
    effectiveness and safety in real-word practice
    \cite{black1996we,hernan_methods_2021} or off-label drug usage
    \cite{radley2006off} for potential repurposing
    \cite{hurle2013computational,dudley2011exploiting}. Epidemiology has
    historically focused on methods that model treatment assignment
    \cite{austin_moving_2015,grose_use_2020}, based on the propensity score
    \cite{rosenbaum_central_1983} but recent empirical results
    \cite{wendling_comparing_2018,dorie_automated_2019} show benefits of
    outcome modeling to estimate average treatment effect (ATE). A major
    benefit of using outcome modeling for causal inference is that these
    methods naturally go beyond average effects,
    estimating individualized or conditional average treatment effects
    (CATE), important for precision medicine.
    %
    For this purpose, such methods are also invaluable on randomized trials
    \cite{su2018random,lamont2018identification,hoogland2021tutorial}.

    %\idea{There is a variety of models for g-estimation}
    Recent developments have seen the multiplication of predictive modeling
    methods. Leaving aside the, overwhelming, machine-learning literature,
    even methods specifically designed for causal
    inference are numerous: Bayesian Additive Regression Trees
    \cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
    \cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
    \cite{powers_methods_2018}, causal multivariate adaptive regression
    splines (MARS) \cite{powers_methods_2018}, random forests
    \cite{wager_estimation_2018, athey_generalized_2019},
    Meta-learners \cite{kunzel_metalearners_2019}, R-learners
    \cite{nie_quasioracle_2017}, Doubly robust estimation
    \cite{chernozhukov_double_2018}...
    %\idea{The variety of estimators calls for model selection procedures}
    The wide variety of methods leaves the applied researcher with the difficult choice
    of selecting between different estimators based on the data at hand.
    %
    Indeed, estimates may vary markedly when using different
    models. For instance, Figure \ref{fig:acic_2016_ate_heterogeneity} shows
    large variations obtained across four different outcome estimators on
    2016 semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models
    such as random forests are doing well in most settings except
    for setups where treated and untreated populations differ markedly in
    which case a simple linear model (ridge) is to be preferred.
    However a different choice of hyper-parameters
    (max depth= 2) for random forests yield the poorest performances.
    A simple rule of thumb such as preferring more flexible models does not work in
    general; model selection is necessary.

    %\idea{It is crucial to have tools for model selection.}

    \begin{figure}[b!]
        \begin{minipage}{.4\linewidth}
            \caption{\textbf{Different outcome models lead to different
                    estimation error on the Average Treatment Effects},
                with six different outcome
                models on 77 classic simulations where the true causal effect is
                known \cite{dorie_automated_2019}. The models are random forests, ridge regression with and without
                interaction with the treatment,
                (hyper-parameters detailed in Appendix
                \ref{apd:toy_example:acic_2016_ate_variability}). The different configurations are
                plotted as a function of increasing difference between treated and
                untreated population --detailed in
                \autoref{subsec:measuring_overlap}.  \\[1ex]
                There is no systematic best performer: choosing
                the best model among a family of candidate estimators is
                important.
            }\label{fig:acic_2016_ate_heterogeneity}%
        \end{minipage}\hfill%
        \begin{minipage}{.6\linewidth}
            \centering
            \includegraphics[width=0.95\linewidth]{images/2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
        \end{minipage}%
    \end{figure}

    Standard practices to select models in predictive settings rely on
    cross-validation on the error on the outcome
    \cite{poldrack2020establishment,varoquaux2022evaluating}. However, as we
    will see, these model-selection practices may not pick the best models
    for causal inference, as they can be misled by inhomogeneities due to
    treatment allocation.
    %
    Given complex, potentially noisy, data, which model is to be most trusted to
    yield valid causal estimates? Because there is no single learner that performs
    best on all data sets, there is a pressing need for clear guidelines to select
    outcome models for causal inference.

    \paragraph{Objectives and structure of the paper}

    In this paper, we study \textit{model selection procedures} with a focus
    on practical settings: \textit{finite samples} settings and without
    \textit{well-specification} assumption. One question is whether
    model-selection procedures, that rely on data split, can estimate
    reliably enough the complex risks, theoretically motivated for causal
    inference. Indeed, these risks force of departure from standard model-selection
    procedures as they come with more quantities to estimate, which may
    bring additional variance, leading to worse model selection.

    We first give a simple illustration of the problem of causal model
    selection and quickly review the important prior art. Then, in Section
    \ref{sec:framework}, we set causal model selection in the
    \emph{potential outcome} framework and detail the causal risks and
    model-selection procedure. Section \ref{sec:theory} gives our theoretical
    result. In section \ref{sec:empirical_study}, we run a thorough empirical
    study, with many different settings covered. Finally, we comment our
    findings in Section \ref{sec:discussion}.
    %
    Results outline how to best select outcome models for causal
    inference with an adapted
    cross-validation that estimate the so-called $R\text{-risk}$.
    The $R\text{-risk}$
    modulates observed prediction error to compensate for systematic
    differences between treated and non-treated individuals. It
    relies on the two \emph{nuisance} models,
    themselves estimated from data and thus imperfect; yet these
    imperfections do not undermine the benefit of the $R\text{-risk}$.

    % XXX: merge the two paragraphs, above and below

    \subsection{Illustration: the best predictor may not estimate best causal
        effects}%

    \begin{figure}[b!]
        \begin{minipage}{.4\linewidth}
            \caption[The best predictor may not estimate best causal
                effects]{\textbf{Illustration}: a) a random-forest estimator
                with high performance for standard prediction (high $R^2$) but that
                yields poor causal estimates (large error between true effect $\tau$ and
                estimated $\hat{\tau}$), b) a linear estimator with smaller
                prediction performance leading to better causal estimation. \\[1ex]
                Selecting the estimator with the smallest error to the individual
                treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$
                --the $\tau\text{-risk}$, def.\,\ref{def:tau_risk} -- would lead to
                the best causal estimates; however computing this error is not
                feasible: computing it requires access to unknown quantities:
                $\tau(x)$. \\[1ex]
                While the random forest fits the data better than the linear model, it
                gives worse causal inference because its error is very inhomogeneous between
                the treated and untreated. The $R^2$ score does not capture this
                inhomogeneity.
            }\label{fig:toy_example} \end{minipage}\hfill%
        \begin{minipage}{.577\linewidth}
            %\centering
            {\sffamily\small a) Random forest, good average prediction but bad
                causal inference}

            \hfill%
            \includegraphics[width=1\linewidth]{images/toy_random_forest_high_R2_high_tau_risk.pdf}%

            {\sffamily\small b) Linear model, worse average prediction but better causal inference}

            \hfill%
            \includegraphics[width=1\linewidth]{images/toy_tlinear_model_small_R2_small_tau_risk.pdf}%
        \end{minipage}%
    \end{figure}

    Using a predictor to reason on causal effects relies on contrasting the
    prediction of the outcome for a given individual with and without the
    treatment --as detailed in \autoref{sec:neyman_rubin}.
    %
    Given various predictors of the outcome, we are interested in
    selecting those that estimate best the treatment effect.
    %
    Standard predictive modeling or machine-learning practice selects the
    best predictor, \emph{ie} the one that minimizes the expected error.
    However, the best predictor may not be the best model to reason about
    causal effects of an intervention, as we illustrate below.

    Figure \ref{fig:toy_example} gives a toy example: the
    outcome $Y \in [0, 1]$, the
    probability of death, a binary treatment $A \in \{0, 1\}$ and a covariate
    $X \in \mathbb R$ which summarizes the patient health status (eg. the
    Charlson co-morbidity index \cite{charlson_new_1987}). We simulate a
    situation for which the treatment is beneficial (decreases
    mortality) for patients with high Charlson scores (bad health
    status). On the contrary, the treatment has little effect for patients in good
    condition (small Charlson scores).

    Figure \ref{fig:toy_example}a shows a random forest predictor with a
    counter-intuitive behavior: it predicts well on average the outcome (as
    measured by a regression $R^2$ score) but perform poorly to estimate
    causal quantities: the average treatment effect $\tau$ (as visible via
    the error $|\tau - \hat{\tau}|$) or the individual treatment
    effect (the error $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$).
    %
    On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
    smaller $R^2$ score but better causal inference.%

    Intuitively, the problem is that causal estimation requires controlling
    an error on both treated and non-treated outcome for the same individual:
    the observed outcome, and the non-observed \emph{counterfactual} one.
    The linear model is misspecified --the outcome functions are not
    linear--, leading to poor $R^2$; but it interpolates better to regions
    where there are few untreated individuals --high Charlson score-- and
    thus gives better causal estimates. Conversely, the random forest puts
    weaker assumptions on the data, thus has higher $R^2$ score but is biased
    by the treated population in the region with poor overlap region, leading
    to bad causal estimates.

    This toy example illustrates that the classic minimum Mean Square Error
    criterion is not suited to choosing a model among a family of candidate
    estimators for causal inference.

    \subsection{Prior work: model selection for outcome modeling (g-computation)}%

    A natural way to select a predictive model for causal inference would be
    an error measure between a causal quantity such as the conditional
    average treatment effects (CATE) and models' estimate. But such error is
    not a ``feasible'' risk: it cannot be computed solely from observed data
    and requires oracle knowledge.

    % XXX: I think that I should make the following paragraph shorter

    \paragraph{Simulation studies of causal model selection}

    Using eight simulations setups from \cite{powers_methods_2018}, where
    the oracle CATE is known, \citet{schuler_comparison_2018} compare four
    causal risks, concluding that for CATE estimation the best
    model-selection risk is the so-called $R\text{-risk}$
    \cite{nie_quasioracle_2017} --def.\,\ref{def:r_risk}, below. Their
    empirical results are clear for randomized treatment allocation but less
    convincing for observational settings where both simple Mean Squared
    Error --MSE, $\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
    reweighted MSE --$\mu\text{-risk}_{IPW}$ def.\,\ref{def:mu_ipw_risk}--
    appear to perform better than $R\text{-risk}$ on half of the simulations.
    Another work \cite{alaa_validating_2019} studied empirically both MSE and
    reweighted MSE risks on the semi-synthetic ACIC 2016 datasets
    \cite{dorie_automated_2019}, but did not include the $R\text{-risk}$ and
    looked only at the agreement of the best selected model with the true
    CATE risk --$\tau\text{-risk}(f)$ def.\, \ref{def:tau_risk}--, not on the
    full ranking of methods compared to the true CATE. We complete these
    prior empirical work by studying a wider variety of data generative
    processes and varying the influence of overlap, an important parameter of
    the data generation process which makes a given causal metric appropriate
    \cite{damour_overlap_2020}. We also study how to best adapt
    cross-validation procedures to causal metrics which themselves come with
    models to estimate.

    \paragraph{Theoretical studies of causal model selection}

    Several theoretical works have proposed causal model selection procedures
    that are \emph{consistent}: select the best model in a family given
    asymptotically large data. These work typically rely on introducing a
    CATE estimator in the testing procedure. For instance matching
    \citep{rolling_model_2014}, an IPW estimate
    \citep{gutierrez_causal_2016}, a doubly robust estimator
    \citep{saito_counterfactual_2020}, or debiasing the error with influence
    functions \cite{alaa_validating_2019}. However, for theoretical
    guarantees to hold, the test-set correction needs to converge to the
    oracle: it needs to be flexible enough --or well-posed-- and asymptotic
    data. From a practical perspective, knowing that such requirements are
    met implies having a good CATE estimate, which amounts to having solved
    the original problem of causal model selection. We study how causal
    model-selection procedures behave outside of these settings.

    \paragraph{Statistical guarantees on causal estimation procedures}

    Much work in causal inference has focused on building procedures that
    guarantee asymptotically consistent estimators. Targeted Machine Learning
    Estimation (TMLE) \cite{laan_targeted_2011,schuler_targeted_2017} and
    Double Machine Learning \cite{chernozhukov_double_2018} both provide
    estimators for Average Treatment Effect combining flexible treatment and
    outcome models. Here also, theories require asymptotic regimes and
    models to be \textit{well-specified}.

    By contrast, \citet{johansson_generalization_2021} studies causal
    estimation without assuming that estimators are well specified. They derive an
    upper bound on the oracle error to the CATE ($\tau\text{-risk}$) that involves
    the error on the outcome and the similarity of the distributions between the
    features of treated and control patients. However, they focus on using this
    upper bound for estimation, and do not give insights on model selection. In
    addition, for hyperparameter selection, they rely on a plugin estimate of the $\tau\text{-risk}$ built with
    counterfactual nearest neighbors, which has been shown ineffective
    \cite{schuler_comparison_2018}.

    \section{Formal setting of causal inference and model selection}\label{sec:framework}

    \subsection{The Neyman-Rubin Potential Outcomes framework}%
    \label{sec:neyman_rubin}%

    \paragraph{Settings}

    The Neyman-Rubin Potential Outcomes framework
    \cite{naimi2023defining,imbens_causal_2015} enables statistical
    reasoning on causal treatment effects: Given an outcome $Y \in \mathbb R$ (eg.
    mortality risk or hospitalization length), function of a binary treatment $A \in
        \mathcal{A} = \{0, 1\}$ (eg.~a medical act, a drug administration), and baseline
    covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
    distribution,
    $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However, we want to model the existence of potential observations
    (unobserved ie. counterfactual) that correspond to a different treatment. Thus we want
    quantities on the counterfactual distribution
    $O^{*} = (Y(1), Y(0), X, A) \sim \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.

    Popular quantities of interest (estimands) are:
    at the population level, the
    Average Treatment Effect
    \begin{equation*}
        \text{(ATE)}\qquad
        \tau \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal D^*}[Y(1) - Y(0)];
    \end{equation*}
    to model heterogeneity, the Conditional Average Treatment Effect
    \begin{equation*}
        \text{(CATE)}\qquad
        \tau (x) \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal{D}^\star}[Y(1) - Y(0) | X=x].
    \end{equation*}

    \paragraph{Causal assumptions}

    Some assumptions are necessary to assure identifiability of the causal estimands
    in observational settings \cite{rubin_causal_2005}. We assume the usual
    strong ignorability assumptions, composed of \emph{1)}
    \emph{unconfoundedness} $\{Y(0),
        Y(1) \} \indep A | X$, \emph{~2)} \emph{strong overlap} ie. every patient has a
    strictly positive probability to receive each treatment, \emph{3)}
    \emph{consistency}, and \emph{4)} \emph{generalization} (detailed in Appendix
    \ref{apd:causal_assumptions}). In this work, we investigate the
    role of the overlap \cite{damour_overlap_2020}, which is testable with
    data.

    \paragraph{Estimating treatment effects with outcome models}\label{subsec:estimators}

    Should we know the two expected outcomes for a given $X$,
    we could compute the
    difference between them, which gives the causal effect of the treatment.
    %
    These two expected outcomes can be computed from the observed data:
    the consistency \ref{assumption:consistency} and unconfoundedness
    \ref{assumption:ignorability} assumptions imply the equality of two different
    expectations:
    \begin{equation}\label{eq:mu_identification}
        \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
    \end{equation}
    On the left, the expectation is taken on the counterfactual unobserved
    distribution. On the right, the expectation is taken on the factual observed
    distribution conditionally on the treatment. This equality is referred as the
    g-formula identification \cite{robins_new_1986}. For the rest of the
    paper, the expectations will always be taken on the factual observed
    distribution $\mathcal{D}$, and we will omit
    to explicitly specify the distribution. This identification leads to outcome based estimators (ie.
    g-computation estimators\cite{snowden_implementation_2011}), targeting the
    ATE $\tau$ with outcome modeling:
    \begin{equation}
        \tau = \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x] = \mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
        \label{eq:tau_population}
    \end{equation}
    This equation has two central quantities: the conditional expectancy
    function of the outcome associated to specific covariates and
    treatment or not is, often called the \emph{response function}:
    \begin{equation}
        \text{(Response function)}
        \qquad
        \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a].
        \label{def:mu_a}
    \end{equation}

    Given a sample of data and the oracle response functions $\mu_0, \mu_1$, the
    finite sum version of \autoref{eq:tau_population} leads to an
    estimator of the ATE written:
    \begin{equation}
        \hat \tau = \frac{1}{n} \biggl(\sum_{i=1}^n \mu_{1}(x_i) - \mu_{0}(x_i) \biggr)
        \label{eq:ate_estimate}
    \end{equation}
    This estimator is an oracle \textbf{finite sum estimator} by opposition to the
    population expression of $\tau$, $\mathbb{E}[\mu_{1}(x_i) - \mu_{0}(x_i)]
    $,
    which involves an expectation taken on the full
    distribution $\mathcal D$, which is observable but requires infinite data. For
    each estimator $\ell$ taking an expectation over $\mathcal D$, we use the symbol
    $\hat \ell$ to note its finite sum version.

    Similarly to the ATE, for the CATE, at the individual level:
    \begin{equation}
        \tau(x) = \mu_{1}(x) - \mu_{0}(x)
        \label{eq:cate_estimate}
    \end{equation}

    \paragraph{Robinson decomposition}
    Another decomposition of the outcome model plays in important role,
    the R-decomposition
    \cite{robinson_rootnconsistent_1988}:
    introducing two quantities, the conditional mean outcome
    and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
    \begin{align}
        \text{(Conditional mean outcome)} \qquad  m(x) \myeq \; & \mathbb E_{Y \sim \mathcal{D}} [Y|X=x].
        \label{def:m}
        \\
        \text{(Propensity score)}\qquad
        e(x) \myeq \;                                           & \mathbb P[A=1|X=x],
        \label{def:propensity_score}
    \end{align}
    the outcome can be written
    \begin{equation}\label{eq:r_decomposition}
        \text{(R-decomposition)}\qquad y(a) = m(x) + \big( a - e(x) \big) \tau(x) + \varepsilon(x; a) \quad\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0
    \end{equation}
    $m$ and $e$ are often called
    \emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown in
    general.

    %As noted by \cite{johansson_generalization_2021}, the machine learning
    %community often referred to the CATE by ITE, the Individual Treatment Effect.
    %From a purely causal point of view, the ITE is uniquely defined for each
    %individual and might not be accessible: $ITE(x_i) = Y_i(1) -  Y_i(0)$. On the
    %contrary, the CATE can always be derived by taking conditional expectancies. It
    %is the expected effect of the treatment in the region of the covariate space
    %around X. %Too cultivate

    \subsection{Model-selection risks, oracle and feasible}\label{sec:problem:model_selection}

    \paragraph{Causal model selection}\label{sec:problem:causal_selection}

    We formalize model selection for causal estimation. Thanks to the g-formula
    identification (\autoref{eq:mu_identification}), a given outcome model $f: \mathcal X
        \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
    domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
    \ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
    %
    Let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
    a family of such estimators. Our goal is to select the best candidate in this
    family for the observed dataset $O$ using a risk of interest
    $\ell$:
    \begin{equation}
        f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
        \label{eq:causal_model_selection}
    \end{equation}

    We now detail possible risks $\ell$, risks useful for causal
    model selection, and how to compute them.

    \paragraph{The $\tau\text{-risk}$: an oracle error risk}\label{paragraph:oracle_metrics}
    %
    As we would like to target the CATE, the following
    evaluation risk is natural:
    \begin{definition}[$\tau\text{-risk}(f)$]\label{def:tau_risk} also called PEHE
        \cite{schulam_reliable_2017, hill_bayesian_2011}:
        \begin{equation*}\label{eq:tau_risk}
            \tau\text{-risk}(f) = \mathbb E_{X\sim p(X)}[(\tau(X) - \hat \tau_f(X))^2]
        \end{equation*}
        its finite-sum version over the observed data:
        \begin{equation*}
            \widehat{\tau\text{-risk}}(f) = \sum_{x \in O} \big(\tau(x) - \hat \tau_f(x)\big)^2
        \end{equation*}
    \end{definition}

    However these risks are not feasible because the oracles $\tau(x)$ are
    not accessible, with the observed data $(Y, X, A) \sim \mathcal D$.

    \paragraph{Feasible error risks}\label{paragraph:feasible_metrics}
    \emph{Feasible} risks are based on the prediction error of the outcome model
    and \emph{observable} quantities.

    \begin{definition}[Factual $\mu\text{-risk}$]\label{def:mu_risk}
        \cite{shalit_estimating_2017} This is the usual Mean Squared Error on
        the target y. It is what is typically meant by ``generalization error'' in
        supervised learning and estimated with cross-validation:
        \begin{equation*}\label{eq:mu_risk}
            \mu\text{-risk}(f)=\mathbb{E}_{(Y, X, A) \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]
        \end{equation*}
    \end{definition}

    The following risks use the nuisances $e$
    --propensity score, def \ref{def:propensity_score}-- and $m$ --conditional mean
    outcome, def \ref{def:m}. We give the definitions as \textit{semi-oracles},
    function of the true unknown nuisances, but later instantiate them with estimated
    nuisances, noted $\big(\check e, \check m \big)$. Semi-oracles risks are
    superscripted with the $^{\star}$ symbol.

    \begin{definition}[$\mu\text{-risk}_{IPW}^{\star}$]\label{def:mu_ipw_risk}
        \cite{vanderlaan_unified_2003} Let the inverse propensity weighting
        function $w(x, a) = \frac{a}{e(x)} + \frac{1 - a}{1 - e(x)}$, we define the
        semi-oracle Inverse Propensity Weighting risk,
        \begin{equation*}\label{eq:mu_ipw_risk}
            \mu\text{-risk}_{IPW}^{\star}(f) = \mathbb{E}_{(Y, X, A) \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big) (Y-f(X ; A))^2 \right]
        \end{equation*}
    \end{definition}

    \smallskip

    \begin{definition}[$\tau\text{-risk}^{\star}_{IPW}$]\label{def:tau_ipw_risk}
        \cite{wager_estimation_2018}
        The CATE $\tau(x)$ can be estimated propensity score, with a
        regression against inverse propensity weighted outcomes \cite{
            athey2016recursive,gutierrez_causal_2016,wager_estimation_2018}. From this objective, we can derive
        the $\tau\text{-risk}_{IPW}$.
        \begin{equation*}
            \tau\text{-risk}^{\star}_{IPW}(f) =\mathbb{E}_{(Y, X, A) \sim
                \mathcal D} \left[ \Big(Y \frac{A - e(X)}{e(X)
                    (1-e(X))}-\tau_f\left(X\right)\Big)^2 \right]  =\mathbb{E}_{(Y, X, A)
                \sim \mathcal D} \left[ \left(Y \left( \frac{A}{e(X)} -
                \frac{1-A}{1-e(X)}\right)-\tau_f\left(X\right)\right)^2 \right]
        \end{equation*}
    \end{definition}

    \begin{definition}[$U\text{-risk}^{\star}$]\label{def:u_risk}
        \cite{kunzel_metalearners_2019,nie_quasioracle_2017} Based on
        the Robinson decomposition --eq. \ref{eq:r_decomposition}
        \cite{robinson_rootnconsistent_1988}, the U-learner
        uses the $A-e(X)$ term
        in the denominator. The derived risk is:
        \begin{equation*}
            U\text{-risk}^{\star}(f) =\mathbb{E}_{(Y, X, A) \sim \mathcal D}
            \left[
                \left( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -
                \tau_f\left(X\right)\right)^{2} \right]
        \end{equation*}
        Note that extreme propensity weights in the
        denominator term might inflate errors in the numerator due to imperfect
        estimation of the mean outcome $m$
        the numerator errors, leading to a highly biased metric.
    \end{definition}

    \begin{definition}[$R\text{-risk}^{\star}$]\label{def:r_risk}
        \cite{nie_quasioracle_2017,schuler_comparison_2018}
        The $R\text{-risk}$ also uses two nuisance $m$ and $e$:
        \begin{equation*}
            R\text{-risk}^{\star}(f) =\mathbb{E}_{(Y, X, A) \sim \mathcal D} \big[
                \big(\left(Y-m\left(X\right)\right) -\left(A-e\left(X\right)\right) \tau_f\left(X\right)\big)^{2} \big]
        \end{equation*}
    \end{definition}

    It is also motivated by the Robinson decomposition --eq. \ref{eq:r_decomposition}
    \cite{robinson_rootnconsistent_1988}. It performs well in various simulations
    where using lasso, boosting or kernel ridge regressions for both the nuisances
    $(\check e, \check m)$ and the target $\hat \tau(x)$
    \cite{nie_quasioracle_2017}.

    These risks are summarized in Table \ref{tab:evaluation_metrics}.

    \begin{table}[!h]
        \makebox[\linewidth]{
            \begin{threeparttable}[b]
                \caption{Review of causal risks}
                \centering
                \label{tab:evaluation_metrics}
                \begin{tabular}{llr}
                    \toprule
                    Risk                                                                                           & Equation
                                                                                                                   & Reference                                                                                                                                           \\
                    \midrule
                    $mse(\tau(X), \tau_f(X))=\tau\text{-risk}$                                                     & $\mathbb E_{X\sim
                                p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
                                                                                                                   & Eq. \ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
                    $mse(Y, f(X)) = \mu\text{-risk}$                                                               & $\mathbb{E}_{(Y, X, A)
                            \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
                                                                                                                   & Def. \ref{def:mu_risk} \cite{schuler_comparison_2018}                                                                                               \\
                    $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
                            \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
                    (Y-f(X ; A))^2 \right]$                                                                        & Def.
                    \ref{def:mu_ipw_risk} \cite{vanderlaan_unified_2003}                                                                                                                                                                                                 \\
                    $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
                    Def. \ref{def:tau_ipw_risk} \cite{wager_estimation_2018}
                    \\
                    $U\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A) \sim \mathcal D}  \big[
                    \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
                    Def. \ref{def:u_risk} \cite{nie_quasioracle_2017}
                    \\
                    $R\text{-risk}^*$ \tnote{1}                                                                    & $\mathbb{E}_{(Y, X, A)
                            \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
                    -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
                    Def. \ref{def:r_risk} \cite{nie_quasioracle_2017}
                    \\
                    \bottomrule
                \end{tabular}
                \begin{tablenotes}
                    \item [1] Called $\tau \text{-risk}_R$ in
                    \citet{schuler_comparison_2018}.
                \end{tablenotes}
            \end{threeparttable}
        }
    \end{table}

    \subsection{Estimation and model selection procedure}\label{problem:estimation_procedure}

    Causal model selection (as in
    \autoref{eq:causal_model_selection}) may involve estimating various quantities
    from the observed data: the outcome model $f$, its induced risk as
    introduce in the previous section, and possibly nuisances required by the
    risk.
    Given a dataset with $N$ samples, we split out a train and a test sets
    $(\mathcal{T}, \mathcal{S})$. We
    fit each candidate estimator $f \in \mathcal{F}$ on $\mathcal{T}$. We also fit
    the nuisance models $(\check e, \check m)$ on the train set
    $\mathcal{T}$, setting hyperparameters by a nested
    cross-validation before fitting the nuisance estimators with these parameters
    on the full train set. Causal quantities are then computed by applying the fitted  candidates
    estimators $f \in \mathcal{F}$ on the test set $\mathcal{S}$. Finally, we
    compute the model-selection metrics for
    each candidate model on the test set. This procedure is described in Algorithm
    \ref{problem:estimation_procedure:algo} and Figure
    \ref{problem:estimation_procedure:figure}.

    As extreme inverse propensity weights induce high variance, clipping can be
    useful for numerical stability
    \cite{swaminathan_counterfactual_2015, ionides_truncated_2008}.

    \begin{algorithm}[!htbp]
        \caption{Evaluation of selection procedures for one
            simulation}\label{problem:estimation_procedure:algo} {%
            Given a train and a test sets $(\mathcal{T}, \mathcal{S}) \sim \mathcal{D}$,
            a family of candidate estimators $\{f \in \mathcal F\}$, a set of causal
            metrics $\ell \in \mathcal L$:
            \begin{enumerate}
                \item Prefit: Learn estimators for unknown nuisance quantities $(\check e,\,\check m)$ on the training set $\mathcal{T}$
                \item Fit: $\forall f \in \mathcal{F}$ learn $\hat f(\cdot, a)$ on
                      $\mathcal T$
                \item Model selection: $\forall{x} \in \mathcal{S}$ predict $\big(\hat f(x, 1), \hat f(x, 0)\big)$ and evaluate each candidate estimator with
                      each causal metric $\mathcal M(\hat f, \mathcal{S})$. For each causal
                      metric $\ell  \in \mathcal{L}$ and each candidate estimator $f \in
                          \mathcal{F}$, store the metric value: $\ell(f, \mathcal S)$ -- possibly
                      function of $\check e$ and $\check m$
                      %\item Metric evaluation: return the oracle evaluation metrics
                      %evaluated on $\mathcal{S}$: $\big(\widehat{\tau\text{-risk}}(\hat
                      %f^*_{\ell}); \widehat{\ell}_{ATE}(\hat f^*_{\ell}) \big)$
            \end{enumerate}

        }
    \end{algorithm}

    \begin{figure}[h!]
        \centering
        \includegraphics[width=0.7\linewidth]{images/estimation_procedure_causal_selection_procedure.png}
        \caption{Estimation procedure for causal model
            selection.}\label{problem:estimation_procedure:figure}
    \end{figure}

    \section{Theory: Links between feasible and oracle risks}\label{sec:theory}

    We now relate two feasible risks, $\mu \text{-risk}_{IPW}$ and the
    $R\text{-risk}$ to the oracle $\tau\text{-risk}$. Both results make
    explicit the role of overlap for the performances of causal risks.

    These bounds depend on a specific form of residual that we now define: for each potential outcome, $a \in  \{0; 1\}$, the variance conditionally on $x$
    is\cite{shalit_estimating_2017}:
    \begin{equation*}\label{eq:residuals}
        \sigma_{y}^{2}(x ; a) \overset{\text{def}}{=}
        \int_{y}\left(y-\mu_{a}(x)\right)^{2} p(y \mid x=x ; A=a) \, d y
    \end{equation*}
    Integrating over the population, we get the Bayes squared error:
    $\sigma^2_{B}(a) = \int_{\mathcal X} \sigma_y^2(x;a) p(x)dx$
    and its propensity weighted version:
    $\tilde{\sigma}^2_{B}(a) = \int_{\mathcal X}\sigma_y^2(x;a)\,  p(x;
        a)\,dx$. In case of a purely deterministic link between the
    covariates, the treatment, and the outcome, these residual terms are null.

    \subsection{Upper bound of $\tau\text{-risk}$ with
        $\mu\text{-risk}_{IPW}$}\label{theory:mu_risk_ipw_bound}

    \begin{proposition}[Upper bound with $\mu \text{-risk}_{IPW}$
        ]\label{theory:prop:mu_risk_ipw_bound}
        \cite{johansson_generalization_2021} Given an outcome model $f$, let a
        weighting function $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$ as the
        Inverse Propensity Weight. Then, under overlap (assumption
        \ref{assumption:overlap}), we have:
        \begin{equation*}
            \tau\text{-risk}(f) \leq \; 2 \, \mu\text{-risk}_{IPW}(w, f)
            - 2 \, \big(\sigma^2_{B}(1) +  \sigma^2_{B}(0)\big)
        \end{equation*}
    \end{proposition}
    This result has been derived in previous work
    \cite{johansson_generalization_2021}. It links $\mu\text{-risk}_{IPW}$ to
    the squared residuals of each population thanks to a reweighted mean-variance
    decomposition. For completeness, we provide the proof in .

    The upper-bound comes from the triangular inequality applied to the residuals of
    both populations. Interestingly, the two quantities are equal when the
    absolute residuals on treated and untreated populations are equal on the
    whole covariate space, \emph{ie} for all
    $x \in \mathcal X, |\mu_1(x) - f(x, 1)| = |\mu_0(x) - f(x, 0)|$.
    The main source of difference between the oracle $\tau \text{-risk}$ and the
    reweighted mean squared error, $\mu\text{-risk}_{IPW}$, comes from heterogeneous
    residuals between populations. These quantities are difficult to characterize as
    they are linked both to the estimator and to the data distribution.
    %
    This bound indicates that minimizing the $\mu\text{-risk}_{IPW}$ helps to
    minimize the $\tau\text{-risk}$, which leads to
    interesting optimization procedures \cite{johansson_generalization_2021}. However, there is no
    guarantee that this bound is tight, which makes it less useful for model
    selection.

    Assuming strict overlap (probability of all individuals being treated or not
    bounded away from 0 and 1 by $\eta$, appendix \ref{apd:causal_assumptions}), the
    above bound simplifies into a looser one involving the usual mean squared error:
    $\tau\text{-risk}(f)\leq \frac{2}{\eta}\, \mu\text{-risk}(f) -  2 \, \big(\sigma^2_{B}(1) +  \sigma^2_{B}(0)\big)$. For weak overlap (propensity scores not bounded far from 0
    or 1), this bound is very loose (as shown in Figure \ref{fig:toy_example})
    and is not appropriate to discriminate between models with close performances.

    %\md{Lower bound: I think that there is no lower bound of the tau-risk with the
    %mu-risk. That means that we can have a mu risk of 0 wheras the taurisk is
    %non-null. Todo for myself: develop an example with overfitted 1NN on the
    %observed data.}

    \subsection{Reformulation of the $R\text{-risk}$ as reweighted
        $\tau\text{-risk}$}\label{theory:r_risk_rewrite}

    We now derive a novel rewriting of the $R\text{-risk}$, making explicit its link
    with the oracle $\tau \text{-risk}$.

    \begin{proposition}[$R \text{-risk}$ as reweighted
            $\tau\text{-risk}$]\label{theory:prop:r_risk_rewrite} Given an outcome model
        $f$, its $R\text{-risk}$ appears as weighted version of its $\tau\text{-risk}$
        (Proof in Appendix \ref{apd:proofs:r_risk_rewrite}):
        \begin{equation}
            R\text{-risk}^*(f) = \int_{x} e(x)\big(1-e(x)\big)\big(\tau(x)-\tau_ {f}(x)\big)^{2} p(x) d x \;+\; \tilde{\sigma}_B^2(1) \;+\; \tilde{\sigma}_B^{2}(0)
        \end{equation}
    \end{proposition}

    The $R \text{-risk}$ targets the oracle at the cost of an overlap re-weighting
    and the addition of the reweighted Bayes residuals, which are independent of
    $f$. In good overlap regions the weights $e(x) \big(1-e(x) \big)$ are close to
    $\frac{1}{4}$, hence the $R \text{-risk}$ is close to the desired gold-standard
    $\tau \text{-risk}$. On the contrary, for units with extreme overlap violation,
    these weights goes down to zero with the propensity score.

    % \begin{remark}[Trivial upper bound] Because $e(x)\leq1$ and $(1-e(x))\leq1$,
    %   we see immediately that \begin{equation} R \text{-risk}^* \leqslant
    %   \tau{\text{-risk}}+\tilde{\sigma_{B}}(1)^{2}+\tilde{\sigma_{B}}^{2}(0)
    %   \end{equation} \end{remark}

    % \begin{remark}[Lower bound] Due to strict overlap assumption, \begin{equation}
    %   R \text{-risk}^* \geqslant \eta (1 - \eta)
    %   \tau{\text{-risk}}-\tilde{\sigma_{B}}(1)^{2}-\tilde{\sigma_{B}}^{2}(0)
    %   \end{equation} \end{remark}

    % \begin{remark}[Cases of loose upper bounds] Giving a strict overlap assumption
    %   $\eta$, we can exhibit for all $C \; st. \; \eta^{-1}>C>1$, simple
    %   simulations where: \begin{equation} R \text{-risk}^* \leqslant C \; \eta \;
    %   \tau \text{-risk}(f) \quad \end{equation} \end{remark}

    \subsection{Interesting special cases}

    \paragraph{Randomization special case}\label{remark:rct} If the treatment is
    randomized as in RCTs, $p(A=1 \mid X=x) = p(A=1)=p_A$, thus
    $\mu\text{-risk}_{IPW}$ takes a simpler form:
    \begin{equation*}
        \mu\text{-risk}_{IPW} = \mathbb{E}_{(Y, X, A) \sim \mathcal D}\left[ \Big( \frac{A}{p_A} + \frac{1-A}{1-p_A} \Big) (Y-f(X ; A))^2 \right]
    \end{equation*}
    However, even if we have randomization, we still can have large differences
    between $\tau\text{-risk}$ and $\mu\text{-risk}_{IPW}$ coming from heterogeneous
    errors between populations as noted in Section \ref{theory:mu_risk_ipw_bound}
    and shown experimentally in simulations \cite{schuler_comparison_2018}.

    Concerning the $R\text{-risk}$, replacing $e(x)$ by its randomized value $p_A$
    in Proposition \ref{theory:prop:r_risk_rewrite} yields the oracle
    $\tau\text{-risk}$ up to multiplicative and additive constants:
    \begin{equation}
        R\text{-risk} = p_A \, (1-p_A) \, \tau\text{-risk} \;+\; (1 - p_A) \,\sigma_B^2(0) \;+\; p_A \sigma_B^2(1)
    \end{equation}
    Therefore, optimizing estimators for CATE with $R\text{-risk}^*$ in the
    randomized setting is optimal if we target the $\tau\text{-risk}$. This explains
    the strong performances of $R\text{-risk}$ in randomized setups
    \cite{schuler_comparison_2018} and is a strong argument in favor of this
    risk for heterogeneity estimation in RCTs.

    \paragraph{Oracle Bayes predictor}\label{remark:bayes_oracle} Consider the case
    where we have access to the oracle Bayes predictor for the outcome ie.~$f(x,
        a)=\mu(x, a)$, then all risks are equivalent up to the residual variance:
    \begin{equation}
        \tau\text{-risk}(\mu) = \mathbb E_{X\sim p(X)}[(\tau(X) - \tau_{\mu}(X))^2] = 0
    \end{equation}
    \begin{equation}
        \mu\text{-risk}(\mu) = \mathbb E_{(Y, X, A) \sim p(Y;X;A)}[\big( Y - \mu_A(X)\big)^2] = \int_{\mathcal X, \mathcal A}
        \,\varepsilon(x,a)^2 p(a \mid x) \,p(x) \,dx\,da  \leq \sigma_B^{2}(0) + \sigma_B^{2}(1)
    \end{equation}
    \hspace*{0.5em}
    \begin{equation}
        \mu\text{-risk}_{IPW}(\mu) = \sigma_B^{2}(0) + \sigma_B^{2}(1)  \quad \text{follows from Lemma \ref{apd:proofs:mu_risk_ipw_link_mu}}
        %\notag
    \end{equation}
    \hspace*{0.5em}
    \begin{equation}
        R\text{-risk}(\mu) = \tilde{\sigma}_B^{2}(0) + \tilde{\sigma}_B^{2}(1)
        \leq \sigma_B^{2}(0) + \sigma_B^{2}(1) \quad \text{follows directly from Proposition \ref{theory:prop:r_risk_rewrite}}%\notag
    \end{equation}
    Thus, differences between causal risks only matter in finite sample regimes.
    Universally consistent learners converge to the Bayes risk in asymptotic
    regimes, making all model selection risks equivalent. However, in practice
    choices must be made in non-asymptotic regimes.

    \section{Empirical Study}\label{sec:empirical_study}

    %We now explore empirically the behavior of these different causal selection
    %metrics.
    %\subsection{Causal metrics under evaluation}

    We evaluate the following causal metrics, oracle and feasible
    versions, presented in Table
    \ref{tab:evaluation_metrics}:
    $\widehat{\mu\text{-risk}}_{IPW}^*$,
    $\widehat{R\text{-risk}}^*$,
    $\widehat{U\text{-risk}}^*$,
    $\widehat{\tau\text{-risk}_{IPW}}^*$,
    $\widehat{\mu\text{-risk}}$,
    $\widehat{\mu\text{-risk}}_{IPW}$,
    $\widehat{R\text{-risk}}$,
    $\widehat{U\text{-risk}}$,
    $\widehat{\tau\text{-risk}_{IPW}}$.
    We benchmark the metrics in a variety of settings:
    many different simulated data generation
    processes and three semi-simulated datasets \footnote{Scripts for the simulations and the selection procedure are available at
        \url{https://github.com/strayMat/caussim}.
    }.

    \subsection{Caussim: Extensive simulation settings}\label{subsec:simulations}

    \paragraph{Data Generation Process}

    We use simulated data, on which the ground-truth causal effect is known. Going
    further than prior empirical studies of causal model selection
    \cite{schuler_comparison_2018,alaa_validating_2019}, we use multiple
    generative processes, to reach more general conclusions (as discussed
    in Appendix \ref{apd:results:fig:seed_effect}).

    We generate the response functions using random bases.
    Basis extension methods are common in biostatistics, \emph{eg} functional
    regression with splines
    \cite{howe_splines_2011, perperoglou_review_2019}. By allowing the
    function to vary at specific knots, they give flexible --non-linear--
    models of the studied mechanisms. Taking inspiration from splines, we use
    random approximation of Radial Basis Function (RBF) kernels
    \cite{rahimi_random_2008} to generate the response surfaces. RBF use the
    same process as polynomial splines but replace polynomial by Gaussian
    kernels. Unlike polynomials, Gaussian kernels have
    decreasing influences in the input space.
    This avoids unrealistic divergences of the population response surfaces
    at the ends of the feature space.

    The number of basis functions --\emph{ie. knots}--, controls the complexity of
    the ground-truth response surfaces and treatment. We first use this process to
    draw the non-treated response surface $\mu_0$ and the causal-effect $\tau$. We
    then draw the observations from a mixture two Gaussians, for the treated and non
    treated. We vary the separation between the two Gaussians to control the amount
    of overlap between treated and control populations, as it an important parameter
    for causal inference (related to $\eta$ which appears in section
    \ref{theory:mu_risk_ipw_bound}). Finally, we generate the observed outcomes
    adding Gaussian noise. We generated such datasets 1000 times, with
    uniformly random overlap parameters $\theta \in \left[ 0, 2.5 \right]$. Appendix
    \ref{apd:experiments:generation} gives more details on the data generation.

    \captionsetup[sub]{font=large,labelfont={bf,sf}}%

    \begin{figure}[!t]
        \begin{minipage}{0.48\linewidth}
            \centering
            %\makebox[(\textwidth-9mm)/2]{\textbf{a. Caussim}}
            \includegraphics[width=0.98\linewidth,clip,trim={0 15.3cm 0 0}]{images/caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
        \end{minipage}
        \hfill
        \begin{minipage}{0.48\linewidth}
            \hfill%
            \includegraphics[width=0.98\linewidth,clip,trim={0 0 0 17cm}]{images/caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
        \end{minipage}
        \caption{Example of the simulation setup in the input space with two
            knots --\emph{ie.}basis functions. The left panel gives
            views of the observations in feature space, while the right panel displays the
            two response surfaces on a 1D cut along the black lines drawn on
            the left panel.}
        \label{fig:simulation_examples}
    \end{figure}

    \paragraph{Family of candidate estimators}

    We test model selection on a family of candidate estimators that
    approximate imperfectly the data-generating process. To build such an
    estimator in two steps, we first use a RBF expansion
    similar as the one used for the data-generation generation process. Concretely,
    we choose two random knots and apply a transformation of the raw data features
    with the same Gaussian kernel used for the data-generation mechanism. This step
    is referred as the featurization. Then, we fit a linear regression on
    these
    transformed features. We consider two ways of combining these steps for outcome
    mode; using common nomenclature \cite{kunzel_metalearners_2019}, we refer
    to these regression structures as different meta-learners which differ on how
    they model, jointly or not, the treated and the non treated:
    \begin{itemize}
        \item SLearner: A single learner for both population, taking the treatment as
              a supplementary covariate.

        \item SftLearner: A single set of basis functions is sampled at random for both
              populations, leading to a given feature space used to model both the treat and
              the non treated, then two
              separate different regressors are fitted on this representation.
        \item TLearner: Two completely different learners for each population, hence
              separate featurization and separate regressors.
    \end{itemize}

    We do not include more elaborated meta-learners such as R-learner
    \cite{nie_quasioracle_2017} or X-learner
    \cite{kunzel_metalearners_2019}. Our goal is not to have the best possible
    learner but to have a variety of sub-optimal learners in order to compare the
    different causal metrics. For the same reason, we did not include more powerful
    outcome models such as random forests or boosting trees.

    For the regression step, we fit a Ridge regression on the transformed features
    with 6 different choices of the regularization parameter $\lambda \in [10^{-3},
            10^{-2}, 10^{-1}, 1, 10^{1}, 10^{2}]$, coupled with a TLearner or a SftLearner.
    We sample 10 different random basis for the learning procedure and the
    featurization yielding a family $\mathcal F$ of 120 candidate estimators.
    %

    \subsection{Semi-simulated datasets}

    \paragraph{Datasets}\label{semi_simulated:datasets}

    We also use semi-simulated datasets, where a known synthetic causal effect is
    added to real --non synthetic-- covariate. We study datasets used in previous work
    to evaluate causal inference:
    \begin{description}
        \item[ACIC 2016] \cite{dorie_automated_2019}: The dataset is based on the
            Collaborative Perinatal Project \cite{niswander_women_1972}, a RCT
            conducted on a cohort of pregnant women to identify causes of infants’
            developmental disorders. The initial intervention was a child’s birth
            weight $(A = 1 \text{ if weight} < 2.5 kg)$, and outcome was the child’s
            IQ after a given follow-up period. The study contained $N=4\,802$ data
            points with $D=55$ features (5 binary, 27 count data, and 23
            continuous). They simulated 77 different setups with varying parameters
            for treatment and response generation models, treatment assignment
            probabilities, overlap, and interactions between treatment and
            covariates \footnote{Original R code available at
                \href{https://github.com/vdorie/aciccomp/tree/master/2016}{https://github.com/vdorie/aciccomp/tree/master/2016}
                to generate 77 simulations settings.}. We used 10 different seeds for
            every setup, totalizing 770 dataset instances.

        \item[ACIC 2018] \cite{shimoni_benchmarking_2018}: The raw covariates
            data comes from the Linked Births and Infant Deaths Database (LBIDD)
            \cite{macdorman_infant_1998} with $D=177$ covariates. Treatment and
            outcome models have been simulated with complex models to reflect
            different scenarii. The data do not provide the true propensity
            scores, so we evaluate only feasible metrics, which do not require this
            nuisance parameter. We used all 432 datasets\footnote{Using the scaling part of the data, from
                \href{https://github.com/IBM-HRL-MLHLS/IBM-Causal-Inference-Benchmarking-Framework}{github.com/IBM-HRL-MLHLS/IBM-Causal-Inference-Benchmarking-Framework}} of size $N=5\,000$.

        \item[Twins] \cite{louizos_causal_2017}: It is an augmentation of the
            real data on twin births and mortality rates in the USA from 1989-1991
            \cite{almond_costs_2005}. There are $N=11\,984$ samples (pairs of twins),
            and $D=50$ covariates\footnote{We obtained the dataset from
                \href{https://github.com/AMLab-Amsterdam/CEVAE/tree/master/datasets/TWINS}{https://github.com/AMLab-Amsterdam/CEVAE/tree/master/datasets/TWINS}}, The outcome is the mortality and the treatment is the
            weight of the heavier twin at birth. This is a "true" counterfactual dataset
            --as remarked in \cite{curth_really_2021}-- in the sense that we have
            both potential outcomes with each twin. They simulate the treatment with a
            sigmoid model based on GESTAT10 (number of gestation weeks before birth) and x
            the 45 other covariates:
            \begin{equation}
                \mathbf{t}_{i} \mid \mathbf{x}_{i}, \mathbf{z}_{i} \sim
                \operatorname{Bern}\left(\sigma\left(w_{o}^{\top}
                \mathbf{x}+w_{h}(\mathbf{z} / 10-0.1)\right)\right) \quad \text{with} \;
                w_{o} \sim \mathcal{N}(0,0.1 \cdot I),\; w_{h} \sim \mathcal{N}(5,0.1)
            \end{equation}
            We built upon this equation, adding a non-constant slope in the
            treatment sigmoid, allowing us to control the amount of overlap between
            treated and control populations.
            We sampled uniformly 1\,000 different overlap parameters between 0 and
            2.5, totalizing 1\,000 dataset instances. Unlike the previous datasets,
            only the overlap varies for these instances. The response surfaces are
            fixed by the original twin outcomes.
    \end{description}

    \paragraph{Family of candidate
        estimators}\label{semi_simulated:candidate_estimators}

    For these three datasets, the family of candidate estimators are gradient
    boosting trees for both the response surfaces and the treatment
    \footnote{Scikit-learn regressor,
        \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingRegressor.html}{HistGradientBoostingRegressor},
        and classifier,
        \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingClassifier.html}{HistGradientBoostingClassifier}.}
    with S-learner, learning rate
    in $\{0.01, 0.1, 1\}$, and maximum number of leaf nodes in $\{25, 27, 30, 32,
        35, 40\}$ resulting in a family of size 18.

    \paragraph{Nuisance estimators}

    Drawing inspiration from the TMLE literature that uses combination of flexible
    machine learning methods \cite{schuler_targeted_2017}, we use as models
    for the nuisances $\check e$ (respectively $\check m$) a form of meta-learner: a
    stacked estimator of ridge and boosting classifiers (respectively
    regressions). We select hyper-parameters with randomized search on a validation set
    $\mathcal{V}$ and keep them fix for model selection (detailed of the hyper
    parameters in Appendix \ref{apd:experiments:nuisances_hp}). As extreme inverse
    propensity weights induce high variance, we use clipping
    \cite{swaminathan_counterfactual_2015, ionides_truncated_2008} to bound
    $min(\check e, 1-\check e)$ away from 0 with a fixed $\eta=10^{-10}$, ensuring
    strict overlap for numerical stability.

    %\paragraph{Model selection procedure}\label{semi_simulated:selection_procedure}

    %We used the same training procedure procedure as in
    %\ref{paragraph:selection_procedure} (Details in supp.~mat.
    %\ref{apd:selection_procedure_acic}).

    \subsection{Measuring overlap between treated and non treated}\label{subsec:measuring_overlap}

    %\idea{Overlap is important but not easily measurable}
    Good overlap, or ``positivity'' between treated and control population is
    crucial for causal inference as it is required by the positivity assumption
    \ref{assumption:overlap} for causal identification. It is typically assessed by
    qualitative methods using population histograms (as in Figure
    \ref{fig:toy_example}) or side-by-side box plots, or quantitative approaches
    such as Standardized Mean Difference
    \cite{austin_introduction_2011,austin_moving_2015}. While these methods are
    useful to decide if positivity holds, they do not summarize a dataset's overlap
    in a single measure. Rather, we compute the divergence
    between the population covariate distributions $\mathbb P(X|A=0)$ and $\mathbb
        P(X|A=1)$ to characterize the behavior of causal risk
    \cite{damour_overlap_2020,johansson_generalization_2021}. We introduce the
    Normalized Total Variation (NTV), a divergence based on the sole propensity
    score. Details are given in Appendix \ref{apd:motivation_ntv}.

    \begin{figure}[!b]
        \centering
        \includegraphics[width=\textwidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset.pdf}
        \hfill
        \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$, measured
            as the difference between each metric Kendall's $\tau$ and the mean kendall's
            $\tau$ over all metric:  $\kappa(\ell,\tau\mathrm{{-risk}}) - mean_{\ell} \big(\kappa(\ell,\tau\mathrm{{-rRisk}}) \big)$. Strong and Weak overlap correspond
            to the first and
            last tertiles of the overlap distribution measured with Normalized Total
            Varation eq. \ref{eq:ntv}.
            Appendix \ref{apd:experiments:additional_results} presents the same results
            measured with absolute Kendall's in Figure
            \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with
            $\tau\mathrm{-risk}$ gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table \ref{apd:table:relative_kendalls_all_datasets} displays the median
            and IQR for the relative Kendall's results.}\label{fig:relative_kendalls_all_datasets}
    \end{figure}

    \subsection{Empirical results: factors driving good model selection across
        datasets}\label{empirical_study:results}

    \paragraph{The $R\text{-risk}$ is the best metric}

    Each metric ranks differently the candidate models. Figure
    \ref{fig:relative_kendalls_all_datasets} shows the agreement between the
    ideal ranking of methods given the oracle $\tau\text{-risk}$ and
    the different causal metrics under evaluation. We measure this agreement
    with a relative\footnote{To remove the variance
        across datasets (some datasets lead to easier model selection than
        others), we report values for one metric relative to the mean of all
        metrics for a given dataset instance: $\text{Relative} \, \kappa(\ell,\tau\mathrm{{-risk}})=
            \kappa(\ell,\tau\mathrm{{-risk}}) -
            mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$ (eq.
    \ref{eq:kendall_tau}) \cite{kendall_new_1938}.
    Given the importance of overlap in
    how well metrics approximate the oracle $\tau\text{-risk}$
    (\autoref{apd:proofs:mu_risk_ipw_bound}), we separate
    strong and weak overlap --defined as first and
    last tertile of the Normalized Total Variation, eq. \ref{eq:ntv}.

    Among all metrics the classical mean squared error (ie. factual
    $\mu\text{-risk}$) is worse and reweighting it with propensity score
    ($\mu\text{-risk}_{IPW}$) does not bring much improvements. The
    $R\text{-risk}$, which includes a model of mean outcome and propensity
    scores, leads to the best performances. Interestingly, the
    $U\text{-risk}$, which uses the same nuisances, has good performances for
    strong overlap but deteriorates in weak overlap, probably due variance
    inflation when dividing by extreme propensity scores.

    Beyond the rankings, the differences in terms of absolute
    ability to select the best model are large: The model selected by the R-risk
    achieves a $\tau\text{-risk}$ only 1\% higher
    than that of the best
    possible candidate for strong overlap on Caussim, but selecting with
    the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
    practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
    --as in some causal-inference methods \citep{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
    (Figure
    \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
    other datasets, the $R\text{-risk}$ consistently decreases the
    risk compared to the $\mu\text{-risk}$:
    0.1\% compared to 1\% on ACIC2016,  1\% compared to 20\% on ACIC 2018,
    and 0.05\% compared to 1\% on Twins.

    \paragraph{Model selection is harder in settings of low population
        overlap}

    Model selection for
    causal inference becomes more and more difficult with increasingly different
    treated and control populations
    (Figure \ref{fig:all_datasets_overlap_effect_r_risk}). The
    absolute Kendall's coefficient correlation with $\tau\text{-risk}$
    drops from values around 0.9 (excellent agreement with oracle selection) to 0.6 on
    both Caussim and ACIC 2018
    (Appendix \ref{apd:experiments:additional_results},
    Figure {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}).

    \begin{figure}[!h]
        \centering
        \includegraphics[width=0.9\textwidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only.pdf}
        \hfill
        \caption{\textbf{Model selection is harder in settings of low population
                overlap}:
            Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
            are the tertiles of the overlap distribution measured with NTV eq. \ref{eq:ntv}. Appendix \ref{apd:experiments:additional_results} presents results for all
            metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} and in absolute
            Kendall's and continuous overlap values in Figure
                {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
    \end{figure}

    \paragraph{Nuisances can be estimated on the same data as outcome models}

    Using the train set $\mathcal{T}$ both to fit the candidate estimator and the
    nuisance estimates is a form of double dipping which can lead errors in
    nuisances to be correlated to that of outcome models
    \cite{nie_quasioracle_2017}. In theory, these correlations can bias model
    selection and, strictly speaking, there are theoretical arguments
    to split out a third separated data set --a ``nuisance set''-- to fit the
    nuisance models. The drawback is that it depletes the data available for
    model estimation and selection. However, Figure \ref{fig:procedures_comparison} shows that
    there is no substantial differences between a procedure with a separated
    nuisance set and the simpler shared nuisance-candidate set procedure
    (results for all metrics in Figure
    \ref{apd:fig:procedures_comparison_all_metrics}).

    \begin{figure}[!h]
        %\centering
        \begin{subfigure}[b]{0.48\textwidth}
            %\centering
            %\caption{\textbf{Two sets procedure}}
            \includegraphics[width=2\textwidth]{images/_3_procedure_r_risk_only_3datasets.pdf}
        \end{subfigure}
        \hfill
        \caption{\textbf{Nuisances can be estimated on the same data as outcome
                models}: Results for the R-risk are similar between the
            \textcolor{MidnightBlue}{shared
                nuisances/candidate set} and
            the \textcolor{RedOrange}{separated nuisances set} procedures. Figure
            \ref{apd:fig:procedures_comparison_all_metrics} details results for all metrics.}\label
        {fig:procedures_comparison}
    \end{figure}

    \paragraph{Stacked models are good overall estimators of nuisances}

    Oracle versions of every risks recover more often the best estimator.
    However, in configurations that we investigated,
    stacked nuisances estimators (boosting and linear) lead to feasible
    metrics with close performances to the oracles ones. This suggests that the
    corresponding estimators recover well-enough the true nuisances.
    One may wonder if it may be useful to use simpler models for the
    nuisances, in particular in settings where there are less data or where
    the true models are linear.
    Figure \ref{fig:all_datasets_nuisances_comparison} compares causal model
    selection estimating nuisances with stacked estimators or linear model.
    It comprises the Twins data, where the true propensity model is linear,
    and a downsampled version of this data, to study a situation favorable to
    linear models. In these settings,
    stacked and linear estimations of the nuisances performs equivalently.
    Overall, Figure \ref{apd:fig:nuisances_comparison_twins} suggests that
    to estimate nuisance it suffices to use adaptive models as built by
    stacking linear models and gradient-boosted trees.
    Figure \ref{apd:fig:nuisances_comparison} in the appendices, details
    other causal metrics.

    \begin{figure}[!h]
        \begin{subfigure}[b]{0.48\textwidth}
            \centering
            \includegraphics[width=2\textwidth]{images/_4_nuisance_models_r_risk_only_3datasets.pdf}
        \end{subfigure}
        \hfill
        \caption{\textbf{\textcolor{DarkOrchid}{Stacked
                    models} are good overall estimators of the nuisances}:
            Results are shown only for the
            R-risk. Details for every metrics are provided in Figure
            \ref{apd:fig:nuisances_comparison}. For Twins, where the true propensity
            model is linear, \textcolor{DarkOrchid}{stacked} and
            \textcolor{ForestGreen}{linear}
            estimations of the nuisances performs equivalently, even for a downsampled version
            (N=4794). }\label{fig:all_datasets_nuisances_comparison}
    \end{figure}

    \paragraph{Use 90\% of the data to estimate outcome models, 10\% to
        select them}

    For best causal modeling, the analyst often faces a compromise: given a finite
    data sample, should she allocate most of the data to estimate the outcome model,
    thus maximizing chances of achieving a high-quality outcome model but leaving
    little data for model selection. Or, she could choose a bigger test set for
    model selection and effect estimation. For causal model selection, there
    is no established practice (as reviewed in Appendix \ref{apd:results:k_fold_choices}).

    We investigate such tradeoff varying the ratio between train and test
    data size. For this, we first split out 30\% of the data as a holdout set
    $\mathcal{V}$ on which we use the oracle response functions to derive
    silver-standard estimates of the causal quantities of interest. We then
    use the standard estimation procedure on the remaining 70\% of the data,
    splitting it into train $\mathcal{T}$ and test $\mathcal{S}$ of varying
    sizes. We finally measure the error between this estimate and the
    silver-standard one.

    We consider two different analytic goals: estimating a average
    treatment effect --a single number used for policy making-- and a
    CATE --A full model of the treatment effect as a function of covariates
    $X$. Given that the latter is a much more complex object than the former,
    the optimal train/test ratio might vary. To measure errors, we use for
    the ATE the relative absolute ATE bias between the ATE computed with the
    selected outcome model on the test set, and the true ATE as evaluated on
    the holdout set $\mathcal{V}$. For the CATE, we compare the
    $\tau\text{-risk}$
    of the best selected model applied on the holdout set $\mathcal{V}$. We explore this trade-off for the ACIC 2016 dataset and the R-risk.

    \begin{figure}[!t]
        %\centering
        \begin{minipage}{.5\textwidth}
            \centerline{\textbf{a) CATE estimation error}}

            \includegraphics[height=.35\textwidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16.pdf}
        \end{minipage}
        \begin{minipage}{.5\textwidth}
            %\centering
            \centerline{\textbf{b) ATE estimation error}}
            \includegraphics[height=.35\textwidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16.pdf}
        \end{minipage}%
        %\hspace{0.3\textwidth}
        \caption{\textbf{a) For CATE, a train/test ratio of 0.9/0.1 appears as a good
                trade-off.} b) For ATE, there is a small signal pointing also to
            0.9/0.1 (K=10).
            for ATE. Experiences on 10 replications of all 78 instances of the ACIC 2016
            data.}\label{fig:train_test_ratio}
    \end{figure}

    Figure \ref{fig:train_test_ratio} shows that a train/test ratio of
    0.9/0.1 (K=10) appears as a good trade-off for CATE and ATE estimation,
    thought there is little difference with a split of 0.8/0.2 (K=5).

    \section{Discussion and conclusion}\label{sec:discussion}

    Predictive models are increasingly used to reason about causal effects.
    Our results highlight that they should be selected, validated, and tuned
    using different procedures and error measures than those classically used
    to assess prediction (estimating the so-called $\mu\text{-risk}$).
    Rather, selecting the best outcome model according to the $R\text{-risk}$
    (eq.\,\autoref{def:r_risk}) leads to more valid causal estimates.
    %
    Estimating this risk requires a markedly more complex procedure than
    standard cross-validation used \emph{e.g.} in machine learning: it involves
    fitting nuisance models necessary for model evaluation, though our
    empirical results show that these can be learned on the same set of data
    as the outcome model evaluated.
    %
    A poor estimation of the nuisance models may compromise the benefits of
    the more complex $R\text{-risk}$ (as shown in Figure
    \ref{fig:all_datasets_nuisances_comparison}). However controlling and selecting
    these latter models is easier because they are associated to errors on
    observed distributions. Our empirical results show that when selecting
    these models in a flexible family of models the $R\text{-risk}$ dominates
    simpler risks for model selection.
    %
    Results show that going from an oracle $R\text{-risk}$ --where the
    nuisances are known-- to a feasible $R\text{-risk}$ --where the nuisances
    are estimated-- decreases only very slightly the model-selection
    performance of the $R\text{-risk}$. This may be explained by
    theoretical results that suggest that estimation errors on both
    nuisances partly compensate out in the
    $R\text{-risk}$\cite{daniel2018double,naimi2021challenges,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018}.
    % 
    %The usage of the $R\text{-risk}$ can also be understood as a
    %$\tau \text{-risk}$ reweighted by the propensity score
    %(prop \autoref{theory:prop:r_risk_rewrite}).
    %

    For strong overlap, the conventional procedure ($\mu\text{-risk}$) appears theoretically motivated
    (\autoref{theory:mu_risk_ipw_bound}), however empirical results show that
    even in this regime the $R\text{-risk}$ brings a sizeable benefit,
    in agreement with \citet{schuler_comparison_2018}.

    Our results points to K=5 or 10 for the choice of the number of folds used in
    cross-validation for CATE and ATE, aligned with previous empirical evidence for
    ATE \cite{chernozhukov_double_2018}.

    \paragraph{Extension to binary outcome}
    While we focused on continuous outcomes, in medicine, the target outcome
    is often a categorical variable such as mortality status or diagnosis. In
    this case, it may be interesting to focus on other estimands than the
    Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
    $\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
    or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
        =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
    \cite{austin2017estimating}. While the odds ratio is natural for
    case-control studies \cite{rothman2008case}, a good choice of measure
    can reduce heterogeneity \cite{colnet2023risk}.
    Using as an estimand the log of
    these values is suitable to additive models (for reasoning or noise
    assumptions). In the log domain, the relative risk or the odds ratio are
    written as a difference, as the ATE: $\log \mathbb{P}(Y(1)=1) - \log
        \mathbb{P}(Y(0=1)$ or $\log (\mathbb{P}(Y(1)=1) / [1 -\mathbb{P}(Y(1)=1)]
        ) - \log \mathbb{P}(Y(0=1) / [1 -\mathbb{P}(Y(0)=1)]$. Hence, the
    framework studied here (\autoref{sec:neyman_rubin}) can directly apply.
    It is particularly easy for the log odds ratio, as it is the output of a
    logistic regression or any model with a cross-entropy loss.

    \paragraph{Going further}
    % Estimation of the nuisances, calibration
    The $R\text{-risk}$ needs good estimation of nuisance models. The
    propensity score $e$ calls for a control on the estimation of the
    individual posterior probability. We have used the Brier score to select
    these models, as it is minimized by the true individual probability. Regarding model-selection for
    propensity score, an easy mistake is to use expected calibration errors
    popular in machine learning
    \cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
    as these select not for the individual posterior probability but for an
    aggregate error rate \cite{perez2022beyond}. An
    open question is whether a better metric than the brier score can be designed that controls for $e \, (1 - e)$, the quantity used in
    the $R\text{-risk}$, rather than $e$.

    %\idea{Treatment heterogeneity seems important.}
    The quality of model selection varies substantially from one
    data-generating mechanism to another. The overlap appears as an important
    parameter: when the treated and untreated, causal model selection is very
    hard. However, remaining variance in the empirical results suggests that
    other parameters of the data generation processes come into play.
    Intuitively, the complexity of the response surfaces and the treatment
    heterogeneity interact with overlap violations: when extrapolations to
    weak-overlap regions is hard, causal model selection is hard.

    Nevertheless, from a practical perspective, our study establishes that
    the $R\text{-risk}$ is the best option to select predictive models for
    causal inference, without requiring assumptions on the data-generating
    mechanism, the amount of data at hand, or the specific estimators used to
    build predictive models.

    % XXX: need a final positive
    % remark
    % Maybe that as our work has no
    % requirement on the learning procedure,
    % it contributes to ensuring the
    % validity of causal inference on complex databases such as clinical
    % records with a lot of open-ended text that must be analyzed with
    % language models.

    % Sentence stolen, adapt
    % unlike parametric regression, machine learning (ML) methods do not
    % generally require precise
    % knowledge of the true data generating mechanisms

\end{multicols}{2}

\section*{Acknowledgments}

We acknowledge fruitful discussions with Bénédicte Colnet.

\subsection*{Financial disclosure}

None reported.

\subsection*{Conflict of interest}

The authors declare no potential conflict of interests.

%\if@Mybibrefstyle%
\printbibliography
%\fi if using njdnatbib.sty \bibliography{bibliography}%

\clearpage

\appendix

%The following supporting information is available as part of the online article:

\section{Variability of ATE estimation on ACIC
  2016}\label{apd:toy_example:acic_2016_ate_variability}

Figure \ref{fig:acic_2016_ate_heterogeneity} shows ATE estimations for six
different models used in g-computation estimators on the 76 configurations of
the ACIC 2016 dataset. Outcome models are fitted on half of the data and
inference is done on the other half --ie. train/test with a split ratio of 0.5.
For each configuration, and each model, this train test split was repeated ten
times, yielding non parametric variance estimates
\cite{bouthillier_accounting_2021}.

Outcome models are implemented with
\href{https://scikit-learn.org/stable/}{scikit-learn}
\cite{pedregosa_scikitlearn_2011} and the following hyper-parameters:

\begin{table}[h!]
    \centering
    \begin{tabular}{llll}
        \toprule
        Outcome Model                                  & Hyper-parameters grid
        \\
        \midrule
        Random Forests                                 & Max depth: [2,
        10]                                                                    \\

        Ridge regression without treatment interaction & Ridge regularization:
        [0.1]                                                                  \\

        Ridge regression with treatment interaction    & Ridge regularization:
        [0.1]                                                                  \\
        \bottomrule
    \end{tabular}
    \caption{Hyper-parameters grid used for ACIC 2016 ATE variability}
    \label{apd:toy_example:acic_2016_ate_variability:table}
\end{table}

\section{Causal assumptions}\label{apd:causal_assumptions}

We assume the following four assumptions, referred as strong ignorability and
necessary to assure identifiability of the causal estimands with observational
data \cite{rubin_causal_2005}:
\begin{assumption}[Unconfoundedness]\label{assumption:ignorability}
    \begin{equation*}\label{eq:ignorability}
        \{Y(0), Y(1) \} \indep A | X
    \end{equation*}
    This condition --also called ignorability-- is equivalent to the conditional
    independence on $e(X)$ \cite{rosenbaum_central_1983}: $\{Y(0), Y(1) \}
        \indep  A | e(X)$.
\end{assumption}

\begin{assumption}[Overlap, also known as Positivity)]\label{assumption:overlap}
    \begin{equation*}\label{eq:overlap}
        \eta < e(x) < 1 - \eta \quad \forall x \in \mathcal X \text{ and some } \eta > 0
    \end{equation*}
    The treatment is not perfectly predictable. Or with different words, every
    patient has a chance to be treated and not to be treated. For a given set of
    covariates, we need examples of both to recover the ATE.
\end{assumption}

As noted by \cite{damour_overlap_2020}, the choice of covariates $X$ can
be viewed as a trade-off between these two central assumptions. A bigger
covariates set generally reinforces the ignorability assumption. In the
contrary, overlap can be weakened by large $\mathcal{X}$ because of the
potential inclusion of instruments: variables only linked to the treatment which
could lead to arbitrarily small propensity scores.

% remark: There is a major counter example to these colliders (variables which
% are caused by both the outcome and the treatment),

\begin{assumption}[Consistency]\label{assumption:consistency} The observed
    outcome is the potential outcome of the assigned treatment:
    \begin{equation*}\label{eq:consistancy}
        Y = A \, Y(1) + (1-A) \, Y(0)
    \end{equation*}
    Here, we assume that the intervention $A$ has been well defined. This
    assumption focuses on the design of the experiment. It clearly states the link
    between the observed outcome and the potential outcomes through the
    intervention \cite{hernan_causal_2020}.
\end{assumption}

\begin{assumption}[Generalization]\label{assumption:generalization} The training
    data on which we build the estimator and the test data on which we make the
    estimation are drawn from the same distribution $\mathcal D^*$, also known as
    the ``no covariate shift'' assumption \cite{jesson_identifying_2020}.
\end{assumption}

\section{Proofs: Links between feasible and oracle risks}\label{apd:proofs}

\subsection{Upper bound of $\tau\text{-risk}$ with
    $\mu\text{-risk}_{IPW}$}\label{apd:proofs:mu_risk_ipw_bound}

For the bound with the $\mu\text{-risk}_{IPW}$, we will decompose the CATE risk
on each factual population risks:

\begin{definition}[Population Factual $\mu\text{-risk}$]\label{mu_risk_a}
    \cite{shalit_estimating_2017}
    \begin{equation*}
        \mu\text{-risk}_{a}(f)= \int_{\mathcal Y \times \mathcal X} (y-f(x ; A=a))^{2}  p(y ; x=x \mid A=a) \; dy dx
    \end{equation*}
\end{definition}

Applying Bayes rule, we can decompose the $\mu\text{-risk}$ on each
intervention:
\begin{equation*}
    \mu\text{-risk}(f)
    =p_{A} \,\mu\text{-risk}_{1}(f)+\left(1-p_{A}\right) \,\mu\text{-risk}_{0}(f)
    \text{with } p_A=\mathbb P(A=1)
\end{equation*}

These definitions allows to state a intermediary result on each population:
\begin{lemma}[Mean-variance decomposition]\label{apd:proofs:mu_risk_ipw_link_mu}
    We need a reweighted version of the classical mean-variance decomposition.

    For an outcome model $f: x \times A \rightarrow \mathcal X$. Let the inverse
    propensity weighting function $w(a ; x)=a e(x)^{-1}+(1-a)(1-e(x))^{-1}$.
    \begin{align*}
         & \int_{\mathcal X}(\mu_{1}(x)-f(x ; 1))^{2} p(x) dx  = p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  -\sigma^{2}_{Bayes}(1)
    \end{align*}
    And
    \begin{align*}
         & \int_{\mathcal X}(\mu_{0}(x)-f(x; 0))^{2} p(x) dx   = (1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)  -\sigma^{2}_{Bayes}(0)
    \end{align*}

    \begin{proof}
        \begin{align*}
             & p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  = \int_{\mathcal X \times \mathcal Y} \frac{1}{e(x)}(y-f(x ; 1))^{2} p(y \mid x ; A=1) p(x ; A=1) d y d x                                                      \\
             & = \int_{\mathcal X \times \mathcal Y} (y-f(x ; 1))^{2} p(y \mid x ; A=1) \frac{p(x ; A=1)}{p(x ; A=1)}p(x)dy dx                                                                                      \\
             & = \int_{\mathcal X \times \mathcal Y} \big[(y-\mu_1(x))^{2}+\left(\mu_{1}(x)-f(x ; 1)\right)^{2} + 2\left(y-\mu_{1}(x)\right)\left(\mu_{1}(x)-f(x, 1)\right) \big] p(y \mid x ; A=1) p(x) d y d x    \\
             & =\int_{\mathcal X} \big [ \int_{\mathcal Y} (y-\mu_1(x))^{2} p(y \mid x ; A=1) dy\big ] p(x)dx + \int_{\mathcal X \times \mathcal Y} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x)p(y \mid x ; A=1)dx dy \\
             & + \qquad 2 \int_{\mathcal X} \big [ \int_{\mathcal Y} \left(y-\mu_{1}(x)\right) p(y \mid x ; A=1) dy \big ] \left(\mu_{1}(x)-f(x, 1)\right)p(x)dx                                                    \\
             & =\int_{\mathcal X} \sigma_{y}^{2}(x, 1) p(x) d x +\int_{\mathcal X} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x) d x+0
        \end{align*}
    \end{proof}
\end{lemma}

\begin{proposition*}[Upper bound with mu-IPW]\label{apd:proofs:prop:upper_bound}
    Let f be a given outcome model, let the weighting function $w$ be the Inverse
    Propensity Weight $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$. Then, under
    overlap (assumption \ref{assumption:overlap}),
    \begin{align*}
        \tau\text{-risk}(f) \leq & \; 2 \, \mu\text{-risk}_{IPW}(w, f)  \; - 2 \, (\sigma^2_{Bayes}(1) +  \sigma^2_{Bayes}(0))
    \end{align*}

    \begin{proof}
        \begin{align*}
             & \tau\text{-risk}(f) =\int_{\mathcal X}(\mu_{1}(x)-\mu_{0}(x)-(f(x ; 1)-f(x ; 0))^{2} p(x) d x
        \end{align*}
        By the triangle inequality $(u+v)^2 \leq 2(u^2 + v^2)$:
        \begin{align*}
             & \tau\text{-risk}(f) \leq
            2 \int_{\mathcal X}\big[\left(\mu_{1}(x)-f(x ; 1)\right)^{2}+\left(\mu_{0}(x)-f(x ; 0)\right)^{2}\big] p(x) d x \\
        \end{align*}
        Applying Lemma \ref{apd:proofs:mu_risk_ipw_link_mu},
        \begin{align*}
             & \tau\text{-risk}(f) \leq 2\big[p_A \mu\text{-risk}_{IPW, 1}(w, f) +(1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)(w, f)\big] -2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1)) \\
             & = 2 \mu\text{-risk}_{IPW}(w, f)-2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1))
        \end{align*}
    \end{proof}
\end{proposition*}

\subsection{Reformulation of the $R\text{-risk}$ as reweighted
    $\tau\text{-risk}$}\label{apd:proofs:r_risk_rewrite}

\begin{proposition*}[$R\text{-risk}$ as reweighted $\tau
            \text{-risk}$]\label{apd:proofs:prop:r_risk_rewrite}

    \begin{proof}

        We consider the R-decomposition: \cite{robinson_rootnconsistent_1988},
        \begin{equation}\label{apd:eq:r_decomposition}
            y(a) = m(x) + \big( a - e(x) \big) \tau(x) + \varepsilon(x; a)
        \end{equation}
        Where $\mathbb E[\varepsilon(X; A)|X, A] = 0$ We can use it as plug in the
        $R\text{-risk}$ formula:

        \begin{align*}
             & R\text {-risk}(f) =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}[(y-m(x))-\big(a-e(x)\big) \tau_f(x)]^{2} p(y ; x ; a) d y d x d a                                     \\
             & =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}} \left[\big(a-e(x)\big)\tau(x)+\varepsilon(x ; a)-\big(a-e(x)\big) \tau_f(x)\right]^{2} p(y ; x ; a) d y d x da          \\
             & =\int_{\mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)^{2}\big(\tau(x)- \tau_f(x)\big)^{2} p(x ; a) d x d a                                                                       \\
             & + 2  \int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)\big(\tau(x)-\tau_f(x)\big)  \int_{\mathcal{Y}} \varepsilon(x ; a) p(y \mid x ; a) d y p(x ; a)dx da \\
             & +\int_{\mathcal{X} \times \mathcal{A}} \int_{\mathcal{Y}} \varepsilon^{2}(x ; a) p(y \mid x ; a) d y p(x ; a) d x d a
        \end{align*}

        The first term can be decomposed on control and treated populations to force
        $e(x)$ to appear:
        \begin{align*}
             & \int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}\left[e(x)^{2}p(x;0) + \big(1-e(x)\big)^{2} p(x;1)\right] d x                    \\
             & =\int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}  \left[e(x)^{2}\big(1-e(x)\big)p(x) + \big(1-e(x)\big)^{2}e(x) p(x)\right] d x \\
             & =\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x)[1-e(x)+e(x)] p(x) d x                                                     \\ &=\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x) p(x) d x.
        \end{align*}

        The second term is null since, $\mathbb E[\varepsilon(x, a) |X, A]=0$.

        The third term corresponds to the modulated residuals \ref{eq:residuals} :
        $\tilde{\sigma}_B^2(0) + \tilde{\sigma}_B^2(1)$

    \end{proof}
\end{proposition*}

\section{Measuring overlap}\label{apd:motivation_ntv}

\paragraph{Motivation of the Normalized Total Variation}
%\idea{A simplier measure, NTV seems to capture what we want}
Computing overlap when working only on samples of the observed distribution,
outside of simulation, requires a sophisticated estimator of discrepancy
between distributions, as two data points never have the same exact set of
features. Maximum Mean Discrepancy \cite{gretton2012kernel} is typically
used in the context of causal inference
\cite{shalit_estimating_2017,johansson_generalization_2021}. However it
needs a kernel, typically Gaussian, to extrapolate across neighboring
observations. We prefer avoiding the need to specify such a kernel, as it must
be adapted to the data which is tricky with categorical or non-Gaussian
features, a common situation for medical data.

For simulated and some semi-simulated data, we have access to the probability of
treatment for each data point, which sample both densities in the same data
point. Thus, we can directly use distribution discrepancy measures and rely on
the Normalized Total Variation (NTV) distance to measure the overlap between the
treated and control propensities. This is the empirical measure of the total
variation distance \cite{sriperumbudur_integral_2009} between the distributions,
$TV(\mathbb{P}(X|A=1), \mathbb{P}(X|A=0))$. As we have both distribution sampled
on the same points, we can rewrite it a sole function of the propensity score, a
low dimensional score more tractable than the full distribution $\mathbb
    P(X|A)$:

\begin{equation}\label{eq:ntv}
    \widehat{NTV}(e, 1-e) = \frac{1}{2N} \sum_{i =1}^{N} \big |\frac{e(x_i)}{p_A}-\frac{1-e(x_i)}{1-{p_A}} \big |
\end{equation}

Formally, we can rewrite NTV as the Total Variation distance
between the two population distributions. For a population $O = (Y(A), X, A)
    \sim \mathcal{D}$:

\begin{align*}
    NTV(O) & = \frac{1}{2N} \sum_{i =1}^{N} \big | \frac{e(x_i)}{p_A}-\frac{1-e(x_i)}{1-{p_A}} \big|         \\
           & = \frac{1}{2N} \sum_{i =1}^{N} \big|\frac{P(A=1|X=x_i)}{p_A}-\frac{P(A=0|X=x_i)}{1-{p_A}} \big|
\end{align*}

Thus NTV approximates the following quantity in expectation over the data
distribution $\mathcal{D}$:

\begin{align*}
    NTV(\mathcal{D}) & = \int_{\mathcal{X}} \big | \frac{p(A=1|X=x)}{p_A}-\frac{p(A=0|X=x)}{1-{p_A}} \big | p(x)dx \\
                     & = \int_{\mathcal{X}} \big | \frac{p(A=1, X=x)}{p_A}-\frac{p(A=0, X=x)}{1-{p_A}} \big |dx    \\
                     & = \int_{\mathcal{X}} \big | p(X=x|A=1)- p(X=x|A=0) \big |dx
\end{align*}

For countable sets, this expression corresponds to the Total Variation distance
between treated and control populations covariate distributions : $TV(p_0(x),
    p_1(x))$.

\paragraph{Measuring overlap without the oracle propensity scores:} For ACIC
2018, or for non-simulated data, the true propensity scores are not known. To
measure overlap, we rely on flexible estimations of the Normalized Total
Variation, using gradient boosting trees to approximate the propensity score.
Empirical arguments for this plug-in approach is given in Figure
\ref{apd:overlap:ntv_approximation}.

\paragraph{Empirical arguments}

We show empirically that NTV is an appropriate measure of overlap by :
\begin{itemize}
    \item Comparing the NTV distance with the MMD for Caussim which is gaussian
          distributed in Figure \ref{apd:overlap:caussim:mmd_vs_ntv},
    \item Verifying that setups with penalized overlap from ACIC 2016 have a
          higher total variation distance than unpenalized setups in Figure
          \ref{apd:overlap:penalized_overlap}.
    \item Verifying that the Inverse Propensity Weights extrema (the inverse of
          the $\nu$ overlap constant appearing in the overlap Assumption
          \ref{assumption:overlap}) positevely correlates with NTV for Caussim,
          ACIC 2016 and Twins in Figure \ref{apd:ntv_vs_max_ipw}. Even if the same
          value of the maximum IPW could lead to different values of NTV, we
          expect both measures to be correlated : the higher the extrem propensity
          weights, the higher the NTV.
\end{itemize}

\paragraph{Estimating NTV in practice}

Finally, we verify that approximating the NTV distance with a learned plug-in
estimates of e(x) is reasonnable. We used either a logistic regression or a
gradient boosting classifier to learn the propensity models for the three
datasets where we have access to the ground truth propensity scores: Caussim,
Twins and ACIC 2016. We respectively sampled 1000, 1000 and 770 instances of
these datasets with different seeds and overlap settings. We first run a
hyperparameter search with cross-validation on the train set, then select the
best estimator. We refit on the train set this estimator with or without
calibration by cross validation and finally estimate the normalized TV with the
obtained model. This training procedure reflects the one described in Algorithm
\ref{problem:estimation_procedure:algo} where nuisance models are fitted only on
the train set.

The hyper parameters are : learning rate $ \in [1e-3, 1e-2, 1e-1, 1]$, minimum
samples leaf $\in [2, 10, 50, 100, 200]$ for boosting and L2 regularization $\in
    [1e-3, 1e-2, 1e-1, 1]$ for logistic regression.

Results in Figure \ref{apd:overlap:ntv_approximation} comparing bias to the true
normalized Total Variation of each dataset instances versus growing true NTV
indicate that calibration of the propensity model is crucial to recover a good
approximation of the NTV.

\begin{figure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \caption{\textbf{Uncalibrated classifiers}}
        \includegraphics[width=\linewidth]{images/overlap_measure_diff_oracle_ntv_to_n_tv_calibration=False_vs_oracle_n_tv.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \caption{\textbf{Calibrated
                classifiers}}
        \includegraphics[width=\linewidth]{images/overlap_measure_diff_oracle_ntv_to_n_tv_calibration=True_vs_oracle_n_tv.pdf}
    \end{subfigure}
    \caption{a) Without calibration, estimation of NTV is not trivial even for
        boosting models. b) Calibrated classifiers are able to recover the true
        Normalized Total Variation for all datasets where it is
        available.}\label{apd:overlap:ntv_approximation}
\end{figure}

\begin{figure}
    \centering
    \caption{NTV recovers well the overlap settings described in the ACIC paper
        \cite{dorie_automated_2019}}\label{apd:overlap:penalized_overlap}
    \includegraphics[width=0.7\linewidth]{images/overlap_measure_acic_2016_recovery_overlap_setup.pdf}
\end{figure}

\begin{figure}[htbp]
    \centering
    \caption{Good correlation between overlap measured as normalized Total
        Variation and Maximum Mean Discrepancy (200 sampled Caussim
        datasets)}\label{apd:overlap:caussim:mmd_vs_ntv}
    \includegraphics[width=0.7\linewidth]{images/overlap_measure_caussim_transformed_mmd_vs_ntv.pdf}
\end{figure}

\begin{figure}[ht]
    \centering
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \caption{\textbf{Caussim}}
        \includegraphics[width=\textwidth]{images/overlap_measure_caussim_max_ipw_vs_ntv.pdf}
        \label{apd:caussim:max_ipw_vs_ntv}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \caption{\textbf{ACIC 2016}}
        \includegraphics[width=\textwidth]{images/overlap_measure_acic_2016_max_ipw_vs_ntv.pdf}
        \label{apd:acic_2016:ntv_vs_max_ipw}
    \end{subfigure}
    \begin{subfigure}[b]{0.47\textwidth}
        \centering
        \caption{\textbf{ACIC 2018}}
        \includegraphics[width=\textwidth]{images/overlap_measure_acic_2018_max_ipw_vs_ntv.pdf}
        \label{apd:acic_2018:ntv_vs_max_ipw}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{\textbf{TWINS}}
        \includegraphics[width=\textwidth]{images/overlap_measure_twins_max_ipw_vs_ntv.pdf}
        \label{apd:twins:ntv_vs_max_ipw}
    \end{subfigure}
    \caption{Maximal value of Inverse Propensity Weights increases exponentially with the overlap as measure by Normalized Total Variation.}
    \label{apd:ntv_vs_max_ipw}
\end{figure}

\section{Experiments}

\subsection{Details on the data generation process}
\label{apd:experiments:generation}

We use Gaussian-distributed covariates and random basis expansion based on
Radial Basis Function kernels. A random basis of RBF kernel enables modeling
non-linear and complex relationships between covariates in a similar way to the
well known spline expansion. The estimators of the response function are learned
with a linear model on another random basis (which can be seen as a stochastic
approximation of the full data kernel \cite{rahimi_random_2008}). We
carefully control the amount of overlap between treated and control populations,
a crucial assumption for causal inference.

\begin{itemize}
    \item The raw features for both populations are drawn from a mixture of
          Gaussians:
          $\mathbb P(X) = p_A \mathbb P(X|A=1) + (1- p_A) \mathbb P(X|A=0)$
          where $\mathbb P(x|A=a)$ is a rotated Gaussian:
          \begin{equation}
              \mathbb P(x|A=a) = W \cdot \mathcal N \Big( \begin{bmatrix} (1-2a) \theta \\ 0\end{bmatrix} ; \begin{bmatrix} \sigma_0 & 0 \\ 0 & \sigma_1\end{bmatrix} \Big)
          \end{equation}
          with $\theta$ a parameter controlling overlap (bigger yields poorer
          overlap), $W$ a random rotation matrix and $\sigma_0^2=2;\sigma_1^2=5$.

          This generation process allows to analytically compute the oracle
          propensity scores $e(x)$, to simply control for overlap with the
          parameter $\theta$, the distance between the two Gaussian main axes and
          to  visualize response surfaces.

    \item A basis expansion of the raw features increases the problem dimension.
          Using Radial Basis Function (RBF) Nystroem transformation \footnote{We use the
              \href{https://scikit-learn.org/stable/modules/generated/sklearn.kernel_approximation.Nystroem.html}{Sklearn
                  implementation, \cite{pedregosa_scikitlearn_2011}}}, we expand the raw
          features into a transformed space. The basis expansion samples randomly a
          small number of representers in the raw data. Then,  it computes an
          approximation of the full N-dimensional kernel with these basis components,
          yielding the transformed features $z(x)$.

          We generate the basis following the original data distribution, $\left [
                  b_1 .. b_D \right ] \sim \mathbb P(x)$, with D=2 in our simulations. Then, we
          compute an approximation of the full kernel of the data generation
          process $RBF(x, \cdot) \;  with \; x \sim \mathbb P(x)$ with these
          representers: $z(x) = [RBF_{\gamma}(x, b_d)]_{d=1..D}
              \cdot Z^T \in \mathbb{R}^D$
          with $RBF_{\gamma}$ being the Gaussian kernel $K(x, y) = exp(-\gamma
              ||x-y||^2)$ and Z the normalization constant of the kernel basis,
          computed as the root inverse of the basis kernel $Z=[K(b_i, b_j)]_{i, j
              \in {1..D}}^{-1/2}$

    \item Functions $\mu_0$, $\tau$ are distinct linear functions of the
          transformed features:
          \begin{equation*}
              \mu_0(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\mu}^T
          \end{equation*}
          \begin{equation*}
              \tau(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\tau}^T
          \end{equation*}
    \item Adding a Gaussian noise, $\varepsilon \sim \mathcal N(0, \sigma(x;a))$,
          we construct the potential outcomes:
          $y(a) = \mu_0(x) + a\,\tau(x) + \varepsilon(x, a)$
\end{itemize}
We generated 1000 instances of this dataset with uniformly random overlap
parameters $\theta \in \left[ 0, 2.5 \right]$.

\subsection{Model selection procedures}

\paragraph{Nuisances estimation}\label{apd:experiments:nuisances_hp}

The nuisances are estimated with a stacked regressor inspired by the Super
Learner framework, \cite{laan_super_2007}). The hyper-parameters are
optimized with a random search with following search grid detailed in Table
\ref{apd:experiments:nuisances_hp_grid}. All implementations come from
\href{https://scikit-learn.org/stable/}{scikit-learn}
\cite{pedregosa_scikitlearn_2011}.

\begin{table}[h!]
    \begin{tabular}{llll}
        \toprule
        Model                & Estimator
                             & Hyper-parameters grid                                         \\
        \midrule
        Outcome, m           & StackedRegressor
                             & ridge regularization: [0.0001, 0.001, 0.01, 0.1, 1, 10, 100]  \\
        \multirow[c]{3}{*}{} & (HistGradientBoostingRegressor, ridge)
                             & HistGradientBoostingRegressor  learning rate: [0.01, 0.1, 1]  \\
                             &
                             & HistGradientBoostingRegressor  max leaf nodes: [10,
        20, 30, 50]                                                                          \\
        \midrule
        Treatment, e         & StackedClassifier
                             & LogisticRegression  C: [0.0001, 0.001, 0.01, 0.1, 1, 10, 100] \\
        \multirow[c]{3}{*}{} & (HistgradientBoostingClassifier, LogisticRegression)
                             & HistGradientBoostingClassifier  learning rate: [0.01, 0.1, 1] \\
                             &
                             & HistGradientBoostingClassifier  max leaf nodes: [10,
        20, 30, 50]                                                                          \\
        \bottomrule
    \end{tabular}
    \caption{Hyper-parameters grid used for nuisance models}
    \label{apd:experiments:nuisances_hp_grid}
\end{table}

\subsection{Additional Results}\label{apd:experiments:additional_results}

\paragraph{Definition of the Kendall's tau, $\kappa$}

The Kendall's tau  is a widely used statistics to measure the rank correlation
between two set of observations. It measures the number of concordant pairs
minus the discordant pairs normalized by the total number of pairs. It takes values in the
$[-1, 1]$ range.
\begin{equation}\label{eq:kendall_tau}
    \kappa=\frac{\text { (number of concordant pairs })-\text { (number of discordant pairs) }}{\text { (number of pairs) }}
\end{equation}

\paragraph{Values of relative $\kappa(\ell,\tau\mathrm{{-risk}})$ compared to
    the mean over all metrics Kendall's as shown in the boxplots of Figure \ref{fig:relative_kendalls_all_datasets}}

\begin{table}
    \centering
    \resizebox{0.7\textwidth}{!}{
        \input{images/_1_r_risk_domination_boxplot_df.tex}
    }
    \caption{Values of relative $\kappa(\ell,\tau\mathrm{{-risk}})$ compared to
        the mean over all metrics Kendall's as shown in the boxplots of Figure
        \ref{fig:relative_kendalls_all_datasets}}\label{apd:table:relative_kendalls_all_datasets}
\end{table}

\paragraph{Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} -
    Results measured in absolute Kendall's}

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.44\textwidth}
        \centering
        \caption{\textbf{Caussim}}
        \includegraphics[width=\textwidth]{images/kendalls_tau_caussim__nuisance_non_linear__candidates_ridge__overlap_01-247.pdf}
        \label{fig:ranking_agreement_w_tau_risk_caussim}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.44\textwidth}
        \centering
        \caption{\textbf{ACIC 2016}}
        \includegraphics[width=\textwidth]{images/kendalls_tau_acic_2016__nuisance_non_linear__candidates_hist_gradient_boosting__dgp_1-77__rs_1-10.pdf}
        \label{fig:ranking_agreement_tau_risk_acic_2016}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.10\textwidth}
        \centering
        \includegraphics[width=1.5\textwidth]{images/legend_metrics.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.44\textwidth}
        \centering
        \caption{\textbf{ACIC 2018}}
        \includegraphics[width=\textwidth]{images/kendalls_tau_acic_2018__nuisance_stacking__candidates_hist_gradient_boosting__tset_50.pdf}
        \label{fig:ranking_agreement_w_tau_risk_acic_2018}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.44\textwidth}
        \centering
        \caption{\textbf{TWINS}}
        \includegraphics[width=\textwidth]{images/kendalls_tau_twins__nuisance_stacking__candidates_hist_gradient_boosting__tset_50__overlap_11-296__rs_0-9_noise10.pdf}
        \label{fig:ranking_agreement_tau_risk_twins}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.10\textwidth}
        ~
    \end{subfigure}
    % \begin{subfigure}[b]{\textwidth}
    %   \centering
    %   \includegraphics[width=0.5\textwidth]{images/legend_metrics_horizontal.pdf}
    % \end{subfigure}
    \caption{Agreement with $\tau\text{-risk}$ ranking of methods function
        of overlap violation. The lines represent medians, estimated with a
        lowess. The transparent
        bands denote the 5\% and 95\% confidence intervals.}\label{apd:fig:all_datasets_tau_risk_ranking_agreement}
\end{figure}

\paragraph{Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}
    - Results measured as distance to the oracle tau-risk}

To see practical gain in term of $\tau\text{-risk}$, we plot the results as the
normalized distance between the estimator selected by the oracle
$\tau\text{-risk}$ and the estimator selected by each causal metric.

Then, $\widehat{R\text{-risk}}^*$ is more efficient than all other metrics. The
gain are substantial for every datasets.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.44\textwidth}
        \centering
        \caption{\textbf{Caussim}}
        \includegraphics[width=\textwidth]{images/normalized_bias_tau_risk_to_best_method_caussim__nuisance_non_linear__candidates_ridge__overlap_01-247.pdf}
        \label{fig:normalized_bias_tau_risk_to_best_method_caussim}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.44\textwidth}
        \centering
        \caption{\textbf{ACIC 2016}}
        \includegraphics[width=\textwidth]{images/normalized_bias_tau_risk_to_best_method_acic_2016__nuisance_non_linear__candidates_hist_gradient_boosting__dgp_1-77__rs_1-10.pdf}
        \label{fig:normalized_bias_tau_risk_to_best_method_acic_2016}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.10\textwidth}
        \centering
        \includegraphics[width=1.5\textwidth]{images/legend_metrics.pdf}
    \end{subfigure}
    \begin{subfigure}[b]{0.44\textwidth}
        \centering
        \caption{\textbf{ACIC 2018}}
        \includegraphics[width=\textwidth]{images/normalized_bias_tau_risk_to_best_method_acic_2018__nuisance_stacking__candidates_hist_gradient_boosting__tset_50.pdf}
        \label{fig:normalized_bias_tau_risk_to_best_method_acic_2018}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.44\textwidth}
        \centering
        \caption{\textbf{TWINS}}
        \includegraphics[width=\textwidth]{images/normalized_bias_tau_risk_to_best_method_twins__nuisance_stacking__candidates_hist_gradient_boosting__tset_50__overlap_11-296__rs_0-9_noise10.pdf}
        \label{fig:normalized_bias_tau_risk_to_best_method_twins}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.10\textwidth}
        ~
    \end{subfigure}
    \caption{Metric performances by normalized tau-risk distance to the best
        method selected with $\tau\text{-risk}$. All nuisances are learned with the
        same estimator stacking gradient boosting and ridge regression. Doted and
        plain lines corresponds to 60\% lowess quantile estimates. This choice of
        quantile allows to see better the oracle metrics lines for which outliers with a value
        of 0 distord the curves.}
    \label
    {apd:all_datasets_normalized_bias_tau_risk_to_best_method}
\end{figure}

\paragraph{Figure \ref{apd:fig:procedures_comparison_all_metrics} - Stacked models for the nuisances is more efficient}
For each metrics the benefit of
using a stacked model of linear and boosting estimators for nuisances compared
to a linear model. The evaluation measure is Kendall's tau relative to the
oracle $R\text{-risk}^{\star}$ to have a stable reference between exepriments.
Thus, we do not include in this analysis the ACIC 2018 dataset since
$R\text{-risk}^{\star}$ is not available due to the lack of the true propensity
score.

\begin{figure}
    \begin{subfigure}[b]{0.9\textwidth}
        %\centering
        \caption{\textbf{Caussim}}
        \includegraphics[width=1.15\textwidth]{images/_3_procedure_Caussim_N5000__ref_metric_oracle_r_risk_training_procedure.pdf}
        \label{fig:experiments:procedures_comparison:caussim}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.9\textwidth}
        \centering
        \caption{\textbf{ACIC 2016}}
        \includegraphics[width=1\textwidth]{images/_3_procedure_ACIC2016_N4802__ref_metric_oracle_r_risk_training_procedure.pdf}
        \label{fig:experiments:procedures_comparison:acic_2016}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.9\textwidth}
        \centering
        \caption{\textbf{Twins}}
        \includegraphics[width=1\textwidth]{images/_3_procedure_Twins_N11984__ref_metric_oracle_r_risk_training_procedure.pdf}
        \label{fig:experiments:procedures_comparison:twins}
    \end{subfigure}
    \hfill
    \caption{Results are similar between the \textcolor{MidnightBlue}{Shared
            nuisances/candidate set} and
        the \textcolor{RedOrange}{Separated nuisances set} procedure. The
        experience has not been run on the full metrics for Caussim due to computation costs.}\label
    {apd:fig:procedures_comparison_all_metrics}
\end{figure}

\paragraph{Figure \ref{apd:fig:all_datasets_overlap_effect} Low population
    overlap hinders model selection for all metrics}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset.pdf}
    \hfill
    \caption{\textbf{Low population overlap hinders causal model selection for all
            metrics}:
        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
        correspond
        to the tertiles of the overlap distribution measured with Normalized Total
        Varation eq. \ref{eq:ntv}.}\label{apd:fig:all_datasets_overlap_effect}
\end{figure}

\paragraph{Figure \ref{apd:fig:nuisances_comparison} - Stacked models for the nuisances is more efficient}
For each metrics the benefit of
using a stacked model of linear and boosting estimators for nuisances compared
to a linear model. The evaluation measure is Kendall's tau relative to the
oracle $R\text{-risk}^{\star}$ to have a stable reference between exepriments.
Thus, we do not include in this analysis the ACIC 2018 dataset since
$R\text{-risk}^{\star}$ is not available due to the lack of the true propensity
score.

\begin{figure}
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{\textbf{Twins}}
        \includegraphics[width=1\textwidth]{images/_4_nuisance_models_Twins_N11984__ref_metric_oracle_r_risk_nuisance_models.pdf}
        \label{fig:experiments:nuisance_comparison:twins}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{\textbf{Twins downsampled}}
        \includegraphics[width=1\textwidth]{images/_4_nuisance_models_Twinsdownsampled_N4794__ref_metric_oracle_r_risk_nuisance_models.pdf}
        \label{fig:experiments:nuisance_comparison:twins_ds}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        %\centering
        \caption{\textbf{Caussim}}
        \includegraphics[width=1.15\textwidth]{images/_4_nuisance_models_Caussim_N5000__ref_metric_oracle_r_risk_nuisance_models.pdf}
        \label{fig:experiments:nuisance_comparison:caussim}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.49\textwidth}
        \centering
        \caption{\textbf{ACIC 2016}}
        \includegraphics[width=1\textwidth]{images/_4_nuisance_models_ACIC2016_N4802__ref_metric_oracle_r_risk_nuisance_models.pdf}
        \label{fig:experiments:nuisance_comparison:acic_2016}
    \end{subfigure}
    \hfill
    \caption{Learning the nuisances with \textcolor{DarkOrchid}{stacked models} (linear and
        gradient boosting) is important for successful model selection with R-risk.
        For Twins dataset, there is no improvement for \textcolor{DarkOrchid}{stacked models} compared to
        \textcolor{ForestGreen}{linear models} because of the linearity of the propensity model.}\label
    {apd:fig:nuisances_comparison}
\end{figure}

\paragraph{Figure \ref{apd:fig:nuisances_comparison_twins} - Flexible models are performant in recovering nuisances even
    in linear setups}

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{images/_4_nuisance_models_r_risk_only_twins_nuisances.pdf}
    \hfill
    \caption{\textbf{Flexible models are performant in recovering nuisances
            in the downsampled Twins dataset.} The propensity score is linear in this
        setup, making it particularly challenging for flexible models compared to
        linear methods.}\label{apd:fig:nuisances_comparison_twins}
\end{figure}

\paragraph{Selecting different seeds and parameters is crucial to draw
    conclucions}\label{{apd:results:seed_effect}}

One strength of our study is the various number of different simulated and
semi-simulated datasets. We are convinced that the usual practice of using only
a small number of generation processes does not allow to draw statistically
significant conclusions.

Figure \ref{apd:results:fig:seed_effect} illustrate the dependence of the
results on the generation process for caussim simulations. We highlighted the
different trajectories induced by three different seeds for data generation and
three different treatment ratio instead of 1000 different seeds. The result
curves are relatively stable from one setup to another for $R{-risk}$, but vary
strongly for $\mu\text{-risk}$ and $\mu\text{-risk}_{IPW}$.

\begin{figure}
    \centering
    \caption{Kendall correlation coefficients for each causal metric. Each (color,
        shape) pair indicates a different (treatment ratio, seed) of the generation
        process.}\label {apd:results:fig:seed_effect}
    \includegraphics[width=\linewidth]{images/caussim_seed_effect.pdf}
\end{figure}

\FloatBarrier

\section{Heterogeneity in practices for data split}\label{apd:results:k_fold_choices}

Splitting the data is common when using machine learning for causal
inference, but practices vary widely in terms of the fraction of data to
allocate to train models, outcomes and nuisances, and to evaluate them.

Before even model selection, data splitting is often required for
estimation of the treatment effect, ATE or CATE, for instance to compute
the nuisances required to optimize the outcome model (as the
$R\text{-risk}$, definition \autoref{def:r_risk}).
%
The most frequent choice is use 80\% of the data to fit the models,
and 20\% to evaluate them.
%
For instance, for CATE estimation, the R-learner has been introduced using K-folds with K = 5
and K = 10: 80\% of the data (4 folds) to train the nuisances and the remaining
fold to minimize the corresponding R-loss \cite{nie_quasioracle_2017}.
Yet, it has been implemented with K=5 in causallib
\cite{causalevaluations} or K=3 in econML \cite{econml}.
Likewise, for ATE estimation, \citet{chernozhukov_double_2018}
introduce doubly-robust machine learning,
recommending K=5 based on an empirical comparison K=2. However,
subsequent works use doubly robust ML with varying choices
of K: \citet{loiseau_external_2022} use K=3, \citet{gao_assessment_2021} use
K=2. In the econML implementation, K is set to 3 \cite{econml}.
\citet{naimi2021challenges} evaluate various machine-learning approaches
--including R-learners-- using K=5 and 10, drawing inspiration from the
TMLE literature which sets
K=5 in the TMLE package \cite{tmle_package_2012}.

Causal model selection has been much less discussed. The only study that
we are aware of, \citet{schuler_comparison_2018}, use a different data
split: a 2-folds train/test procedure,
training the nuisances on the first half of the data, and using the
second half to estimate the $R\text{-risk}$ and select the best treatment
effect model.

% \paragraph{Simulations: naive reweighting of the $R \text{-risk}$}

% Applying a naive reweighting, $w(x, a)=\frac{1}{e(x)\big(1-e(x) \big )}$ to the
% $R \text{-risk}$ to recover the $\tau \text{-risk}$ in the first part of
% \ref{theory:prop:r_risk_rewrite} makes the residuals explode in case of noise as
% shown in Figure \ref{apd:simu_noised}.

% \begin{figure}[htbp]
%   \centering
%   \caption{Caussim simulations (500 repetitions):
%   $R\text{-risk}_{IPW^*_{naive}}$ (in green) is exploding}\label
%   {apd:simu_noised}
%   \includegraphics[width=0.7\linewidth]{images/noised_0.001_simu_best_normalized_abs_bias_ate_vs_mmd.pdf}
% \end{figure}

\end{document}

### 2023-11-16 how_to_select_causal_models.tex
wip: get back alst version of text

@@
 \setlength{\columnsep}{1cm}
+\usepackage{wrapfig}
+%\usepackage{wraptable}
 \usepackage{authblk}% for author affiliations
@@
 \usepackage{comment}
-%\usepackage{snaptodo}
+\setcounter{secnumdepth}{0}
 
@@
 \setlength\textwidth{177.8mm}
-
-% \snaptodoset{block rise=1em}
-% \snaptodoset{margin block/.style={font=\tiny}} \newcommand{\gv}[1]{%
-%     \snaptodo[margin block/.append style=green!50!black]{%
-% 	\sloppy\textbf{Gael}: #1}}
-
-% \newcommand{\md}[1]{\snaptodo[margin block/.append style=blue]{\textbf{Matt}:
-% #1}} \newcommand{\idea}[1]{\snaptodo[margin block/.append
-% style=red]{\textbf{Idea}: #1}}
 
@@
 \begin{abstract}
+    \textbf{Background:}
     As predictive models --\emph{eg} from machine learning-- give likely
@@
     Pandora box of model selection: which of these models yield the most
-    valid causal estimates?
-    Here we highlight that classic machine-learning model selection
-    does not select the best outcome models for causal inference.
-    Indeed, causal model selection should control both outcome errors
-    for each individual, treated or not
-    treated, whereas only one outcome is observed.
-    Theoretically, simple risks used in machine learning do not control
-    causal effects when treated and non-treated population differ too much.
-    More elaborate risks build proxies of the causal error using
-    ``nuisance'' re-weighting to compute it on the observed data. But
-    does computing these nuisance adds noise to model selection? Drawing
-    from an extensive empirical study, we outline a good causal
-    model-selection procedure: using the so-called
-    $R\text{-risk}$; using flexible estimators to
-    compute the nuisance models on the train set; and splitting out 10\% of
-    the data to compute risks.
+    valid causal estimates? Causal model selection should control both outcome errors for each
+    individual, treated or not treated, whereas only one outcome is observed.
+    Theoretically, simple risks used in machine learning do not control causal
+    effects when treated and non-treated population differ too much. \\
+    \textbf{Methods:}
+    More elaborate risks build proxies of the causal error using ``nuisance''
+    re-weighting to compute it on the observed data. But does computing these
+    nuisance adds noise to model selection? Drawing from an extensive empirical
+    study based on a simulation and three semi-simulated datasets based on real
+    covariates, we compare different metrics for causal model selection.\\
+    \textbf{Results:} Among all metrics the classical mean squared error (ie.
+    factual $\mu\mathrm{-risk}$) is worse and re-weighting it with propensity score
+    ($\mu\mathrm{-risk}_{IPW}$) does not bring much improvements. The
+    $R\text{-risk}$, which includes a model of mean outcome and propensity
+    scores, leads to the best performances.\\
+    \textbf{Conclusions:} We outline a good causal model-selection procedure
+    using the so-called $R\text{-risk}$; using flexible estimators to compute
+    the nuisance models on the train set; and splitting out 10\% of the data to
+    compute risks.
 \end{abstract}
@@
 
-    \subsection{Extending prediction to prescription requires causal model selection}
-
-    % Framing:
-    % - Increasing rich observational, ``real-life'' data.
-    %   - routinely collected health care
-    %   - convenient to build prognostic model (eg using claims or
-    %   administrative data
-    %   \cite{desai2020comparison} ) \cite{yurkovich2015systematic},
-    %   electronic health records \cite{fontana2019can}, encompassing notes and questionnaires
-    %   \cite{horng2017creating,simon2018predicting}
-    % - Motivates building increasingly complex predictive model, such as
-    %   prognostic models
-    % - Plethora of predictive models
-    % - From prediction to prescription There is often a desire to reason on intervention => causal inference is necessary
-
-    Increasingly rich data drives new predictive models. In health, new risks
-    or prognostic models leverage routinely-collected data, sometimes with
-    machine learning \cite{mooney2018bigdata}: predicting morbidity-related
-    outcomes from administrative data \cite{yurkovich2015systematic}, heart
-    failure from claims \cite{desai2020comparison}, sepsis from clinical
-    records \cite{horng2017creating}, suicide attempts from patient records
-    and questionnaires \cite{simon2018predicting}... Data may be difficult to
-    control and model, but claims of accurate prediction can be established
+    \subsection{Extending prediction to prescription needs causality}
+
+    Progress in machine learning brings predictive models to new health
+    data \cite{beam2018big,rajkomar2019machine}, with the promise of
+    precision medicine.
+    Automated analysis of
+    medical images is increasingly accurate, \emph{eg} for brain images,
+    \cite{khojaste2022deep,zhang2019radiological} or mammography
+    \cite{yala2019deep,shen2019deep,nassif2022breast}.
+    New prognostic models leverage routinely-collected
+    patient records
+    \cite{mooney2018bigdata}: predicting heart
+    failure from claims \cite{desai2020comparison}, suicide attempts
+    from questionnaires \cite{simon2018predicting}...
+    Clinical notes contain much prognostic information
+    but require text modeling  \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
+    Data may be difficult to
+    control and model, but the accuracy of the prediction can be verified
     on left-out data
     \cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
-    Given a model predicting of an outcome of interest, it is tempting to use
-    it to guide decisions: will an individual benefit or not from an
-    intervention such as surgery \cite{fontana2019can}? This is a
-    causal-inference task and principled approaches can build
-    on contrasting the
-    prediction of the outcome with and without the treatment
+    Given a model predicting a health outcome, precision medicine
+    would like it to guide decisions: will an individual benefit from an
+    intervention such as surgery \cite{fontana2019can}?
+    Contrasting predictions with and without the treatment gives an answer,
+    but statistical validity requires causal inference
     \cite{snowden_implementation_2011,blakely2020reflection}.
@@
 
-    Outcome modeling is an integral part of the causal modeling toolkit,
-    under the names of
-    G-computation, G-formula \cite{robins_role_1986}, Q-model
-    \cite{snowden_implementation_2011}, or conditional mean
-    regression \cite{wendling_comparing_2018}. On observational data, causal
-    inference of treatment effects is brittle to un-accounted for
-    confounding, but it can assess
-    effectiveness and safety in real-word practice
-    \cite{black1996we,hernan_methods_2021} or off-label drug usage
-    \cite{radley2006off} for potential repurposing
-    \cite{hurle2013computational,dudley2011exploiting}. Epidemiology has
-    historically focused on methods that model treatment assignment
-    \cite{austin_moving_2015,grose_use_2020}, based on the propensity score
-    \cite{rosenbaum_central_1983} but recent empirical results
-    \cite{wendling_comparing_2018,dorie_automated_2019} show benefits of
-    outcome modeling to estimate average treatment effect (ATE). A major
-    benefit of using outcome modeling for causal inference is that these
-    methods naturally go beyond average effects,
-    estimating individualized or conditional average treatment effects
-    (CATE), important for precision medicine.
+    There is a rich causal-inference statistical literature, which bridges to
+    predictive modeling via so-called \emph{outcome models}, or G-computation,
+    G-formula \cite{robins_role_1986}, Q-model \cite{snowden_implementation_2011},
+    conditional mean regression \cite{wendling_comparing_2018}. A central challenge
+    of inference of treatment effects is that of confounding: spurious associations
+    between treatment allocation and baseline health, \emph{eg} only prescribing a
+    drug to mild cases \citep{hernan_causal_2020,vanderweele2019principles}.
+    Controlled allocation of treatment, as in randomized controlled trials (RCTs),
+    alleviate this concern. Yet most machine-learning models are trained on
+    \emph{observational} data, close to real-word practice
+    \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
+    Causal inference has been central to epidemiology, typically with methods that
+    model treatment assignment \cite{austin_moving_2015,grose_use_2020}, based on
+    propensity scores \cite{rosenbaum_central_1983}. Recent empirical results
+    \cite{wendling_comparing_2018,dorie_automated_2019} show benefits of outcome
+    modeling to estimate average treatment effects. Maybe a greater benefit is
+    that these methods naturally go beyond average effects, estimating
+    individualized or conditional average treatment effects (CATE), central to
+    precision medicine.
     %
@@
     %\idea{There is a variety of models for g-estimation}
-    Recent developments have seen the multiplication of predictive modeling
-    methods. Leaving aside the, overwhelming, machine-learning literature,
-    even methods specifically designed for causal
+    There is an explosion of outcome modeling or machine learning methods.
+    For instance, many deep-learning methods have been developed for medical
+    image analysis \cite{shen2017deep,monshi2020deep}.
+    Even outcome-modeling methods specifically designed for causal
     inference are numerous: Bayesian Additive Regression Trees
@@
     \cite{powers_methods_2018}, causal multivariate adaptive regression
-    splines (MARS) \cite{powers_methods_2018}, random forests
+    splines \cite{powers_methods_2018}, random forests
     \cite{wager_estimation_2018, athey_generalized_2019},
@@
     %\idea{The variety of estimators calls for model selection procedures}
-    The wide variety of methods leaves the applied researcher with the difficult choice
+    The wide variety of methods raises the problem
     of selecting between different estimators based on the data at hand.
     %
-    Indeed, estimates may vary markedly when using different
-    models. For instance, Figure \ref{fig:acic_2016_ate_heterogeneity} shows
-    large variations obtained across four different outcome estimators on
-    2016 semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models
+    Indeed, estimates of treatment effects can vary markedly across different
+    predictive models. For instance, Figure shows
+    large variations obtained across different outcome estimators on
+    semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models
     such as random forests are doing well in most settings except
-    for setups where treated and untreated populations differ markedly in
-    which case a simple linear model (ridge) is to be preferred.
-    However a different choice of hyper-parameters
-    (max depth= 2) for random forests yield the poorest performances.
-    A simple rule of thumb such as preferring more flexible models does not work in
-    general; model selection is necessary.
+    when treated and untreated populations differ noticeably, in
+    which case a linear model (ridge) is to be preferred.
+    However random forests with different hyper-parameters
+    (max depth= 2) yield poor estimates.
+    A simple rule of thumb such as preferring flexible models does not work in
+    general; model selection is needed.
 
@@
 
-    \begin{figure}[b!]
-        \begin{minipage}{.4\linewidth}
+    \begin{figure}
+        \centering
+        \includegraphics[width=0.95\linewidth]{images/2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
         \caption{\textbf{Different outcome models lead to different
-                    estimation error on the Average Treatment Effects},
-                with six different outcome
-                models on 77 classic simulations where the true causal effect is
-                known \cite{dorie_automated_2019}. The models are random forests, ridge regression with and without
-                interaction with the treatment,
-                (hyper-parameters detailed in Appendix
+                estimation errors on the Average Treatment Effects},
+            on 77 classic simulations with known true causal effect
+            \cite{dorie_automated_2019}. The different models are ridge regression
+            and random forests with different hyper-parameters
+            (details
             \ref{apd:toy_example:acic_2016_ate_variability}). The different configurations are
             plotted as a function of increasing difference between treated and
-                untreated population --detailed in
-                \autoref{subsec:measuring_overlap}.  \\[1ex]
-                There is no systematic best performer: choosing
-                the best model among a family of candidate estimators is
-                important.
-            }\label{fig:acic_2016_ate_heterogeneity}%
-        \end{minipage}\hfill%
-        \begin{minipage}{.6\linewidth}
-            \centering
-            \includegraphics[width=0.95\linewidth]{images/2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
-        \end{minipage}%
+            untreated population --see
+            \autoref{subsec:measuring_overlap}.
+            There is no systematic best performer; data-driven model
+            selection is important.
+            \label{fig:acic_2016_ate_heterogeneity}%
+        }
     \end{figure}
@@
     Standard practices to select models in predictive settings rely on
-    cross-validation on the error on the outcome
+    the error on the outcome
     \cite{poldrack2020establishment,varoquaux2022evaluating}. However, as we
-    will see, these model-selection practices may not pick the best models
+    will see, these practices may not pick the best models
     for causal inference, as they can be misled by inhomogeneities due to
@@
     Given complex, potentially noisy, data, which model is to be most trusted to
-    yield valid causal estimates? Because there is no single learner that performs
+    yield valid causal estimates? As no single learner performs
     best on all data sets, there is a pressing need for clear guidelines to select
@@
 
-    In this paper, we study \textit{model selection procedures} with a focus
-    on practical settings: \textit{finite samples} settings and without
-    \textit{well-specification} assumption. One question is whether
-    model-selection procedures, that rely on data split, can estimate
-    reliably enough the complex risks, theoretically motivated for causal
-    inference. Indeed, these risks force of departure from standard model-selection
-    procedures as they come with more quantities to estimate, which may
+    In this paper, we study \textit{model selection procedures}
+    in practical settings: \textit{finite samples} settings and without
+    \textit{well-specification} assumption. Asymptotic causal-inference
+    theory calls for complex risks, but a practical question is
+    whether model-selection procedures, that rely on data split, can estimate
+    these risks reliably enough. Indeed, they
+    come with more quantities to estimate, which may
     bring additional variance, leading to worse model selection.
 
-    We first give a simple illustration of the problem of causal model
-    selection and quickly review the important prior art. Then, in Section
-    \ref{sec:framework}, we set causal model selection in the
-    \emph{potential outcome} framework and detail the causal risks and
-    model-selection procedure. Section \ref{sec:theory} gives our theoretical
-    result. In section \ref{sec:empirical_study}, we run a thorough empirical
-    study, with many different settings covered. Finally, we comment our
-    findings in Section \ref{sec:discussion}.
+    We first illustrate the problem of causal model
+    selection and briefly review prior art. Then, Section
+    \ref{sec:framework} sets causal model selection in the
+    \emph{potential outcome} framework and details the causal risks and
+    model-selection procedure. Section \ref{sec:theory} gives theoretical
+    results. Section \ref{sec:empirical_study} details a thorough empirical
+    study, covering many different settings. Finally, Section \ref{sec:discussion}
+    discusses the findings.
     %
@@
     inference with an adapted
-    cross-validation that estimate the so-called $R\text{-risk}$.
-    The $R\text{-risk}$
-    modulates observed prediction error to compensate for systematic
-    differences between treated and non-treated individuals. It
-    relies on the two \emph{nuisance} models,
+    cross-validation to estimate the so-called $R\text{-risk}$.
+    This risk compensates for systematic
+    differences between treated and non-treated individuals using
+    two \emph{nuisance} models,
     themselves estimated from data and thus imperfect; yet these
-    imperfections do not undermine the benefit of the $R\text{-risk}$.
-
-    % XXX: merge the two paragraphs, above and below
+    imperfections do not undermine the $R\text{-risk}$.
 
@@
     prediction of the outcome for a given individual with and without the
-    treatment --as detailed in \autoref{sec:neyman_rubin}.
+    treatment --as detailed in the Methods \autoref{sec:neyman_rubin}.
     %
@@
 
-    \section{Formal setting of causal inference and model selection}\label{sec:framework}
-
+    %\section{Formal setting of causal inference and model selection}\label{sec:framework}
+    \section{Methods}\label{sec:framework}
     \subsection{The Neyman-Rubin Potential Outcomes framework}%
@@
     \end{equation}
+    \begin{align}
+        \mu\text{-risk}(\mu) & = \mathbb E_{(Y, X, A) \sim p(Y;X;A)}[\big( Y - \mu_A(X)\big)^2] \\
+                             & = \int_{\mathcal X, \mathcal A}
+        \,\varepsilon(x,a)^2 p(a \mid x) \,p(x) \,dx\,da  \leq \sigma_B^{2}(0) + \sigma_B^{2}(1) \nonumber
+    \end{align}
     \begin{equation}
-        \mu\text{-risk}(\mu) = \mathbb E_{(Y, X, A) \sim p(Y;X;A)}[\big( Y - \mu_A(X)\big)^2] = \int_{\mathcal X, \mathcal A}
-        \,\varepsilon(x,a)^2 p(a \mid x) \,p(x) \,dx\,da  \leq \sigma_B^{2}(0) + \sigma_B^{2}(1)
-    \end{equation}
-    \hspace*{0.5em}
-    \begin{equation}
-        \mu\text{-risk}_{IPW}(\mu) = \sigma_B^{2}(0) + \sigma_B^{2}(1)  \quad \text{follows from Lemma \ref{apd:proofs:mu_risk_ipw_link_mu}}
+        \mu\text{-risk}_{IPW}(\mu) = \sigma_B^{2}(0) + \sigma_B^{2}(1)  \quad \text{from Lemma \ref{apd:proofs:mu_risk_ipw_link_mu}}
         %\notag
     \end{equation}
-    \hspace*{0.5em}
-    \begin{equation}
+    \begin{multline}
         R\text{-risk}(\mu) = \tilde{\sigma}_B^{2}(0) + \tilde{\sigma}_B^{2}(1)
-        \leq \sigma_B^{2}(0) + \sigma_B^{2}(1) \quad \text{follows directly from Proposition \ref{theory:prop:r_risk_rewrite}}%\notag
-    \end{equation}
+        \leq \sigma_B^{2}(0) + \sigma_B^{2}(1)  \\  \text{from Proposition \ref{theory:prop:r_risk_rewrite}}%\notag
+    \end{multline}
     Thus, differences between causal risks only matter in finite sample regimes.

### 2023-11-16 how_to_select_causal_models.tex
adapt images to two cols

@@
 %\documentclass[MYBIB,STIX1COL]{WileyNJD-v2}
-\documentclass[10pt]{article}
-
+\documentclass[10pt,twocolumn]{article}
+
+\usepackage[a4paper, left=15mm, right=15mm, top=20mm,]{geometry}
 % declared functions and additional packages
@@
 % associated arxiv class
-\usepackage{arxiv}
+%\usepackage{arxiv}
 \usepackage{multicol}
-\setlength{\columnsep}{1cm}
+\setlength{\columnsep}{0.5cm}
 \usepackage{wrapfig}
@@
 
-\setlength{\marginparwidth}{1.5cm}
-\setlength\textwidth{177.8mm}
+\setlength{\marginparwidth}{0.5cm}
+%\setlength\textwidth{177.8mm}
 
@@
 \newtheorem{proof}{Proof}
+%
+\def\keywordname{{\bfseries \emph{Keywords}}}%
+\def\keywords#1{\par\addvspace\medskipamount{\rightskip=0pt plus1cm
+\def\and{\ifhmode\unskip\nobreak\fi\ $\cdot$
+}\noindent\keywordname\enspace\ignorespaces#1\par}}
 
@@
 \affil[2]{Mission Data, Haute Autorité de Santé, Saint-Denis, France}
-\affil[*]{Corresponding author: matthieu.doutreligne@inria.fr}
+\affil[*]{Corresponding author: m.doutreligne@has-sante.fr}
 
@@
 
+\onecolumn
 \maketitle
-
 \begin{abstract}
@@
 
-\begin{multicols}{2}
-
+\twocolumn
 \section{Background}\label{sec:intro}
@@
 
-    \begin{figure}[b!]
-        \begin{minipage}{.4\linewidth}
+\begin{figure}[t!]
+    %\centering
+    {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
+            causal inference}}
+
+    \includegraphics[width=1\linewidth]{images/toy_random_forest_high_R2_high_tau_risk.pdf}%
+
+    {\sffamily\footnotesize\scalebox{1.03}{b) Linear model, worse average
+            prediction but better causal inference}}
+
+    \hfill%
+    \includegraphics[width=1\linewidth]{images/toy_tlinear_model_small_R2_small_tau_risk.pdf}%
+
     \caption[The best predictor may not estimate best causal
@@
         the best causal estimates; however computing this error is not
-                feasible: computing it requires access to unknown quantities:
+        feasible: it requires access to unknown quantities:
         $\tau(x)$. \\[1ex]
         While the random forest fits the data better than the linear model, it
-                gives worse causal inference because its error is very inhomogeneous between
-                the treated and untreated. The $R^2$ score does not capture this
+        gives worse causal inference because its error is inhomogeneous between
+        treated and untreated. The $R^2$ score does not capture this
         inhomogeneity.
-            }\label{fig:toy_example} \end{minipage}\hfill%
-        \begin{minipage}{.577\linewidth}
-            %\centering
-            {\sffamily\small a) Random forest, good average prediction but bad
-                causal inference}
-
-            \hfill%
-            \includegraphics[width=1\linewidth]{images/toy_random_forest_high_R2_high_tau_risk.pdf}%
-
-            {\sffamily\small b) Linear model, worse average prediction but better causal inference}
-
-            \hfill%
-            \includegraphics[width=1\linewidth]{images/toy_tlinear_model_small_R2_small_tau_risk.pdf}%
-        \end{minipage}%
+        \label{fig:toy_example}
+    }
 \end{figure}
@@
 prediction of the outcome for a given individual with and without the
-    treatment --as detailed in the Methods \autoref{sec:neyman_rubin}.
+treatment --as detailed in \autoref{sec:neyman_rubin}.
 %
-    Given various predictors of the outcome, we are interested in
-    selecting those that estimate best the treatment effect.
+Given various predictors of the outcome, which one should we use?
 %
 Standard predictive modeling or machine-learning practice selects the
-    best predictor, \emph{ie} the one that minimizes the expected error.
-    However, the best predictor may not be the best model to reason about
+predictor that minimizes the expected error.
+However, this predictor may not be the best model to reason about
 causal effects of an intervention, as we illustrate below.
 
-    Figure \ref{fig:toy_example} gives a toy example: the
-    outcome $Y \in [0, 1]$, the
-    probability of death, a binary treatment $A \in \{0, 1\}$ and a covariate
-    $X \in \mathbb R$ which summarizes the patient health status (eg. the
-    Charlson co-morbidity index \cite{charlson_new_1987}). We simulate a
-    situation for which the treatment is beneficial (decreases
-    mortality) for patients with high Charlson scores (bad health
-    status). On the contrary, the treatment has little effect for patients in good
-    condition (small Charlson scores).
+Figure \ref{fig:toy_example} gives a toy example: the probability $Y$ of
+an undesirable
+outcome (\emph{eg} death), a binary treatment $A \in \{0, 1\}$, and a covariate
+$X \in \mathbb R$ summarizing the patient health status (eg. the
+Charlson index \cite{charlson_new_1987}). We simulate a
+treatment beneficial (decreases $Y$) for patients with high Charlson scores (bad health
+status) but with little effect for patients in good
+condition (low Charlson scores).
 
 Figure \ref{fig:toy_example}a shows a random forest predictor with a
-    counter-intuitive behavior: it predicts well on average the outcome (as
-    measured by a regression $R^2$ score) but perform poorly to estimate
-    causal quantities: the average treatment effect $\tau$ (as visible via
-    the error $|\tau - \hat{\tau}|$) or the individual treatment
-    effect (the error $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$).
+counter-intuitive behavior: it predicts well on average the outcome (as measured
+by a regression $R^2$ score) but perform poorly to estimate causal quantities:
+the average treatment effect $\tau$ (as visible via the error $|\tau -
+    \hat{\tau}|$) or the conditional average treatment effect (the error
+$\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$, called CATE).
 %
@@
 
-    Intuitively, the problem is that causal estimation requires controlling
+The problem is that causal estimation requires controlling
 an error on both treated and non-treated outcome for the same individual:
@@
 weaker assumptions on the data, thus has higher $R^2$ score but is biased
-    by the treated population in the region with poor overlap region, leading
+by the treated population in the poor-overlap region, leading
 to bad causal estimates.
@@
 This toy example illustrates that the classic minimum Mean Square Error
-    criterion is not suited to choosing a model among a family of candidate
+criterion is not suited to choosing a model among candidate
 estimators for causal inference.
@@
 A natural way to select a predictive model for causal inference would be
-    an error measure between a causal quantity such as the conditional
-    average treatment effects (CATE) and models' estimate. But such error is
+an error measure between a causal quantity such as the CATE and models' estimate. But such error is
 not a ``feasible'' risk: it cannot be computed solely from observed data
@@
 reweighted MSE risks on the semi-synthetic ACIC 2016 datasets
-    \cite{dorie_automated_2019}, but did not include the $R\text{-risk}$ and
-    looked only at the agreement of the best selected model with the true
-    CATE risk --$\tau\text{-risk}(f)$ def.\, \ref{def:tau_risk}--, not on the
-    full ranking of methods compared to the true CATE. We complete these
+\cite{dorie_automated_2019}, but did not include the $R\text{-risk}$. We complete these
 prior empirical work by studying a wider variety of data generative
@@
 that are \emph{consistent}: select the best model in a family given
-    asymptotically large data. These work typically rely on introducing a
-    CATE estimator in the testing procedure. For instance matching
+asymptotically large data. These work rely on introducing a
+CATE estimator in the testing procedure: matching
 \citep{rolling_model_2014}, an IPW estimate
@@
 guarantees to hold, the test-set correction needs to converge to the
-    oracle: it needs to be flexible enough --or well-posed-- and asymptotic
-    data. From a practical perspective, knowing that such requirements are
-    met implies having a good CATE estimate, which amounts to having solved
-    the original problem of causal model selection. We study how causal
-    model-selection procedures behave outside of these settings.
+oracle: it needs to be flexible enough --well-posed-- and asymptotic
+data. From a practical perspective, meeting such requirements
+implies having a good CATE estimate, thus having solved
+the original problem of causal model selection.
 
@@
 
-    Much work in causal inference has focused on building procedures that
-    guarantee asymptotically consistent estimators. Targeted Machine Learning
-    Estimation (TMLE) \cite{laan_targeted_2011,schuler_targeted_2017} and
-    Double Machine Learning \cite{chernozhukov_double_2018} both provide
-    estimators for Average Treatment Effect combining flexible treatment and
-    outcome models. Here also, theories require asymptotic regimes and
+Much work in causal inference has focused on procedures that
+guarantee asymptotically consistent estimators, such as Targeted
+Machine Learning
+Estimation (TMLE) \cite{laan_targeted_2011,schuler_targeted_2017} or
+Double Machine Learning \cite{chernozhukov_double_2018}. Here also, theories require asymptotic regimes and
 models to be \textit{well-specified}.
 
-    By contrast, \citet{johansson_generalization_2021} studies causal
-    estimation without assuming that estimators are well specified. They derive an
-    upper bound on the oracle error to the CATE ($\tau\text{-risk}$) that involves
-    the error on the outcome and the similarity of the distributions between the
-    features of treated and control patients. However, they focus on using this
-    upper bound for estimation, and do not give insights on model selection. In
-    addition, for hyperparameter selection, they rely on a plugin estimate of the $\tau\text{-risk}$ built with
+By contrast, \citet{johansson2022generalization} studies causal estimation
+without assuming that estimators are well specified. They derive an upper bound
+on the oracle error to the CATE ($\tau\text{-risk}$) that involves the error on
+the outcome and the similarity of the distributions of treated and control
+patients. However, they use this upper bound for model optimization,
+and do not give insights on model selection. In addition, for hyperparameter
+selection, they rely on a plugin estimate of the $\tau\text{-risk}$ built with
 counterfactual nearest neighbors, which has been shown ineffective
@@
 The Neyman-Rubin Potential Outcomes framework
-    \cite{naimi2023defining,imbens_causal_2015} enables statistical
-    reasoning on causal treatment effects: Given an outcome $Y \in \mathbb R$ (eg.
-    mortality risk or hospitalization length), function of a binary treatment $A \in
-        \mathcal{A} = \{0, 1\}$ (eg.~a medical act, a drug administration), and baseline
+\cite{naimi2023defining,imbens_causal_2015} enables statistical reasoning on
+causal treatment effects: Given an outcome $Y \in \mathbb R$ (eg. mortality risk
+or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
+    \{0, 1\}$ (eg.~a medical procedure, a drug administration), and baseline
 covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
-    distribution,
-    $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However, we want to model the existence of potential observations
-    (unobserved ie. counterfactual) that correspond to a different treatment. Thus we want
-    quantities on the counterfactual distribution
-    $O^{*} = (Y(1), Y(0), X, A) \sim \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.
+distribution, $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However,
+we want to model the existence of potential observations (unobserved ie.
+counterfactual) that correspond to a different treatment. Thus we want
+quantities on the counterfactual distribution $O^{*} = (Y(1), Y(0), X, A) \sim
+    \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.
 
@@
 Average Treatment Effect
-    \begin{equation*}
-        \text{(ATE)}\qquad
+\begin{flalign*}
+    \text{ATE} &  &
     \tau \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal D^*}[Y(1) - Y(0)];
-    \end{equation*}
-    to model heterogeneity, the Conditional Average Treatment Effect
-    \begin{equation*}
-        \text{(CATE)}\qquad
+               &  &
+\end{flalign*}
+at the individual level, to model heterogeneity, the Conditional Average Treatment Effect
+\begin{flalign*}
+    \text{CATE} &  &
     \tau (x) \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal{D}^\star}[Y(1) - Y(0) | X=x].
-    \end{equation*}
+                &  &
+\end{flalign*}
 
@@
 
-    Some assumptions are necessary to assure identifiability of the causal estimands
+Assumptions are necessary for causal estimands to be
+identifiability
 in observational settings \cite{rubin_causal_2005}. We assume the usual
-    strong ignorability assumptions, composed of \emph{1)}
-    \emph{unconfoundedness} $\{Y(0),
-        Y(1) \} \indep A | X$, \emph{~2)} \emph{strong overlap} ie. every patient has a
+strong ignorability assumptions: \emph{1)}
+\emph{unconfoundedness} \mbox{$\{Y(0),
+        Y(1) \} \indep A | X$}, \emph{~2)} \emph{strong overlap} ie. every patient has a
 strictly positive probability to receive each treatment, \emph{3)}
-    \emph{consistency}, and \emph{4)} \emph{generalization} (detailed in Appendix
-    \ref{apd:causal_assumptions}). In this work, we investigate the
-    role of the overlap \cite{damour_overlap_2020}, which is testable with
-    data.
+\emph{consistency}, and \emph{4)} \emph{generalization} (detailed in
+\ref{apd:causal_assumptions}).
 
@@
 paper, the expectations will always be taken on the factual observed
-    distribution $\mathcal{D}$, and we will omit
-    to explicitly specify the distribution. This identification leads to outcome based estimators (ie.
-    g-computation estimators\cite{snowden_implementation_2011}), targeting the
+distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
+g-computation estimators \cite{snowden_implementation_2011}), targeting the
 ATE $\tau$ with outcome modeling:
-    \begin{equation}
-        \tau = \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x] = \mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
+\begin{eqnarray}
+    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x]
+    \notag
+    \\
+    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
     \label{eq:tau_population}
-    \end{equation}
-    This equation has two central quantities: the conditional expectancy
-    function of the outcome associated to specific covariates and
-    treatment or not is, often called the \emph{response function}:
-    \begin{equation}
-        \text{(Response function)}
-        \qquad
-        \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a].
-        \label{def:mu_a}
-    \end{equation}
+\end{eqnarray}
+This equation builds on two quantities: the conditional expectancy
+of the outcome given the covariates and either
+treatment or no no treatment, called \emph{response function}:
+\begin{flalign*}
+    \text{Response function}
+     &  &
+    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
+     &  &
+\end{flalign*}
 
@@
 
-    Similarly to the ATE, for the CATE, at the individual level:
+Similarly to the ATE, at the individual level, the CATE:
 \begin{equation}
@@
 \paragraph{Robinson decomposition}
-    Another decomposition of the outcome model plays in important role,
-    the R-decomposition
+The \emph{R-decomposition}
+of the outcome model plays an important role,
 \cite{robinson_rootnconsistent_1988}:
@@
 and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
-    \begin{align}
-        \text{(Conditional mean outcome)} \qquad  m(x) \myeq \; & \mathbb E_{Y \sim \mathcal{D}} [Y|X=x].
+\begin{flalign}
+    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
+    \mathcal{D}} [Y|X=x]            &                    &
     \label{def:m}
     \\
-        \text{(Propensity score)}\qquad
-        e(x) \myeq \;                                           & \mathbb P[A=1|X=x],
+    \text{Propensity score}         &                    &
+    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
     \label{def:propensity_score}
-    \end{align}
+\end{flalign}
 the outcome can be written
-    \begin{equation}\label{eq:r_decomposition}
-        \text{(R-decomposition)}\qquad y(a) = m(x) + \big( a - e(x) \big) \tau(x) + \varepsilon(x; a) \quad\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0
-    \end{equation}
+\begin{flalign}\label{eq:r_decomposition}
+    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
+    \tau(x) + \varepsilon(x; a)\notag &  &
+    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
+\end{flalign}
 $m$ and $e$ are often called
-    \emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown in
-    general.
-
-    %As noted by \cite{johansson_generalization_2021}, the machine learning
+\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown.
+
+%As noted by \cite{johansson2022generalization}, the machine learning
 %community often referred to the CATE by ITE, the Individual Treatment Effect.
@@
 
-    \subsection{Model-selection risks, oracle and feasible}\label{sec:problem:model_selection}
-
-    \paragraph{Causal model selection}\label{sec:problem:causal_selection}
-
-    We formalize model selection for causal estimation. Thanks to the g-formula
-    identification (\autoref{eq:mu_identification}), a given outcome model $f: \mathcal X
-        \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
-    domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
-    \ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
-    %
-    Let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
-    a family of such estimators. Our goal is to select the best candidate in this
-    family for the observed dataset $O$ using a risk of interest
-    $\ell$:
-    \begin{equation}
-        f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
-        \label{eq:causal_model_selection}
-    \end{equation}
-
-    We now detail possible risks $\ell$, risks useful for causal
-    model selection, and how to compute them.
-
-    \paragraph{The $\tau\text{-risk}$: an oracle error risk}\label{paragraph:oracle_metrics}
-    %
-    As we would like to target the CATE, the following
-    evaluation risk is natural:
-    \begin{definition}[$\tau\text{-risk}(f)$]\label{def:tau_risk} also called PEHE
-        \cite{schulam_reliable_2017, hill_bayesian_2011}:
-        \begin{equation*}\label{eq:tau_risk}
-            \tau\text{-risk}(f) = \mathbb E_{X\sim p(X)}[(\tau(X) - \hat \tau_f(X))^2]
-        \end{equation*}
-        its finite-sum version over the observed data:
-        \begin{equation*}
-            \widehat{\tau\text{-risk}}(f) = \sum_{x \in O} \big(\tau(x) - \hat \tau_f(x)\big)^2
-        \end{equation*}
-    \end{definition}
-
-    However these risks are not feasible because the oracles $\tau(x)$ are
-    not accessible, with the observed data $(Y, X, A) \sim \mathcal D$.
-
-    \paragraph{Feasible error risks}\label{paragraph:feasible_metrics}
-    \emph{Feasible} risks are based on the prediction error of the outcome model
-    and \emph{observable} quantities.
-
-    \begin{definition}[Factual $\mu\text{-risk}$]\label{def:mu_risk}
-        \cite{shalit_estimating_2017} This is the usual Mean Squared Error on
-        the target y. It is what is typically meant by ``generalization error'' in
-        supervised learning and estimated with cross-validation:
-        \begin{equation*}\label{eq:mu_risk}
-            \mu\text{-risk}(f)=\mathbb{E}_{(Y, X, A) \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]
-        \end{equation*}
-    \end{definition}
-
-    The following risks use the nuisances $e$
-    --propensity score, def \ref{def:propensity_score}-- and $m$ --conditional mean
-    outcome, def \ref{def:m}. We give the definitions as \textit{semi-oracles},
-    function of the true unknown nuisances, but later instantiate them with estimated
-    nuisances, noted $\big(\check e, \check m \big)$. Semi-oracles risks are
-    superscripted with the $^{\star}$ symbol.
-
-    \begin{definition}[$\mu\text{-risk}_{IPW}^{\star}$]\label{def:mu_ipw_risk}
-        \cite{vanderlaan_unified_2003} Let the inverse propensity weighting
-        function $w(x, a) = \frac{a}{e(x)} + \frac{1 - a}{1 - e(x)}$, we define the
-        semi-oracle Inverse Propensity Weighting risk,
-        \begin{equation*}\label{eq:mu_ipw_risk}
-            \mu\text{-risk}_{IPW}^{\star}(f) = \mathbb{E}_{(Y, X, A) \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big) (Y-f(X ; A))^2 \right]
-        \end{equation*}
-    \end{definition}
-
-    \smallskip
-
-    \begin{definition}[$\tau\text{-risk}^{\star}_{IPW}$]\label{def:tau_ipw_risk}
-        \cite{wager_estimation_2018}
-        The CATE $\tau(x)$ can be estimated propensity score, with a
-        regression against inverse propensity weighted outcomes \cite{
-            athey2016recursive,gutierrez_causal_2016,wager_estimation_2018}. From this objective, we can derive
-        the $\tau\text{-risk}_{IPW}$.
-        \begin{equation*}
-            \tau\text{-risk}^{\star}_{IPW}(f) =\mathbb{E}_{(Y, X, A) \sim
-                \mathcal D} \left[ \Big(Y \frac{A - e(X)}{e(X)
-                    (1-e(X))}-\tau_f\left(X\right)\Big)^2 \right]  =\mathbb{E}_{(Y, X, A)
-                \sim \mathcal D} \left[ \left(Y \left( \frac{A}{e(X)} -
-                \frac{1-A}{1-e(X)}\right)-\tau_f\left(X\right)\right)^2 \right]
-        \end{equation*}
-    \end{definition}
-
-    \begin{definition}[$U\text{-risk}^{\star}$]\label{def:u_risk}
-        \cite{kunzel_metalearners_2019,nie_quasioracle_2017} Based on
-        the Robinson decomposition --eq. \ref{eq:r_decomposition}
-        \cite{robinson_rootnconsistent_1988}, the U-learner
-        uses the $A-e(X)$ term
-        in the denominator. The derived risk is:
-        \begin{equation*}
-            U\text{-risk}^{\star}(f) =\mathbb{E}_{(Y, X, A) \sim \mathcal D}
-            \left[
-                \left( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -
-                \tau_f\left(X\right)\right)^{2} \right]
-        \end{equation*}
-        Note that extreme propensity weights in the
-        denominator term might inflate errors in the numerator due to imperfect
-        estimation of the mean outcome $m$
-        the numerator errors, leading to a highly biased metric.
-    \end{definition}
-
-    \begin{definition}[$R\text{-risk}^{\star}$]\label{def:r_risk}
-        \cite{nie_quasioracle_2017,schuler_comparison_2018}
-        The $R\text{-risk}$ also uses two nuisance $m$ and $e$:
-        \begin{equation*}
-            R\text{-risk}^{\star}(f) =\mathbb{E}_{(Y, X, A) \sim \mathcal D} \big[
-                \big(\left(Y-m\left(X\right)\right) -\left(A-e\left(X\right)\right) \tau_f\left(X\right)\big)^{2} \big]
-        \end{equation*}
-    \end{definition}
-
-    It is also motivated by the Robinson decomposition --eq. \ref{eq:r_decomposition}
-    \cite{robinson_rootnconsistent_1988}. It performs well in various simulations
-    where using lasso, boosting or kernel ridge regressions for both the nuisances
-    $(\check e, \check m)$ and the target $\hat \tau(x)$
-    \cite{nie_quasioracle_2017}.
-
-    These risks are summarized in Table \ref{tab:evaluation_metrics}.
-
-    \begin{table}[!h]
-        \makebox[\linewidth]{
+\begin{table*}[!tb]
+    \makebox[\textwidth]{
         \begin{threeparttable}[b]
-                \caption{Review of causal risks}
-                \centering
-                \label{tab:evaluation_metrics}
+            \caption{Review of causal risks
+                ---
+                The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
+                \citet{schuler_comparison_2018}.
+                \label{tab:evaluation_metrics}}
+            \centering
             \begin{tabular}{llr}
@@
                 \\
-                    $R\text{-risk}^*$ \tnote{1}                                                                    & $\mathbb{E}_{(Y, X, A)
+                $R\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A)
                         \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
@@
             \end{tabular}
-                \begin{tablenotes}
-                    \item [1] Called $\tau \text{-risk}_R$ in
-                    \citet{schuler_comparison_2018}.
-                \end{tablenotes}
         \end{threeparttable}
     }
-    \end{table}
+\end{table*}
+
+\subsection{Model-selection risks, oracle and feasible}\label{sec:problem:model_selection}
+
+\paragraph{Causal model selection}\label{sec:problem:causal_selection}
+
+We formalize model selection for causal estimation. Thanks to the g-formula
+identification (\autoref{eq:mu_identification}), a given outcome model $f: \mathcal X
+    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
+domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
+\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
+%
+Let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
+a family of such estimators. Our goal is to select the best candidate in this
+family for the observed dataset $O$ using a risk
+$\ell$:
+\begin{equation}
+    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
+    \label{eq:causal_model_selection}
+\end{equation}
+
+We now detail possible risks $\ell$, risks useful for causal
+model selection, and how to compute them.
+
+\paragraph{The $\tau\text{-risk}$: an oracle error risk}\label{paragraph:oracle_metrics}
+%
+As we would like to target the CATE, the following
+evaluation risk is natural:
+\begin{definition}[$\tau\text{-risk}(f)$]\label{def:tau_risk} also called PEHE
+    \cite{schulam_reliable_2017, hill_bayesian_2011}:
+    \begin{equation*}\label{eq:tau_risk}
+        \tau\text{-risk}(f) = \mathbb E_{X\sim p(X)}[(\tau(X) - \hat \tau_f(X))^2]
+    \end{equation*}
+\end{definition}
+
+Given observed data from $p(X)$, the expectation is computed with a
+finite sum, as in eq.\,\ref{eq:ate_estimate}, to give an estimated
+value $\widehat{\tau\text{-risk}}(f)$.
+%
+However this risk is not feasible as the oracles $\tau(x)$ are
+not accessible with the observed data $(Y, X, A) \sim \mathcal D$.
+
+\paragraph{Feasible error risks}\label{paragraph:feasible_metrics}
+\emph{Feasible} risks are based on the prediction error of the outcome model
+and \emph{observable} quantities.
+
+All expectations below are on observed distribution:
+$(Y, X, A) \sim \mathcal D$.
+
+\begin{definition}[Factual $\mu\text{-risk}$]\label{def:mu_risk}
+    \cite{shalit_estimating_2017} This is the usual Mean Squared Error on
+    the target y. It is what is typically meant by ``generalization error'' in
+    supervised learning:
+    \begin{equation*}\label{eq:mu_risk}
+        \mu\text{-risk}(f)=\mathbb{E}\left[(Y-f(X ; A))^2 \right]
+    \end{equation*}
+\end{definition}
+
+We now detail risks that use the nuisances $e$
+--propensity score, def \ref{def:propensity_score}-- and $m$ --conditional mean
+outcome, def \ref{def:m}. We give the definitions as \textit{semi-oracles},
+function of the true unknown nuisances, but later instantiate them with estimated
+nuisances, noted $\big(\check e, \check m \big)$. Semi-oracles risks are
+superscripted with the $^{\star}$ symbol.
+
+\begin{definition}[$\mu\text{-risk}_{IPW}^{\star}$]\label{def:mu_ipw_risk}
+    \cite{vanderlaan_unified_2003} Let the inverse propensity weighting
+    function $w(x, a) = \frac{a}{e(x)} + \frac{1 - a}{1 - e(x)}$, we define the
+    semi-oracle Inverse Propensity Weighting risk,
+    \begin{equation*}\label{eq:mu_ipw_risk}
+        \mu\text{-risk}_{IPW}^{\star}(f) = \mathbb{E}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big) (Y-f(X ; A))^2 \right]
+    \end{equation*}
+\end{definition}
+
+\smallskip
+
+\begin{definition}[$\tau\text{-risk}^{\star}_{IPW}$]\label{def:tau_ipw_risk}
+    \cite{wager_estimation_2018}
+    The CATE $\tau(x)$ can be estimated with a
+    regression against inverse propensity weighted outcomes \cite{
+        athey2016recursive,gutierrez_causal_2016,wager_estimation_2018},
+    the $\tau\text{-risk}_{IPW}$.
+    \begin{equation*}
+        \tau\text{-risk}^{\star}_{IPW}(f) =\mathbb{E}
+        \left[ \Big(Y \frac{A - e(X)}{e(X)
+                (1-e(X))}-\tau_f\left(X\right)\Big)^2 \right]
+        %  =\mathbb{E}_{(Y, X, A)
+        %            \sim \mathcal D} \left[ \left(Y \left( \frac{A}{e(X)} -
+        %            \frac{1-A}{1-e(X)}\right)-\tau_f\left(X\right)\right)^2 \right]
+    \end{equation*}
+\end{definition}
+
+\begin{definition}[$U\text{-risk}^{\star}$]\label{def:u_risk}
+    \cite{kunzel_metalearners_2019,nie_quasioracle_2017} Based on
+    the Robinson decomposition --eq. \ref{eq:r_decomposition}, the U-learner
+    uses the $A-e(X)$ term
+    in the denominator. The derived risk is:
+    \begin{equation*}
+        U\text{-risk}^{\star}(f) =\mathbb{E}
+        \left[
+            \left( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -
+            \tau_f\left(X\right)\right)^{2} \right]
+    \end{equation*}
+    Note that extreme propensity weights in the
+    denominator term might inflate errors in the numerator due to imperfect
+    estimation of the mean outcome $m$.
+\end{definition}
+
+\begin{definition}[$R\text{-risk}^{\star}$]\label{def:r_risk}
+    \cite{nie_quasioracle_2017,schuler_comparison_2018}
+    The $R\text{-risk}$ also uses two nuisance $m$ and $e$:
+    \begin{equation*}
+        R\text{-risk}^{\star}(f) =\mathbb{E} \big[
+            \big(\left(Y-m\left(X\right)\right) -\left(A-e\left(X\right)\right) \tau_f\left(X\right)\big)^{2} \big]
+    \end{equation*}
+\end{definition}
+
+It is also based on the Robinson decomposition --eq. \ref{eq:r_decomposition}. %It performs well in various
+%simulations, estimating the nuisances
+%$(\check e, \check m)$ with
+%lasso, boosting or kernel ridge regressions
+%\cite{nie_quasioracle_2017}.
+
+These risks are summarized in Table \ref{tab:evaluation_metrics}.
 
@@
 \begin{algorithm}[!htbp]
-        \caption{Evaluation of selection procedures for one
-            simulation}\label{problem:estimation_procedure:algo} {%
-            Given a train and a test sets $(\mathcal{T}, \mathcal{S}) \sim \mathcal{D}$,
-            a family of candidate estimators $\{f \in \mathcal F\}$, a set of causal
-            metrics $\ell \in \mathcal L$:
+    \caption{Model selection procedure}\label{problem:estimation_procedure:algo} {%
+        Given train and test sets $(\mathcal{T}, \mathcal{S}) \sim \mathcal{D}$,
+        a candidate estimators $f$, a causal
+        metrics $\ell$:
         \begin{enumerate}
             \item Prefit: Learn estimators for unknown nuisance quantities $(\check e,\,\check m)$ on the training set $\mathcal{T}$
-                \item Fit: $\forall f \in \mathcal{F}$ learn $\hat f(\cdot, a)$ on
+            \item Fit: learn $\hat f(\cdot, a)$ on
                   $\mathcal T$
-                \item Model selection: $\forall{x} \in \mathcal{S}$ predict $\big(\hat f(x, 1), \hat f(x, 0)\big)$ and evaluate each candidate estimator with
-                      each causal metric $\mathcal M(\hat f, \mathcal{S})$. For each causal
-                      metric $\ell  \in \mathcal{L}$ and each candidate estimator $f \in
-                          \mathcal{F}$, store the metric value: $\ell(f, \mathcal S)$ -- possibly
+            \item Model selection: $\forall{x} \in \mathcal{S}$ predict
+                  $\big(\hat f(x, 1), \hat f(x, 0)\big)$ and evaluate the estimator storing the metric value: $\ell(f, \mathcal S)$ -- possibly
                   function of $\check e$ and $\check m$
@@
 \begin{figure}[h!]
-        \centering
-        \includegraphics[width=0.7\linewidth]{images/estimation_procedure_causal_selection_procedure.png}
+    \includegraphics[width=\linewidth]{images/estimation_procedure_causal_selection_procedure.png}
     \caption{Estimation procedure for causal model
@@
 These bounds depend on a specific form of residual that we now define: for each potential outcome, $a \in  \{0; 1\}$, the variance conditionally on $x$
-    is\cite{shalit_estimating_2017}:
+is \cite{shalit_estimating_2017}:
 \begin{equation*}\label{eq:residuals}
@@
     ]\label{theory:prop:mu_risk_ipw_bound}
-        \cite{johansson_generalization_2021} Given an outcome model $f$, let a
-        weighting function $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$ as the
+    \cite{johansson2022generalization} Given an outcome model $f$, let a
+    weighting function\\ $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$ as the
     Inverse Propensity Weight. Then, under overlap (assumption
@@
 This result has been derived in previous work
-    \cite{johansson_generalization_2021}. It links $\mu\text{-risk}_{IPW}$ to
-    the squared residuals of each population thanks to a reweighted mean-variance
-    decomposition. For completeness, we provide the proof in .
+\cite{johansson2022generalization}. It links $\mu\text{-risk}_{IPW}$ to
+the squared residuals of each population. For completeness, we provide the proof in \ref{apd:proofs}.
 
 The upper-bound comes from the triangular inequality applied to the residuals of
-    both populations. Interestingly, the two quantities are equal when the
+both populations. The two quantities are equal when the
 absolute residuals on treated and untreated populations are equal on the
-    whole covariate space, \emph{ie} for all
-    $x \in \mathcal X, |\mu_1(x) - f(x, 1)| = |\mu_0(x) - f(x, 0)|$.
-    The main source of difference between the oracle $\tau \text{-risk}$ and the
+whole covariate space:
+$\forall x \in \mathcal X, |\mu_1(x) - f(x, 1)| = |\mu_0(x) - f(x, 0)|$.
+The main difference between the oracle $\tau \text{-risk}$ and the
 reweighted mean squared error, $\mu\text{-risk}_{IPW}$, comes from heterogeneous
-    residuals between populations. These quantities are difficult to characterize as
-    they are linked both to the estimator and to the data distribution.
+residuals between populations.
+% These quantities are difficult to characterize as
+%they are linked both to the estimator and to the data distribution.
 %
-    This bound indicates that minimizing the $\mu\text{-risk}_{IPW}$ helps to
+This bound shows that minimizing the $\mu\text{-risk}_{IPW}$ helps to
 minimize the $\tau\text{-risk}$, which leads to
-    interesting optimization procedures \cite{johansson_generalization_2021}. However, there is no
-    guarantee that this bound is tight, which makes it less useful for model
+interesting optimization procedures \cite{johansson2022generalization}. However, there is no
+guarantee that this bound is tight, which makes it fragile for model
 selection.
@@
 Assuming strict overlap (probability of all individuals being treated or not
-    bounded away from 0 and 1 by $\eta$, appendix \ref{apd:causal_assumptions}), the
+bounded away from 0 and 1 by $\eta$, \ref{apd:causal_assumptions}), the
 above bound simplifies into a looser one involving the usual mean squared error:
@@
     $f$, its $R\text{-risk}$ appears as weighted version of its $\tau\text{-risk}$
-        (Proof in Appendix \ref{apd:proofs:r_risk_rewrite}):
-        \begin{equation}
-            R\text{-risk}^*(f) = \int_{x} e(x)\big(1-e(x)\big)\big(\tau(x)-\tau_ {f}(x)\big)^{2} p(x) d x \;+\; \tilde{\sigma}_B^2(1) \;+\; \tilde{\sigma}_B^{2}(0)
-        \end{equation}
+    (Proof in \ref{apd:proofs:r_risk_rewrite}):
+    \begin{align}
+        R\text{-risk}^*(f) & = \int_{x} e(x)\big(1-e(x)\big)\big(\tau(x)-\tau_ {f}(x)\big)^{2} p(x) d x \nonumber \\
+                           & \quad\; + \tilde{\sigma}_B^2(1) + \tilde{\sigma}_B^{2}(0)
+    \end{align}
 \end{proposition}
@@
 $\tau \text{-risk}$. On the contrary, for units with extreme overlap violation,
-    these weights goes down to zero with the propensity score.
+these weights go down to zero with the propensity score.
 
@@
 \end{equation*}
-    However, even if we have randomization, we still can have large differences
+However, we still can have large differences
 between $\tau\text{-risk}$ and $\mu\text{-risk}_{IPW}$ coming from heterogeneous
 errors between populations as noted in Section \ref{theory:mu_risk_ipw_bound}
-    and shown experimentally in simulations \cite{schuler_comparison_2018}.
+and shown experimentally in \citet{schuler_comparison_2018} and our
+results below.
 
@@
 $\tau\text{-risk}$ up to multiplicative and additive constants:
-    \begin{equation}
+\begin{equation*}
     R\text{-risk} = p_A \, (1-p_A) \, \tau\text{-risk} \;+\; (1 - p_A) \,\sigma_B^2(0) \;+\; p_A \sigma_B^2(1)
-    \end{equation}
-    Therefore, optimizing estimators for CATE with $R\text{-risk}^*$ in the
-    randomized setting is optimal if we target the $\tau\text{-risk}$. This explains
+\end{equation*}
+Thus, selecting estimators with $R\text{-risk}^*$ in
+randomized setting controls the $\tau\text{-risk}$. This explains
 the strong performances of $R\text{-risk}$ in randomized setups
-    \cite{schuler_comparison_2018} and is a strong argument in favor of this
-    risk for heterogeneity estimation in RCTs.
-
-    \paragraph{Oracle Bayes predictor}\label{remark:bayes_oracle} Consider the case
-    where we have access to the oracle Bayes predictor for the outcome ie.~$f(x,
+\cite{schuler_comparison_2018} and is a strong argument to use it
+to estimate heterogeneity in RCTs.
+
+\paragraph{Oracle Bayes predictor}\label{remark:bayes_oracle} If we
+have access to the oracle Bayes predictor for the outcome ie.~$f(x,
     a)=\mu(x, a)$, then all risks are equivalent up to the residual variance:
@@
 \end{multline}
+
 Thus, differences between causal risks only matter in finite sample regimes.
 Universally consistent learners converge to the Bayes risk in asymptotic
-    regimes, making all model selection risks equivalent. However, in practice
+regimes, making all model selection risks equivalent. In practice however,
 choices must be made in non-asymptotic regimes.
@@
 versions, presented in Table
-    \ref{tab:evaluation_metrics}:
+\ref{tab:evaluation_metrics}:\\
 $\widehat{\mu\text{-risk}}_{IPW}^*$,
@@
 processes and three semi-simulated datasets \footnote{Scripts for the simulations and the selection procedure are available at
-        \url{https://github.com/strayMat/caussim}.
+    \url{https://github.com/soda-inria/caussim}.
 }.
@@
 
-    \paragraph{Data Generation Process}
+\paragraph{Data Generation}
 
 We use simulated data, on which the ground-truth causal effect is known. Going
-    further than prior empirical studies of causal model selection
-    \cite{schuler_comparison_2018,alaa_validating_2019}, we use multiple
+beyond prior empirical studies of causal model selection
+\cite{schuler_comparison_2018,alaa_validating_2019}, we use many
 generative processes, to reach more general conclusions (as discussed
-    in Appendix \ref{apd:results:fig:seed_effect}).
-
-    We generate the response functions using random bases.
-    Basis extension methods are common in biostatistics, \emph{eg} functional
-    regression with splines
-    \cite{howe_splines_2011, perperoglou_review_2019}. By allowing the
-    function to vary at specific knots, they give flexible --non-linear--
-    models of the studied mechanisms. Taking inspiration from splines, we use
-    random approximation of Radial Basis Function (RBF) kernels
-    \cite{rahimi_random_2008} to generate the response surfaces. RBF use the
-    same process as polynomial splines but replace polynomial by Gaussian
-    kernels. Unlike polynomials, Gaussian kernels have
-    decreasing influences in the input space.
-    This avoids unrealistic divergences of the population response surfaces
-    at the ends of the feature space.
+in \ref{apd:results:fig:seed_effect}).
+
+We generate the response functions using random bases. Basis extension methods
+are common in biostatistics, \emph{eg} functional regression with splines
+\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
+vary at specific knots, they give flexible non-linear models. We use random approximation of Radial Basis Function (RBF) kernels
+\cite{rahimi_random_2008} to generate the response functions. RBF use the same
+process as polynomial splines but replace polynomial by Gaussian kernels. Unlike
+polynomials, Gaussian kernels have decreasing influences in the input space.
+This avoids unrealistic divergences of the response surfaces at the
+ends of the feature space.
 
@@
 the ground-truth response surfaces and treatment. We first use this process to
-    draw the non-treated response surface $\mu_0$ and the causal-effect $\tau$. We
+draw the non-treated response surface $\mu_0$ and the causal effect $\tau$. We
 then draw the observations from a mixture two Gaussians, for the treated and non
-    treated. We vary the separation between the two Gaussians to control the amount
-    of overlap between treated and control populations, as it an important parameter
-    for causal inference (related to $\eta$ which appears in section
-    \ref{theory:mu_risk_ipw_bound}). Finally, we generate the observed outcomes
-    adding Gaussian noise. We generated such datasets 1000 times, with
-    uniformly random overlap parameters $\theta \in \left[ 0, 2.5 \right]$. Appendix
-    \ref{apd:experiments:generation} gives more details on the data generation.
+treated. We vary the separation between the two Gaussians to control the
+overlap between treated and non-treated populations, an important parameter
+for causal inference (related to $\eta$ in section
+\ref{theory:mu_risk_ipw_bound}). Finally, we generate observed outcomes
+adding Gaussian noise. We generate 1\,000 of such datasets, with
+uniformly random overlap parameters. Details in
+\ref{apd:experiments:generation}.
 
@@
 \begin{figure}[!t]
-        \begin{minipage}{0.48\linewidth}
-            \centering
-            %\makebox[(\textwidth-9mm)/2]{\textbf{a. Caussim}}
-            \includegraphics[width=0.98\linewidth,clip,trim={0 15.3cm 0 0}]{images/caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
-        \end{minipage}
-        \hfill
-        \begin{minipage}{0.48\linewidth}
-            \hfill%
-            \includegraphics[width=0.98\linewidth,clip,trim={0 0 0 17cm}]{images/caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
-        \end{minipage}
+    \centering
+    \includegraphics[width=\linewidth]{images/caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
+
     \caption{Example of the simulation setup in the input space with two
-            knots --\emph{ie.}basis functions. The left panel gives
-            views of the observations in feature space, while the right panel displays the
+        knots --\emph{ie.}basis functions. The top panel
+        shows the observations in feature space, while the bottom panel displays the
         two response surfaces on a 1D cut along the black lines drawn on
-            the left panel.}
+        the top panel.}
     \label{fig:simulation_examples}
@@
 approximate imperfectly the data-generating process. To build such an
-    estimator in two steps, we first use a RBF expansion
-    similar as the one used for the data-generation generation process. Concretely,
-    we choose two random knots and apply a transformation of the raw data features
-    with the same Gaussian kernel used for the data-generation mechanism. This step
+estimator, we first use a RBF expansion
+similar to the one used for data generation.
+We choose two random knots and apply a transformation of the raw data features
+with a Gaussian kernel. This step
 is referred as the featurization. Then, we fit a linear regression on
@@
 transformed features. We consider two ways of combining these steps for outcome
-    mode; using common nomenclature \cite{kunzel_metalearners_2019}, we refer
-    to these regression structures as different meta-learners which differ on how
+mode; we use common nomenclature \cite{kunzel_metalearners_2019,shen2023rctrep} to refer
+to these different meta-learners that differ on how
 they model, jointly or not, the treated and the non treated:
@@
           the non treated, then two
-              separate different regressors are fitted on this representation.
+          separate different regressors are fitted on this shared representation.
     \item TLearner: Two completely different learners for each population, hence
-              separate featurization and separate regressors.
+          separate feature representations and regressors.
 \end{itemize}
@@
 \cite{kunzel_metalearners_2019}. Our goal is not to have the best possible
-    learner but to have a variety of sub-optimal learners in order to compare the
+learner but to have a variety of sub-optimal learners to compare the
 different causal metrics. For the same reason, we did not include more powerful
@@
         10^{-2}, 10^{-1}, 1, 10^{1}, 10^{2}]$, coupled with a TLearner or a SftLearner.
-    We sample 10 different random basis for the learning procedure and the
+We sample 10 different random basis for learning and
 featurization yielding a family $\mathcal F$ of 120 candidate estimators.
@@
 
-    We also use semi-simulated datasets, where a known synthetic causal effect is
-    added to real --non synthetic-- covariate. We study datasets used in previous work
-    to evaluate causal inference:
+We also use classic benchmarks of the causal-inference
+literature, semi-simulated data adding a known synthetic causal effect to real --non synthetic-- covariate:
 \begin{description}
@@
         Collaborative Perinatal Project \cite{niswander_women_1972}, a RCT
-            conducted on a cohort of pregnant women to identify causes of infants’
+        studying infants’
         developmental disorders. The initial intervention was a child’s birth
         weight $(A = 1 \text{ if weight} < 2.5 kg)$, and outcome was the child’s
-            IQ after a given follow-up period. The study contained $N=4\,802$ data
+        IQ after a follow-up period. The study contained $N=4\,802$ data
         points with $D=55$ features (5 binary, 27 count data, and 23
-            continuous). They simulated 77 different setups with varying parameters
-            for treatment and response generation models, treatment assignment
-            probabilities, overlap, and interactions between treatment and
+        continuous). They simulated 77 different setups varying parameters
+        for treatment and response models, overlap, and interactions between treatment and
         covariates \footnote{Original R code available at
-                \href{https://github.com/vdorie/aciccomp/tree/master/2016}{https://github.com/vdorie/aciccomp/tree/master/2016}
+            \url{https://github.com/vdorie/aciccomp/tree/master/2016}
             to generate 77 simulations settings.}. We used 10 different seeds for
-            every setup, totalizing 770 dataset instances.
-
-        \item[ACIC 2018] \cite{shimoni_benchmarking_2018}: The raw covariates
-            data comes from the Linked Births and Infant Deaths Database (LBIDD)
-            \cite{macdorman_infant_1998} with $D=177$ covariates. Treatment and
-            outcome models have been simulated with complex models to reflect
+        each setup, totaling 770 dataset instances.
+
+    \item[ACIC 2018] \cite{shimoni_benchmarking_2018}: Starting from data
+        from the Linked Births and Infant Deaths Database (LBIDD)
+        \cite{macdorman_infant_1998} with $D=177$ covariates, treatment and
+        outcome models are simulated with complex models to reflect
         different scenarii. The data do not provide the true propensity
@@
 
-        \item[Twins] \cite{louizos_causal_2017}: It is an augmentation of the
-            real data on twin births and mortality rates in the USA from 1989-1991
+    \item[Twins] \cite{louizos_causal_2017}: It is an augmentation of
+        real data on twin births and mortality rates
         \cite{almond_costs_2005}. There are $N=11\,984$ samples (pairs of twins),
@@
         weight of the heavier twin at birth. This is a "true" counterfactual dataset
-            --as remarked in \cite{curth_really_2021}-- in the sense that we have
+        \cite{curth_really_2021} in the sense that we have
         both potential outcomes with each twin. They simulate the treatment with a
@@
         the 45 other covariates:
-            \begin{equation}
-                \mathbf{t}_{i} \mid \mathbf{x}_{i}, \mathbf{z}_{i} \sim
+        \begin{align}
+            \mathbf{t}_{i} & \mid \mathbf{x}_{i}, \mathbf{z}_{i} \sim
             \operatorname{Bern}\left(\sigma\left(w_{o}^{\top}
-                \mathbf{x}+w_{h}(\mathbf{z} / 10-0.1)\right)\right) \quad \text{with} \;
-                w_{o} \sim \mathcal{N}(0,0.1 \cdot I),\; w_{h} \sim \mathcal{N}(5,0.1)
-            \end{equation}
-            We built upon this equation, adding a non-constant slope in the
-            treatment sigmoid, allowing us to control the amount of overlap between
+            \mathbf{x}+w_{h}(\mathbf{z} / 10-0.1)\right)\right)       \\ & \text{with} \;
+            w_{o} \sim \mathcal{N}(0,0.1 \cdot I),\; w_{h} \sim \mathcal{N}(5,0.1) \nonumber
+        \end{align}
+        We add a non-constant slope in the
+        sigmoid to control the overlap between
         treated and control populations.
         We sampled uniformly 1\,000 different overlap parameters between 0 and
-            2.5, totalizing 1\,000 dataset instances. Unlike the previous datasets,
+        2.5, totaling 1\,000 dataset instances. Unlike the previous datasets,
         only the overlap varies for these instances. The response surfaces are
-            fixed by the original twin outcomes.
+        set by the original outcomes.
 \end{description}
@@
 regressions). We select hyper-parameters with randomized search on a validation set
-    $\mathcal{V}$ and keep them fix for model selection (detailed of the hyper
-    parameters in Appendix \ref{apd:experiments:nuisances_hp}). As extreme inverse
+$\mathcal{V}$ and keep them fix for model selection
+(\ref{apd:experiments:nuisances_hp} lists hyperparameters). As extreme inverse
 propensity weights induce high variance, we use clipping
@@
 %\idea{Overlap is important but not easily measurable}
-    Good overlap, or ``positivity'' between treated and control population is
+Good overlap between treated and control population is
 crucial for causal inference as it is required by the positivity assumption
-    \ref{assumption:overlap} for causal identification. It is typically assessed by
-    qualitative methods using population histograms (as in Figure
-    \ref{fig:toy_example}) or side-by-side box plots, or quantitative approaches
-    such as Standardized Mean Difference
+\ref{assumption:overlap}. It is often assessed by
+comparing visually population distributions (as in Figure
+\ref{fig:toy_example}) or computing standardized difference on each
+feature
 \cite{austin_introduction_2011,austin_moving_2015}. While these methods are
-    useful to decide if positivity holds, they do not summarize a dataset's overlap
-    in a single measure. Rather, we compute the divergence
+useful to decide if positivity holds, they do not yield a
+single measure. Rather, we compute the divergence
 between the population covariate distributions $\mathbb P(X|A=0)$ and $\mathbb
-        P(X|A=1)$ to characterize the behavior of causal risk
-    \cite{damour_overlap_2020,johansson_generalization_2021}. We introduce the
+    P(X|A=1)$
+\cite{damour_overlap_2020,johansson2022generalization}. We introduce the
 Normalized Total Variation (NTV), a divergence based on the sole propensity
-    score. Details are given in Appendix \ref{apd:motivation_ntv}.
+score (see \ref{apd:motivation_ntv}).
 
@@
     \centering
-        \includegraphics[width=\textwidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset.pdf}
-        \hfill
-        \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$, measured
-            as the difference between each metric Kendall's $\tau$ and the mean kendall's
-            $\tau$ over all metric:  $\kappa(\ell,\tau\mathrm{{-risk}}) - mean_{\ell} \big(\kappa(\ell,\tau\mathrm{{-rRisk}}) \big)$. Strong and Weak overlap correspond
-            to the first and
-            last tertiles of the overlap distribution measured with Normalized Total
-            Varation eq. \ref{eq:ntv}.
-            Appendix \ref{apd:experiments:additional_results} presents the same results
-            measured with absolute Kendall's in Figure
-            \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with
-            $\tau\mathrm{-risk}$ gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table \ref{apd:table:relative_kendalls_all_datasets} displays the median
-            and IQR for the relative Kendall's results.}\label{fig:relative_kendalls_all_datasets}
+    \includegraphics[width=\linewidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
+    %\hfill
+    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$.
+        Strong and Weak overlap correspond to the first and last tertiles of the overlap distribution measured with
+        Normalized Total Variation eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents the same results
+        by adding semi-oracle risks in Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics}, measured with
+        absolute Kendall's in Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with $\tau\mathrm{-risk}$
+        gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table
+        \ref{apd:table:relative_kendalls_all_datasets} gives median and
+        IQR of the relative Kendall.}\label{fig:relative_kendalls_all_datasets}
 \end{figure}
 
-    \subsection{Empirical results: factors driving good model selection across
-        datasets}\label{empirical_study:results}
+\subsection{Results: factors driving good model selection}\label{empirical_study:results}
 
@@
 ideal ranking of methods given the oracle $\tau\text{-risk}$ and
-    the different causal metrics under evaluation. We measure this agreement
+the different feasible causal metrics. We measure this agreement
 with a relative\footnote{To remove the variance
@@
 how well metrics approximate the oracle $\tau\text{-risk}$
-    (\autoref{apd:proofs:mu_risk_ipw_bound}), we separate
-    strong and weak overlap --defined as first and
-    last tertile of the Normalized Total Variation, eq. \ref{eq:ntv}.
-
-    Among all metrics the classical mean squared error (ie. factual
+(\ref{apd:proofs:mu_risk_ipw_bound}), we separate
+strong and weak overlap.
+
+Among all metrics, the classical mean squared error (ie. factual
 $\mu\text{-risk}$) is worse and reweighting it with propensity score
@@
 scores, leads to the best performances. Interestingly, the
-    $U\text{-risk}$, which uses the same nuisances, has good performances for
-    strong overlap but deteriorates in weak overlap, probably due variance
+$U\text{-risk}$, which uses the same nuisances, deteriorates in weak overlap, probably due to variance
 inflation when dividing by extreme propensity scores.
 
-    Beyond the rankings, the differences in terms of absolute
-    ability to select the best model are large: The model selected by the R-risk
-    achieves a $\tau\text{-risk}$ only 1\% higher
-    than that of the best
+Beyond rankings, the differences in terms of absolute
+ability to select the best model are large: The R-risk selects a model
+with a $\tau\text{-risk}$ only 1\% higher
+than the best
 possible candidate for strong overlap on Caussim, but selecting with
@@
 \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
-    other datasets, the $R\text{-risk}$ consistently decreases the
+datasets, the $R\text{-risk}$ consistently decreases the
 risk compared to the $\mu\text{-risk}$:
-    0.1\% compared to 1\% on ACIC2016,  1\% compared to 20\% on ACIC 2018,
+0.1\% compared to 1\% on ACIC2016,  1\% compared to 20\% on ACIC2018,
 and 0.05\% compared to 1\% on Twins.
 
-    \paragraph{Model selection is harder in settings of low population
+\paragraph{Model selection is harder for low population
     overlap}
@@
 both Caussim and ACIC 2018
-    (Appendix \ref{apd:experiments:additional_results},
-    Figure {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}).
+(\ref{apd:experiments:additional_results}).
 
@@
     \centering
-        \includegraphics[width=0.9\textwidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only.pdf}
-        \hfill
-        \caption{\textbf{Model selection is harder in settings of low population
+    \includegraphics[width=\linewidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
+    %\hfill
+    \caption{\textbf{Model selection is harder for low population
             overlap}:
         Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
-            are the tertiles of the overlap distribution measured with NTV eq. \ref{eq:ntv}. Appendix \ref{apd:experiments:additional_results} presents results for all
-            metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} and in absolute
+        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents results for all
+        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
         Kendall's and continuous overlap values in Figure
@@
 nuisance estimates is a form of double dipping which can lead errors in
-    nuisances to be correlated to that of outcome models
+nuisances correlated to that of outcome models
 \cite{nie_quasioracle_2017}. In theory, these correlations can bias model
-    selection and, strictly speaking, there are theoretical arguments
+selection and, strictly speaking, push
 to split out a third separated data set --a ``nuisance set''-- to fit the
 nuisance models. The drawback is that it depletes the data available for
-    model estimation and selection. However, Figure \ref{fig:procedures_comparison} shows that
-    there is no substantial differences between a procedure with a separated
-    nuisance set and the simpler shared nuisance-candidate set procedure
-    (results for all metrics in Figure
-    \ref{apd:fig:procedures_comparison_all_metrics}).
-
-    \begin{figure}[!h]
-        %\centering
-        \begin{subfigure}[b]{0.48\textwidth}
-            %\centering
-            %\caption{\textbf{Two sets procedure}}
-            \includegraphics[width=2\textwidth]{images/_3_procedure_r_risk_only_3datasets.pdf}
-        \end{subfigure}
-        \hfill
+model estimation and selection. However, Figure
+\ref{fig:procedures_comparison} shows no substantial difference between a procedure with a separated
+nuisance set and the simpler shared nuisance-candidate set procedure.
+
+\begin{figure}[!tb]
+    \includegraphics[width=\linewidth]{images/_3_procedure_r_risk_only_3datasets_twocols.pdf}
+
     \caption{\textbf{Nuisances can be estimated on the same data as outcome
@@
         the \textcolor{RedOrange}{separated nuisances set} procedures. Figure
-            \ref{apd:fig:procedures_comparison_all_metrics} details results for all metrics.}\label
-        {fig:procedures_comparison}
+        \ref{apd:fig:procedures_comparison_all_metrics} details results for all metrics.}\label{fig:procedures_comparison}
 \end{figure}
@@
 
-    Oracle versions of every risks recover more often the best estimator.
-    However, in configurations that we investigated,
+For every risk, the oracle version recovers better the best estimator.
+However,
 stacked nuisances estimators (boosting and linear) lead to feasible
-    metrics with close performances to the oracles ones. This suggests that the
+metrics with close performances to the oracles ones: the
 corresponding estimators recover well-enough the true nuisances.
-    One may wonder if it may be useful to use simpler models for the
-    nuisances, in particular in settings where there are less data or where
-    the true models are linear.
+One may wonder if simpler models for the nuisance could be useful,
+in particular in data-poor settings or when the true models are linear.
 Figure \ref{fig:all_datasets_nuisances_comparison} compares causal model
@@
 stacked and linear estimations of the nuisances performs equivalently.
-    Overall, Figure \ref{apd:fig:nuisances_comparison_twins} suggests that
-    to estimate nuisance it suffices to use adaptive models as built by
-    stacking linear models and gradient-boosted trees.
-    Figure \ref{apd:fig:nuisances_comparison} in the appendices, details
-    other causal metrics.
-
-    \begin{figure}[!h]
-        \begin{subfigure}[b]{0.48\textwidth}
-            \centering
-            \includegraphics[width=2\textwidth]{images/_4_nuisance_models_r_risk_only_3datasets.pdf}
-        \end{subfigure}
-        \hfill
+Detailed analysis (Figure \ref{apd:fig:nuisances_comparison_twins})
+confirms that using adaptive models --as built by
+stacking linear models and gradient-boosted trees-- suffices to estimate nuisance.
+
+\begin{figure}[!tb]
+    \includegraphics[width=\linewidth]{images/_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
+
     \caption{\textbf{\textcolor{DarkOrchid}{Stacked
@@
         Results are shown only for the
-            R-risk. Details for every metrics are provided in Figure
-            \ref{apd:fig:nuisances_comparison}. For Twins, where the true propensity
+        R-risk; Figure \ref{apd:fig:nuisances_comparison}
+        details every metrics. For Twins, where the true propensity
         model is linear, \textcolor{DarkOrchid}{stacked} and
@@
 
-    For best causal modeling, the analyst often faces a compromise: given a finite
-    data sample, should she allocate most of the data to estimate the outcome model,
-    thus maximizing chances of achieving a high-quality outcome model but leaving
+The analyst faces a compromise: given a finite
+data sample, should she allocate more data to estimate the outcome model,
+thus improving the quality of the outcome model but leaving
 little data for model selection. Or, she could choose a bigger test set for
 model selection and effect estimation. For causal model selection, there
-    is no established practice (as reviewed in Appendix \ref{apd:results:k_fold_choices}).
+is no established practice (as reviewed in \ref{apd:results:k_fold_choices}).
 
@@
 $\mathcal{V}$ on which we use the oracle response functions to derive
-    silver-standard estimates of the causal quantities of interest. We then
+silver-standard estimates of causal quantities. We then
 use the standard estimation procedure on the remaining 70\% of the data,
@@
 sizes. We finally measure the error between this estimate and the
-    silver-standard one.
+silver standard.
 
@@
 treatment effect --a single number used for policy making-- and a
-    CATE --A full model of the treatment effect as a function of covariates
+CATE --a full model of the treatment effect as a function of covariates
 $X$. Given that the latter is a much more complex object than the former,
@@
 \begin{figure}[!t]
-        %\centering
     \begin{minipage}{.5\textwidth}
         \centerline{\textbf{a) CATE estimation error}}
-
-            \includegraphics[height=.35\textwidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16.pdf}
+        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16_twocols.pdf}
     \end{minipage}
+    %\hspace{0.3\textwidth}
     \begin{minipage}{.5\textwidth}
-            %\centering
         \centerline{\textbf{b) ATE estimation error}}
-            \includegraphics[height=.35\textwidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16.pdf}
+        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16_twocols.pdf}
     \end{minipage}%
-        %\hspace{0.3\textwidth}
-        \caption{\textbf{a) For CATE, a train/test ratio of 0.9/0.1 appears as a good
+    \caption{\textbf{a) For CATE, a train/test ratio of 0.9/0.1 appears a good
             trade-off.} b) For ATE, there is a small signal pointing also to
@@
 Figure \ref{fig:train_test_ratio} shows that a train/test ratio of
-    0.9/0.1 (K=10) appears as a good trade-off for CATE and ATE estimation,
-    thought there is little difference with a split of 0.8/0.2 (K=5).
+0.9/0.1 (K=10) or 0.8/0.2 (K=5) appears best to estimate CATE and
+ATE.
 
@@
 
-    Predictive models are increasingly used to reason about causal effects.
+Predictive models are increasingly used to reason about causal effects,
+for instance in precision medicine to drive individualized decision.
 Our results highlight that they should be selected, validated, and tuned
@@
 (eq.\,\autoref{def:r_risk}) leads to more valid causal estimates.
+
+\paragraph{Nuisance models: more gain than pain}
 %
-    Estimating this risk requires a markedly more complex procedure than
-    standard cross-validation used \emph{e.g.} in machine learning: it involves
+Estimating the $R\text{-risk}$ requires a more complex procedure than
+standard cross-validation used \emph{e.g.}~in machine learning: it involves
 fitting nuisance models necessary for model evaluation, though our
-    empirical results show that these can be learned on the same set of data
+results show that these can be learned on the same set of data
 as the outcome model evaluated.
+
+The nuisance models must be well estimated (Figure
+\ref{fig:all_datasets_nuisances_comparison}). However these models are
+easier to select and control than a causally-valid outcome model,
+as they are associated to errors on
+observed distributions. Our results show that using for nuisance models
+a flexible stacking-based family of estimator suffices for good model selection.
 %
-    A poor estimation of the nuisance models may compromise the benefits of
-    the more complex $R\text{-risk}$ (as shown in Figure
-    \ref{fig:all_datasets_nuisances_comparison}). However controlling and selecting
-    these latter models is easier because they are associated to errors on
-    observed distributions. Our empirical results show that when selecting
-    these models in a flexible family of models the $R\text{-risk}$ dominates
-    simpler risks for model selection.
-    %
-    Results show that going from an oracle $R\text{-risk}$ --where the
-    nuisances are known-- to a feasible $R\text{-risk}$ --where the nuisances
-    are estimated-- decreases only very slightly the model-selection
-    performance of the $R\text{-risk}$. This may be explained by
-    theoretical results that suggest that estimation errors on both
+In fact, a feasible $R\text{-risk}$ --where the nuisances
+are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
+nuisances are known. This may be explained by
+results that suggest that estimation errors on both
 nuisances partly compensate out in the
-    $R\text{-risk}$\cite{daniel2018double,naimi2021challenges,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018}.
-    % 
-    %The usage of the $R\text{-risk}$ can also be understood as a
-    %$\tau \text{-risk}$ reweighted by the propensity score
-    %(prop \autoref{theory:prop:r_risk_rewrite}).
-    %
-
-    For strong overlap, the conventional procedure ($\mu\text{-risk}$) appears theoretically motivated
-    (\autoref{theory:mu_risk_ipw_bound}), however empirical results show that
-    even in this regime the $R\text{-risk}$ brings a sizeable benefit,
-    in agreement with \citet{schuler_comparison_2018}.
-
-    Our results points to K=5 or 10 for the choice of the number of folds used in
-    cross-validation for CATE and ATE, aligned with previous empirical evidence for
-    ATE \cite{chernozhukov_double_2018}.
-
-    \paragraph{Extension to binary outcome}
+$R\text{-risk} $\cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.
+
+Note that propensity score models must be selected to estimate
+the individual posterior probability. For this, we used the Brier score,
+is minimized by the true individual probability. An easy mistake is to
+use calibration errors popular in machine learning
+\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
+as these select not for the individual posterior probability but for an
+aggregate error rate \cite{perez2022beyond}.
+
+\paragraph{Extension to binary outcomes}
 While we focused on continuous outcomes, in medicine, the target outcome
@@
 \cite{austin2017estimating}. While the odds ratio is natural for
-    case-control studies \cite{rothman2008case}, a good choice of measure
+case-control studies \cite{rothman2008case}, other measures
 can reduce heterogeneity \cite{colnet2023risk}.
-    Using as an estimand the log of
-    these values is suitable to additive models (for reasoning or noise
-    assumptions). In the log domain, the relative risk or the odds ratio are
-    written as a difference, as the ATE: $\log \mathbb{P}(Y(1)=1) - \log
-        \mathbb{P}(Y(0=1)$ or $\log (\mathbb{P}(Y(1)=1) / [1 -\mathbb{P}(Y(1)=1)]
-        ) - \log \mathbb{P}(Y(0=1) / [1 -\mathbb{P}(Y(0)=1)]$. Hence, the
+In the log domain, the ratios are written as a difference, the
 framework studied here (\autoref{sec:neyman_rubin}) can directly apply.
-    It is particularly easy for the log odds ratio, as it is the output of a
-    logistic regression or any model with a cross-entropy loss.
-
-    \paragraph{Going further}
-    % Estimation of the nuisances, calibration
-    The $R\text{-risk}$ needs good estimation of nuisance models. The
-    propensity score $e$ calls for a control on the estimation of the
-    individual posterior probability. We have used the Brier score to select
-    these models, as it is minimized by the true individual probability. Regarding model-selection for
-    propensity score, an easy mistake is to use expected calibration errors
-    popular in machine learning
-    \cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
-    as these select not for the individual posterior probability but for an
-    aggregate error rate \cite{perez2022beyond}. An
-    open question is whether a better metric than the brier score can be designed that controls for $e \, (1 - e)$, the quantity used in
-    the $R\text{-risk}$, rather than $e$.
-
-    %\idea{Treatment heterogeneity seems important.}
-    The quality of model selection varies substantially from one
-    data-generating mechanism to another. The overlap appears as an important
-    parameter: when the treated and untreated, causal model selection is very
-    hard. However, remaining variance in the empirical results suggests that
-    other parameters of the data generation processes come into play.
-    Intuitively, the complexity of the response surfaces and the treatment
-    heterogeneity interact with overlap violations: when extrapolations to
-    weak-overlap regions is hard, causal model selection is hard.
-
-    Nevertheless, from a practical perspective, our study establishes that
-    the $R\text{-risk}$ is the best option to select predictive models for
-    causal inference, without requiring assumptions on the data-generating
-    mechanism, the amount of data at hand, or the specific estimators used to
-    build predictive models.
-
-    % XXX: need a final positive
-    % remark
-    % Maybe that as our work has no
-    % requirement on the learning procedure,
-    % it contributes to ensuring the
-    % validity of causal inference on complex databases such as clinical
-    % records with a lot of open-ended text that must be analyzed with
-    % language models.
-
-    % Sentence stolen, adapt
-    % unlike parametric regression, machine learning (ML) methods do not
-    % generally require precise
-    % knowledge of the true data generating mechanisms
-
-\end{multicols}{2}
+In particular, the log odds ratio is estimated by the common
+cross-entropy loss (or log loss) as in logistic regression.
+
+\paragraph{More $R\text{-risk}$ to select models driving decisions}
+
+Prediction models have flourished because their predictions can be easily
+demonstrated and validated on left-out data. But they require more
+careful validation for decision making, using a metric accounting for the
+putative intervention, the $R\text{-risk}$. Even when treated and
+untreated population differ little, as in RCTs, the $R\text{-risk}$
+brings a sizeable benefit. To facilitate better model selection, we provide Python
+code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
+Using the $R\text{-risk}$ does make evaluation more
+complicated not only because the procedure is more involved, but also
+because each intervention requires a dedicated evaluation. However, such
+off-policy evaluation remains much less costly than the recommended good
+practice of impact evaluation testing the ability of a prediction model
+to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
+the model-selection procedure puts no constraints on the models used to
+build predictive models: it opens the door to evaluating a wide range of
+models, from gradient boosting to convolutional neutral, or language
+models.
 
@@
 
+\onecolumn
+
 \appendix
+%\setcounter{secnumdepth}{0}
+\setcounter{secnumdepth}{3}
 

### 2023-11-16 how_to_select_causal_models.tex
finish to adpat paper

@@
 Indeed, estimates of treatment effects can vary markedly across different
-predictive models. For instance, Figure shows
+predictive models. For instance, Figure \ref{fig:acic_2016_ate_heterogeneity} shows
 large variations obtained across different outcome estimators on
@@
 We first illustrate the problem of causal model
-selection and briefly review prior art. Then, Section
-\ref{sec:framework} sets causal model selection in the
+selection and briefly review prior art. Then the methods
+\autoref{sec:framework} sets causal model selection in the
 \emph{potential outcome} framework and details the causal risks and
-model-selection procedure. Section \ref{sec:theory} gives theoretical
-results. Section \ref{sec:empirical_study} details a thorough empirical
-study, covering many different settings. Finally, Section \ref{sec:discussion}
-discusses the findings.
+model-selection procedure. The next \autoref{sec:theory} gives theoretical
+results. The results \autoref{sec:empirical_study} details a thorough empirical
+study, covering many different settings. Finally, we
+discuss the findings in the last \autoref{sec:discussion}.
 %
@@
 
-\begin{figure}[t!]
+\begin{figure}[b!]
     %\centering
@@
 
-\begin{table*}[!tb]
+\begin{table*}[bt!]
     \makebox[\textwidth]{
@@
 between $\tau\text{-risk}$ and $\mu\text{-risk}_{IPW}$ coming from heterogeneous
-errors between populations as noted in Section \ref{theory:mu_risk_ipw_bound}
+errors between populations as noted by Proposition \ref{theory:prop:mu_risk_ipw_bound}
 and shown experimentally in \citet{schuler_comparison_2018} and our
@@
 overlap between treated and non-treated populations, an important parameter
-for causal inference (related to $\eta$ in section
-\ref{theory:mu_risk_ipw_bound}). Finally, we generate observed outcomes
+for causal inference (related to $\eta$ in Proposition
+\ref{theory:prop:mu_risk_ipw_bound}). Finally, we generate observed outcomes
 adding Gaussian noise. We generate 1\,000 of such datasets, with
-uniformly random overlap parameters. Details in
+uniformly random overlap parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
 \ref{apd:experiments:generation}.
@@
 Rather, selecting the best outcome model according to the $R\text{-risk}$
-(eq.\,\autoref{def:r_risk}) leads to more valid causal estimates.
+(eq.\, Definition \ref{def:r_risk}) leads to more valid causal estimates.
 
@@
 nuisances partly compensate out in the
-$R\text{-risk} $\cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.
+$R\text{-risk}$ \cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.
 
@@
 used in the context of causal inference
-\cite{shalit_estimating_2017,johansson_generalization_2021}. However it
+\cite{shalit_estimating_2017,johansson2022generalization}. However it
 needs a kernel, typically Gaussian, to extrapolate across neighboring
@@
 
+\paragraph{Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics} -
+    Results measured in relative Kendall's for feasible and semi-oracle risks}
+
+Because of extreme propensity scores in the denominator and bayes error residuals in the numerator, the semi-oracle
+$U$-risk has poor performances at bad overlap. Estimating these propensity scores in the is feasible $U$-risk reduces
+the variance since clipping is performed.
+
+\begin{figure}[!b]
+    \centering
+    \includegraphics[width=\linewidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset.pdf}
+    %\hfill
+    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$
+        agreement with $\tau\text{-risk}$. Strong and Weak overlap correspond to
+        the first and last tertiles of the overlap distribution measured with
+        Normalized Total Variation eq. \ref{eq:ntv}.
+    }\label{apd:fig:relative_kendalls_all_datasets_all_metrics}
+\end{figure}
+
 \paragraph{Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} -
@@
 the nuisances required to optimize the outcome model (as the
-$R\text{-risk}$, definition \autoref{def:r_risk}).
+$R\text{-risk}$, Definition \ref{def:r_risk}).
 %

### 2023-11-19 how_to_select_causal_models.tex
ignore auxilliary files
updated cover letteer
condense the paper

@@
-%\documentclass[MYBIB,STIX1COL]{WileyNJD-v2}
 \documentclass[10pt,twocolumn]{article}
@@
 
-\title{How to select predictive models for causal inference?}
+\title{How to select predictive models for decision-making or causal inference?}
 \addbibresource{biblio.bib}
@@
 \begin{abstract}
-    \textbf{Background:}
+    \textbf{Objective:}
     As predictive models --\emph{eg} from machine learning-- give likely
-    outcomes, they may be used to reason on the effect of an intervention,
-    a causal-inference task. The increasing complexity of
-    health data has opened the door to a plethora of models, but also the
-    Pandora box of model selection: which of these models yield the most
-    valid causal estimates? Causal model selection should control both outcome errors for each
-    individual, treated or not treated, whereas only one outcome is observed.
-    Theoretically, simple risks used in machine learning do not control causal
-    effects when treated and non-treated population differ too much. \\
+    outcomes, they may be used to reason on the effect of an intervention, a
+    causal-inference task. The increasing complexity of health data has opened
+    the door to a plethora of models, but also the Pandora box of model
+    selection: which of these models yield the most valid causal estimates?  In
+    this paper, we study model selection procedures in practical settings:
+    finite samples settings and without well-specification assumption.
     \textbf{Methods:}
-    More elaborate risks build proxies of the causal error using ``nuisance''
+    Elaborate risks build proxies of the causal error using ``nuisance''
     re-weighting to compute it on the observed data. But does computing these
     nuisance adds noise to model selection? Drawing from an extensive empirical
-    study based on a simulation and three semi-simulated datasets based on real
-    covariates, we compare different metrics for causal model selection.\\
+    study based on a simulation and three semi-simulated healthcare datasets
+    based on real covariates, we compare different metrics for causal model
+    selection.\\
     \textbf{Results:} Among all metrics the classical mean squared error (ie.
@@
 Indeed, estimates of treatment effects can vary markedly across different
-predictive models. For instance, Figure \ref{fig:acic_2016_ate_heterogeneity} shows
-large variations obtained across different outcome estimators on
-semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models
-such as random forests are doing well in most settings except
-when treated and untreated populations differ noticeably, in
-which case a linear model (ridge) is to be preferred.
-However random forests with different hyper-parameters
-(max depth= 2) yield poor estimates.
-A simple rule of thumb such as preferring flexible models does not work in
-general; model selection is needed.
-
-%\idea{It is crucial to have tools for model selection.}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{images/2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
-    \caption{\textbf{Different outcome models lead to different
-            estimation errors on the Average Treatment Effects},
-        on 77 classic simulations with known true causal effect
-        \cite{dorie_automated_2019}. The different models are ridge regression
-        and random forests with different hyper-parameters
-        (details
-        \ref{apd:toy_example:acic_2016_ate_variability}). The different configurations are
-        plotted as a function of increasing difference between treated and
-        untreated population --see
-        \autoref{subsec:measuring_overlap}.
-        There is no systematic best performer; data-driven model
-        selection is important.
-        \label{fig:acic_2016_ate_heterogeneity}%
-    }
-\end{figure}
+predictive models (see example from \cite{dorie_automated_2019} developed in
+Appendix \ref{apd:toy_example:acic_2016_ate_variability}).
 
@@
 %
-Standard predictive modeling or machine-learning practice selects the
-predictor that minimizes the expected error.
-However, this predictor may not be the best model to reason about
-causal effects of an intervention, as we illustrate below.
-
-Figure \ref{fig:toy_example} gives a toy example: the probability $Y$ of
-an undesirable
-outcome (\emph{eg} death), a binary treatment $A \in \{0, 1\}$, and a covariate
-$X \in \mathbb R$ summarizing the patient health status (eg. the
-Charlson index \cite{charlson_new_1987}). We simulate a
-treatment beneficial (decreases $Y$) for patients with high Charlson scores (bad health
-status) but with little effect for patients in good
-condition (low Charlson scores).
+Standard predictive modeling or machine-learning practice selects the predictor
+that minimizes the expected error. However, this predictor may not be the best
+model to reason about causal effects of an intervention as Figure
+\ref{fig:toy_example} illustrates. Consider the probability $Y$ of an undesirable outcome
+(\emph{eg} death), a binary treatment $A \in \{0, 1\}$, and a covariate $X \in
+\mathbb R$ summarizing the patient health status (eg. the Charlson index
+\cite{charlson_new_1987}). We simulate a treatment beneficial (decreases $Y$)
+for patients with high Charlson scores (bad health status) but with little
+effect for patients in good condition (low Charlson scores).
 
@@
 \cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
-vary at specific knots, they give flexible non-linear models. We use random approximation of Radial Basis Function (RBF) kernels
-\cite{rahimi_random_2008} to generate the response functions. RBF use the same
-process as polynomial splines but replace polynomial by Gaussian kernels. Unlike
-polynomials, Gaussian kernels have decreasing influences in the input space.
-This avoids unrealistic divergences of the response surfaces at the
-ends of the feature space.
-
-The number of basis functions --\emph{ie. knots}--, controls the complexity of
-the ground-truth response surfaces and treatment. We first use this process to
-draw the non-treated response surface $\mu_0$ and the causal effect $\tau$. We
-then draw the observations from a mixture two Gaussians, for the treated and non
-treated. We vary the separation between the two Gaussians to control the
-overlap between treated and non-treated populations, an important parameter
-for causal inference (related to $\eta$ in Proposition
-\ref{theory:prop:mu_risk_ipw_bound}). Finally, we generate observed outcomes
-adding Gaussian noise. We generate 1\,000 of such datasets, with
-uniformly random overlap parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
+vary at specific knots, they give flexible non-linear models. The number of
+knots controls the complexity of the function. We use random approximation of
+Radial Basis Function (RBF) kernels \cite{rahimi_random_2008} to generate the
+outcome and treatment functions. RBF use the same process as polynomial splines
+but replace polynomial by Gaussian kernels. Unlike polynomials, Gaussian kernels
+have decreasing influences in the input space. This avoids unrealistic
+divergences of the response surfaces at the ends of the feature space. We
+generate 1\,000 datasets based on this functions, with random overlap
+parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
 \ref{apd:experiments:generation}.
@@
 
-We test model selection on a family of candidate estimators that
-approximate imperfectly the data-generating process. To build such an
-estimator, we first use a RBF expansion
-similar to the one used for data generation.
-We choose two random knots and apply a transformation of the raw data features
-with a Gaussian kernel. This step
-is referred as the featurization. Then, we fit a linear regression on
-these
-transformed features. We consider two ways of combining these steps for outcome
-mode; we use common nomenclature \cite{kunzel_metalearners_2019,shen2023rctrep} to refer
-to these different meta-learners that differ on how
-they model, jointly or not, the treated and the non treated:
+We compare different candidate estimators that approximate imperfectly the
+data-generating process. To build such estimators, we first use a RBF
+expansion similar to the one used for data generation. We choose two random
+knots and apply a transformation of the raw data features with a Gaussian
+kernel. This step is referred as the featurization. Then, we fit a linear
+regression on these transformed features. We consider two ways of combining
+these steps for outcome mode; we use common nomenclature
+\cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different
+meta-learners that differ on how they model, jointly or not, the treated and the
+non treated:
 \begin{itemize}
@@
           a supplementary covariate.
-
     \item SftLearner: A single set of basis functions is sampled at random for both
@@
 
-We also use classic benchmarks of the causal-inference
-literature, semi-simulated data adding a known synthetic causal effect to real --non synthetic-- covariate:
+We also use three benchmarks of the causal-inference literature, semi-simulated
+data adding a known synthetic causal effect to real --non synthetic-- healthcare
+covariate. ACIC 2016 \cite{dorie_automated_2019} is based on the Collaborative
+Perinatal Project \cite{niswander_women_1972}, a RCT studying infants’
+developmental disorders containing 4,802 indivduals and 55 features. We used 770
+dataset instances: 10 random seeds for each of the 77 simulated settings for the
+treatment and outcomes.  ACIC 2018 \cite{shimoni_benchmarking_2018} simulated
+treatment and outcomes for the Linked Births and Infant Deaths Database (LBIDD)
+\cite{macdorman_infant_1998} with $D=177$ covariates. We used all 432 datasets
+of size $N=5\,000$.  Twins \cite{louizos_causal_2017} is an augmentation of real
+data on twin births and mortality rates \cite{almond_costs_2005}. There are
+$N=11\,984$ samples, and $D=50$ covariates for which we simulated 1,000
+different treatment allocations. Datasets details are given in Appendix
+\ref{apd:experiments:datasets}.
+
+\paragraph{Family of candidate
+    estimators}\label{semi_simulated:candidate_estimators}
+
+For these three datasets, the family of candidate estimators are gradient
+boosting trees for both the response surfaces and the treatment
+\footnote{Scikit-learn regressor,
+    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingRegressor.html}{HistGradientBoostingRegressor},
+    and classifier,
+    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingClassifier.html}{HistGradientBoostingClassifier}.}
+with S-learner, learning rate
+in $\{0.01, 0.1, 1\}$, and maximum number of leaf nodes in $\{25, 27, 30, 32,
+    35, 40\}$ resulting in a family of size 18.
+
+\paragraph{Nuisance estimators}
+
+Drawing inspiration from the TMLE literature that uses combination of flexible
+machine learning methods \cite{schuler_targeted_2017}, we use a meta-learner as models
+for the nuisances $\check e$ (respectively $\check m$): a
+stacked estimator of ridge and boosting classifiers (respectively
+regressions). We select hyper-parameters with randomized search on a validation set
+$\mathcal{V}$ and keep them fix for model selection
+(\ref{apd:experiments:nuisances_hp} lists hyperparameters).
+
+\subsection{Measuring overlap between treated and non treated}\label{subsec:measuring_overlap}
+
+%\idea{Overlap is important but not easily measurable}
+Good overlap between treated and control population is
+crucial for causal inference as it is required by the positivity assumption
+\ref{assumption:overlap}. It is often assessed by
+comparing visually population distributions (as in Figure
+\ref{fig:toy_example}) or computing standardized difference on each
+feature
+\cite{austin_introduction_2011,austin_moving_2015}. While these methods are
+useful to decide if positivity holds, they do not yield a
+single measure. Rather, we compute the divergence
+between the population covariate distributions $\mathbb P(X|A=0)$ and $\mathbb
+    P(X|A=1)$
+\cite{damour_overlap_2020,johansson2022generalization}. We introduce the
+Normalized Total Variation (NTV), a divergence based on the sole propensity
+score (see \ref{apd:motivation_ntv}).
+
+\begin{figure}[!b]
+    \centering
+    \includegraphics[width=\linewidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
+    %\hfill
+    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$.
+        Strong and Weak overlap correspond to the first and last tertiles of the overlap distribution measured with
+        Normalized Total Variation eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents the same results
+        by adding semi-oracle risks in Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics}, measured with
+        absolute Kendall's in Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with $\tau\mathrm{-risk}$
+        gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table
+        \ref{apd:table:relative_kendalls_all_datasets} gives median and
+        IQR of the relative Kendall.}\label{fig:relative_kendalls_all_datasets}
+\end{figure}
+
+\subsection{Results: factors driving good model selection}\label{empirical_study:results}
+
+\paragraph{The $R\text{-risk}$ is the best metric}
+
+Each metric ranks differently the candidate models. Figure
+\ref{fig:relative_kendalls_all_datasets} shows the agreement between the
+ideal ranking of methods given the oracle $\tau\text{-risk}$ and
+the different feasible causal metrics. We measure this agreement
+with a relative\footnote{To remove the variance
+    across datasets (some datasets lead to easier model selection than
+    others), we report values for one metric relative to the mean of all
+    metrics for a given dataset instance: $\text{Relative} \, \kappa(\ell,\tau\mathrm{{-risk}})=
+        \kappa(\ell,\tau\mathrm{{-risk}}) -
+        mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$ (eq.
+\ref{eq:kendall_tau}) \cite{kendall_new_1938}.
+Given the importance of overlap in
+how well metrics approximate the oracle $\tau\text{-risk}$
+(\ref{apd:proofs:mu_risk_ipw_bound}), we separate
+strong and weak overlap.
+
+Among all metrics, the classical mean squared error (ie. factual
+$\mu\text{-risk}$) is worse and reweighting it with propensity score
+($\mu\text{-risk}_{IPW}$) does not bring much improvements. The
+$R\text{-risk}$, which includes a model of mean outcome and propensity
+scores, leads to the best performances. Interestingly, the
+$U\text{-risk}$, which uses the same nuisances, deteriorates in weak overlap, probably due to variance
+inflation when dividing by extreme propensity scores.
+
+Beyond rankings, the differences in terms of absolute
+ability to select the best model are large: The R-risk selects a model
+with a $\tau\text{-risk}$ only 1\% higher
+than the best
+possible candidate for strong overlap on Caussim, but selecting with
+the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
+practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
+--as in some causal-inference methods \citep{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
+(Figure
+\ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
+datasets, the $R\text{-risk}$ consistently decreases the
+risk compared to the $\mu\text{-risk}$:
+0.1\% compared to 1\% on ACIC2016,  1\% compared to 20\% on ACIC2018,
+and 0.05\% compared to 1\% on Twins.
+
+\paragraph{Model selection is harder for low population
+    overlap}
+
+Model selection for causal inference becomes more and more difficult with
+increasingly different treated and control populations (Figure
+\ref{fig:all_datasets_overlap_effect_r_risk}). The absolute Kendall's
+coefficient correlation with $\tau\text{-risk}$ drops from values around 0.9
+(excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
+(\ref{apd:experiments:additional_results}).
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=\linewidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
+    %\hfill
+    \caption{\textbf{Model selection is harder for low population
+            overlap}:
+        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
+        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents results for all
+        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
+        Kendall's and continuous overlap values in Figure
+            {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
+\end{figure}
+
+\paragraph{Nuisances can be estimated on the same data as outcome models}
+
+Using the train set $\mathcal{T}$ both to fit the candidate estimator and the
+nuisance estimates is a form of double dipping which can lead errors in
+nuisances correlated to that of outcome models
+\cite{nie_quasioracle_2017}. In theory, these correlations can bias model
+selection and, strictly speaking, push
+to split out a third separated data set --a ``nuisance set''-- to fit the
+nuisance models. The drawback is that it depletes the data available for
+model estimation and selection. However, Figure
+\ref{fig:procedures_comparison} shows no substantial difference between a procedure with a separated
+nuisance set and the simpler shared nuisance-candidate set procedure.
+
+\begin{figure}[!tb]
+    \includegraphics[width=\linewidth]{images/_3_procedure_r_risk_only_3datasets_twocols.pdf}
+
+    \caption{\textbf{Nuisances can be estimated on the same data as outcome
+            models}: Results for the R-risk are similar between the
+        \textcolor{MidnightBlue}{shared
+            nuisances/candidate set} and
+        the \textcolor{RedOrange}{separated nuisances set} procedures. Figure
+        \ref{apd:fig:procedures_comparison_all_metrics} details results for all metrics.}\label{fig:procedures_comparison}
+\end{figure}
+
+\paragraph{Stacked models are good overall estimators of nuisances}
+
+For every risk, the oracle version recovers better the best estimator.
+However,
+stacked nuisances estimators (boosting and linear) lead to feasible
+metrics with close performances to the oracles ones: the
+corresponding estimators recover well-enough the true nuisances.
+One may wonder if simpler models for the nuisance could be useful,
+in particular in data-poor settings or when the true models are linear.
+Figure \ref{fig:all_datasets_nuisances_comparison} compares causal model
+selection estimating nuisances with stacked estimators or linear model.
+It comprises the Twins data, where the true propensity model is linear,
+and a downsampled version of this data, to study a situation favorable to
+linear models. In these settings,
+stacked and linear estimations of the nuisances performs equivalently.
+Detailed analysis (Figure \ref{apd:fig:nuisances_comparison_twins})
+confirms that using adaptive models --as built by
+stacking linear models and gradient-boosted trees-- suffices to estimate nuisance.
+
+\begin{figure}[!tb]
+    \includegraphics[width=\linewidth]{images/_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
+
+    \caption{\textbf{\textcolor{DarkOrchid}{Stacked
+                models} are good overall estimators of the nuisances}:
+        Results are shown only for the
+        R-risk; Figure \ref{apd:fig:nuisances_comparison}
+        details every metrics. For Twins, where the true propensity
+        model is linear, \textcolor{DarkOrchid}{stacked} and
+        \textcolor{ForestGreen}{linear}
+        estimations of the nuisances performs equivalently, even for a downsampled version
+        (N=4794). }\label{fig:all_datasets_nuisances_comparison}
+\end{figure}
+
+\paragraph{Use 90\% of the data to estimate outcome models, 10\% to
+    select them}
+
+The analyst faces a compromise: given a finite
+data sample, should she allocate more data to estimate the outcome model,
+thus improving the quality of the outcome model but leaving
+little data for model selection. Or, she could choose a bigger test set for
+model selection and effect estimation. For causal model selection, there
+is no established practice (as reviewed in \ref{apd:results:k_fold_choices}).
+
+We investigate such tradeoff varying the ratio between train and test
+data size. For this, we first split out 30\% of the data as a holdout set
+$\mathcal{V}$ on which we use the oracle response functions to derive
+silver-standard estimates of causal quantities. We then
+use the standard estimation procedure on the remaining 70\% of the data,
+splitting it into train $\mathcal{T}$ and test $\mathcal{S}$ of varying
+sizes. We finally measure the error between this estimate and the
+silver standard.
+
+We consider two different analytic goals: estimating a average
+treatment effect --a single number used for policy making-- and a
+CATE --a full model of the treatment effect as a function of covariates
+$X$. Given that the latter is a much more complex object than the former,
+the optimal train/test ratio might vary. To measure errors, we use for
+the ATE the relative absolute ATE bias between the ATE computed with the
+selected outcome model on the test set, and the true ATE as evaluated on
+the holdout set $\mathcal{V}$. For the CATE, we compare the
+$\tau\text{-risk}$
+of the best selected model applied on the holdout set $\mathcal{V}$. We explore this trade-off for the ACIC 2016 dataset and the R-risk.
+
+\begin{figure}[!t]
+    \begin{minipage}{.5\textwidth}
+        \centerline{\textbf{a) CATE estimation error}}
+        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16_twocols.pdf}
+    \end{minipage}
+    %\hspace{0.3\textwidth}
+    \begin{minipage}{.5\textwidth}
+        \centerline{\textbf{b) ATE estimation error}}
+        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16_twocols.pdf}
+    \end{minipage}%
+    \caption{\textbf{a) For CATE, a train/test ratio of 0.9/0.1 appears a good
+            trade-off.} b) For ATE, there is a small signal pointing also to
+        0.9/0.1 (K=10).
+        for ATE. Experiences on 10 replications of all 78 instances of the ACIC 2016
+        data.}\label{fig:train_test_ratio}
+\end{figure}
+
+Figure \ref{fig:train_test_ratio} shows that a train/test ratio of
+0.9/0.1 (K=10) or 0.8/0.2 (K=5) appears best to estimate CATE and
+ATE.
+
+\section{Discussion and conclusion}\label{sec:discussion}
+
+Predictive models are increasingly used to reason about causal effects,
+for instance in precision medicine to drive individualized decision.
+Our results highlight that they should be selected, validated, and tuned
+using different procedures and error measures than those classically used
+to assess prediction (estimating the so-called $\mu\text{-risk}$).
+Rather, selecting the best outcome model according to the $R\text{-risk}$
+(eq.\, Definition \ref{def:r_risk}) leads to more valid causal estimates.
+
+\paragraph{Nuisance models: more gain than pain}
+%
+Estimating the $R\text{-risk}$ requires a more complex procedure than
+standard cross-validation used \emph{e.g.}~in machine learning: it involves
+fitting nuisance models necessary for model evaluation, though our
+results show that these can be learned on the same set of data
+as the outcome model evaluated.
+
+The nuisance models must be well estimated (Figure
+\ref{fig:all_datasets_nuisances_comparison}). However these models are
+easier to select and control than a causally-valid outcome model,
+as they are associated to errors on
+observed distributions. Our results show that using for nuisance models
+a flexible stacking-based family of estimator suffices for good model selection.
+%
+In fact, a feasible $R\text{-risk}$ --where the nuisances
+are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
+nuisances are known. This may be explained by
+results that suggest that estimation errors on both
+nuisances partly compensate out in the
+$R\text{-risk}$ \cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.
+
+Note that propensity score models must be selected to estimate
+the individual posterior probability. For this, we used the Brier score,
+is minimized by the true individual probability. An easy mistake is to
+use calibration errors popular in machine learning
+\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
+as these select not for the individual posterior probability but for an
+aggregate error rate \cite{perez2022beyond}.
+
+\paragraph{Extension to binary outcomes}
+While we focused on continuous outcomes, in medicine, the target outcome
+is often a categorical variable such as mortality status or diagnosis. In
+this case, it may be interesting to focus on other estimands than the
+Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
+$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
+or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
+    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
+\cite{austin2017estimating}. While the odds ratio is natural for
+case-control studies \cite{rothman2008case}, other measures
+can reduce heterogeneity \cite{colnet2023risk}.
+In the log domain, the ratios are written as a difference, the
+framework studied here (\autoref{sec:neyman_rubin}) can directly apply.
+In particular, the log odds ratio is estimated by the common
+cross-entropy loss (or log loss) as in logistic regression.
+
+\paragraph{More $R\text{-risk}$ to select models driving decisions}
+
+Prediction models have flourished because their predictions can be easily
+demonstrated and validated on left-out data. But they require more
+careful validation for decision making, using a metric accounting for the
+putative intervention, the $R\text{-risk}$. Even when treated and
+untreated population differ little, as in RCTs, the $R\text{-risk}$
+brings a sizeable benefit. To facilitate better model selection, we provide Python
+code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
+Using the $R\text{-risk}$ does make evaluation more
+complicated not only because the procedure is more involved, but also
+because each intervention requires a dedicated evaluation. However, such
+off-policy evaluation remains much less costly than the recommended good
+practice of impact evaluation testing the ability of a prediction model
+to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
+the model-selection procedure puts no constraints on the models used to
+build predictive models: it opens the door to evaluating a wide range of
+models, from gradient boosting to convolutional neutral, or language
+models.
+
+\section*{Acknowledgments}
+
+We acknowledge fruitful discussions with Bénédicte Colnet.
+
+\subsection*{Financial disclosure}
+
+None reported.
+
+\subsection*{Conflict of interest}
+
+The authors declare no potential conflict of interests.
+
+%\if@Mybibrefstyle%
+\printbibliography
+%\fi if using njdnatbib.sty \bibliography{bibliography}%
+
+\clearpage
+
+\onecolumn
+
+\appendix
+%\setcounter{secnumdepth}{0}
+\setcounter{secnumdepth}{3}
+
+%The following supporting information is available as part of the online article:
+
+\section{Variability of ATE estimation on ACIC
+  2016}\label{apd:toy_example:acic_2016_ate_variability}
+
+Figure \ref{fig:acic_2016_ate_heterogeneity} shows ATE estimations for six
+different models used in g-computation estimators on the 76 configurations of
+the ACIC 2016 dataset. Outcome models are fitted on half of the data and
+inference is done on the other half --ie. train/test with a split ratio of 0.5.
+For each configuration, and each model, this train test split was repeated ten
+times, yielding non parametric variance estimates
+\cite{bouthillier_accounting_2021}. Figure \ref{fig:acic_2016_ate_heterogeneity}
+shows large variations obtained across different outcome estimators on
+semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models such as
+random forests are doing well in most settings except when treated and untreated
+populations differ noticeably, in which case a linear model (ridge) is to be
+preferred. However random forests with different hyper-parameters (max depth= 2)
+yield poor estimates. A simple rule of thumb such as preferring flexible models
+does not work in general; model selection is needed.
+
+%\idea{It is crucial to have tools for model selection.}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.95\linewidth]{images/2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
+    \caption{\textbf{Different outcome models lead to different
+            estimation errors on the Average Treatment Effects},
+        on 77 classic simulations with known true causal effect
+        \cite{dorie_automated_2019}. The different models are ridge regression
+        and random forests with different hyper-parameters
+        (details
+        \ref{apd:toy_example:acic_2016_ate_variability}). The different configurations are
+        plotted as a function of increasing difference between treated and
+        untreated population --see
+        \autoref{subsec:measuring_overlap}.
+        There is no systematic best performer; data-driven model
+        selection is important.
+        \label{fig:acic_2016_ate_heterogeneity}%
+    }
+\end{figure}
+
+Outcome models are implemented with
+\href{https://scikit-learn.org/stable/}{scikit-learn}
+\cite{pedregosa_scikitlearn_2011} and the following hyper-parameters:
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{llll}
+        \toprule
+        Outcome Model                                  & Hyper-parameters grid
+        \\
+        \midrule
+        Random Forests                                 & Max depth: [2,
+        10]                                                                    \\
+
+        Ridge regression without treatment interaction & Ridge regularization:
+        [0.1]                                                                  \\
+
+        Ridge regression with treatment interaction    & Ridge regularization:
+        [0.1]                                                                  \\
+        \bottomrule
+    \end{tabular}
+    \caption{Hyper-parameters grid used for ACIC 2016 ATE variability}
+    \label{apd:toy_example:acic_2016_ate_variability:table}
+\end{table}
+
+\section{Causal assumptions}\label{apd:causal_assumptions}
+
+We assume the following four assumptions, referred as strong ignorability and
+necessary to assure identifiability of the causal estimands with observational
+data \cite{rubin_causal_2005}:
+\begin{assumption}[Unconfoundedness]\label{assumption:ignorability}
+    \begin{equation*}\label{eq:ignorability}
+        \{Y(0), Y(1) \} \indep A | X
+    \end{equation*}
+    This condition --also called ignorability-- is equivalent to the conditional
+    independence on $e(X)$ \cite{rosenbaum_central_1983}: $\{Y(0), Y(1) \}
+        \indep  A | e(X)$.
+\end{assumption}
+
+\begin{assumption}[Overlap, also known as Positivity)]\label{assumption:overlap}
+    \begin{equation*}\label{eq:overlap}
+        \eta < e(x) < 1 - \eta \quad \forall x \in \mathcal X \text{ and some } \eta > 0
+    \end{equation*}
+    The treatment is not perfectly predictable. Or with different words, every
+    patient has a chance to be treated and not to be treated. For a given set of
+    covariates, we need examples of both to recover the ATE.
+\end{assumption}
+
+As noted by \cite{damour_overlap_2020}, the choice of covariates $X$ can
+be viewed as a trade-off between these two central assumptions. A bigger
+covariates set generally reinforces the ignorability assumption. In the
+contrary, overlap can be weakened by large $\mathcal{X}$ because of the
+potential inclusion of instruments: variables only linked to the treatment which
+could lead to arbitrarily small propensity scores.
+
+% remark: There is a major counter example to these colliders (variables which
+% are caused by both the outcome and the treatment),
+
+\begin{assumption}[Consistency]\label{assumption:consistency} The observed
+    outcome is the potential outcome of the assigned treatment:
+    \begin{equation*}\label{eq:consistancy}
+        Y = A \, Y(1) + (1-A) \, Y(0)
+    \end{equation*}
+    Here, we assume that the intervention $A$ has been well defined. This
+    assumption focuses on the design of the experiment. It clearly states the link
+    between the observed outcome and the potential outcomes through the
+    intervention \cite{hernan_causal_2020}.
+\end{assumption}
+
+\begin{assumption}[Generalization]\label{assumption:generalization} The training
+    data on which we build the estimator and the test data on which we make the
+    estimation are drawn from the same distribution $\mathcal D^*$, also known as
+    the ``no covariate shift'' assumption \cite{jesson_identifying_2020}.
+\end{assumption}
+
+\section{Proofs: Links between feasible and oracle risks}\label{apd:proofs}
+
+\subsection{Upper bound of $\tau\text{-risk}$ with
+    $\mu\text{-risk}_{IPW}$}\label{apd:proofs:mu_risk_ipw_bound}
+
+For the bound with the $\mu\text{-risk}_{IPW}$, we will decompose the CATE risk
+on each factual population risks:
+
+\begin{definition}[Population Factual $\mu\text{-risk}$]\label{mu_risk_a}
+    \cite{shalit_estimating_2017}
+    \begin{equation*}
+        \mu\text{-risk}_{a}(f)= \int_{\mathcal Y \times \mathcal X} (y-f(x ; A=a))^{2}  p(y ; x=x \mid A=a) \; dy dx
+    \end{equation*}
+\end{definition}
+
+Applying Bayes rule, we can decompose the $\mu\text{-risk}$ on each
+intervention:
+\begin{equation*}
+    \mu\text{-risk}(f)
+    =p_{A} \,\mu\text{-risk}_{1}(f)+\left(1-p_{A}\right) \,\mu\text{-risk}_{0}(f)
+    \text{with } p_A=\mathbb P(A=1)
+\end{equation*}
+
+These definitions allows to state a intermediary result on each population:
+\begin{lemma}[Mean-variance decomposition]\label{apd:proofs:mu_risk_ipw_link_mu}
+    We need a reweighted version of the classical mean-variance decomposition.
+
+    For an outcome model $f: x \times A \rightarrow \mathcal X$. Let the inverse
+    propensity weighting function $w(a ; x)=a e(x)^{-1}+(1-a)(1-e(x))^{-1}$.
+    \begin{align*}
+         & \int_{\mathcal X}(\mu_{1}(x)-f(x ; 1))^{2} p(x) dx  = p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  -\sigma^{2}_{Bayes}(1)
+    \end{align*}
+    And
+    \begin{align*}
+         & \int_{\mathcal X}(\mu_{0}(x)-f(x; 0))^{2} p(x) dx   = (1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)  -\sigma^{2}_{Bayes}(0)
+    \end{align*}
+
+    \begin{proof}
+        \begin{align*}
+             & p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  = \int_{\mathcal X \times \mathcal Y} \frac{1}{e(x)}(y-f(x ; 1))^{2} p(y \mid x ; A=1) p(x ; A=1) d y d x                                                      \\
+             & = \int_{\mathcal X \times \mathcal Y} (y-f(x ; 1))^{2} p(y \mid x ; A=1) \frac{p(x ; A=1)}{p(x ; A=1)}p(x)dy dx                                                                                      \\
+             & = \int_{\mathcal X \times \mathcal Y} \big[(y-\mu_1(x))^{2}+\left(\mu_{1}(x)-f(x ; 1)\right)^{2} + 2\left(y-\mu_{1}(x)\right)\left(\mu_{1}(x)-f(x, 1)\right) \big] p(y \mid x ; A=1) p(x) d y d x    \\
+             & =\int_{\mathcal X} \big [ \int_{\mathcal Y} (y-\mu_1(x))^{2} p(y \mid x ; A=1) dy\big ] p(x)dx + \int_{\mathcal X \times \mathcal Y} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x)p(y \mid x ; A=1)dx dy \\
+             & + \qquad 2 \int_{\mathcal X} \big [ \int_{\mathcal Y} \left(y-\mu_{1}(x)\right) p(y \mid x ; A=1) dy \big ] \left(\mu_{1}(x)-f(x, 1)\right)p(x)dx                                                    \\
+             & =\int_{\mathcal X} \sigma_{y}^{2}(x, 1) p(x) d x +\int_{\mathcal X} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x) d x+0
+        \end{align*}
+    \end{proof}
+\end{lemma}
+
+\begin{proposition*}[Upper bound with mu-IPW]\label{apd:proofs:prop:upper_bound}
+    Let f be a given outcome model, let the weighting function $w$ be the Inverse
+    Propensity Weight $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$. Then, under
+    overlap (assumption \ref{assumption:overlap}),
+    \begin{align*}
+        \tau\text{-risk}(f) \leq & \; 2 \, \mu\text{-risk}_{IPW}(w, f)  \; - 2 \, (\sigma^2_{Bayes}(1) +  \sigma^2_{Bayes}(0))
+    \end{align*}
+
+    \begin{proof}
+        \begin{align*}
+             & \tau\text{-risk}(f) =\int_{\mathcal X}(\mu_{1}(x)-\mu_{0}(x)-(f(x ; 1)-f(x ; 0))^{2} p(x) d x
+        \end{align*}
+        By the triangle inequality $(u+v)^2 \leq 2(u^2 + v^2)$:
+        \begin{align*}
+             & \tau\text{-risk}(f) \leq
+            2 \int_{\mathcal X}\big[\left(\mu_{1}(x)-f(x ; 1)\right)^{2}+\left(\mu_{0}(x)-f(x ; 0)\right)^{2}\big] p(x) d x \\
+        \end{align*}
+        Applying Lemma \ref{apd:proofs:mu_risk_ipw_link_mu},
+        \begin{align*}
+             & \tau\text{-risk}(f) \leq 2\big[p_A \mu\text{-risk}_{IPW, 1}(w, f) +(1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)(w, f)\big] -2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1)) \\
+             & = 2 \mu\text{-risk}_{IPW}(w, f)-2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1))
+        \end{align*}
+    \end{proof}
+\end{proposition*}
+
+\subsection{Reformulation of the $R\text{-risk}$ as reweighted
+    $\tau\text{-risk}$}\label{apd:proofs:r_risk_rewrite}
+
+\begin{proposition*}[$R\text{-risk}$ as reweighted $\tau
+            \text{-risk}$]\label{apd:proofs:prop:r_risk_rewrite}
+
+    \begin{proof}
+
+        We consider the R-decomposition: \cite{robinson_rootnconsistent_1988},
+        \begin{equation}\label{apd:eq:r_decomposition}
+            y(a) = m(x) + \big( a - e(x) \big) \tau(x) + \varepsilon(x; a)
+        \end{equation}
+        Where $\mathbb E[\varepsilon(X; A)|X, A] = 0$ We can use it as plug in the
+        $R\text{-risk}$ formula:
+
+        \begin{align*}
+             & R\text {-risk}(f) =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}[(y-m(x))-\big(a-e(x)\big) \tau_f(x)]^{2} p(y ; x ; a) d y d x d a                                     \\
+             & =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}} \left[\big(a-e(x)\big)\tau(x)+\varepsilon(x ; a)-\big(a-e(x)\big) \tau_f(x)\right]^{2} p(y ; x ; a) d y d x da          \\
+             & =\int_{\mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)^{2}\big(\tau(x)- \tau_f(x)\big)^{2} p(x ; a) d x d a                                                                       \\
+             & + 2  \int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)\big(\tau(x)-\tau_f(x)\big)  \int_{\mathcal{Y}} \varepsilon(x ; a) p(y \mid x ; a) d y p(x ; a)dx da \\
+             & +\int_{\mathcal{X} \times \mathcal{A}} \int_{\mathcal{Y}} \varepsilon^{2}(x ; a) p(y \mid x ; a) d y p(x ; a) d x d a
+        \end{align*}
+
+        The first term can be decomposed on control and treated populations to force
+        $e(x)$ to appear:
+        \begin{align*}
+             & \int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}\left[e(x)^{2}p(x;0) + \big(1-e(x)\big)^{2} p(x;1)\right] d x                    \\
+             & =\int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}  \left[e(x)^{2}\big(1-e(x)\big)p(x) + \big(1-e(x)\big)^{2}e(x) p(x)\right] d x \\
+             & =\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x)[1-e(x)+e(x)] p(x) d x                                                     \\ &=\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x) p(x) d x.
+        \end{align*}
+
+        The second term is null since, $\mathbb E[\varepsilon(x, a) |X, A]=0$.
+
+        The third term corresponds to the modulated residuals \ref{eq:residuals} :
+        $\tilde{\sigma}_B^2(0) + \tilde{\sigma}_B^2(1)$
+
+    \end{proof}
+\end{proposition*}
+
+\section{Measuring overlap}\label{apd:motivation_ntv}
+
+\paragraph{Motivation of the Normalized Total Variation}
+%\idea{A simplier measure, NTV seems to capture what we want}
+Computing overlap when working only on samples of the observed distribution,
+outside of simulation, requires a sophisticated estimator of discrepancy
+between distributions, as two data points never have the same exact set of
+features. Maximum Mean Discrepancy \cite{gretton2012kernel} is typically
+used in the context of causal inference
+\cite{shalit_estimating_2017,johansson2022generalization}. However it
+needs a kernel, typically Gaussian, to extrapolate across neighboring
+observations. We prefer avoiding the need to specify such a kernel, as it must
+be adapted to the data which is tricky with categorical or non-Gaussian
+features, a common situation for medical data.
+
+For simulated and some semi-simulated data, we have access to the probability of
+treatment for each data point, which sample both densities in the same data
+point. Thus, we can directly use distribution discrepancy measures and rely on
+the Normalized Total Variation (NTV) distance to measure the overlap between the
+treated and control propensities. This is the empirical measure of the total
+variation distance \cite{sriperumbudur_integral_2009} between the distributions,
+$TV(\mathbb{P}(X|A=1), \mathbb{P}(X|A=0))$. As we have both distribution sampled
+on the same points, we can rewrite it a sole function of the propensity score, a
+low dimensional score more tractable than the full distribution $\mathbb
+    P(X|A)$:
+
+\begin{equation}\label{eq:ntv}
+    \widehat{NTV}(e, 1-e) = \frac{1}{2N} \sum_{i =1}^{N} \big |\frac{e(x_i)}{p_A}-\frac{1-e(x_i)}{1-{p_A}} \big |
+\end{equation}
+
+Formally, we can rewrite NTV as the Total Variation distance
+between the two population distributions. For a population $O = (Y(A), X, A)
+    \sim \mathcal{D}$:
+
+\begin{align*}
+    NTV(O) & = \frac{1}{2N} \sum_{i =1}^{N} \big | \frac{e(x_i)}{p_A}-\frac{1-e(x_i)}{1-{p_A}} \big|         \\
+           & = \frac{1}{2N} \sum_{i =1}^{N} \big|\frac{P(A=1|X=x_i)}{p_A}-\frac{P(A=0|X=x_i)}{1-{p_A}} \big|
+\end{align*}
+
+Thus NTV approximates the following quantity in expectation over the data
+distribution $\mathcal{D}$:
+
+\begin{align*}
+    NTV(\mathcal{D}) & = \int_{\mathcal{X}} \big | \frac{p(A=1|X=x)}{p_A}-\frac{p(A=0|X=x)}{1-{p_A}} \big | p(x)dx \\
+                     & = \int_{\mathcal{X}} \big | \frac{p(A=1, X=x)}{p_A}-\frac{p(A=0, X=x)}{1-{p_A}} \big |dx    \\
+                     & = \int_{\mathcal{X}} \big | p(X=x|A=1)- p(X=x|A=0) \big |dx
+\end{align*}
+
+For countable sets, this expression corresponds to the Total Variation distance
+between treated and control populations covariate distributions : $TV(p_0(x),
+    p_1(x))$.
+
+\paragraph{Measuring overlap without the oracle propensity scores:} For ACIC
+2018, or for non-simulated data, the true propensity scores are not known. To
+measure overlap, we rely on flexible estimations of the Normalized Total
+Variation, using gradient boosting trees to approximate the propensity score.
+Empirical arguments for this plug-in approach is given in Figure
+\ref{apd:overlap:ntv_approximation}.
+
+\paragraph{Empirical arguments}
+
+We show empirically that NTV is an appropriate measure of overlap by :
+\begin{itemize}
+    \item Comparing the NTV distance with the MMD for Caussim which is gaussian
+          distributed in Figure \ref{apd:overlap:caussim:mmd_vs_ntv},
+    \item Verifying that setups with penalized overlap from ACIC 2016 have a
+          higher total variation distance than unpenalized setups in Figure
+          \ref{apd:overlap:penalized_overlap}.
+    \item Verifying that the Inverse Propensity Weights extrema (the inverse of
+          the $\nu$ overlap constant appearing in the overlap Assumption
+          \ref{assumption:overlap}) positevely correlates with NTV for Caussim,
+          ACIC 2016 and Twins in Figure \ref{apd:ntv_vs_max_ipw}. Even if the same
+          value of the maximum IPW could lead to different values of NTV, we
+          expect both measures to be correlated : the higher the extrem propensity
+          weights, the higher the NTV.
+\end{itemize}
+
+\paragraph{Estimating NTV in practice}
+
+Finally, we verify that approximating the NTV distance with a learned plug-in
+estimates of e(x) is reasonnable. We used either a logistic regression or a
+gradient boosting classifier to learn the propensity models for the three
+datasets where we have access to the ground truth propensity scores: Caussim,
+Twins and ACIC 2016. We respectively sampled 1000, 1000 and 770 instances of
+these datasets with different seeds and overlap settings. We first run a
+hyperparameter search with cross-validation on the train set, then select the
+best estimator. We refit on the train set this estimator with or without
+calibration by cross validation and finally estimate the normalized TV with the
+obtained model. This training procedure reflects the one described in Algorithm
+\ref{problem:estimation_procedure:algo} where nuisance models are fitted only on
+the train set.
+
+The hyper parameters are : learning rate $ \in [1e-3, 1e-2, 1e-1, 1]$, minimum
+samples leaf $\in [2, 10, 50, 100, 200]$ for boosting and L2 regularization $\in
+    [1e-3, 1e-2, 1e-1, 1]$ for logistic regression.
+
+Results in Figure \ref{apd:overlap:ntv_approximation} comparing bias to the true
+normalized Total Variation of each dataset instances versus growing true NTV
+indicate that calibration of the propensity model is crucial to recover a good
+approximation of the NTV.
+
+\begin{figure}
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \caption{\textbf{Uncalibrated classifiers}}
+        \includegraphics[width=\linewidth]{images/overlap_measure_diff_oracle_ntv_to_n_tv_calibration=False_vs_oracle_n_tv.pdf}
+    \end{subfigure}
+    \begin{subfigure}[b]{\textwidth}
+        \centering
+        \caption{\textbf{Calibrated
+                classifiers}}
+        \includegraphics[width=\linewidth]{images/overlap_measure_diff_oracle_ntv_to_n_tv_calibration=True_vs_oracle_n_tv.pdf}
+    \end{subfigure}
+    \caption{a) Without calibration, estimation of NTV is not trivial even for
+        boosting models. b) Calibrated classifiers are able to recover the true
+        Normalized Total Variation for all datasets where it is
+        available.}\label{apd:overlap:ntv_approximation}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \caption{NTV recovers well the overlap settings described in the ACIC paper
+        \cite{dorie_automated_2019}}\label{apd:overlap:penalized_overlap}
+    \includegraphics[width=0.7\linewidth]{images/overlap_measure_acic_2016_recovery_overlap_setup.pdf}
+\end{figure}
+
+\begin{figure}[htbp]
+    \centering
+    \caption{Good correlation between overlap measured as normalized Total
+        Variation and Maximum Mean Discrepancy (200 sampled Caussim
+        datasets)}\label{apd:overlap:caussim:mmd_vs_ntv}
+    \includegraphics[width=0.7\linewidth]{images/overlap_measure_caussim_transformed_mmd_vs_ntv.pdf}
+\end{figure}
+
+\begin{figure}[ht]
+    \centering
+    \begin{subfigure}[b]{0.47\textwidth}
+        \centering
+        \caption{\textbf{Caussim}}
+        \includegraphics[width=\textwidth]{images/overlap_measure_caussim_max_ipw_vs_ntv.pdf}
+        \label{apd:caussim:max_ipw_vs_ntv}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.47\textwidth}
+        \centering
+        \caption{\textbf{ACIC 2016}}
+        \includegraphics[width=\textwidth]{images/overlap_measure_acic_2016_max_ipw_vs_ntv.pdf}
+        \label{apd:acic_2016:ntv_vs_max_ipw}
+    \end{subfigure}
+    \begin{subfigure}[b]{0.47\textwidth}
+        \centering
+        \caption{\textbf{ACIC 2018}}
+        \includegraphics[width=\textwidth]{images/overlap_measure_acic_2018_max_ipw_vs_ntv.pdf}
+        \label{apd:acic_2018:ntv_vs_max_ipw}
+    \end{subfigure}
+    \hfill
+    \begin{subfigure}[b]{0.49\textwidth}
+        \centering
+        \caption{\textbf{TWINS}}
+        \includegraphics[width=\textwidth]{images/overlap_measure_twins_max_ipw_vs_ntv.pdf}
+        \label{apd:twins:ntv_vs_max_ipw}
+    \end{subfigure}
+    \caption{Maximal value of Inverse Propensity Weights increases exponentially with the overlap as measure by Normalized Total Variation.}
+    \label{apd:ntv_vs_max_ipw}
+\end{figure}
+
+\section{Experiments}
+
+\subsection{Details on the data generation process}
+\label{apd:experiments:generation}
+
+We use Gaussian-distributed covariates and random basis expansion based on
+Radial Basis Function kernels. A random basis of RBF kernel enables modeling
+non-linear and complex relationships between covariates in a similar way to the
+well known spline expansion. The estimators of the response function are learned
+with a linear model on another random basis (which can be seen as a stochastic
+approximation of the full data kernel \cite{rahimi_random_2008}). We
+carefully control the amount of overlap between treated and control populations,
+a crucial assumption for causal inference.
+
+\begin{itemize}
+    \item The raw features for both populations are drawn from a mixture of
+          Gaussians:
+          $\mathbb P(X) = p_A \mathbb P(X|A=1) + (1- p_A) \mathbb P(X|A=0)$
+          where $\mathbb P(x|A=a)$ is a rotated Gaussian:
+          \begin{equation}
+              \mathbb P(x|A=a) = W \cdot \mathcal N \Big( \begin{bmatrix} (1-2a) \theta \\ 0\end{bmatrix} ; \begin{bmatrix} \sigma_0 & 0 \\ 0 & \sigma_1\end{bmatrix} \Big)
+          \end{equation}
+          with $\theta$ a parameter controlling overlap (bigger yields poorer
+          overlap), $W$ a random rotation matrix and $\sigma_0^2=2;\sigma_1^2=5$.
+
+          This generation process allows to analytically compute the oracle
+          propensity scores $e(x)$, to simply control for overlap with the
+          parameter $\theta$, the distance between the two Gaussian main axes and
+          to  visualize response surfaces.
+
+    \item A basis expansion of the raw features increases the problem dimension.
+          Using Radial Basis Function (RBF) Nystroem transformation \footnote{We
+          use the
+          \href{https://scikit-learn.org/stable/modules/generated/sklearn.kernel_approximation.Nystroem.html}{Sklearn
+          implementation, \cite{pedregosa_scikitlearn_2011}}}, we expand the raw
+          features into a transformed space. The basis expansion samples
+          randomly a small number of representers in the raw data. Then,  it
+          computes an approximation of the full N-dimensional kernel with these
+          basis components, yielding the transformed features $z(x)$. The number
+          of basis functions --\emph{ie. knots}--, controls the complexity of
+          the ground-truth response surfaces and treatment. We first use this
+          process to draw the non-treated response surface $\mu_0$ and the
+          causal effect $\tau$. We then draw the observations from a mixture two
+          Gaussians, for the treated and non treated. We vary the separation
+          between the two Gaussians to control the overlap between treated and
+          non-treated populations, an important parameter for causal inference
+          (related to $\eta$ in Proposition
+          \ref{theory:prop:mu_risk_ipw_bound}). Finally, we generate observed
+          outcomes adding Gaussian noise.
+
+          More formally, we generate the basis following the original data
+                  distribution, $\left [ b_1 .. b_D \right ] \sim \mathbb P(x)$,
+                  with D=2 in our simulations. Then, we compute an approximation
+                  of the full kernel of the data generation process $RBF(x,
+                  \cdot) \;  with \; x \sim \mathbb P(x)$ with these
+                  representers: $z(x) = [RBF_{\gamma}(x, b_d)]_{d=1..D} \cdot
+                  Z^T \in \mathbb{R}^D$ with $RBF_{\gamma}$ being the Gaussian
+                  kernel $K(x, y) = exp(-\gamma ||x-y||^2)$ and Z the
+                  normalization constant of the kernel basis, computed as the
+                  root inverse of the basis kernel $Z=[K(b_i, b_j)]_{i, j \in
+                  {1..D}}^{-1/2}$
+
+    \item Functions $\mu_0$, $\tau$ are distinct linear functions of the
+          transformed features:
+          \begin{equation*}
+              \mu_0(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\mu}^T
+          \end{equation*}
+          \begin{equation*}
+              \tau(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\tau}^T
+          \end{equation*}
+    \item Adding a Gaussian noise, $\varepsilon \sim \mathcal N(0, \sigma(x;a))$,
+          we construct the potential outcomes:
+          $y(a) = \mu_0(x) + a\,\tau(x) + \varepsilon(x, a)$
+\end{itemize}
+We generated 1000 instances of this dataset with uniformly random overlap
+parameters $\theta \in \left[ 0, 2.5 \right]$.
+
+\subsection{ Details on the semi-simulated datasets}\label{apd:experiments:datasets}
+
 \begin{description}
-    \item[ACIC 2016] \cite{dorie_automated_2019}: The dataset is based on the
-        Collaborative Perinatal Project \cite{niswander_women_1972}, a RCT
-        studying infants’
-        developmental disorders. The initial intervention was a child’s birth
+    \item[ACIC 2018] \cite{dorie_automated_2019}:  The initial intervention was a child’s birth
         weight $(A = 1 \text{ if weight} < 2.5 kg)$, and outcome was the child’s
@@
             \url{https://github.com/vdorie/aciccomp/tree/master/2016}
-            to generate 77 simulations settings.}. We used 10 different seeds for
-        each setup, totaling 770 dataset instances.
-
-    \item[ACIC 2018] \cite{shimoni_benchmarking_2018}: Starting from data
-        from the Linked Births and Infant Deaths Database (LBIDD)
+            to generate 77 simulations settings.}. We used 10 different seeds
+        for each setup, totaling 770 dataset instances.
+
+    \item[ACIC 2018] \cite{shimoni_benchmarking_2018}: Starting from data from
+        the Linked Births and Infant Deaths Database (LBIDD)
         \cite{macdorman_infant_1998} with $D=177$ covariates, treatment and
-        outcome models are simulated with complex models to reflect
-        different scenarii. The data do not provide the true propensity
-        scores, so we evaluate only feasible metrics, which do not require this
-        nuisance parameter. We used all 432 datasets\footnote{Using the scaling part of the data, from
-            \href{https://github.com/IBM-HRL-MLHLS/IBM-Causal-Inference-Benchmarking-Framework}{github.com/IBM-HRL-MLHLS/IBM-Causal-Inference-Benchmarking-Framework}} of size $N=5\,000$.
+        outcome models are simulated with complex models to reflect different
+        scenarii. The data do not provide the true propensity scores, so we
+        evaluate only feasible metrics, which do not require this nuisance
+        parameter. We used all 432 datasets\footnote{Using the scaling part of
+            the data, from
+            \href{https://github.com/IBM-HRL-MLHLS/IBM-Causal-Inference-Benchmarking-Framework}{github.com/IBM-HRL-MLHLS/IBM-Causal-Inference-Benchmarking-Framework}}
+        of size $N=5\,000$.
 
@@
 
-\paragraph{Family of candidate
-    estimators}\label{semi_simulated:candidate_estimators}
-
-For these three datasets, the family of candidate estimators are gradient
-boosting trees for both the response surfaces and the treatment
-\footnote{Scikit-learn regressor,
-    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingRegressor.html}{HistGradientBoostingRegressor},
-    and classifier,
-    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingClassifier.html}{HistGradientBoostingClassifier}.}
-with S-learner, learning rate
-in $\{0.01, 0.1, 1\}$, and maximum number of leaf nodes in $\{25, 27, 30, 32,
-    35, 40\}$ resulting in a family of size 18.
-
-\paragraph{Nuisance estimators}
-
-Drawing inspiration from the TMLE literature that uses combination of flexible
-machine learning methods \cite{schuler_targeted_2017}, we use as models
-for the nuisances $\check e$ (respectively $\check m$) a form of meta-learner: a
-stacked estimator of ridge and boosting classifiers (respectively
-regressions). We select hyper-parameters with randomized search on a validation set
-$\mathcal{V}$ and keep them fix for model selection
-(\ref{apd:experiments:nuisances_hp} lists hyperparameters). As extreme inverse
-propensity weights induce high variance, we use clipping
-\cite{swaminathan_counterfactual_2015, ionides_truncated_2008} to bound
-$min(\check e, 1-\check e)$ away from 0 with a fixed $\eta=10^{-10}$, ensuring
-strict overlap for numerical stability.
-
-%\paragraph{Model selection procedure}\label{semi_simulated:selection_procedure}
-
-%We used the same training procedure procedure as in
-%\ref{paragraph:selection_procedure} (Details in supp.~mat.
-%\ref{apd:selection_procedure_acic}).
-
-\subsection{Measuring overlap between treated and non treated}\label{subsec:measuring_overlap}
-
-%\idea{Overlap is important but not easily measurable}
-Good overlap between treated and control population is
-crucial for causal inference as it is required by the positivity assumption
-\ref{assumption:overlap}. It is often assessed by
-comparing visually population distributions (as in Figure
-\ref{fig:toy_example}) or computing standardized difference on each
-feature
-\cite{austin_introduction_2011,austin_moving_2015}. While these methods are
-useful to decide if positivity holds, they do not yield a
-single measure. Rather, we compute the divergence
-between the population covariate distributions $\mathbb P(X|A=0)$ and $\mathbb
-    P(X|A=1)$
-\cite{damour_overlap_2020,johansson2022generalization}. We introduce the
-Normalized Total Variation (NTV), a divergence based on the sole propensity
-score (see \ref{apd:motivation_ntv}).
-
-\begin{figure}[!b]
-    \centering
-    \includegraphics[width=\linewidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
-    %\hfill
-    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$.
-        Strong and Weak overlap correspond to the first and last tertiles of the overlap distribution measured with
-        Normalized Total Variation eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents the same results
-        by adding semi-oracle risks in Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics}, measured with
-        absolute Kendall's in Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with $\tau\mathrm{-risk}$
-        gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table
-        \ref{apd:table:relative_kendalls_all_datasets} gives median and
-        IQR of the relative Kendall.}\label{fig:relative_kendalls_all_datasets}
-\end{figure}
-
-\subsection{Results: factors driving good model selection}\label{empirical_study:results}
-
-\paragraph{The $R\text{-risk}$ is the best metric}
-
-Each metric ranks differently the candidate models. Figure
-\ref{fig:relative_kendalls_all_datasets} shows the agreement between the
-ideal ranking of methods given the oracle $\tau\text{-risk}$ and
-the different feasible causal metrics. We measure this agreement
-with a relative\footnote{To remove the variance
-    across datasets (some datasets lead to easier model selection than
-    others), we report values for one metric relative to the mean of all
-    metrics for a given dataset instance: $\text{Relative} \, \kappa(\ell,\tau\mathrm{{-risk}})=
-        \kappa(\ell,\tau\mathrm{{-risk}}) -
-        mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$ (eq.
-\ref{eq:kendall_tau}) \cite{kendall_new_1938}.
-Given the importance of overlap in
-how well metrics approximate the oracle $\tau\text{-risk}$
-(\ref{apd:proofs:mu_risk_ipw_bound}), we separate
-strong and weak overlap.
-
-Among all metrics, the classical mean squared error (ie. factual
-$\mu\text{-risk}$) is worse and reweighting it with propensity score
-($\mu\text{-risk}_{IPW}$) does not bring much improvements. The
-$R\text{-risk}$, which includes a model of mean outcome and propensity
-scores, leads to the best performances. Interestingly, the
-$U\text{-risk}$, which uses the same nuisances, deteriorates in weak overlap, probably due to variance
-inflation when dividing by extreme propensity scores.
-
-Beyond rankings, the differences in terms of absolute
-ability to select the best model are large: The R-risk selects a model
-with a $\tau\text{-risk}$ only 1\% higher
-than the best
-possible candidate for strong overlap on Caussim, but selecting with
-the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
-practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
---as in some causal-inference methods \citep{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
-(Figure
-\ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
-datasets, the $R\text{-risk}$ consistently decreases the
-risk compared to the $\mu\text{-risk}$:
-0.1\% compared to 1\% on ACIC2016,  1\% compared to 20\% on ACIC2018,
-and 0.05\% compared to 1\% on Twins.
-
-\paragraph{Model selection is harder for low population
-    overlap}
-
-Model selection for
-causal inference becomes more and more difficult with increasingly different
-treated and control populations
-(Figure \ref{fig:all_datasets_overlap_effect_r_risk}). The
-absolute Kendall's coefficient correlation with $\tau\text{-risk}$
-drops from values around 0.9 (excellent agreement with oracle selection) to 0.6 on
-both Caussim and ACIC 2018
-(\ref{apd:experiments:additional_results}).
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=\linewidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
-    %\hfill
-    \caption{\textbf{Model selection is harder for low population
-            overlap}:
-        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
-        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents results for all
-        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
-        Kendall's and continuous overlap values in Figure
-            {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
-\end{figure}
-
-\paragraph{Nuisances can be estimated on the same data as outcome models}
-
-Using the train set $\mathcal{T}$ both to fit the candidate estimator and the
-nuisance estimates is a form of double dipping which can lead errors in
-nuisances correlated to that of outcome models
-\cite{nie_quasioracle_2017}. In theory, these correlations can bias model
-selection and, strictly speaking, push
-to split out a third separated data set --a ``nuisance set''-- to fit the
-nuisance models. The drawback is that it depletes the data available for
-model estimation and selection. However, Figure
-\ref{fig:procedures_comparison} shows no substantial difference between a procedure with a separated
-nuisance set and the simpler shared nuisance-candidate set procedure.
-
-\begin{figure}[!tb]
-    \includegraphics[width=\linewidth]{images/_3_procedure_r_risk_only_3datasets_twocols.pdf}
-
-    \caption{\textbf{Nuisances can be estimated on the same data as outcome
-            models}: Results for the R-risk are similar between the
-        \textcolor{MidnightBlue}{shared
-            nuisances/candidate set} and
-        the \textcolor{RedOrange}{separated nuisances set} procedures. Figure
-        \ref{apd:fig:procedures_comparison_all_metrics} details results for all metrics.}\label{fig:procedures_comparison}
-\end{figure}
-
-\paragraph{Stacked models are good overall estimators of nuisances}
-
-For every risk, the oracle version recovers better the best estimator.
-However,
-stacked nuisances estimators (boosting and linear) lead to feasible
-metrics with close performances to the oracles ones: the
-corresponding estimators recover well-enough the true nuisances.
-One may wonder if simpler models for the nuisance could be useful,
-in particular in data-poor settings or when the true models are linear.
-Figure \ref{fig:all_datasets_nuisances_comparison} compares causal model
-selection estimating nuisances with stacked estimators or linear model.
-It comprises the Twins data, where the true propensity model is linear,
-and a downsampled version of this data, to study a situation favorable to
-linear models. In these settings,
-stacked and linear estimations of the nuisances performs equivalently.
-Detailed analysis (Figure \ref{apd:fig:nuisances_comparison_twins})
-confirms that using adaptive models --as built by
-stacking linear models and gradient-boosted trees-- suffices to estimate nuisance.
-
-\begin{figure}[!tb]
-    \includegraphics[width=\linewidth]{images/_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
-
-    \caption{\textbf{\textcolor{DarkOrchid}{Stacked
-                models} are good overall estimators of the nuisances}:
-        Results are shown only for the
-        R-risk; Figure \ref{apd:fig:nuisances_comparison}
-        details every metrics. For Twins, where the true propensity
-        model is linear, \textcolor{DarkOrchid}{stacked} and
-        \textcolor{ForestGreen}{linear}
-        estimations of the nuisances performs equivalently, even for a downsampled version
-        (N=4794). }\label{fig:all_datasets_nuisances_comparison}
-\end{figure}
-
-\paragraph{Use 90\% of the data to estimate outcome models, 10\% to
-    select them}
-
-The analyst faces a compromise: given a finite
-data sample, should she allocate more data to estimate the outcome model,
-thus improving the quality of the outcome model but leaving
-little data for model selection. Or, she could choose a bigger test set for
-model selection and effect estimation. For causal model selection, there
-is no established practice (as reviewed in \ref{apd:results:k_fold_choices}).
-
-We investigate such tradeoff varying the ratio between train and test
-data size. For this, we first split out 30\% of the data as a holdout set
-$\mathcal{V}$ on which we use the oracle response functions to derive
-silver-standard estimates of causal quantities. We then
-use the standard estimation procedure on the remaining 70\% of the data,
-splitting it into train $\mathcal{T}$ and test $\mathcal{S}$ of varying
-sizes. We finally measure the error between this estimate and the
-silver standard.
-
-We consider two different analytic goals: estimating a average
-treatment effect --a single number used for policy making-- and a
-CATE --a full model of the treatment effect as a function of covariates
-$X$. Given that the latter is a much more complex object than the former,
-the optimal train/test ratio might vary. To measure errors, we use for
-the ATE the relative absolute ATE bias between the ATE computed with the
-selected outcome model on the test set, and the true ATE as evaluated on
-the holdout set $\mathcal{V}$. For the CATE, we compare the
-$\tau\text{-risk}$
-of the best selected model applied on the holdout set $\mathcal{V}$. We explore this trade-off for the ACIC 2016 dataset and the R-risk.
-
-\begin{figure}[!t]
-    \begin{minipage}{.5\textwidth}
-        \centerline{\textbf{a) CATE estimation error}}
-        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16_twocols.pdf}
-    \end{minipage}
-    %\hspace{0.3\textwidth}
-    \begin{minipage}{.5\textwidth}
-        \centerline{\textbf{b) ATE estimation error}}
-        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16_twocols.pdf}
-    \end{minipage}%
-    \caption{\textbf{a) For CATE, a train/test ratio of 0.9/0.1 appears a good
-            trade-off.} b) For ATE, there is a small signal pointing also to
-        0.9/0.1 (K=10).
-        for ATE. Experiences on 10 replications of all 78 instances of the ACIC 2016
-        data.}\label{fig:train_test_ratio}
-\end{figure}
-
-Figure \ref{fig:train_test_ratio} shows that a train/test ratio of
-0.9/0.1 (K=10) or 0.8/0.2 (K=5) appears best to estimate CATE and
-ATE.
-
-\section{Discussion and conclusion}\label{sec:discussion}
-
-Predictive models are increasingly used to reason about causal effects,
-for instance in precision medicine to drive individualized decision.
-Our results highlight that they should be selected, validated, and tuned
-using different procedures and error measures than those classically used
-to assess prediction (estimating the so-called $\mu\text{-risk}$).
-Rather, selecting the best outcome model according to the $R\text{-risk}$
-(eq.\, Definition \ref{def:r_risk}) leads to more valid causal estimates.
-
-\paragraph{Nuisance models: more gain than pain}
-%
-Estimating the $R\text{-risk}$ requires a more complex procedure than
-standard cross-validation used \emph{e.g.}~in machine learning: it involves
-fitting nuisance models necessary for model evaluation, though our
-results show that these can be learned on the same set of data
-as the outcome model evaluated.
-
-The nuisance models must be well estimated (Figure
-\ref{fig:all_datasets_nuisances_comparison}). However these models are
-easier to select and control than a causally-valid outcome model,
-as they are associated to errors on
-observed distributions. Our results show that using for nuisance models
-a flexible stacking-based family of estimator suffices for good model selection.
-%
-In fact, a feasible $R\text{-risk}$ --where the nuisances
-are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
-nuisances are known. This may be explained by
-results that suggest that estimation errors on both
-nuisances partly compensate out in the
-$R\text{-risk}$ \cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.
-
-Note that propensity score models must be selected to estimate
-the individual posterior probability. For this, we used the Brier score,
-is minimized by the true individual probability. An easy mistake is to
-use calibration errors popular in machine learning
-\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
-as these select not for the individual posterior probability but for an
-aggregate error rate \cite{perez2022beyond}.
-
-\paragraph{Extension to binary outcomes}
-While we focused on continuous outcomes, in medicine, the target outcome
-is often a categorical variable such as mortality status or diagnosis. In
-this case, it may be interesting to focus on other estimands than the
-Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
-$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
-or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
-    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
-\cite{austin2017estimating}. While the odds ratio is natural for
-case-control studies \cite{rothman2008case}, other measures
-can reduce heterogeneity \cite{colnet2023risk}.
-In the log domain, the ratios are written as a difference, the
-framework studied here (\autoref{sec:neyman_rubin}) can directly apply.
-In particular, the log odds ratio is estimated by the common
-cross-entropy loss (or log loss) as in logistic regression.
-
-\paragraph{More $R\text{-risk}$ to select models driving decisions}
-
-Prediction models have flourished because their predictions can be easily
-demonstrated and validated on left-out data. But they require more
-careful validation for decision making, using a metric accounting for the
-putative intervention, the $R\text{-risk}$. Even when treated and
-untreated population differ little, as in RCTs, the $R\text{-risk}$
-brings a sizeable benefit. To facilitate better model selection, we provide Python
-code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
-Using the $R\text{-risk}$ does make evaluation more
-complicated not only because the procedure is more involved, but also
-because each intervention requires a dedicated evaluation. However, such
-off-policy evaluation remains much less costly than the recommended good
-practice of impact evaluation testing the ability of a prediction model
-to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
-the model-selection procedure puts no constraints on the models used to
-build predictive models: it opens the door to evaluating a wide range of
-models, from gradient boosting to convolutional neutral, or language
-models.
-
-\section*{Acknowledgments}
-
-We acknowledge fruitful discussions with Bénédicte Colnet.
-
-\subsection*{Financial disclosure}
-
-None reported.
-
-\subsection*{Conflict of interest}
-
-The authors declare no potential conflict of interests.
-
-%\if@Mybibrefstyle%
-\printbibliography
-%\fi if using njdnatbib.sty \bibliography{bibliography}%
-
-\clearpage
-
-\onecolumn
-
-\appendix
-%\setcounter{secnumdepth}{0}
-\setcounter{secnumdepth}{3}
-
-%The following supporting information is available as part of the online article:
-
-\section{Variability of ATE estimation on ACIC
-  2016}\label{apd:toy_example:acic_2016_ate_variability}
-
-Figure \ref{fig:acic_2016_ate_heterogeneity} shows ATE estimations for six
-different models used in g-computation estimators on the 76 configurations of
-the ACIC 2016 dataset. Outcome models are fitted on half of the data and
-inference is done on the other half --ie. train/test with a split ratio of 0.5.
-For each configuration, and each model, this train test split was repeated ten
-times, yielding non parametric variance estimates
-\cite{bouthillier_accounting_2021}.
-
-Outcome models are implemented with
-\href{https://scikit-learn.org/stable/}{scikit-learn}
-\cite{pedregosa_scikitlearn_2011} and the following hyper-parameters:
-
-\begin{table}[h!]
-    \centering
-    \begin{tabular}{llll}
-        \toprule
-        Outcome Model                                  & Hyper-parameters grid
-        \\
-        \midrule
-        Random Forests                                 & Max depth: [2,
-        10]                                                                    \\
-
-        Ridge regression without treatment interaction & Ridge regularization:
-        [0.1]                                                                  \\
-
-        Ridge regression with treatment interaction    & Ridge regularization:
-        [0.1]                                                                  \\
-        \bottomrule
-    \end{tabular}
-    \caption{Hyper-parameters grid used for ACIC 2016 ATE variability}
-    \label{apd:toy_example:acic_2016_ate_variability:table}
-\end{table}
-
-\section{Causal assumptions}\label{apd:causal_assumptions}
-
-We assume the following four assumptions, referred as strong ignorability and
-necessary to assure identifiability of the causal estimands with observational
-data \cite{rubin_causal_2005}:
-\begin{assumption}[Unconfoundedness]\label{assumption:ignorability}
-    \begin{equation*}\label{eq:ignorability}
-        \{Y(0), Y(1) \} \indep A | X
-    \end{equation*}
-    This condition --also called ignorability-- is equivalent to the conditional
-    independence on $e(X)$ \cite{rosenbaum_central_1983}: $\{Y(0), Y(1) \}
-        \indep  A | e(X)$.
-\end{assumption}
-
-\begin{assumption}[Overlap, also known as Positivity)]\label{assumption:overlap}
-    \begin{equation*}\label{eq:overlap}
-        \eta < e(x) < 1 - \eta \quad \forall x \in \mathcal X \text{ and some } \eta > 0
-    \end{equation*}
-    The treatment is not perfectly predictable. Or with different words, every
-    patient has a chance to be treated and not to be treated. For a given set of
-    covariates, we need examples of both to recover the ATE.
-\end{assumption}
-
-As noted by \cite{damour_overlap_2020}, the choice of covariates $X$ can
-be viewed as a trade-off between these two central assumptions. A bigger
-covariates set generally reinforces the ignorability assumption. In the
-contrary, overlap can be weakened by large $\mathcal{X}$ because of the
-potential inclusion of instruments: variables only linked to the treatment which
-could lead to arbitrarily small propensity scores.
-
-% remark: There is a major counter example to these colliders (variables which
-% are caused by both the outcome and the treatment),
-
-\begin{assumption}[Consistency]\label{assumption:consistency} The observed
-    outcome is the potential outcome of the assigned treatment:
-    \begin{equation*}\label{eq:consistancy}
-        Y = A \, Y(1) + (1-A) \, Y(0)
-    \end{equation*}
-    Here, we assume that the intervention $A$ has been well defined. This
-    assumption focuses on the design of the experiment. It clearly states the link
-    between the observed outcome and the potential outcomes through the
-    intervention \cite{hernan_causal_2020}.
-\end{assumption}
-
-\begin{assumption}[Generalization]\label{assumption:generalization} The training
-    data on which we build the estimator and the test data on which we make the
-    estimation are drawn from the same distribution $\mathcal D^*$, also known as
-    the ``no covariate shift'' assumption \cite{jesson_identifying_2020}.
-\end{assumption}
-
-\section{Proofs: Links between feasible and oracle risks}\label{apd:proofs}
-
-\subsection{Upper bound of $\tau\text{-risk}$ with
-    $\mu\text{-risk}_{IPW}$}\label{apd:proofs:mu_risk_ipw_bound}
-
-For the bound with the $\mu\text{-risk}_{IPW}$, we will decompose the CATE risk
-on each factual population risks:
-
-\begin{definition}[Population Factual $\mu\text{-risk}$]\label{mu_risk_a}
-    \cite{shalit_estimating_2017}
-    \begin{equation*}
-        \mu\text{-risk}_{a}(f)= \int_{\mathcal Y \times \mathcal X} (y-f(x ; A=a))^{2}  p(y ; x=x \mid A=a) \; dy dx
-    \end{equation*}
-\end{definition}
-
-Applying Bayes rule, we can decompose the $\mu\text{-risk}$ on each
-intervention:
-\begin{equation*}
-    \mu\text{-risk}(f)
-    =p_{A} \,\mu\text{-risk}_{1}(f)+\left(1-p_{A}\right) \,\mu\text{-risk}_{0}(f)
-    \text{with } p_A=\mathbb P(A=1)
-\end{equation*}
-
-These definitions allows to state a intermediary result on each population:
-\begin{lemma}[Mean-variance decomposition]\label{apd:proofs:mu_risk_ipw_link_mu}
-    We need a reweighted version of the classical mean-variance decomposition.
-
-    For an outcome model $f: x \times A \rightarrow \mathcal X$. Let the inverse
-    propensity weighting function $w(a ; x)=a e(x)^{-1}+(1-a)(1-e(x))^{-1}$.
-    \begin{align*}
-         & \int_{\mathcal X}(\mu_{1}(x)-f(x ; 1))^{2} p(x) dx  = p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  -\sigma^{2}_{Bayes}(1)
-    \end{align*}
-    And
-    \begin{align*}
-         & \int_{\mathcal X}(\mu_{0}(x)-f(x; 0))^{2} p(x) dx   = (1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)  -\sigma^{2}_{Bayes}(0)
-    \end{align*}
-
-    \begin{proof}
-        \begin{align*}
-             & p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  = \int_{\mathcal X \times \mathcal Y} \frac{1}{e(x)}(y-f(x ; 1))^{2} p(y \mid x ; A=1) p(x ; A=1) d y d x                                                      \\
-             & = \int_{\mathcal X \times \mathcal Y} (y-f(x ; 1))^{2} p(y \mid x ; A=1) \frac{p(x ; A=1)}{p(x ; A=1)}p(x)dy dx                                                                                      \\
-             & = \int_{\mathcal X \times \mathcal Y} \big[(y-\mu_1(x))^{2}+\left(\mu_{1}(x)-f(x ; 1)\right)^{2} + 2\left(y-\mu_{1}(x)\right)\left(\mu_{1}(x)-f(x, 1)\right) \big] p(y \mid x ; A=1) p(x) d y d x    \\
-             & =\int_{\mathcal X} \big [ \int_{\mathcal Y} (y-\mu_1(x))^{2} p(y \mid x ; A=1) dy\big ] p(x)dx + \int_{\mathcal X \times \mathcal Y} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x)p(y \mid x ; A=1)dx dy \\
-             & + \qquad 2 \int_{\mathcal X} \big [ \int_{\mathcal Y} \left(y-\mu_{1}(x)\right) p(y \mid x ; A=1) dy \big ] \left(\mu_{1}(x)-f(x, 1)\right)p(x)dx                                                    \\
-             & =\int_{\mathcal X} \sigma_{y}^{2}(x, 1) p(x) d x +\int_{\mathcal X} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x) d x+0
-        \end{align*}
-    \end{proof}
-\end{lemma}
-
-\begin{proposition*}[Upper bound with mu-IPW]\label{apd:proofs:prop:upper_bound}
-    Let f be a given outcome model, let the weighting function $w$ be the Inverse
-    Propensity Weight $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$. Then, under
-    overlap (assumption \ref{assumption:overlap}),
-    \begin{align*}
-        \tau\text{-risk}(f) \leq & \; 2 \, \mu\text{-risk}_{IPW}(w, f)  \; - 2 \, (\sigma^2_{Bayes}(1) +  \sigma^2_{Bayes}(0))
-    \end{align*}
-
-    \begin{proof}
-        \begin{align*}
-             & \tau\text{-risk}(f) =\int_{\mathcal X}(\mu_{1}(x)-\mu_{0}(x)-(f(x ; 1)-f(x ; 0))^{2} p(x) d x
-        \end{align*}
-        By the triangle inequality $(u+v)^2 \leq 2(u^2 + v^2)$:
-        \begin{align*}
-             & \tau\text{-risk}(f) \leq
-            2 \int_{\mathcal X}\big[\left(\mu_{1}(x)-f(x ; 1)\right)^{2}+\left(\mu_{0}(x)-f(x ; 0)\right)^{2}\big] p(x) d x \\
-        \end{align*}
-        Applying Lemma \ref{apd:proofs:mu_risk_ipw_link_mu},
-        \begin{align*}
-             & \tau\text{-risk}(f) \leq 2\big[p_A \mu\text{-risk}_{IPW, 1}(w, f) +(1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)(w, f)\big] -2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1)) \\
-             & = 2 \mu\text{-risk}_{IPW}(w, f)-2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1))
-        \end{align*}
-    \end{proof}
-\end{proposition*}
-
-\subsection{Reformulation of the $R\text{-risk}$ as reweighted
-    $\tau\text{-risk}$}\label{apd:proofs:r_risk_rewrite}
-
-\begin{proposition*}[$R\text{-risk}$ as reweighted $\tau
-            \text{-risk}$]\label{apd:proofs:prop:r_risk_rewrite}
-
-    \begin{proof}
-
-        We consider the R-decomposition: \cite{robinson_rootnconsistent_1988},
-        \begin{equation}\label{apd:eq:r_decomposition}
-            y(a) = m(x) + \big( a - e(x) \big) \tau(x) + \varepsilon(x; a)
-        \end{equation}
-        Where $\mathbb E[\varepsilon(X; A)|X, A] = 0$ We can use it as plug in the
-        $R\text{-risk}$ formula:
-
-        \begin{align*}
-             & R\text {-risk}(f) =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}[(y-m(x))-\big(a-e(x)\big) \tau_f(x)]^{2} p(y ; x ; a) d y d x d a                                     \\
-             & =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}} \left[\big(a-e(x)\big)\tau(x)+\varepsilon(x ; a)-\big(a-e(x)\big) \tau_f(x)\right]^{2} p(y ; x ; a) d y d x da          \\
-             & =\int_{\mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)^{2}\big(\tau(x)- \tau_f(x)\big)^{2} p(x ; a) d x d a                                                                       \\
-             & + 2  \int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)\big(\tau(x)-\tau_f(x)\big)  \int_{\mathcal{Y}} \varepsilon(x ; a) p(y \mid x ; a) d y p(x ; a)dx da \\
-             & +\int_{\mathcal{X} \times \mathcal{A}} \int_{\mathcal{Y}} \varepsilon^{2}(x ; a) p(y \mid x ; a) d y p(x ; a) d x d a
-        \end{align*}
-
-        The first term can be decomposed on control and treated populations to force
-        $e(x)$ to appear:
-        \begin{align*}
-             & \int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}\left[e(x)^{2}p(x;0) + \big(1-e(x)\big)^{2} p(x;1)\right] d x                    \\
-             & =\int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}  \left[e(x)^{2}\big(1-e(x)\big)p(x) + \big(1-e(x)\big)^{2}e(x) p(x)\right] d x \\
-             & =\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x)[1-e(x)+e(x)] p(x) d x                                                     \\ &=\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x) p(x) d x.
-        \end{align*}
-
-        The second term is null since, $\mathbb E[\varepsilon(x, a) |X, A]=0$.
-
-        The third term corresponds to the modulated residuals \ref{eq:residuals} :
-        $\tilde{\sigma}_B^2(0) + \tilde{\sigma}_B^2(1)$
-
-    \end{proof}
-\end{proposition*}
-
-\section{Measuring overlap}\label{apd:motivation_ntv}
-
-\paragraph{Motivation of the Normalized Total Variation}
-%\idea{A simplier measure, NTV seems to capture what we want}
-Computing overlap when working only on samples of the observed distribution,
-outside of simulation, requires a sophisticated estimator of discrepancy
-between distributions, as two data points never have the same exact set of
-features. Maximum Mean Discrepancy \cite{gretton2012kernel} is typically
-used in the context of causal inference
-\cite{shalit_estimating_2017,johansson2022generalization}. However it
-needs a kernel, typically Gaussian, to extrapolate across neighboring
-observations. We prefer avoiding the need to specify such a kernel, as it must
-be adapted to the data which is tricky with categorical or non-Gaussian
-features, a common situation for medical data.
-
-For simulated and some semi-simulated data, we have access to the probability of
-treatment for each data point, which sample both densities in the same data
-point. Thus, we can directly use distribution discrepancy measures and rely on
-the Normalized Total Variation (NTV) distance to measure the overlap between the
-treated and control propensities. This is the empirical measure of the total
-variation distance \cite{sriperumbudur_integral_2009} between the distributions,
-$TV(\mathbb{P}(X|A=1), \mathbb{P}(X|A=0))$. As we have both distribution sampled
-on the same points, we can rewrite it a sole function of the propensity score, a
-low dimensional score more tractable than the full distribution $\mathbb
-    P(X|A)$:
-
-\begin{equation}\label{eq:ntv}
-    \widehat{NTV}(e, 1-e) = \frac{1}{2N} \sum_{i =1}^{N} \big |\frac{e(x_i)}{p_A}-\frac{1-e(x_i)}{1-{p_A}} \big |
-\end{equation}
-
-Formally, we can rewrite NTV as the Total Variation distance
-between the two population distributions. For a population $O = (Y(A), X, A)
-    \sim \mathcal{D}$:
-
-\begin{align*}
-    NTV(O) & = \frac{1}{2N} \sum_{i =1}^{N} \big | \frac{e(x_i)}{p_A}-\frac{1-e(x_i)}{1-{p_A}} \big|         \\
-           & = \frac{1}{2N} \sum_{i =1}^{N} \big|\frac{P(A=1|X=x_i)}{p_A}-\frac{P(A=0|X=x_i)}{1-{p_A}} \big|
-\end{align*}
-
-Thus NTV approximates the following quantity in expectation over the data
-distribution $\mathcal{D}$:
-
-\begin{align*}
-    NTV(\mathcal{D}) & = \int_{\mathcal{X}} \big | \frac{p(A=1|X=x)}{p_A}-\frac{p(A=0|X=x)}{1-{p_A}} \big | p(x)dx \\
-                     & = \int_{\mathcal{X}} \big | \frac{p(A=1, X=x)}{p_A}-\frac{p(A=0, X=x)}{1-{p_A}} \big |dx    \\
-                     & = \int_{\mathcal{X}} \big | p(X=x|A=1)- p(X=x|A=0) \big |dx
-\end{align*}
-
-For countable sets, this expression corresponds to the Total Variation distance
-between treated and control populations covariate distributions : $TV(p_0(x),
-    p_1(x))$.
-
-\paragraph{Measuring overlap without the oracle propensity scores:} For ACIC
-2018, or for non-simulated data, the true propensity scores are not known. To
-measure overlap, we rely on flexible estimations of the Normalized Total
-Variation, using gradient boosting trees to approximate the propensity score.
-Empirical arguments for this plug-in approach is given in Figure
-\ref{apd:overlap:ntv_approximation}.
-
-\paragraph{Empirical arguments}
-
-We show empirically that NTV is an appropriate measure of overlap by :
-\begin{itemize}
-    \item Comparing the NTV distance with the MMD for Caussim which is gaussian
-          distributed in Figure \ref{apd:overlap:caussim:mmd_vs_ntv},
-    \item Verifying that setups with penalized overlap from ACIC 2016 have a
-          higher total variation distance than unpenalized setups in Figure
-          \ref{apd:overlap:penalized_overlap}.
-    \item Verifying that the Inverse Propensity Weights extrema (the inverse of
-          the $\nu$ overlap constant appearing in the overlap Assumption
-          \ref{assumption:overlap}) positevely correlates with NTV for Caussim,
-          ACIC 2016 and Twins in Figure \ref{apd:ntv_vs_max_ipw}. Even if the same
-          value of the maximum IPW could lead to different values of NTV, we
-          expect both measures to be correlated : the higher the extrem propensity
-          weights, the higher the NTV.
-\end{itemize}
-
-\paragraph{Estimating NTV in practice}
-
-Finally, we verify that approximating the NTV distance with a learned plug-in
-estimates of e(x) is reasonnable. We used either a logistic regression or a
-gradient boosting classifier to learn the propensity models for the three
-datasets where we have access to the ground truth propensity scores: Caussim,
-Twins and ACIC 2016. We respectively sampled 1000, 1000 and 770 instances of
-these datasets with different seeds and overlap settings. We first run a
-hyperparameter search with cross-validation on the train set, then select the
-best estimator. We refit on the train set this estimator with or without
-calibration by cross validation and finally estimate the normalized TV with the
-obtained model. This training procedure reflects the one described in Algorithm
-\ref{problem:estimation_procedure:algo} where nuisance models are fitted only on
-the train set.
-
-The hyper parameters are : learning rate $ \in [1e-3, 1e-2, 1e-1, 1]$, minimum
-samples leaf $\in [2, 10, 50, 100, 200]$ for boosting and L2 regularization $\in
-    [1e-3, 1e-2, 1e-1, 1]$ for logistic regression.
-
-Results in Figure \ref{apd:overlap:ntv_approximation} comparing bias to the true
-normalized Total Variation of each dataset instances versus growing true NTV
-indicate that calibration of the propensity model is crucial to recover a good
-approximation of the NTV.
-
-\begin{figure}
-    \begin{subfigure}[b]{\textwidth}
-        \centering
-        \caption{\textbf{Uncalibrated classifiers}}
-        \includegraphics[width=\linewidth]{images/overlap_measure_diff_oracle_ntv_to_n_tv_calibration=False_vs_oracle_n_tv.pdf}
-    \end{subfigure}
-    \begin{subfigure}[b]{\textwidth}
-        \centering
-        \caption{\textbf{Calibrated
-                classifiers}}
-        \includegraphics[width=\linewidth]{images/overlap_measure_diff_oracle_ntv_to_n_tv_calibration=True_vs_oracle_n_tv.pdf}
-    \end{subfigure}
-    \caption{a) Without calibration, estimation of NTV is not trivial even for
-        boosting models. b) Calibrated classifiers are able to recover the true
-        Normalized Total Variation for all datasets where it is
-        available.}\label{apd:overlap:ntv_approximation}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \caption{NTV recovers well the overlap settings described in the ACIC paper
-        \cite{dorie_automated_2019}}\label{apd:overlap:penalized_overlap}
-    \includegraphics[width=0.7\linewidth]{images/overlap_measure_acic_2016_recovery_overlap_setup.pdf}
-\end{figure}
-
-\begin{figure}[htbp]
-    \centering
-    \caption{Good correlation between overlap measured as normalized Total
-        Variation and Maximum Mean Discrepancy (200 sampled Caussim
-        datasets)}\label{apd:overlap:caussim:mmd_vs_ntv}
-    \includegraphics[width=0.7\linewidth]{images/overlap_measure_caussim_transformed_mmd_vs_ntv.pdf}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \begin{subfigure}[b]{0.47\textwidth}
-        \centering
-        \caption{\textbf{Caussim}}
-        \includegraphics[width=\textwidth]{images/overlap_measure_caussim_max_ipw_vs_ntv.pdf}
-        \label{apd:caussim:max_ipw_vs_ntv}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.47\textwidth}
-        \centering
-        \caption{\textbf{ACIC 2016}}
-        \includegraphics[width=\textwidth]{images/overlap_measure_acic_2016_max_ipw_vs_ntv.pdf}
-        \label{apd:acic_2016:ntv_vs_max_ipw}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.47\textwidth}
-        \centering
-        \caption{\textbf{ACIC 2018}}
-        \includegraphics[width=\textwidth]{images/overlap_measure_acic_2018_max_ipw_vs_ntv.pdf}
-        \label{apd:acic_2018:ntv_vs_max_ipw}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.49\textwidth}
-        \centering
-        \caption{\textbf{TWINS}}
-        \includegraphics[width=\textwidth]{images/overlap_measure_twins_max_ipw_vs_ntv.pdf}
-        \label{apd:twins:ntv_vs_max_ipw}
-    \end{subfigure}
-    \caption{Maximal value of Inverse Propensity Weights increases exponentially with the overlap as measure by Normalized Total Variation.}
-    \label{apd:ntv_vs_max_ipw}
-\end{figure}
-
-\section{Experiments}
-
-\subsection{Details on the data generation process}
-\label{apd:experiments:generation}
-
-We use Gaussian-distributed covariates and random basis expansion based on
-Radial Basis Function kernels. A random basis of RBF kernel enables modeling
-non-linear and complex relationships between covariates in a similar way to the
-well known spline expansion. The estimators of the response function are learned
-with a linear model on another random basis (which can be seen as a stochastic
-approximation of the full data kernel \cite{rahimi_random_2008}). We
-carefully control the amount of overlap between treated and control populations,
-a crucial assumption for causal inference.
-
-\begin{itemize}
-    \item The raw features for both populations are drawn from a mixture of
-          Gaussians:
-          $\mathbb P(X) = p_A \mathbb P(X|A=1) + (1- p_A) \mathbb P(X|A=0)$
-          where $\mathbb P(x|A=a)$ is a rotated Gaussian:
-          \begin{equation}
-              \mathbb P(x|A=a) = W \cdot \mathcal N \Big( \begin{bmatrix} (1-2a) \theta \\ 0\end{bmatrix} ; \begin{bmatrix} \sigma_0 & 0 \\ 0 & \sigma_1\end{bmatrix} \Big)
-          \end{equation}
-          with $\theta$ a parameter controlling overlap (bigger yields poorer
-          overlap), $W$ a random rotation matrix and $\sigma_0^2=2;\sigma_1^2=5$.
-
-          This generation process allows to analytically compute the oracle
-          propensity scores $e(x)$, to simply control for overlap with the
-          parameter $\theta$, the distance between the two Gaussian main axes and
-          to  visualize response surfaces.
-
-    \item A basis expansion of the raw features increases the problem dimension.
-          Using Radial Basis Function (RBF) Nystroem transformation \footnote{We use the
-              \href{https://scikit-learn.org/stable/modules/generated/sklearn.kernel_approximation.Nystroem.html}{Sklearn
-                  implementation, \cite{pedregosa_scikitlearn_2011}}}, we expand the raw
-          features into a transformed space. The basis expansion samples randomly a
-          small number of representers in the raw data. Then,  it computes an
-          approximation of the full N-dimensional kernel with these basis components,
-          yielding the transformed features $z(x)$.
-
-          We generate the basis following the original data distribution, $\left [
-                  b_1 .. b_D \right ] \sim \mathbb P(x)$, with D=2 in our simulations. Then, we
-          compute an approximation of the full kernel of the data generation
-          process $RBF(x, \cdot) \;  with \; x \sim \mathbb P(x)$ with these
-          representers: $z(x) = [RBF_{\gamma}(x, b_d)]_{d=1..D}
-              \cdot Z^T \in \mathbb{R}^D$
-          with $RBF_{\gamma}$ being the Gaussian kernel $K(x, y) = exp(-\gamma
-              ||x-y||^2)$ and Z the normalization constant of the kernel basis,
-          computed as the root inverse of the basis kernel $Z=[K(b_i, b_j)]_{i, j
-              \in {1..D}}^{-1/2}$
-
-    \item Functions $\mu_0$, $\tau$ are distinct linear functions of the
-          transformed features:
-          \begin{equation*}
-              \mu_0(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\mu}^T
-          \end{equation*}
-          \begin{equation*}
-              \tau(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\tau}^T
-          \end{equation*}
-    \item Adding a Gaussian noise, $\varepsilon \sim \mathcal N(0, \sigma(x;a))$,
-          we construct the potential outcomes:
-          $y(a) = \mu_0(x) + a\,\tau(x) + \varepsilon(x, a)$
-\end{itemize}
-We generated 1000 instances of this dataset with uniformly random overlap
-parameters $\theta \in \left[ 0, 2.5 \right]$.
-
 \subsection{Model selection procedures}
@@
 \href{https://scikit-learn.org/stable/}{scikit-learn}
-\cite{pedregosa_scikitlearn_2011}.
+\cite{pedregosa_scikitlearn_2011}.  As extreme inverse propensity weights induce
+high variance, we use clipping \cite{swaminathan_counterfactual_2015,
+ionides_truncated_2008} to bound $min(\check e, 1-\check e)$ away from 0 with a
+fixed $\eta=10^{-10}$, ensuring strict overlap for numerical stability.
 

### 2023-11-19 how_to_select_causal_models.tex
iter condense

@@
 
-Progress in machine learning brings predictive models to new health
-data \cite{beam2018big,rajkomar2019machine}, with the promise of
-precision medicine.
-Automated analysis of
-medical images is increasingly accurate, \emph{eg} for brain images,
-\cite{khojaste2022deep,zhang2019radiological} or mammography
-\cite{yala2019deep,shen2019deep,nassif2022breast}.
-New prognostic models leverage routinely-collected
-patient records
-\cite{mooney2018bigdata}: predicting heart
-failure from claims \cite{desai2020comparison}, suicide attempts
-from questionnaires \cite{simon2018predicting}...
-Clinical notes contain much prognostic information
-but require text modeling  \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
-Data may be difficult to
-control and model, but the accuracy of the prediction can be verified
-on left-out data
+Progress in machine learning brings predictive models to new health data
+\cite{beam2018big,rajkomar2019machine}, with the promise of precision medicine.
+Predictive models are increaslingly used for image analysis
+\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
+routinely-collected patient records
+\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
+notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}. Data may
+be difficult to control and model, but the accuracy of the prediction can be
+verified on left-out data
 \cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
-Given a model predicting a health outcome, precision medicine
-would like it to guide decisions: will an individual benefit from an
-intervention such as surgery \cite{fontana2019can}?
-Contrasting predictions with and without the treatment gives an answer,
-but statistical validity requires causal inference
+Given a model predicting a health outcome, precision medicine would like it to
+guide decisions: will an individual benefit from an intervention such as surgery
+\cite{fontana2019can}? Contrasting predictions with and without the treatment
+gives an answer, but statistical validity requires causal inference
 \cite{snowden_implementation_2011,blakely2020reflection}.
@@
 drug to mild cases \citep{hernan_causal_2020,vanderweele2019principles}.
-Controlled allocation of treatment, as in randomized controlled trials (RCTs),
-alleviate this concern. Yet most machine-learning models are trained on
-\emph{observational} data, close to real-word practice
-\cite{black1996we,hernan_methods_2021} but challenging for causal inference.
+% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
+% alleviate this concern. Yet most machine-learning models are trained on
+% \emph{observational} data, close to real-word practice
+% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
 Causal inference has been central to epidemiology, typically with methods that
 model treatment assignment \cite{austin_moving_2015,grose_use_2020}, based on
-propensity scores \cite{rosenbaum_central_1983}. Recent empirical results
-\cite{wendling_comparing_2018,dorie_automated_2019} show benefits of outcome
-modeling to estimate average treatment effects. Maybe a greater benefit is
-that these methods naturally go beyond average effects, estimating
-individualized or conditional average treatment effects (CATE), central to
-precision medicine.
+propensity scores \cite{rosenbaum_central_1983}.  However, outcome modeling go
+beyond average effects, estimating individualized or conditional average
+treatment effects (CATE), central to precision medicine.
 %
@@
 
+% Recent empirical results \cite{wendling_comparing_2018,dorie_automated_2019}
+% show benefits of outcome modeling to estimate average treatment effects.
+
 %\idea{There is a variety of models for g-estimation}
-There is an explosion of outcome modeling or machine learning methods.
-For instance, many deep-learning methods have been developed for medical
-image analysis \cite{shen2017deep,monshi2020deep}.
-Even outcome-modeling methods specifically designed for causal
+% There is an explosion of outcome modeling or machine learning methods, either 
+% developed for
+% For instance, many deep-learning methods have been developed for medical
+% image analysis \cite{shen2017deep,monshi2020deep}.
+Outcome-modeling methods, even when specifically designed for causal
 inference are numerous: Bayesian Additive Regression Trees
@@
 Indeed, estimates of treatment effects can vary markedly across different
-predictive models (see example from \cite{dorie_automated_2019} developed in
+predictive models (as shown by \cite{dorie_automated_2019} and developed in
 Appendix \ref{apd:toy_example:acic_2016_ate_variability}).
 
+Given complex,
+potentially noisy, data, which model is to be most trusted to yield valid causal
+estimates? As no single learner performs best on all data sets, there is a
+pressing need for clear guidelines to select outcome models for causal
+inference.
+
+% TODO:move below
 Standard practices to select models in predictive settings rely on
@@
 %
-Given complex, potentially noisy, data, which model is to be most trusted to
-yield valid causal estimates? As no single learner performs
-best on all data sets, there is a pressing need for clear guidelines to select
-outcome models for causal inference.
 
@@
 
-In this paper, we study \textit{model selection procedures}
+We study \textit{model selection procedures}
 in practical settings: \textit{finite samples} settings and without
@@
 
-We first illustrate the problem of causal model
-selection and briefly review prior art. Then the methods
-\autoref{sec:framework} sets causal model selection in the
-\emph{potential outcome} framework and details the causal risks and
-model-selection procedure. The next \autoref{sec:theory} gives theoretical
-results. The results \autoref{sec:empirical_study} details a thorough empirical
-study, covering many different settings. Finally, we
-discuss the findings in the last \autoref{sec:discussion}.
-%
-Results outline how to best select outcome models for causal
-inference with an adapted
-cross-validation to estimate the so-called $R\text{-risk}$.
-This risk compensates for systematic
-differences between treated and non-treated individuals using
-two \emph{nuisance} models,
-themselves estimated from data and thus imperfect; yet these
-imperfections do not undermine the $R\text{-risk}$.
+We first illustrate the problem of causal model selection and briefly review
+prior art. Then we anchors causal model selection in the \emph{potential
+outcome} framework and details the causal risks and model-selection procedure.
+We then rewrite the so-called $R\text{-risk}$, as a reweighted version of mean
+squared difference between the true and estimated individualized treatment
+effect. Finally, we conduct a thorough empirical study comparing the different
+metrics on a wide variety of data-generative processes based on simulations and
+real data.
+
+% not the place for results
+% Results outline how to best select outcome models for causal
+% inference with an adapted
+% cross-validation to estimate the so-called $R\text{-risk}$.
+% This risk compensates for systematic
+% differences between treated and non-treated individuals using
+% two \emph{nuisance} models,
+% themselves estimated from data and thus imperfect; yet these
+% imperfections do not undermine the $R\text{-risk}$.
 

### 2023-11-19 how_to_select_causal_models.tex
iter condense

@@
 
-There is a rich causal-inference statistical literature, which bridges to
-predictive modeling via so-called \emph{outcome models}, or G-computation,
-G-formula \cite{robins_role_1986}, Q-model \cite{snowden_implementation_2011},
-conditional mean regression \cite{wendling_comparing_2018}. A central challenge
-of inference of treatment effects is that of confounding: spurious associations
-between treatment allocation and baseline health, \emph{eg} only prescribing a
-drug to mild cases \citep{hernan_causal_2020,vanderweele2019principles}.
+Predictive modeling bridges to causal-inference with the so-called \emph{outcome
+    models}, or G-computation, G-formula \cite{robins_role_1986}, Q-model
+\cite{snowden_implementation_2011}, conditional mean regression
+\cite{wendling_comparing_2018}. A central challenge of inference of treatment
+effects is that of confounding: spurious associations between treatment
+allocation and baseline health, \emph{eg} only prescribing a drug to mild cases
+\citep{hernan_causal_2020,vanderweele2019principles}.
 % Controlled allocation of treatment, as in randomized controlled trials (RCTs),
@@
 model treatment assignment \cite{austin_moving_2015,grose_use_2020}, based on
-propensity scores \cite{rosenbaum_central_1983}.  However, outcome modeling go
+propensity scores \cite{rosenbaum_central_1983}. But outcome modeling go
 beyond average effects, estimating individualized or conditional average
@@
 The wide variety of methods raises the problem
-of selecting between different estimators based on the data at hand.
+of selecting between different estimators based on the data at hand.%TODO: might be removed since repeated two sentences below
 %
@@
 Using a predictor to reason on causal effects relies on contrasting the
-prediction of the outcome for a given individual with and without the
-treatment --as detailed in \autoref{sec:neyman_rubin}.
+prediction of the outcome for a given individual with and without the treatment.
 %
@@
     \mathbb R$ summarizing the patient health status (eg. the Charlson index
-\cite{charlson_new_1987}). We simulate a treatment beneficial (decreases $Y$)
+\cite{charlson_new_1987}). We simulate a treatment beneficial (decreases mortality)
 for patients with high Charlson scores (bad health status) but with little
@@
 
-\paragraph{Estimating treatment effects with outcome models}\label{subsec:estimators}
+\paragraph{Estimating treatment effects with outcome models -- g-computation \cite{robins_new_1986}}\label{subsec:estimators}
 
@@
 distribution. On the right, the expectation is taken on the factual observed
-distribution conditionally on the treatment. This equality is referred as the
-g-formula identification \cite{robins_new_1986}. For the rest of the
+distribution conditionally on the treatment. For the rest of the
 paper, the expectations will always be taken on the factual observed
 distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
-g-computation estimators \cite{snowden_implementation_2011}), targeting the
-ATE $\tau$ with outcome modeling:
+g-computation estimators \cite{snowden_implementation_2011}):
 \begin{eqnarray}
@@
 
-As extreme inverse propensity weights induce high variance, clipping can be
-useful for numerical stability
-\cite{swaminathan_counterfactual_2015, ionides_truncated_2008}.
+% As extreme inverse propensity weights induce high variance, clipping can be
+% useful for numerical stability
+% \cite{swaminathan_counterfactual_2015, ionides_truncated_2008}.
 
@@
 
-\section{Theory: Links between feasible and oracle risks}\label{sec:theory}
-
-We now relate two feasible risks, $\mu \text{-risk}_{IPW}$ and the
-$R\text{-risk}$ to the oracle $\tau\text{-risk}$. Both results make
-explicit the role of overlap for the performances of causal risks.
-
-These bounds depend on a specific form of residual that we now define: for each potential outcome, $a \in  \{0; 1\}$, the variance conditionally on $x$
-is \cite{shalit_estimating_2017}:
+\section{A theoretical argument favoring the R-risk}\label{theory:r_risk_rewrite}
+
+We now derive a novel rewriting of the $R\text{-risk}$, making explicit its link
+with the oracle $\tau \text{-risk}$.
+
+% We now one feasible risks, $\mu \text{-risk}_{IPW}$ and the
+% $R\text{-risk}$ to the oracle $\tau\text{-risk}$. Both results make
+% explicit the role of overlap for the performances of causal risks.
+
+This rewriting involves reweighted residuals: for each potential outcome, $a \in
+    \{0; 1\}$, the variance conditionally on $x$ is \cite{shalit_estimating_2017}:
 \begin{equation*}\label{eq:residuals}
@@
 
-\subsection{Upper bound of $\tau\text{-risk}$ with
-    $\mu\text{-risk}_{IPW}$}\label{theory:mu_risk_ipw_bound}
-
-\begin{proposition}[Upper bound with $\mu \text{-risk}_{IPW}$
-    ]\label{theory:prop:mu_risk_ipw_bound}
-    \cite{johansson2022generalization} Given an outcome model $f$, let a
-    weighting function\\ $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$ as the
-    Inverse Propensity Weight. Then, under overlap (assumption
-    \ref{assumption:overlap}), we have:
-    \begin{equation*}
-        \tau\text{-risk}(f) \leq \; 2 \, \mu\text{-risk}_{IPW}(w, f)
-        - 2 \, \big(\sigma^2_{B}(1) +  \sigma^2_{B}(0)\big)
-    \end{equation*}
-\end{proposition}
-This result has been derived in previous work
-\cite{johansson2022generalization}. It links $\mu\text{-risk}_{IPW}$ to
-the squared residuals of each population. For completeness, we provide the proof in \ref{apd:proofs}.
-
-The upper-bound comes from the triangular inequality applied to the residuals of
-both populations. The two quantities are equal when the
-absolute residuals on treated and untreated populations are equal on the
-whole covariate space:
-$\forall x \in \mathcal X, |\mu_1(x) - f(x, 1)| = |\mu_0(x) - f(x, 0)|$.
-The main difference between the oracle $\tau \text{-risk}$ and the
-reweighted mean squared error, $\mu\text{-risk}_{IPW}$, comes from heterogeneous
-residuals between populations.
-% These quantities are difficult to characterize as
-%they are linked both to the estimator and to the data distribution.
-%
-This bound shows that minimizing the $\mu\text{-risk}_{IPW}$ helps to
-minimize the $\tau\text{-risk}$, which leads to
-interesting optimization procedures \cite{johansson2022generalization}. However, there is no
-guarantee that this bound is tight, which makes it fragile for model
-selection.
-
-Assuming strict overlap (probability of all individuals being treated or not
-bounded away from 0 and 1 by $\eta$, \ref{apd:causal_assumptions}), the
-above bound simplifies into a looser one involving the usual mean squared error:
-$\tau\text{-risk}(f)\leq \frac{2}{\eta}\, \mu\text{-risk}(f) -  2 \, \big(\sigma^2_{B}(1) +  \sigma^2_{B}(0)\big)$. For weak overlap (propensity scores not bounded far from 0
-or 1), this bound is very loose (as shown in Figure \ref{fig:toy_example})
-and is not appropriate to discriminate between models with close performances.
-
 %\md{Lower bound: I think that there is no lower bound of the tau-risk with the
@@
 %observed data.}
-
-\subsection{Reformulation of the $R\text{-risk}$ as reweighted
-    $\tau\text{-risk}$}\label{theory:r_risk_rewrite}
-
-We now derive a novel rewriting of the $R\text{-risk}$, making explicit its link
-with the oracle $\tau \text{-risk}$.
 
@@
 $\frac{1}{4}$, hence the $R \text{-risk}$ is close to the desired gold-standard
-$\tau \text{-risk}$. On the contrary, for units with extreme overlap violation,
-these weights go down to zero with the propensity score.
-
-% \begin{remark}[Trivial upper bound] Because $e(x)\leq1$ and $(1-e(x))\leq1$,
-%   we see immediately that \begin{equation} R \text{-risk}^* \leqslant
-%   \tau{\text{-risk}}+\tilde{\sigma_{B}}(1)^{2}+\tilde{\sigma_{B}}^{2}(0)
-%   \end{equation} \end{remark}
-
-% \begin{remark}[Lower bound] Due to strict overlap assumption, \begin{equation}
-%   R \text{-risk}^* \geqslant \eta (1 - \eta)
-%   \tau{\text{-risk}}-\tilde{\sigma_{B}}(1)^{2}-\tilde{\sigma_{B}}^{2}(0)
-%   \end{equation} \end{remark}
-
-% \begin{remark}[Cases of loose upper bounds] Giving a strict overlap assumption
-%   $\eta$, we can exhibit for all $C \; st. \; \eta^{-1}>C>1$, simple
-%   simulations where: \begin{equation} R \text{-risk}^* \leqslant C \; \eta \;
-%   \tau \text{-risk}(f) \quad \end{equation} \end{remark}
-
-\subsection{Interesting special cases}
+$\tau \text{-risk}$. For randomized control trials, this weight is constant
+making the $R\text{-risk}$ particularly suited for exploring
+heterogeneity (Details in Appendix \ref{apd:theory:special_cases})
+
+\section{Empirical Study}\label{sec:empirical_study}
+
+%We now explore empirically the behavior of these different causal selection
+%metrics.
+%\subsection{Causal metrics under evaluation}
+
+We evaluate the following causal metrics, oracle and feasible
+versions, presented in Table
+\ref{tab:evaluation_metrics}:\\
+$\widehat{\mu\text{-risk}}_{IPW}^*$,
+$\widehat{R\text{-risk}}^*$,
+$\widehat{U\text{-risk}}^*$,
+$\widehat{\tau\text{-risk}_{IPW}}^*$,
+$\widehat{\mu\text{-risk}}$,
+$\widehat{\mu\text{-risk}}_{IPW}$,
+$\widehat{R\text{-risk}}$,
+$\widehat{U\text{-risk}}$,
+$\widehat{\tau\text{-risk}_{IPW}}$.
+We benchmark the metrics in a variety of settings:
+many different simulated data generation
+processes and three semi-simulated datasets \footnote{Scripts for the simulations and the selection procedure are available at
+    \url{https://github.com/soda-inria/caussim}.
+}.
+
+\subsection{Caussim: Extensive simulation settings}\label{subsec:simulations}
+
+\paragraph{Data Generation}
+
+We use simulated data, on which the ground-truth causal effect is known. Going
+beyond prior empirical studies of causal model selection
+\cite{schuler_comparison_2018,alaa_validating_2019}, we use many
+generative processes, to reach more general conclusions (as discussed
+in \ref{apd:results:fig:seed_effect}).
+
+We generate the response functions using random bases extension, a common method in
+biostatistics, \emph{eg} functional regression with splines
+\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
+vary at specific knots, they give flexible non-linear models. The number of
+knots controls the complexity of the function. We use random approximation of
+Radial Basis Function (RBF) kernels \cite{rahimi_random_2008} to generate the
+outcome and treatment functions. RBF use the same process as polynomial splines
+but replace polynomial by Gaussian kernels. Unlike polynomials, Gaussian kernels
+have decreasing influences in the input space. This avoids unrealistic
+divergences of the response surfaces at the ends of the feature space. We
+generate 1\,000 datasets based on these functions, with random overlap
+parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
+\ref{apd:experiments:generation}.
+
+\captionsetup[sub]{font=large,labelfont={bf,sf}}%
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\linewidth]{images/caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
+
+    \caption{Example of the simulation setup in the input space with two
+        knots --\emph{ie.}basis functions. The top panel
+        shows the observations in feature space, while the bottom panel displays the
+        two response surfaces on a 1D cut along the black lines drawn on
+        the top panel.}
+    \label{fig:simulation_examples}
+\end{figure}
+
+\paragraph{Family of candidate estimators}
+
+We compare different candidate estimators that approximate imperfectly the
+data-generating process. To build such estimators, we first use a RBF
+expansion similar to the one used for data generation. We choose two random
+knots and apply a transformation of the raw data features with a Gaussian
+kernel. This step is referred as the featurization. Then, we fit a linear
+regression on these transformed features. We consider two ways of combining
+these steps for outcome mode; we use common nomenclature
+\cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different
+meta-learners that differ on how they model, jointly or not, the treated and the
+non treated:
+\begin{itemize}
+    \item SLearner: A single learner for both population, taking the treatment as
+          a supplementary covariate.
+    \item SftLearner: A single set of basis functions is sampled at random for both
+          populations, leading to a given feature space used to model both the treated and
+          controls, then two
+          separate different regressors are fitted on this shared representation.
+    \item TLearner: Two completely different learners for each population, hence
+          separate feature representations and regressors.
+\end{itemize}
+
+% We do not include more elaborated meta-learners such as R-learner
+% \cite{nie_quasioracle_2017} or X-learner
+% \cite{kunzel_metalearners_2019}. Our goal is not to have the best possible
+% learner but to have a variety of sub-optimal learners to compare the
+% different causal metrics. For the same reason, we did not include more powerful
+% outcome models such as random forests or boosting trees.
+
+For the regression step, we fit a Ridge regression on the transformed features
+with 6 different choices of the regularization parameter $\lambda \in [10^{-3},
+        10^{-2}, 10^{-1}, 1, 10^{1}, 10^{2}]$, coupled with a TLearner or a SftLearner.
+We sample 10 different random basis for learning and
+featurization yielding a family $\mathcal F$ of 120 candidate estimators.
+%
+
+\subsection{Semi-simulated datasets}
+
+\paragraph{Datasets}\label{semi_simulated:datasets}
+
+We also use three benchmarks of the causal-inference literature, semi-simulated
+data adding a known synthetic causal effect to real --non synthetic-- healthcare
+covariate. ACIC 2016 \cite{dorie_automated_2019} is based on the Collaborative
+Perinatal Project \cite{niswander_women_1972}, a RCT studying infants’
+developmental disorders containing 4,802 indivduals and 55 features. We used 770
+dataset instances: 10 random seeds for each of the 77 simulated settings for the
+treatment and outcomes.  ACIC 2018 \cite{shimoni_benchmarking_2018} simulated
+treatment and outcomes for the Linked Births and Infant Deaths Database (LBIDD)
+\cite{macdorman_infant_1998} with $D=177$ covariates. We used all 432 datasets
+of size $N=5\,000$.  Twins \cite{louizos_causal_2017} is an augmentation of real
+data on twin births and mortality rates \cite{almond_costs_2005}. There are
+$N=11\,984$ samples, and $D=50$ covariates for which we simulated 1,000
+different treatment allocations. Datasets details are given in Appendix
+\ref{apd:experiments:datasets}.
+
+\paragraph{Family of candidate
+    estimators}\label{semi_simulated:candidate_estimators}
+
+For these three datasets, the family of candidate estimators are gradient
+boosting trees for both the response surfaces and the treatment
+\footnote{Scikit-learn regressor,
+    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingRegressor.html}{HistGradientBoostingRegressor},
+    and classifier,
+    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingClassifier.html}{HistGradientBoostingClassifier}.}
+with S-learner, learning rate
+in $\{0.01, 0.1, 1\}$, and maximum number of leaf nodes in $\{25, 27, 30, 32,
+    35, 40\}$ resulting in a family of size 18.
+
+\paragraph{Nuisance estimators}
+
+Drawing inspiration from the TMLE literature that uses combination of flexible
+machine learning methods \cite{schuler_targeted_2017}, we use a meta-learner as models
+for the nuisances $\check e$ (respectively $\check m$): a
+stacked estimator of ridge and boosting classifiers (respectively
+regressions). We select hyper-parameters with randomized search on a validation set
+$\mathcal{V}$ and keep them fix for model selection
+(\ref{apd:experiments:nuisances_hp} lists hyperparameters).
+
+\subsection{Measuring overlap between treated and non treated}\label{subsec:measuring_overlap}
+
+%\idea{Overlap is important but not easily measurable}
+Good overlap between treated and control population is
+crucial for causal inference as it is required by the positivity assumption
+\ref{assumption:overlap}. It is often assessed by
+comparing visually population distributions (as in Figure
+\ref{fig:toy_example}) or computing standardized difference on each
+feature
+\cite{austin_introduction_2011,austin_moving_2015}. While these methods are
+useful to decide if positivity holds, they do not yield a
+single measure. Rather, we compute the divergence
+between the population covariate distributions $\mathbb P(X|A=0)$ and $\mathbb
+    P(X|A=1)$
+\cite{damour_overlap_2020,johansson2022generalization}. We introduce the
+Normalized Total Variation (NTV), a divergence based on the sole propensity
+score (see \ref{apd:motivation_ntv}).
+
+\begin{figure}[!b]
+    \centering
+    \includegraphics[width=\linewidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
+    %\hfill
+    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$.
+        Strong and Weak overlap correspond to the first and last tertiles of the overlap distribution measured with
+        Normalized Total Variation eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents the same results
+        by adding semi-oracle risks in Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics}, measured with
+        absolute Kendall's in Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with $\tau\mathrm{-risk}$
+        gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table
+        \ref{apd:table:relative_kendalls_all_datasets} gives median and
+        IQR of the relative Kendall.}\label{fig:relative_kendalls_all_datasets}
+\end{figure}
+
+\subsection{Results: factors driving good model selection}\label{empirical_study:results}
+
+\paragraph{The $R\text{-risk}$ is the best metric}
+
+Figure \ref{fig:relative_kendalls_all_datasets} shows the agreement between the
+ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the different
+feasible causal metrics. We measure this agreement with a relative\footnote{To
+    remove the variance across datasets (some datasets lead to easier model
+    selection than others), we report values for one metric relative to the mean of
+    all metrics for a given dataset instance: $\text{Relative} \,
+        \kappa(\ell,\tau\mathrm{{-risk}})= \kappa(\ell,\tau\mathrm{{-risk}}) -
+        mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$
+(eq. \ref{eq:kendall_tau}) \cite{kendall_new_1938}. Given the importance of
+overlap in how well metrics approximate the oracle $\tau\text{-risk}$, we
+separate strong and weak overlap.
+
+Among all metrics, the classical mean squared error (ie. factual
+$\mu\text{-risk}$) is worse and reweighting it with propensity score
+($\mu\text{-risk}_{IPW}$) does not bring much improvements. The
+$R\text{-risk}$, which includes a model of mean outcome and propensity
+scores, leads to the best performances. Interestingly, the
+$U\text{-risk}$, which uses the same nuisances, deteriorates in weak overlap, probably due to variance
+inflation when dividing by extreme propensity scores.
+
+Beyond rankings, the differences in terms of absolute
+ability to select the best model are large: The R-risk selects a model
+with a $\tau\text{-risk}$ only 1\% higher
+than the best
+possible candidate for strong overlap on Caussim, but selecting with
+the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
+practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
+--as in some causal-inference methods \citep{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
+(Figure
+\ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
+datasets, the $R\text{-risk}$ consistently decreases the
+risk compared to the $\mu\text{-risk}$:
+0.1\% compared to 1\% on ACIC2016,  1\% compared to 20\% on ACIC2018,
+and 0.05\% compared to 1\% on Twins.
+
+\paragraph{Model selection is harder for low population
+    overlap}
+
+Model selection for causal inference becomes more and more difficult with
+increasingly different treated and control populations (Figure
+\ref{fig:all_datasets_overlap_effect_r_risk}). The absolute Kendall's
+coefficient correlation with $\tau\text{-risk}$ drops from values around 0.9
+(excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
+(\ref{apd:experiments:additional_results}).
+
+\begin{figure}[!h]
+    \centering
+    \includegraphics[width=\linewidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
+    %\hfill
+    \caption{\textbf{Model selection is harder for low population
+            overlap}:
+        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
+        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents results for all
+        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
+        Kendall's and continuous overlap values in Figure
+            {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
+\end{figure}
+
+\paragraph{Nuisances can be estimated on the same data as outcome models}
+
+Using the train set $\mathcal{T}$ both to fit the candidate estimator and the
+nuisance estimates is a form of double dipping which can lead errors in
+nuisances correlated to that of outcome models
+\cite{nie_quasioracle_2017}. In theory, these correlations can bias model
+selection and, strictly speaking, push
+to split out a third separated data set --a ``nuisance set''-- to fit the
+nuisance models. The drawback is that it depletes the data available for
+model estimation and selection. However, Figure
+\ref{fig:procedures_comparison} shows no substantial difference between a procedure with a separated
+nuisance set and the simpler shared nuisance-candidate set procedure.
+
+\begin{figure}[!tb]
+    \includegraphics[width=\linewidth]{images/_3_procedure_r_risk_only_3datasets_twocols.pdf}
+
+    \caption{\textbf{Nuisances can be estimated on the same data as outcome
+            models}: Results for the R-risk are similar between the
+        \textcolor{MidnightBlue}{shared
+            nuisances/candidate set} and
+        the \textcolor{RedOrange}{separated nuisances set} procedures. Figure
+        \ref{apd:fig:procedures_comparison_all_metrics} details results for all metrics.}\label{fig:procedures_comparison}
+\end{figure}
+
+\paragraph{Stacked models are good overall estimators of nuisances}
+
+For every risk, the oracle version recovers better the best estimator.
+However,
+stacked nuisances estimators (boosting and linear) lead to feasible
+metrics with close performances to the oracles ones: the
+corresponding estimators recover well-enough the true nuisances.
+One may wonder if simpler models for the nuisance could be useful,
+in particular in data-poor settings or when the true models are linear.
+Figure \ref{fig:all_datasets_nuisances_comparison} compares causal model
+selection estimating nuisances with stacked estimators or linear model.
+It comprises the Twins data, where the true propensity model is linear,
+and a downsampled version of this data, to study a situation favorable to
+linear models. In these settings,
+stacked and linear estimations of the nuisances performs equivalently.
+Detailed analysis (Figure \ref{apd:fig:nuisances_comparison_twins})
+confirms that using adaptive models --as built by
+stacking linear models and gradient-boosted trees-- suffices to estimate nuisance.
+
+\begin{figure}[!tb]
+    \includegraphics[width=\linewidth]{images/_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
+
+    \caption{\textbf{\textcolor{DarkOrchid}{Stacked
+                models} are good overall estimators of the nuisances}:
+        Results are shown only for the
+        R-risk; Figure \ref{apd:fig:nuisances_comparison}
+        details every metrics. For Twins, where the true propensity
+        model is linear, \textcolor{DarkOrchid}{stacked} and
+        \textcolor{ForestGreen}{linear}
+        estimations of the nuisances performs equivalently, even for a downsampled version
+        (N=4794). }\label{fig:all_datasets_nuisances_comparison}
+\end{figure}
+
+\paragraph{Use 90\% of the data to estimate outcome models, 10\% to
+    select them}
+
+The analyst faces a compromise: given a finite
+data sample, should she allocate more data to estimate the outcome model,
+thus improving the quality of the outcome model but leaving
+little data for model selection. Or, she could choose a bigger test set for
+model selection and effect estimation. For causal model selection, there
+is no established practice (as reviewed in \ref{apd:results:k_fold_choices}).
+
+We investigate such tradeoff varying the ratio between train and test
+data size. For this, we first split out 30\% of the data as a holdout set
+$\mathcal{V}$ on which we use the oracle response functions to derive
+silver-standard estimates of causal quantities. We then
+use the standard estimation procedure on the remaining 70\% of the data,
+splitting it into train $\mathcal{T}$ and test $\mathcal{S}$ of varying
+sizes. We finally measure the error between this estimate and the
+silver standard.
+
+We consider two different analytic goals: estimating a average
+treatment effect --a single number used for policy making-- and a
+CATE --a full model of the treatment effect as a function of covariates
+$X$. Given that the latter is a much more complex object than the former,
+the optimal train/test ratio might vary. To measure errors, we use for
+the ATE the relative absolute ATE bias between the ATE computed with the
+selected outcome model on the test set, and the true ATE as evaluated on
+the holdout set $\mathcal{V}$. For the CATE, we compare the
+$\tau\text{-risk}$
+of the best selected model applied on the holdout set $\mathcal{V}$. We explore this trade-off for the ACIC 2016 dataset and the R-risk.
+
+\begin{figure}[!t]
+    \begin{minipage}{.5\textwidth}
+        \centerline{\textbf{a) CATE estimation error}}
+        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16_twocols.pdf}
+    \end{minipage}
+    %\hspace{0.3\textwidth}
+    \begin{minipage}{.5\textwidth}
+        \centerline{\textbf{b) ATE estimation error}}
+        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16_twocols.pdf}
+    \end{minipage}%
+    \caption{\textbf{a) For CATE, a train/test ratio of 0.9/0.1 appears a good
+            trade-off.} b) For ATE, there is a small signal pointing also to
+        0.9/0.1 (K=10).
+        for ATE. Experiences on 10 replications of all 78 instances of the ACIC 2016
+        data.}\label{fig:train_test_ratio}
+\end{figure}
+
+Figure \ref{fig:train_test_ratio} shows that a train/test ratio of
+0.9/0.1 (K=10) or 0.8/0.2 (K=5) appears best to estimate CATE and
+ATE.
+
+\section{Discussion and conclusion}\label{sec:discussion}
+
+Predictive models are increasingly used to reason about causal effects,
+for instance in precision medicine to drive individualized decision.
+Our results highlight that they should be selected, validated, and tuned
+using different procedures and error measures than those classically used
+to assess prediction (estimating the so-called $\mu\text{-risk}$).
+Rather, selecting the best outcome model according to the $R\text{-risk}$
+(eq.\, Definition \ref{def:r_risk}) leads to more valid causal estimates.
+
+\paragraph{Nuisance models: more gain than pain}
+%
+Estimating the $R\text{-risk}$ requires a more complex procedure than
+standard cross-validation used \emph{e.g.}~in machine learning: it involves
+fitting nuisance models necessary for model evaluation, though our
+results show that these can be learned on the same set of data
+as the outcome model evaluated.
+
+The nuisance models must be well estimated (Figure
+\ref{fig:all_datasets_nuisances_comparison}). However these models are
+easier to select and control than a causally-valid outcome model,
+as they are associated to errors on
+observed distributions. Our results show that using for nuisance models
+a flexible stacking-based family of estimator suffices for good model selection.
+%
+In fact, a feasible $R\text{-risk}$ --where the nuisances
+are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
+nuisances are known. This may be explained by
+results that suggest that estimation errors on both
+nuisances partly compensate out in the
+$R\text{-risk}$ \cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.
+
+Note that propensity score models must be selected to estimate
+the individual posterior probability. For this, we used the Brier score,
+is minimized by the true individual probability. An easy mistake is to
+use calibration errors popular in machine learning
+\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
+as these select not for the individual posterior probability but for an
+aggregate error rate \cite{perez2022beyond}.
+
+\paragraph{Extension to binary outcomes}
+While we focused on continuous outcomes, in medicine, the target outcome
+is often a categorical variable such as mortality status or diagnosis. In
+this case, it may be interesting to focus on other estimands than the
+Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
+$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
+or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
+    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
+\cite{austin2017estimating}. While the odds ratio is natural for
+case-control studies \cite{rothman2008case}, other measures
+can reduce heterogeneity \cite{colnet2023risk}.
+In the log domain, the ratios are written as a difference, the
+framework studied here (\autoref{sec:neyman_rubin}) can directly apply.
+In particular, the log odds ratio is estimated by the common
+cross-entropy loss (or log loss) as in logistic regression.
+
+\paragraph{More $R\text{-risk}$ to select models driving decisions}
+
+Prediction models have flourished because their predictions can be easily
+demonstrated and validated on left-out data. But they require more
+careful validation for decision making, using a metric accounting for the
+putative intervention, the $R\text{-risk}$. Even when treated and
+untreated population differ little, as in RCTs, the $R\text{-risk}$
+brings a sizeable benefit. To facilitate better model selection, we provide Python
+code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
+Using the $R\text{-risk}$ does make evaluation more
+complicated not only because the procedure is more involved, but also
+because each intervention requires a dedicated evaluation. However, such
+off-policy evaluation remains much less costly than the recommended good
+practice of impact evaluation testing the ability of a prediction model
+to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
+the model-selection procedure puts no constraints on the models used to
+build predictive models: it opens the door to evaluating a wide range of
+models, from gradient boosting to convolutional neutral, or language
+models.
+
+\section*{Acknowledgments}
+
+We acknowledge fruitful discussions with Bénédicte Colnet.
+
+\subsection*{Financial disclosure}
+
+None reported.
+
+\subsection*{Conflict of interest}
+
+The authors declare no potential conflict of interests.
+
+%\if@Mybibrefstyle%
+\printbibliography
+%\fi if using njdnatbib.sty \bibliography{bibliography}%
+
+\clearpage
+
+\onecolumn
+
+\appendix
+%\setcounter{secnumdepth}{0}
+\setcounter{secnumdepth}{3}
+
+%The following supporting information is available as part of the online article:
+
+\section{Variability of ATE estimation on ACIC
+  2016}\label{apd:toy_example:acic_2016_ate_variability}
+
+Figure \ref{fig:acic_2016_ate_heterogeneity} shows ATE estimations for six
+different models used in g-computation estimators on the 76 configurations of
+the ACIC 2016 dataset. Outcome models are fitted on half of the data and
+inference is done on the other half --ie. train/test with a split ratio of 0.5.
+For each configuration, and each model, this train test split was repeated ten
+times, yielding non parametric variance estimates
+\cite{bouthillier_accounting_2021}. Figure \ref{fig:acic_2016_ate_heterogeneity}
+shows large variations obtained across different outcome estimators on
+semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models such as
+random forests are doing well in most settings except when treated and untreated
+populations differ noticeably, in which case a linear model (ridge) is to be
+preferred. However random forests with different hyper-parameters (max depth= 2)
+yield poor estimates. A simple rule of thumb such as preferring flexible models
+does not work in general; model selection is needed.
+
+%\idea{It is crucial to have tools for model selection.}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.95\linewidth]{images/2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
+    \caption{\textbf{Different outcome models lead to different
+            estimation errors on the Average Treatment Effects},
+        on 77 classic simulations with known true causal effect
+        \cite{dorie_automated_2019}. The different models are ridge regression
+        and random forests with different hyper-parameters
+        (details
+        \ref{apd:toy_example:acic_2016_ate_variability}). The different configurations are
+        plotted as a function of increasing difference between treated and
+        untreated population --see
+        \autoref{subsec:measuring_overlap}.
+        There is no systematic best performer; data-driven model
+        selection is important.
+        \label{fig:acic_2016_ate_heterogeneity}%
+    }
+\end{figure}
+
+Outcome models are implemented with
+\href{https://scikit-learn.org/stable/}{scikit-learn}
+\cite{pedregosa_scikitlearn_2011} and the following hyper-parameters:
+
+\begin{table}[h!]
+    \centering
+    \begin{tabular}{llll}
+        \toprule
+        Outcome Model                                  & Hyper-parameters grid
+        \\
+        \midrule
+        Random Forests                                 & Max depth: [2,
+        10]                                                                    \\
+
+        Ridge regression without treatment interaction & Ridge regularization:
+        [0.1]                                                                  \\
+
+        Ridge regression with treatment interaction    & Ridge regularization:
+        [0.1]                                                                  \\
+        \bottomrule
+    \end{tabular}
+    \caption{Hyper-parameters grid used for ACIC 2016 ATE variability}
+    \label{apd:toy_example:acic_2016_ate_variability:table}
+\end{table}
+
+\section{Causal assumptions}\label{apd:causal_assumptions}
+
+We assume the following four assumptions, referred as strong ignorability and
+necessary to assure identifiability of the causal estimands with observational
+data \cite{rubin_causal_2005}:
+\begin{assumption}[Unconfoundedness]\label{assumption:ignorability}
+    \begin{equation*}\label{eq:ignorability}
+        \{Y(0), Y(1) \} \indep A | X
+    \end{equation*}
+    This condition --also called ignorability-- is equivalent to the conditional
+    independence on $e(X)$ \cite{rosenbaum_central_1983}: $\{Y(0), Y(1) \}
+        \indep  A | e(X)$.
+\end{assumption}
+
+\begin{assumption}[Overlap, also known as Positivity)]\label{assumption:overlap}
+    \begin{equation*}\label{eq:overlap}
+        \eta < e(x) < 1 - \eta \quad \forall x \in \mathcal X \text{ and some } \eta > 0
+    \end{equation*}
+    The treatment is not perfectly predictable. Or with different words, every
+    patient has a chance to be treated and not to be treated. For a given set of
+    covariates, we need examples of both to recover the ATE.
+\end{assumption}
+
+As noted by \cite{damour_overlap_2020}, the choice of covariates $X$ can
+be viewed as a trade-off between these two central assumptions. A bigger
+covariates set generally reinforces the ignorability assumption. In the
+contrary, overlap can be weakened by large $\mathcal{X}$ because of the
+potential inclusion of instruments: variables only linked to the treatment which
+could lead to arbitrarily small propensity scores.
+
+% remark: There is a major counter example to these colliders (variables which
+% are caused by both the outcome and the treatment),
+
+\begin{assumption}[Consistency]\label{assumption:consistency} The observed
+    outcome is the potential outcome of the assigned treatment:
+    \begin{equation*}\label{eq:consistancy}
+        Y = A \, Y(1) + (1-A) \, Y(0)
+    \end{equation*}
+    Here, we assume that the intervention $A$ has been well defined. This
+    assumption focuses on the design of the experiment. It clearly states the link
+    between the observed outcome and the potential outcomes through the
+    intervention \cite{hernan_causal_2020}.
+\end{assumption}
+
+\begin{assumption}[Generalization]\label{assumption:generalization} The training
+    data on which we build the estimator and the test data on which we make the
+    estimation are drawn from the same distribution $\mathcal D^*$, also known as
+    the ``no covariate shift'' assumption \cite{jesson_identifying_2020}.
+\end{assumption}
+
+\section{Proofs: Links between feasible and oracle risks}\label{apd:proofs}
+
+% \subsection{Upper bound of $\tau\text{-risk}$ with
+%     $\mu\text{-risk}_{IPW}$}\label{apd:proofs:mu_risk_ipw_bound}
+
+% For the bound with the $\mu\text{-risk}_{IPW}$, we will decompose the CATE risk
+% on each factual population risks:
+
+% \begin{definition}[Population Factual $\mu\text{-risk}$]\label{mu_risk_a}
+%     \cite{shalit_estimating_2017}
+%     \begin{equation*}
+%         \mu\text{-risk}_{a}(f)= \int_{\mathcal Y \times \mathcal X} (y-f(x ; A=a))^{2}  p(y ; x=x \mid A=a) \; dy dx
+%     \end{equation*}
+% \end{definition}
+
+% Applying Bayes rule, we can decompose the $\mu\text{-risk}$ on each
+% intervention:
+% \begin{equation*}
+%     \mu\text{-risk}(f)
+%     =p_{A} \,\mu\text{-risk}_{1}(f)+\left(1-p_{A}\right) \,\mu\text{-risk}_{0}(f)
+%     \text{with } p_A=\mathbb P(A=1)
+% \end{equation*}
+
+% These definitions allows to state a intermediary result on each population:
+% \begin{lemma}[Mean-variance decomposition]\label{apd:proofs:mu_risk_ipw_link_mu}
+%     We need a reweighted version of the classical mean-variance decomposition.
+
+%     For an outcome model $f: x \times A \rightarrow \mathcal X$. Let the inverse
+%     propensity weighting function $w(a ; x)=a e(x)^{-1}+(1-a)(1-e(x))^{-1}$.
+%     \begin{align*}
+%          & \int_{\mathcal X}(\mu_{1}(x)-f(x ; 1))^{2} p(x) dx  = p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  -\sigma^{2}_{Bayes}(1)
+%     \end{align*}
+%     And
+%     \begin{align*}
+%          & \int_{\mathcal X}(\mu_{0}(x)-f(x; 0))^{2} p(x) dx   = (1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)  -\sigma^{2}_{Bayes}(0)
+%     \end{align*}
+
+%     \begin{proof}
+%         \begin{align*}
+%              & p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  = \int_{\mathcal X \times \mathcal Y} \frac{1}{e(x)}(y-f(x ; 1))^{2} p(y \mid x ; A=1) p(x ; A=1) d y d x                                                      \\
+%              & = \int_{\mathcal X \times \mathcal Y} (y-f(x ; 1))^{2} p(y \mid x ; A=1) \frac{p(x ; A=1)}{p(x ; A=1)}p(x)dy dx                                                                                      \\
+%              & = \int_{\mathcal X \times \mathcal Y} \big[(y-\mu_1(x))^{2}+\left(\mu_{1}(x)-f(x ; 1)\right)^{2} + 2\left(y-\mu_{1}(x)\right)\left(\mu_{1}(x)-f(x, 1)\right) \big] p(y \mid x ; A=1) p(x) d y d x    \\
+%              & =\int_{\mathcal X} \big [ \int_{\mathcal Y} (y-\mu_1(x))^{2} p(y \mid x ; A=1) dy\big ] p(x)dx + \int_{\mathcal X \times \mathcal Y} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x)p(y \mid x ; A=1)dx dy \\
+%              & + \qquad 2 \int_{\mathcal X} \big [ \int_{\mathcal Y} \left(y-\mu_{1}(x)\right) p(y \mid x ; A=1) dy \big ] \left(\mu_{1}(x)-f(x, 1)\right)p(x)dx                                                    \\
+%              & =\int_{\mathcal X} \sigma_{y}^{2}(x, 1) p(x) d x +\int_{\mathcal X} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x) d x+0
+%         \end{align*}
+%     \end{proof}
+% \end{lemma}
+
+% \begin{proposition*}[Upper bound with mu-IPW]\label{apd:proofs:prop:upper_bound}
+%     Let f be a given outcome model, let the weighting function $w$ be the Inverse
+%     Propensity Weight $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$. Then, under
+%     overlap (assumption \ref{assumption:overlap}),
+%     \begin{align*}
+%         \tau\text{-risk}(f) \leq & \; 2 \, \mu\text{-risk}_{IPW}(w, f)  \; - 2 \, (\sigma^2_{Bayes}(1) +  \sigma^2_{Bayes}(0))
+%     \end{align*}
+
+%     \begin{proof}
+%         \begin{align*}
+%              & \tau\text{-risk}(f) =\int_{\mathcal X}(\mu_{1}(x)-\mu_{0}(x)-(f(x ; 1)-f(x ; 0))^{2} p(x) d x
+%         \end{align*}
+%         By the triangle inequality $(u+v)^2 \leq 2(u^2 + v^2)$:
+%         \begin{align*}
+%              & \tau\text{-risk}(f) \leq
+%             2 \int_{\mathcal X}\big[\left(\mu_{1}(x)-f(x ; 1)\right)^{2}+\left(\mu_{0}(x)-f(x ; 0)\right)^{2}\big] p(x) d x \\
+%         \end{align*}
+%         Applying Lemma \ref{apd:proofs:mu_risk_ipw_link_mu},
+%         \begin{align*}
+%              & \tau\text{-risk}(f) \leq 2\big[p_A \mu\text{-risk}_{IPW, 1}(w, f) +(1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)(w, f)\big] -2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1)) \\
+%              & = 2 \mu\text{-risk}_{IPW}(w, f)-2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1))
+%         \end{align*}
+%     \end{proof}
+% \end{proposition*}
+
+\subsection{Reformulation of the $R\text{-risk}$ as reweighted
+    $\tau\text{-risk}$}\label{apd:proofs:r_risk_rewrite}
+
+\begin{proposition*}[$R\text{-risk}$ as reweighted $\tau
+            \text{-risk}$]\label{apd:proofs:prop:r_risk_rewrite}
+
+    \begin{proof}
+
+        We consider the R-decomposition: \cite{robinson_rootnconsistent_1988},
+        \begin{equation}\label{apd:eq:r_decomposition}
+            y(a) = m(x) + \big( a - e(x) \big) \tau(x) + \varepsilon(x; a)
+        \end{equation}
+        Where $\mathbb E[\varepsilon(X; A)|X, A] = 0$ We can use it as plug in the
+        $R\text{-risk}$ formula:
+
+        \begin{align*}
+             & R\text {-risk}(f) =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}[(y-m(x))-\big(a-e(x)\big) \tau_f(x)]^{2} p(y ; x ; a) d y d x d a                                     \\
+             & =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}} \left[\big(a-e(x)\big)\tau(x)+\varepsilon(x ; a)-\big(a-e(x)\big) \tau_f(x)\right]^{2} p(y ; x ; a) d y d x da          \\
+             & =\int_{\mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)^{2}\big(\tau(x)- \tau_f(x)\big)^{2} p(x ; a) d x d a                                                                       \\
+             & + 2  \int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)\big(\tau(x)-\tau_f(x)\big)  \int_{\mathcal{Y}} \varepsilon(x ; a) p(y \mid x ; a) d y p(x ; a)dx da \\
+             & +\int_{\mathcal{X} \times \mathcal{A}} \int_{\mathcal{Y}} \varepsilon^{2}(x ; a) p(y \mid x ; a) d y p(x ; a) d x d a
+        \end{align*}
+
+        The first term can be decomposed on control and treated populations to force
+        $e(x)$ to appear:
+        \begin{align*}
+             & \int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}\left[e(x)^{2}p(x;0) + \big(1-e(x)\big)^{2} p(x;1)\right] d x                    \\
+             & =\int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}  \left[e(x)^{2}\big(1-e(x)\big)p(x) + \big(1-e(x)\big)^{2}e(x) p(x)\right] d x \\
+             & =\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x)[1-e(x)+e(x)] p(x) d x                                                     \\ &=\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x) p(x) d x.
+        \end{align*}
+
+        The second term is null since, $\mathbb E[\varepsilon(x, a) |X, A]=0$.
+
+        The third term corresponds to the modulated residuals \ref{eq:residuals} :
+        $\tilde{\sigma}_B^2(0) + \tilde{\sigma}_B^2(1)$
+
+    \end{proof}
+\end{proposition*}
+
+\subsection{Interesting special cases}\label{apd:theory:special_cases}
 
@@
 between $\tau\text{-risk}$ and $\mu\text{-risk}_{IPW}$ coming from heterogeneous
-errors between populations as noted by Proposition \ref{theory:prop:mu_risk_ipw_bound}
-and shown experimentally in \citet{schuler_comparison_2018} and our
+errors between populations as shown experimentally in \citet{schuler_comparison_2018} and our
 results below.
@@
 choices must be made in non-asymptotic regimes.
-
-\section{Empirical Study}\label{sec:empirical_study}
-
-%We now explore empirically the behavior of these different causal selection
-%metrics.
-%\subsection{Causal metrics under evaluation}
-
-We evaluate the following causal metrics, oracle and feasible
-versions, presented in Table
-\ref{tab:evaluation_metrics}:\\
-$\widehat{\mu\text{-risk}}_{IPW}^*$,
-$\widehat{R\text{-risk}}^*$,
-$\widehat{U\text{-risk}}^*$,
-$\widehat{\tau\text{-risk}_{IPW}}^*$,
-$\widehat{\mu\text{-risk}}$,
-$\widehat{\mu\text{-risk}}_{IPW}$,
-$\widehat{R\text{-risk}}$,
-$\widehat{U\text{-risk}}$,
-$\widehat{\tau\text{-risk}_{IPW}}$.
-We benchmark the metrics in a variety of settings:
-many different simulated data generation
-processes and three semi-simulated datasets \footnote{Scripts for the simulations and the selection procedure are available at
-    \url{https://github.com/soda-inria/caussim}.
-}.
-
-\subsection{Caussim: Extensive simulation settings}\label{subsec:simulations}
-
-\paragraph{Data Generation}
-
-We use simulated data, on which the ground-truth causal effect is known. Going
-beyond prior empirical studies of causal model selection
-\cite{schuler_comparison_2018,alaa_validating_2019}, we use many
-generative processes, to reach more general conclusions (as discussed
-in \ref{apd:results:fig:seed_effect}).
-
-We generate the response functions using random bases. Basis extension methods
-are common in biostatistics, \emph{eg} functional regression with splines
-\cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
-vary at specific knots, they give flexible non-linear models. The number of
-knots controls the complexity of the function. We use random approximation of
-Radial Basis Function (RBF) kernels \cite{rahimi_random_2008} to generate the
-outcome and treatment functions. RBF use the same process as polynomial splines
-but replace polynomial by Gaussian kernels. Unlike polynomials, Gaussian kernels
-have decreasing influences in the input space. This avoids unrealistic
-divergences of the response surfaces at the ends of the feature space. We
-generate 1\,000 datasets based on this functions, with random overlap
-parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
-\ref{apd:experiments:generation}.
-
-\captionsetup[sub]{font=large,labelfont={bf,sf}}%
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\linewidth]{images/caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
-
-    \caption{Example of the simulation setup in the input space with two
-        knots --\emph{ie.}basis functions. The top panel
-        shows the observations in feature space, while the bottom panel displays the
-        two response surfaces on a 1D cut along the black lines drawn on
-        the top panel.}
-    \label{fig:simulation_examples}
-\end{figure}
-
-\paragraph{Family of candidate estimators}
-
-We compare different candidate estimators that approximate imperfectly the
-data-generating process. To build such estimators, we first use a RBF
-expansion similar to the one used for data generation. We choose two random
-knots and apply a transformation of the raw data features with a Gaussian
-kernel. This step is referred as the featurization. Then, we fit a linear
-regression on these transformed features. We consider two ways of combining
-these steps for outcome mode; we use common nomenclature
-\cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different
-meta-learners that differ on how they model, jointly or not, the treated and the
-non treated:
-\begin{itemize}
-    \item SLearner: A single learner for both population, taking the treatment as
-          a supplementary covariate.
-    \item SftLearner: A single set of basis functions is sampled at random for both
-          populations, leading to a given feature space used to model both the treat and
-          the non treated, then two
-          separate different regressors are fitted on this shared representation.
-    \item TLearner: Two completely different learners for each population, hence
-          separate feature representations and regressors.
-\end{itemize}
-
-We do not include more elaborated meta-learners such as R-learner
-\cite{nie_quasioracle_2017} or X-learner
-\cite{kunzel_metalearners_2019}. Our goal is not to have the best possible
-learner but to have a variety of sub-optimal learners to compare the
-different causal metrics. For the same reason, we did not include more powerful
-outcome models such as random forests or boosting trees.
-
-For the regression step, we fit a Ridge regression on the transformed features
-with 6 different choices of the regularization parameter $\lambda \in [10^{-3},
-        10^{-2}, 10^{-1}, 1, 10^{1}, 10^{2}]$, coupled with a TLearner or a SftLearner.
-We sample 10 different random basis for learning and
-featurization yielding a family $\mathcal F$ of 120 candidate estimators.
-%
-
-\subsection{Semi-simulated datasets}
-
-\paragraph{Datasets}\label{semi_simulated:datasets}
-
-We also use three benchmarks of the causal-inference literature, semi-simulated
-data adding a known synthetic causal effect to real --non synthetic-- healthcare
-covariate. ACIC 2016 \cite{dorie_automated_2019} is based on the Collaborative
-Perinatal Project \cite{niswander_women_1972}, a RCT studying infants’
-developmental disorders containing 4,802 indivduals and 55 features. We used 770
-dataset instances: 10 random seeds for each of the 77 simulated settings for the
-treatment and outcomes.  ACIC 2018 \cite{shimoni_benchmarking_2018} simulated
-treatment and outcomes for the Linked Births and Infant Deaths Database (LBIDD)
-\cite{macdorman_infant_1998} with $D=177$ covariates. We used all 432 datasets
-of size $N=5\,000$.  Twins \cite{louizos_causal_2017} is an augmentation of real
-data on twin births and mortality rates \cite{almond_costs_2005}. There are
-$N=11\,984$ samples, and $D=50$ covariates for which we simulated 1,000
-different treatment allocations. Datasets details are given in Appendix
-\ref{apd:experiments:datasets}.
-
-\paragraph{Family of candidate
-    estimators}\label{semi_simulated:candidate_estimators}
-
-For these three datasets, the family of candidate estimators are gradient
-boosting trees for both the response surfaces and the treatment
-\footnote{Scikit-learn regressor,
-    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingRegressor.html}{HistGradientBoostingRegressor},
-    and classifier,
-    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingClassifier.html}{HistGradientBoostingClassifier}.}
-with S-learner, learning rate
-in $\{0.01, 0.1, 1\}$, and maximum number of leaf nodes in $\{25, 27, 30, 32,
-    35, 40\}$ resulting in a family of size 18.
-
-\paragraph{Nuisance estimators}
-
-Drawing inspiration from the TMLE literature that uses combination of flexible
-machine learning methods \cite{schuler_targeted_2017}, we use a meta-learner as models
-for the nuisances $\check e$ (respectively $\check m$): a
-stacked estimator of ridge and boosting classifiers (respectively
-regressions). We select hyper-parameters with randomized search on a validation set
-$\mathcal{V}$ and keep them fix for model selection
-(\ref{apd:experiments:nuisances_hp} lists hyperparameters).
-
-\subsection{Measuring overlap between treated and non treated}\label{subsec:measuring_overlap}
-
-%\idea{Overlap is important but not easily measurable}
-Good overlap between treated and control population is
-crucial for causal inference as it is required by the positivity assumption
-\ref{assumption:overlap}. It is often assessed by
-comparing visually population distributions (as in Figure
-\ref{fig:toy_example}) or computing standardized difference on each
-feature
-\cite{austin_introduction_2011,austin_moving_2015}. While these methods are
-useful to decide if positivity holds, they do not yield a
-single measure. Rather, we compute the divergence
-between the population covariate distributions $\mathbb P(X|A=0)$ and $\mathbb
-    P(X|A=1)$
-\cite{damour_overlap_2020,johansson2022generalization}. We introduce the
-Normalized Total Variation (NTV), a divergence based on the sole propensity
-score (see \ref{apd:motivation_ntv}).
-
-\begin{figure}[!b]
-    \centering
-    \includegraphics[width=\linewidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
-    %\hfill
-    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$.
-        Strong and Weak overlap correspond to the first and last tertiles of the overlap distribution measured with
-        Normalized Total Variation eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents the same results
-        by adding semi-oracle risks in Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics}, measured with
-        absolute Kendall's in Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with $\tau\mathrm{-risk}$
-        gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table
-        \ref{apd:table:relative_kendalls_all_datasets} gives median and
-        IQR of the relative Kendall.}\label{fig:relative_kendalls_all_datasets}
-\end{figure}
-
-\subsection{Results: factors driving good model selection}\label{empirical_study:results}
-
-\paragraph{The $R\text{-risk}$ is the best metric}
-
-Each metric ranks differently the candidate models. Figure
-\ref{fig:relative_kendalls_all_datasets} shows the agreement between the
-ideal ranking of methods given the oracle $\tau\text{-risk}$ and
-the different feasible causal metrics. We measure this agreement
-with a relative\footnote{To remove the variance
-    across datasets (some datasets lead to easier model selection than
-    others), we report values for one metric relative to the mean of all
-    metrics for a given dataset instance: $\text{Relative} \, \kappa(\ell,\tau\mathrm{{-risk}})=
-        \kappa(\ell,\tau\mathrm{{-risk}}) -
-        mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$ (eq.
-\ref{eq:kendall_tau}) \cite{kendall_new_1938}.
-Given the importance of overlap in
-how well metrics approximate the oracle $\tau\text{-risk}$
-(\ref{apd:proofs:mu_risk_ipw_bound}), we separate
-strong and weak overlap.
-
-Among all metrics, the classical mean squared error (ie. factual
-$\mu\text{-risk}$) is worse and reweighting it with propensity score
-($\mu\text{-risk}_{IPW}$) does not bring much improvements. The
-$R\text{-risk}$, which includes a model of mean outcome and propensity
-scores, leads to the best performances. Interestingly, the
-$U\text{-risk}$, which uses the same nuisances, deteriorates in weak overlap, probably due to variance
-inflation when dividing by extreme propensity scores.
-
-Beyond rankings, the differences in terms of absolute
-ability to select the best model are large: The R-risk selects a model
-with a $\tau\text{-risk}$ only 1\% higher
-than the best
-possible candidate for strong overlap on Caussim, but selecting with
-the $\mu\text{-risk}$ or $\mu\text{-risk}_{IPW}$ --as per machine-learning
-practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
---as in some causal-inference methods \citep{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
-(Figure
-\ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
-datasets, the $R\text{-risk}$ consistently decreases the
-risk compared to the $\mu\text{-risk}$:
-0.1\% compared to 1\% on ACIC2016,  1\% compared to 20\% on ACIC2018,
-and 0.05\% compared to 1\% on Twins.
-
-\paragraph{Model selection is harder for low population
-    overlap}
-
-Model selection for causal inference becomes more and more difficult with
-increasingly different treated and control populations (Figure
-\ref{fig:all_datasets_overlap_effect_r_risk}). The absolute Kendall's
-coefficient correlation with $\tau\text{-risk}$ drops from values around 0.9
-(excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
-(\ref{apd:experiments:additional_results}).
-
-\begin{figure}[!h]
-    \centering
-    \includegraphics[width=\linewidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
-    %\hfill
-    \caption{\textbf{Model selection is harder for low population
-            overlap}:
-        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
-        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents results for all
-        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
-        Kendall's and continuous overlap values in Figure
-            {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
-\end{figure}
-
-\paragraph{Nuisances can be estimated on the same data as outcome models}
-
-Using the train set $\mathcal{T}$ both to fit the candidate estimator and the
-nuisance estimates is a form of double dipping which can lead errors in
-nuisances correlated to that of outcome models
-\cite{nie_quasioracle_2017}. In theory, these correlations can bias model
-selection and, strictly speaking, push
-to split out a third separated data set --a ``nuisance set''-- to fit the
-nuisance models. The drawback is that it depletes the data available for
-model estimation and selection. However, Figure
-\ref{fig:procedures_comparison} shows no substantial difference between a procedure with a separated
-nuisance set and the simpler shared nuisance-candidate set procedure.
-
-\begin{figure}[!tb]
-    \includegraphics[width=\linewidth]{images/_3_procedure_r_risk_only_3datasets_twocols.pdf}
-
-    \caption{\textbf{Nuisances can be estimated on the same data as outcome
-            models}: Results for the R-risk are similar between the
-        \textcolor{MidnightBlue}{shared
-            nuisances/candidate set} and
-        the \textcolor{RedOrange}{separated nuisances set} procedures. Figure
-        \ref{apd:fig:procedures_comparison_all_metrics} details results for all metrics.}\label{fig:procedures_comparison}
-\end{figure}
-
-\paragraph{Stacked models are good overall estimators of nuisances}
-
-For every risk, the oracle version recovers better the best estimator.
-However,
-stacked nuisances estimators (boosting and linear) lead to feasible
-metrics with close performances to the oracles ones: the
-corresponding estimators recover well-enough the true nuisances.
-One may wonder if simpler models for the nuisance could be useful,
-in particular in data-poor settings or when the true models are linear.
-Figure \ref{fig:all_datasets_nuisances_comparison} compares causal model
-selection estimating nuisances with stacked estimators or linear model.
-It comprises the Twins data, where the true propensity model is linear,
-and a downsampled version of this data, to study a situation favorable to
-linear models. In these settings,
-stacked and linear estimations of the nuisances performs equivalently.
-Detailed analysis (Figure \ref{apd:fig:nuisances_comparison_twins})
-confirms that using adaptive models --as built by
-stacking linear models and gradient-boosted trees-- suffices to estimate nuisance.
-
-\begin{figure}[!tb]
-    \includegraphics[width=\linewidth]{images/_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
-
-    \caption{\textbf{\textcolor{DarkOrchid}{Stacked
-                models} are good overall estimators of the nuisances}:
-        Results are shown only for the
-        R-risk; Figure \ref{apd:fig:nuisances_comparison}
-        details every metrics. For Twins, where the true propensity
-        model is linear, \textcolor{DarkOrchid}{stacked} and
-        \textcolor{ForestGreen}{linear}
-        estimations of the nuisances performs equivalently, even for a downsampled version
-        (N=4794). }\label{fig:all_datasets_nuisances_comparison}
-\end{figure}
-
-\paragraph{Use 90\% of the data to estimate outcome models, 10\% to
-    select them}
-
-The analyst faces a compromise: given a finite
-data sample, should she allocate more data to estimate the outcome model,
-thus improving the quality of the outcome model but leaving
-little data for model selection. Or, she could choose a bigger test set for
-model selection and effect estimation. For causal model selection, there
-is no established practice (as reviewed in \ref{apd:results:k_fold_choices}).
-
-We investigate such tradeoff varying the ratio between train and test
-data size. For this, we first split out 30\% of the data as a holdout set
-$\mathcal{V}$ on which we use the oracle response functions to derive
-silver-standard estimates of causal quantities. We then
-use the standard estimation procedure on the remaining 70\% of the data,
-splitting it into train $\mathcal{T}$ and test $\mathcal{S}$ of varying
-sizes. We finally measure the error between this estimate and the
-silver standard.
-
-We consider two different analytic goals: estimating a average
-treatment effect --a single number used for policy making-- and a
-CATE --a full model of the treatment effect as a function of covariates
-$X$. Given that the latter is a much more complex object than the former,
-the optimal train/test ratio might vary. To measure errors, we use for
-the ATE the relative absolute ATE bias between the ATE computed with the
-selected outcome model on the test set, and the true ATE as evaluated on
-the holdout set $\mathcal{V}$. For the CATE, we compare the
-$\tau\text{-risk}$
-of the best selected model applied on the holdout set $\mathcal{V}$. We explore this trade-off for the ACIC 2016 dataset and the R-risk.
-
-\begin{figure}[!t]
-    \begin{minipage}{.5\textwidth}
-        \centerline{\textbf{a) CATE estimation error}}
-        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16_twocols.pdf}
-    \end{minipage}
-    %\hspace{0.3\textwidth}
-    \begin{minipage}{.5\textwidth}
-        \centerline{\textbf{b) ATE estimation error}}
-        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16_twocols.pdf}
-    \end{minipage}%
-    \caption{\textbf{a) For CATE, a train/test ratio of 0.9/0.1 appears a good
-            trade-off.} b) For ATE, there is a small signal pointing also to
-        0.9/0.1 (K=10).
-        for ATE. Experiences on 10 replications of all 78 instances of the ACIC 2016
-        data.}\label{fig:train_test_ratio}
-\end{figure}
-
-Figure \ref{fig:train_test_ratio} shows that a train/test ratio of
-0.9/0.1 (K=10) or 0.8/0.2 (K=5) appears best to estimate CATE and
-ATE.
-
-\section{Discussion and conclusion}\label{sec:discussion}
-
-Predictive models are increasingly used to reason about causal effects,
-for instance in precision medicine to drive individualized decision.
-Our results highlight that they should be selected, validated, and tuned
-using different procedures and error measures than those classically used
-to assess prediction (estimating the so-called $\mu\text{-risk}$).
-Rather, selecting the best outcome model according to the $R\text{-risk}$
-(eq.\, Definition \ref{def:r_risk}) leads to more valid causal estimates.
-
-\paragraph{Nuisance models: more gain than pain}
-%
-Estimating the $R\text{-risk}$ requires a more complex procedure than
-standard cross-validation used \emph{e.g.}~in machine learning: it involves
-fitting nuisance models necessary for model evaluation, though our
-results show that these can be learned on the same set of data
-as the outcome model evaluated.
-
-The nuisance models must be well estimated (Figure
-\ref{fig:all_datasets_nuisances_comparison}). However these models are
-easier to select and control than a causally-valid outcome model,
-as they are associated to errors on
-observed distributions. Our results show that using for nuisance models
-a flexible stacking-based family of estimator suffices for good model selection.
-%
-In fact, a feasible $R\text{-risk}$ --where the nuisances
-are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
-nuisances are known. This may be explained by
-results that suggest that estimation errors on both
-nuisances partly compensate out in the
-$R\text{-risk}$ \cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.
-
-Note that propensity score models must be selected to estimate
-the individual posterior probability. For this, we used the Brier score,
-is minimized by the true individual probability. An easy mistake is to
-use calibration errors popular in machine learning
-\cite{platt_probabilistic_1999,zadrozny_obtaining_2001,niculescu-mizil_predicting_2005,minderer_revisiting_2021}
-as these select not for the individual posterior probability but for an
-aggregate error rate \cite{perez2022beyond}.
-
-\paragraph{Extension to binary outcomes}
-While we focused on continuous outcomes, in medicine, the target outcome
-is often a categorical variable such as mortality status or diagnosis. In
-this case, it may be interesting to focus on other estimands than the
-Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
-$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
-or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
-    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
-\cite{austin2017estimating}. While the odds ratio is natural for
-case-control studies \cite{rothman2008case}, other measures
-can reduce heterogeneity \cite{colnet2023risk}.
-In the log domain, the ratios are written as a difference, the
-framework studied here (\autoref{sec:neyman_rubin}) can directly apply.
-In particular, the log odds ratio is estimated by the common
-cross-entropy loss (or log loss) as in logistic regression.
-
-\paragraph{More $R\text{-risk}$ to select models driving decisions}
-
-Prediction models have flourished because their predictions can be easily
-demonstrated and validated on left-out data. But they require more
-careful validation for decision making, using a metric accounting for the
-putative intervention, the $R\text{-risk}$. Even when treated and
-untreated population differ little, as in RCTs, the $R\text{-risk}$
-brings a sizeable benefit. To facilitate better model selection, we provide Python
-code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
-Using the $R\text{-risk}$ does make evaluation more
-complicated not only because the procedure is more involved, but also
-because each intervention requires a dedicated evaluation. However, such
-off-policy evaluation remains much less costly than the recommended good
-practice of impact evaluation testing the ability of a prediction model
-to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
-the model-selection procedure puts no constraints on the models used to
-build predictive models: it opens the door to evaluating a wide range of
-models, from gradient boosting to convolutional neutral, or language
-models.
-
-\section*{Acknowledgments}
-
-We acknowledge fruitful discussions with Bénédicte Colnet.
-
-\subsection*{Financial disclosure}
-
-None reported.
-
-\subsection*{Conflict of interest}
-
-The authors declare no potential conflict of interests.
-
-%\if@Mybibrefstyle%
-\printbibliography
-%\fi if using njdnatbib.sty \bibliography{bibliography}%
-
-\clearpage
-
-\onecolumn
-
-\appendix
-%\setcounter{secnumdepth}{0}
-\setcounter{secnumdepth}{3}
-
-%The following supporting information is available as part of the online article:
-
-\section{Variability of ATE estimation on ACIC
-  2016}\label{apd:toy_example:acic_2016_ate_variability}
-
-Figure \ref{fig:acic_2016_ate_heterogeneity} shows ATE estimations for six
-different models used in g-computation estimators on the 76 configurations of
-the ACIC 2016 dataset. Outcome models are fitted on half of the data and
-inference is done on the other half --ie. train/test with a split ratio of 0.5.
-For each configuration, and each model, this train test split was repeated ten
-times, yielding non parametric variance estimates
-\cite{bouthillier_accounting_2021}. Figure \ref{fig:acic_2016_ate_heterogeneity}
-shows large variations obtained across different outcome estimators on
-semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models such as
-random forests are doing well in most settings except when treated and untreated
-populations differ noticeably, in which case a linear model (ridge) is to be
-preferred. However random forests with different hyper-parameters (max depth= 2)
-yield poor estimates. A simple rule of thumb such as preferring flexible models
-does not work in general; model selection is needed.
-
-%\idea{It is crucial to have tools for model selection.}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.95\linewidth]{images/2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
-    \caption{\textbf{Different outcome models lead to different
-            estimation errors on the Average Treatment Effects},
-        on 77 classic simulations with known true causal effect
-        \cite{dorie_automated_2019}. The different models are ridge regression
-        and random forests with different hyper-parameters
-        (details
-        \ref{apd:toy_example:acic_2016_ate_variability}). The different configurations are
-        plotted as a function of increasing difference between treated and
-        untreated population --see
-        \autoref{subsec:measuring_overlap}.
-        There is no systematic best performer; data-driven model
-        selection is important.
-        \label{fig:acic_2016_ate_heterogeneity}%
-    }
-\end{figure}
-
-Outcome models are implemented with
-\href{https://scikit-learn.org/stable/}{scikit-learn}
-\cite{pedregosa_scikitlearn_2011} and the following hyper-parameters:
-
-\begin{table}[h!]
-    \centering
-    \begin{tabular}{llll}
-        \toprule
-        Outcome Model                                  & Hyper-parameters grid
-        \\
-        \midrule
-        Random Forests                                 & Max depth: [2,
-        10]                                                                    \\
-
-        Ridge regression without treatment interaction & Ridge regularization:
-        [0.1]                                                                  \\
-
-        Ridge regression with treatment interaction    & Ridge regularization:
-        [0.1]                                                                  \\
-        \bottomrule
-    \end{tabular}
-    \caption{Hyper-parameters grid used for ACIC 2016 ATE variability}
-    \label{apd:toy_example:acic_2016_ate_variability:table}
-\end{table}
-
-\section{Causal assumptions}\label{apd:causal_assumptions}
-
-We assume the following four assumptions, referred as strong ignorability and
-necessary to assure identifiability of the causal estimands with observational
-data \cite{rubin_causal_2005}:
-\begin{assumption}[Unconfoundedness]\label{assumption:ignorability}
-    \begin{equation*}\label{eq:ignorability}
-        \{Y(0), Y(1) \} \indep A | X
-    \end{equation*}
-    This condition --also called ignorability-- is equivalent to the conditional
-    independence on $e(X)$ \cite{rosenbaum_central_1983}: $\{Y(0), Y(1) \}
-        \indep  A | e(X)$.
-\end{assumption}
-
-\begin{assumption}[Overlap, also known as Positivity)]\label{assumption:overlap}
-    \begin{equation*}\label{eq:overlap}
-        \eta < e(x) < 1 - \eta \quad \forall x \in \mathcal X \text{ and some } \eta > 0
-    \end{equation*}
-    The treatment is not perfectly predictable. Or with different words, every
-    patient has a chance to be treated and not to be treated. For a given set of
-    covariates, we need examples of both to recover the ATE.
-\end{assumption}
-
-As noted by \cite{damour_overlap_2020}, the choice of covariates $X$ can
-be viewed as a trade-off between these two central assumptions. A bigger
-covariates set generally reinforces the ignorability assumption. In the
-contrary, overlap can be weakened by large $\mathcal{X}$ because of the
-potential inclusion of instruments: variables only linked to the treatment which
-could lead to arbitrarily small propensity scores.
-
-% remark: There is a major counter example to these colliders (variables which
-% are caused by both the outcome and the treatment),
-
-\begin{assumption}[Consistency]\label{assumption:consistency} The observed
-    outcome is the potential outcome of the assigned treatment:
-    \begin{equation*}\label{eq:consistancy}
-        Y = A \, Y(1) + (1-A) \, Y(0)
-    \end{equation*}
-    Here, we assume that the intervention $A$ has been well defined. This
-    assumption focuses on the design of the experiment. It clearly states the link
-    between the observed outcome and the potential outcomes through the
-    intervention \cite{hernan_causal_2020}.
-\end{assumption}
-
-\begin{assumption}[Generalization]\label{assumption:generalization} The training
-    data on which we build the estimator and the test data on which we make the
-    estimation are drawn from the same distribution $\mathcal D^*$, also known as
-    the ``no covariate shift'' assumption \cite{jesson_identifying_2020}.
-\end{assumption}
-
-\section{Proofs: Links between feasible and oracle risks}\label{apd:proofs}
-
-\subsection{Upper bound of $\tau\text{-risk}$ with
-    $\mu\text{-risk}_{IPW}$}\label{apd:proofs:mu_risk_ipw_bound}
-
-For the bound with the $\mu\text{-risk}_{IPW}$, we will decompose the CATE risk
-on each factual population risks:
-
-\begin{definition}[Population Factual $\mu\text{-risk}$]\label{mu_risk_a}
-    \cite{shalit_estimating_2017}
-    \begin{equation*}
-        \mu\text{-risk}_{a}(f)= \int_{\mathcal Y \times \mathcal X} (y-f(x ; A=a))^{2}  p(y ; x=x \mid A=a) \; dy dx
-    \end{equation*}
-\end{definition}
-
-Applying Bayes rule, we can decompose the $\mu\text{-risk}$ on each
-intervention:
-\begin{equation*}
-    \mu\text{-risk}(f)
-    =p_{A} \,\mu\text{-risk}_{1}(f)+\left(1-p_{A}\right) \,\mu\text{-risk}_{0}(f)
-    \text{with } p_A=\mathbb P(A=1)
-\end{equation*}
-
-These definitions allows to state a intermediary result on each population:
-\begin{lemma}[Mean-variance decomposition]\label{apd:proofs:mu_risk_ipw_link_mu}
-    We need a reweighted version of the classical mean-variance decomposition.
-
-    For an outcome model $f: x \times A \rightarrow \mathcal X$. Let the inverse
-    propensity weighting function $w(a ; x)=a e(x)^{-1}+(1-a)(1-e(x))^{-1}$.
-    \begin{align*}
-         & \int_{\mathcal X}(\mu_{1}(x)-f(x ; 1))^{2} p(x) dx  = p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  -\sigma^{2}_{Bayes}(1)
-    \end{align*}
-    And
-    \begin{align*}
-         & \int_{\mathcal X}(\mu_{0}(x)-f(x; 0))^{2} p(x) dx   = (1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)  -\sigma^{2}_{Bayes}(0)
-    \end{align*}
-
-    \begin{proof}
-        \begin{align*}
-             & p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  = \int_{\mathcal X \times \mathcal Y} \frac{1}{e(x)}(y-f(x ; 1))^{2} p(y \mid x ; A=1) p(x ; A=1) d y d x                                                      \\
-             & = \int_{\mathcal X \times \mathcal Y} (y-f(x ; 1))^{2} p(y \mid x ; A=1) \frac{p(x ; A=1)}{p(x ; A=1)}p(x)dy dx                                                                                      \\
-             & = \int_{\mathcal X \times \mathcal Y} \big[(y-\mu_1(x))^{2}+\left(\mu_{1}(x)-f(x ; 1)\right)^{2} + 2\left(y-\mu_{1}(x)\right)\left(\mu_{1}(x)-f(x, 1)\right) \big] p(y \mid x ; A=1) p(x) d y d x    \\
-             & =\int_{\mathcal X} \big [ \int_{\mathcal Y} (y-\mu_1(x))^{2} p(y \mid x ; A=1) dy\big ] p(x)dx + \int_{\mathcal X \times \mathcal Y} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x)p(y \mid x ; A=1)dx dy \\
-             & + \qquad 2 \int_{\mathcal X} \big [ \int_{\mathcal Y} \left(y-\mu_{1}(x)\right) p(y \mid x ; A=1) dy \big ] \left(\mu_{1}(x)-f(x, 1)\right)p(x)dx                                                    \\
-             & =\int_{\mathcal X} \sigma_{y}^{2}(x, 1) p(x) d x +\int_{\mathcal X} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x) d x+0
-        \end{align*}
-    \end{proof}
-\end{lemma}
-
-\begin{proposition*}[Upper bound with mu-IPW]\label{apd:proofs:prop:upper_bound}
-    Let f be a given outcome model, let the weighting function $w$ be the Inverse
-    Propensity Weight $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$. Then, under
-    overlap (assumption \ref{assumption:overlap}),
-    \begin{align*}
-        \tau\text{-risk}(f) \leq & \; 2 \, \mu\text{-risk}_{IPW}(w, f)  \; - 2 \, (\sigma^2_{Bayes}(1) +  \sigma^2_{Bayes}(0))
-    \end{align*}
-
-    \begin{proof}
-        \begin{align*}
-             & \tau\text{-risk}(f) =\int_{\mathcal X}(\mu_{1}(x)-\mu_{0}(x)-(f(x ; 1)-f(x ; 0))^{2} p(x) d x
-        \end{align*}
-        By the triangle inequality $(u+v)^2 \leq 2(u^2 + v^2)$:
-        \begin{align*}
-             & \tau\text{-risk}(f) \leq
-            2 \int_{\mathcal X}\big[\left(\mu_{1}(x)-f(x ; 1)\right)^{2}+\left(\mu_{0}(x)-f(x ; 0)\right)^{2}\big] p(x) d x \\
-        \end{align*}
-        Applying Lemma \ref{apd:proofs:mu_risk_ipw_link_mu},
-        \begin{align*}
-             & \tau\text{-risk}(f) \leq 2\big[p_A \mu\text{-risk}_{IPW, 1}(w, f) +(1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)(w, f)\big] -2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1)) \\
-             & = 2 \mu\text{-risk}_{IPW}(w, f)-2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1))
-        \end{align*}
-    \end{proof}
-\end{proposition*}
-
-\subsection{Reformulation of the $R\text{-risk}$ as reweighted
-    $\tau\text{-risk}$}\label{apd:proofs:r_risk_rewrite}
-
-\begin{proposition*}[$R\text{-risk}$ as reweighted $\tau
-            \text{-risk}$]\label{apd:proofs:prop:r_risk_rewrite}
-
-    \begin{proof}
-
-        We consider the R-decomposition: \cite{robinson_rootnconsistent_1988},
-        \begin{equation}\label{apd:eq:r_decomposition}
-            y(a) = m(x) + \big( a - e(x) \big) \tau(x) + \varepsilon(x; a)
-        \end{equation}
-        Where $\mathbb E[\varepsilon(X; A)|X, A] = 0$ We can use it as plug in the
-        $R\text{-risk}$ formula:
-
-        \begin{align*}
-             & R\text {-risk}(f) =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}[(y-m(x))-\big(a-e(x)\big) \tau_f(x)]^{2} p(y ; x ; a) d y d x d a                                     \\
-             & =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}} \left[\big(a-e(x)\big)\tau(x)+\varepsilon(x ; a)-\big(a-e(x)\big) \tau_f(x)\right]^{2} p(y ; x ; a) d y d x da          \\
-             & =\int_{\mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)^{2}\big(\tau(x)- \tau_f(x)\big)^{2} p(x ; a) d x d a                                                                       \\
-             & + 2  \int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)\big(\tau(x)-\tau_f(x)\big)  \int_{\mathcal{Y}} \varepsilon(x ; a) p(y \mid x ; a) d y p(x ; a)dx da \\
-             & +\int_{\mathcal{X} \times \mathcal{A}} \int_{\mathcal{Y}} \varepsilon^{2}(x ; a) p(y \mid x ; a) d y p(x ; a) d x d a
-        \end{align*}
-
-        The first term can be decomposed on control and treated populations to force
-        $e(x)$ to appear:
-        \begin{align*}
-             & \int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}\left[e(x)^{2}p(x;0) + \big(1-e(x)\big)^{2} p(x;1)\right] d x                    \\
-             & =\int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}  \left[e(x)^{2}\big(1-e(x)\big)p(x) + \big(1-e(x)\big)^{2}e(x) p(x)\right] d x \\
-             & =\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x)[1-e(x)+e(x)] p(x) d x                                                     \\ &=\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x) p(x) d x.
-        \end{align*}
-
-        The second term is null since, $\mathbb E[\varepsilon(x, a) |X, A]=0$.
-
-        The third term corresponds to the modulated residuals \ref{eq:residuals} :
-        $\tilde{\sigma}_B^2(0) + \tilde{\sigma}_B^2(1)$
-
-    \end{proof}
-\end{proposition*}
 

### 2023-12-08 how_to_select_causal_models.tex
move definitions to appendix

@@
 \affil[2]{Mission Data, Haute Autorité de Santé, Saint-Denis, France}
-\affil[*]{Corresponding author: m.doutreligne@has-sante.fr}
+\affil[*]{Corresponding author: matt.dout@gmail.com}
 
@@
         treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$
-        --the $\tau\text{-risk}$, def.\,\ref{def:tau_risk} -- would lead to
+        --the $\tau\text{-risk}$, eq.\,\ref{eq:tau_risk} -- would lead to
         the best causal estimates; however computing this error is not
@@
 \section{Methods}\label{sec:framework}
-\subsection{The Neyman-Rubin Potential Outcomes framework}%
+\subsection{Neyman-Rubin Potential Outcomes framework}%
 \label{sec:neyman_rubin}%
@@
 or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
-    \{0, 1\}$ (eg.~a medical procedure, a drug administration), and baseline
+    \{0, 1\}$ (eg.~a medical procedure), and baseline
 covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
@@
 %
-These two expected outcomes can be computed from the observed data:
+These two expected outcomes can be estimated from observed data:
 the consistency \ref{assumption:consistency} and unconfoundedness
-\ref{assumption:ignorability} assumptions imply the equality of two different
-expectations:
+\ref{assumption:ignorability} assumptions imply the following equality:
 \begin{equation}\label{eq:mu_identification}
@@
 As we would like to target the CATE, the following
-evaluation risk is natural:
-\begin{definition}[$\tau\text{-risk}(f)$]\label{def:tau_risk} also called PEHE
-    \cite{schulam_reliable_2017, hill_bayesian_2011}:
-    \begin{equation*}\label{eq:tau_risk}
-        \tau\text{-risk}(f) = \mathbb E_{X\sim p(X)}[(\tau(X) - \hat \tau_f(X))^2]
-    \end{equation*}
-\end{definition}
+evaluation risk is natural (also called PEHE \cite{schulam_reliable_2017, hill_bayesian_2011}):
+\begin{flalign}\label{eq:tau_risk}
+    \tau\text{-risk}(f) \myeq \mathbb E_{X\sim p(X)}[(\tau(X) - \hat \tau_f(X))^2]
+\end{flalign}
 
@@
 
-\paragraph{Feasible error risks}\label{paragraph:feasible_metrics}
-\emph{Feasible} risks are based on the prediction error of the outcome model
-and \emph{observable} quantities.
-
-All expectations below are on observed distribution:
-$(Y, X, A) \sim \mathcal D$.
-
-\begin{definition}[Factual $\mu\text{-risk}$]\label{def:mu_risk}
-    \cite{shalit_estimating_2017} This is the usual Mean Squared Error on
-    the target y. It is what is typically meant by ``generalization error'' in
-    supervised learning:
-    \begin{equation*}\label{eq:mu_risk}
-        \mu\text{-risk}(f)=\mathbb{E}\left[(Y-f(X ; A))^2 \right]
-    \end{equation*}
-\end{definition}
-
-We now detail risks that use the nuisances $e$
---propensity score, def \ref{def:propensity_score}-- and $m$ --conditional mean
-outcome, def \ref{def:m}. We give the definitions as \textit{semi-oracles},
-function of the true unknown nuisances, but later instantiate them with estimated
-nuisances, noted $\big(\check e, \check m \big)$. Semi-oracles risks are
-superscripted with the $^{\star}$ symbol.
-
-\begin{definition}[$\mu\text{-risk}_{IPW}^{\star}$]\label{def:mu_ipw_risk}
-    \cite{vanderlaan_unified_2003} Let the inverse propensity weighting
-    function $w(x, a) = \frac{a}{e(x)} + \frac{1 - a}{1 - e(x)}$, we define the
-    semi-oracle Inverse Propensity Weighting risk,
-    \begin{equation*}\label{eq:mu_ipw_risk}
-        \mu\text{-risk}_{IPW}^{\star}(f) = \mathbb{E}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big) (Y-f(X ; A))^2 \right]
-    \end{equation*}
-\end{definition}
-
-\smallskip
-
-\begin{definition}[$\tau\text{-risk}^{\star}_{IPW}$]\label{def:tau_ipw_risk}
-    \cite{wager_estimation_2018}
-    The CATE $\tau(x)$ can be estimated with a
-    regression against inverse propensity weighted outcomes \cite{
-        athey2016recursive,gutierrez_causal_2016,wager_estimation_2018},
-    the $\tau\text{-risk}_{IPW}$.
-    \begin{equation*}
-        \tau\text{-risk}^{\star}_{IPW}(f) =\mathbb{E}
-        \left[ \Big(Y \frac{A - e(X)}{e(X)
-                (1-e(X))}-\tau_f\left(X\right)\Big)^2 \right]
-        %  =\mathbb{E}_{(Y, X, A)
-        %            \sim \mathcal D} \left[ \left(Y \left( \frac{A}{e(X)} -
-        %            \frac{1-A}{1-e(X)}\right)-\tau_f\left(X\right)\right)^2 \right]
-    \end{equation*}
-\end{definition}
-
-\begin{definition}[$U\text{-risk}^{\star}$]\label{def:u_risk}
-    \cite{kunzel_metalearners_2019,nie_quasioracle_2017} Based on
-    the Robinson decomposition --eq. \ref{eq:r_decomposition}, the U-learner
-    uses the $A-e(X)$ term
-    in the denominator. The derived risk is:
-    \begin{equation*}
-        U\text{-risk}^{\star}(f) =\mathbb{E}
-        \left[
-            \left( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -
-            \tau_f\left(X\right)\right)^{2} \right]
-    \end{equation*}
-    Note that extreme propensity weights in the
-    denominator term might inflate errors in the numerator due to imperfect
-    estimation of the mean outcome $m$.
-\end{definition}
-
-\begin{definition}[$R\text{-risk}^{\star}$]\label{def:r_risk}
-    \cite{nie_quasioracle_2017,schuler_comparison_2018}
-    The $R\text{-risk}$ also uses two nuisance $m$ and $e$:
-    \begin{equation*}
-        R\text{-risk}^{\star}(f) =\mathbb{E} \big[
-            \big(\left(Y-m\left(X\right)\right) -\left(A-e\left(X\right)\right) \tau_f\left(X\right)\big)^{2} \big]
-    \end{equation*}
-\end{definition}
-
-It is also based on the Robinson decomposition --eq. \ref{eq:r_decomposition}. %It performs well in various
-%simulations, estimating the nuisances
-%$(\check e, \check m)$ with
-%lasso, boosting or kernel ridge regressions
-%\cite{nie_quasioracle_2017}.
-
-These risks are summarized in Table \ref{tab:evaluation_metrics}.
+\paragraph{Feasible error risks}\label{paragraph:feasible_metrics} Table
+\ref{tab:evaluation_metrics} lists \emph{feasible} risks (Detailed in Appendix
+\ref{def:feasible_risks}), based on the prediction error of the outcome model
+and \emph{observable} quantities. These observable, called nuisances are $e$
+--propensity score, eq \ref{def:propensity_score}-- and $m$ --conditional mean
+outcome, eq \ref{def:m}. We give the definitions as \textit{semi-oracles},
+function of the true unknown nuisances, but later instantiate them with
+estimated nuisances, noted $\big(\check e, \check m \big)$. Semi-oracles risks
+are superscripted with the $^{\star}$ symbol.
 
@@
 
-Causal model selection (as in
-\autoref{eq:causal_model_selection}) may involve estimating various quantities
+Causal model selection (eq
+\ref{eq:causal_model_selection}) may involve estimating various quantities
 from the observed data: the outcome model $f$, its induced risk as
@@
 
-\section{A theoretical argument favoring the R-risk}\label{theory:r_risk_rewrite}
-
-We now derive a novel rewriting of the $R\text{-risk}$, making explicit its link
-with the oracle $\tau \text{-risk}$.
+\section{The R-risk, a reweighted oracle metric}\label{theory:r_risk_rewrite}
+
+The $R\text{-risk}$ can be rewritten as a rebalanced $\tau \text{-risk}$.
 
@@
 Good overlap between treated and control population is
-crucial for causal inference as it is required by the positivity assumption
-\ref{assumption:overlap}. It is often assessed by
-comparing visually population distributions (as in Figure
-\ref{fig:toy_example}) or computing standardized difference on each
+crucial for causal inference (positivity assumption
+\ref{assumption:overlap}). It is often assessed by
+comparing visually population distributions as in Figure
+\ref{fig:toy_example} or computing standardized difference on each
 feature
@@
 Normalized Total Variation (NTV), a divergence based on the sole propensity
-score (see \ref{apd:motivation_ntv}).
+score (Appendix \ref{apd:motivation_ntv}).
 
@@
 ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the different
-feasible causal metrics. We measure this agreement with a relative\footnote{To
+feasible causal metrics. We measure this agreement with relative\footnote{To
     remove the variance across datasets (some datasets lead to easier model
@@
 the individual posterior probability. For this, we used the Brier score,
-is minimized by the true individual probability. An easy mistake is to
+which is minimized by the true individual probability. An easy mistake is to
 use calibration errors popular in machine learning
@@
 In the log domain, the ratios are written as a difference, the
-framework studied here (\autoref{sec:neyman_rubin}) can directly apply.
+framework studied here can directly apply.
 In particular, the log odds ratio is estimated by the common
@@
 \end{assumption}
+
+\section{Definitions of feasible risks}\label{def:feasible_risks}
+
+\begin{definition}[Factual $\mu\text{-risk}$]\label{def:mu_risk}
+    \cite{shalit_estimating_2017} This is the usual Mean Squared Error on
+    the target y. It is what is typically meant by ``generalization error'' in
+    supervised learning:
+    \begin{equation*}\label{eq:mu_risk}
+        \mu\text{-risk}(f)=\mathbb{E}\left[(Y-f(X ; A))^2 \right]
+    \end{equation*}
+\end{definition}
+
+\begin{definition}[$\mu\text{-risk}_{IPW}^{\star}$]\label{def:mu_ipw_risk}
+    \cite{vanderlaan_unified_2003} Let the inverse propensity weighting
+    function $w(x, a) = \frac{a}{e(x)} + \frac{1 - a}{1 - e(x)}$, we define the
+    semi-oracle Inverse Propensity Weighting risk,
+    \begin{equation*}\label{eq:mu_ipw_risk}
+        \mu\text{-risk}_{IPW}^{\star}(f) = \mathbb{E}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big) (Y-f(X ; A))^2 \right]
+    \end{equation*}
+\end{definition}
+
+\smallskip
+
+\begin{definition}[$\tau\text{-risk}^{\star}_{IPW}$]\label{def:tau_ipw_risk}
+    \cite{wager_estimation_2018}
+    The CATE $\tau(x)$ can be estimated with a
+    regression against inverse propensity weighted outcomes \cite{
+        athey2016recursive,gutierrez_causal_2016,wager_estimation_2018},
+    the $\tau\text{-risk}_{IPW}$.
+    \begin{equation*}
+        \tau\text{-risk}^{\star}_{IPW}(f) =\mathbb{E}
+        \left[ \Big(Y \frac{A - e(X)}{e(X)
+                (1-e(X))}-\tau_f\left(X\right)\Big)^2 \right]
+        %  =\mathbb{E}_{(Y, X, A)
+        %            \sim \mathcal D} \left[ \left(Y \left( \frac{A}{e(X)} -
+        %            \frac{1-A}{1-e(X)}\right)-\tau_f\left(X\right)\right)^2 \right]
+    \end{equation*}
+\end{definition}
+
+\begin{definition}[$U\text{-risk}^{\star}$]\label{def:u_risk}
+    \cite{kunzel_metalearners_2019,nie_quasioracle_2017} Based on
+    the Robinson decomposition --eq. \ref{eq:r_decomposition}, the U-learner
+    uses the $A-e(X)$ term
+    in the denominator. The derived risk is:
+    \begin{equation*}
+        U\text{-risk}^{\star}(f) =\mathbb{E}
+        \left[
+            \left( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -
+            \tau_f\left(X\right)\right)^{2} \right]
+    \end{equation*}
+    Note that extreme propensity weights in the
+    denominator term might inflate errors in the numerator due to imperfect
+    estimation of the mean outcome $m$.
+\end{definition}
+
+\begin{definition}[$R\text{-risk}^{\star}$]\label{def:r_risk}
+    \cite{nie_quasioracle_2017,schuler_comparison_2018}
+    The $R\text{-risk}$ also uses two nuisance $m$ and $e$:
+    \begin{equation*}
+        R\text{-risk}^{\star}(f) =\mathbb{E} \big[
+            \big(\left(Y-m\left(X\right)\right) -\left(A-e\left(X\right)\right) \tau_f\left(X\right)\big)^{2} \big]
+    \end{equation*}
+\end{definition}
+
+It is also based on the Robinson decomposition --eq. \ref{eq:r_decomposition}. %It performs well in various
+%simulations, estimating the nuisances
+%$(\check e, \check m)$ with
+%lasso, boosting or kernel ridge regressions
+%\cite{nie_quasioracle_2017}.
 

### 2023-12-08 how_to_select_causal_models.tex
iter

@@
 making the $R\text{-risk}$ particularly suited for exploring
-heterogeneity (Details in Appendix \ref{apd:theory:special_cases})
+heterogeneity (Appendix \ref{apd:theory:special_cases})
 
@@
 
-We generate the response functions using random bases extension, a common method in
-biostatistics, \emph{eg} functional regression with splines
+We generate the response functions using random bases extension, a common method
+in biostatistics, \emph{eg} functional regression with splines
 \cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
-vary at specific knots, they give flexible non-linear models. The number of
-knots controls the complexity of the function. We use random approximation of
-Radial Basis Function (RBF) kernels \cite{rahimi_random_2008} to generate the
-outcome and treatment functions. RBF use the same process as polynomial splines
-but replace polynomial by Gaussian kernels. Unlike polynomials, Gaussian kernels
-have decreasing influences in the input space. This avoids unrealistic
-divergences of the response surfaces at the ends of the feature space. We
-generate 1\,000 datasets based on these functions, with random overlap
-parameters. Example shown in Figure \ref{fig:simulation_examples} and details in
-\ref{apd:experiments:generation}.
+vary at specific knots, we control the complexity of the non-linear outcome
+models. We use random approximation of Radial Basis Function (RBF) kernels
+\cite{rahimi_random_2008} to generate the outcome and treatment functions. RBF
+use the same process as polynomial splines but replace polynomial by Gaussian
+kernels. Unlike polynomial, Gaussian kernels have decreasing influences in the
+input space. This avoids unrealistic divergences of the response surfaces at the
+ends of the feature space. We generate 1\,000 datasets based on these functions,
+with random overlap parameters. Example shown in Figure
+\ref{fig:simulation_examples} and details in \ref{apd:experiments:generation}.
 
 \captionsetup[sub]{font=large,labelfont={bf,sf}}%
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=\linewidth]{images/caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
-
-    \caption{Example of the simulation setup in the input space with two
-        knots --\emph{ie.}basis functions. The top panel
-        shows the observations in feature space, while the bottom panel displays the
-        two response surfaces on a 1D cut along the black lines drawn on
-        the top panel.}
-    \label{fig:simulation_examples}
-\end{figure}
 
@@
 with a linear model on another random basis (which can be seen as a stochastic
-approximation of the full data kernel \cite{rahimi_random_2008}). We
-carefully control the amount of overlap between treated and control populations,
-a crucial assumption for causal inference.
+approximation of the full data kernel \cite{rahimi_random_2008}). We carefully
+control the amount of overlap between treated and control populations, a crucial
+assumption for causal inference. Figure \ref{fig:simulation_examples}
+illustrates 2D examples of the simulation.
 
@@
 parameters $\theta \in \left[ 0, 2.5 \right]$.
+
+\begin{figure}[!t]
+    \centering
+    \includegraphics[width=\linewidth]{images/caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
+
+    \caption{Example of the simulation setup in the input space with two
+        knots --\emph{ie.} basis functions. The top panel
+        shows the observations in feature space, while the bottom panel displays the
+        two response surfaces on a 1D cut along the black lines drawn on
+        the top panel.}
+    \label{fig:simulation_examples}
+\end{figure}
 

### 2023-12-08 how_to_select_causal_models.tex
iter

@@
 The wide variety of methods raises the problem
-of selecting between different estimators based on the data at hand.%TODO: might be removed since repeated two sentences below
+of selecting between different estimators based on the data at hand. %TODO: might be removed since repeated two sentences below
 %
@@
 We first illustrate the problem of causal model selection and briefly review
-prior art. Then we anchors causal model selection in the \emph{potential
+prior art. Then we anchor causal model selection in the \emph{potential
     outcome} framework and details the causal risks and model-selection procedure.
@@
         estimations of the nuisances performs equivalently, even for a downsampled version
-        (N=4794). }\label{fig:all_datasets_nuisances_comparison}
+        (N=4,794). }\label{fig:all_datasets_nuisances_comparison}
 \end{figure}
@@
 
-\section{Discussion and conclusion}\label{sec:discussion}
-
-Predictive models are increasingly used to reason about causal effects,
-for instance in precision medicine to drive individualized decision.
-Our results highlight that they should be selected, validated, and tuned
-using different procedures and error measures than those classically used
-to assess prediction (estimating the so-called $\mu\text{-risk}$).
-Rather, selecting the best outcome model according to the $R\text{-risk}$
-(eq.\, Definition \ref{def:r_risk}) leads to more valid causal estimates.
+\section{Discussion}\label{sec:discussion}
 
@@
 %
-Estimating the $R\text{-risk}$ requires a more complex procedure than
-standard cross-validation used \emph{e.g.}~in machine learning: it involves
-fitting nuisance models necessary for model evaluation, though our
-results show that these can be learned on the same set of data
-as the outcome model evaluated.
-
-The nuisance models must be well estimated (Figure
-\ref{fig:all_datasets_nuisances_comparison}). However these models are
-easier to select and control than a causally-valid outcome model,
-as they are associated to errors on
-observed distributions. Our results show that using for nuisance models
-a flexible stacking-based family of estimator suffices for good model selection.
-%
-In fact, a feasible $R\text{-risk}$ --where the nuisances
-are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
-nuisances are known. This may be explained by
-results that suggest that estimation errors on both
-nuisances partly compensate out in the
-$R\text{-risk}$ \cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.
-
-Note that propensity score models must be selected to estimate
-the individual posterior probability. For this, we used the Brier score,
+Predictive models are increasingly used to reason about causal effects, for
+instance in precision medicine to drive individualized decision. Our results
+highlight that they should be selected, validated, and tuned using different
+procedures and error measures than those classically used to assess prediction
+(estimating the so-called $\mu\text{-risk}$). Rather, selecting the best outcome
+model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
+to more valid causal estimates. Estimating the $R\text{-risk}$ requires a more
+complex procedure than standard cross-validation used \emph{e.g.}~in machine
+learning: it involves fitting nuisance models necessary for model evaluation,
+though our results show that these can be learned on the same set of data as the
+outcome model evaluated. The nuisance models must be well estimated (Figure
+\ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
+nuisance models a flexible stacking-based family of estimator suffices for good
+model selection. For propensity score, we used the Brier score,
 which is minimized by the true individual probability. An easy mistake is to
@@
 aggregate error rate \cite{perez2022beyond}.
+
+%
+% However these models are easier to select and control than a causally-valid
+% outcome model, as they are associated to errors on observed distributions. 
+% In fact, a feasible $R\text{-risk}$ --where the nuisances
+% are estimated-- performs almost as well as an oracle $R\text{-risk}$ --where the
+% nuisances are known. This may be explained by
+% results that suggest that estimation errors on both
+% nuisances partly compensate out in the
+% $R\text{-risk}$ \cite{daniel2018double,kennedy2020optimal,nie_quasioracle_2017,chernozhukov_double_2018,zivich2021machine,naimi2021challenges}.
 
@@
 
+\section{Conclusion}\label{sec:conclusion}
 \paragraph{More $R\text{-risk}$ to select models driving decisions}

### 2023-12-08 how_to_select_causal_models.tex
adding jamia refs on causal inference

@@
 gives an answer, but statistical validity requires causal inference
-\cite{snowden_implementation_2011,blakely2020reflection}.
+\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.
 
@@
 Causal inference has been central to epidemiology, typically with methods that
-model treatment assignment \cite{austin_moving_2015,grose_use_2020}, based on
+model treatment assignment \cite{austin_moving_2015,casucci2018estimating,grose_use_2020}, based on
 propensity scores \cite{rosenbaum_central_1983}. But outcome modeling go
@@
 Indeed, estimates of treatment effects can vary markedly across different
-predictive models (as shown by \cite{dorie_automated_2019} and developed in
+predictive models \cite{fang2019applying,dorie_automated_2019} (illustration in
 Appendix \ref{apd:toy_example:acic_2016_ate_variability}).

### 2023-12-08 how_to_select_causal_models.tex
shorter overlap measurement

@@
 %\idea{Overlap is important but not easily measurable}
-Good overlap between treated and control population is
-crucial for causal inference (positivity assumption
-\ref{assumption:overlap}). It is often assessed by
-comparing visually population distributions as in Figure
-\ref{fig:toy_example} or computing standardized difference on each
-feature
-\cite{austin_introduction_2011,austin_moving_2015}. While these methods are
-useful to decide if positivity holds, they do not yield a
-single measure. Rather, we compute the divergence
-between the population covariate distributions $\mathbb P(X|A=0)$ and $\mathbb
-    P(X|A=1)$
-\cite{damour_overlap_2020,johansson2022generalization}. We introduce the
-Normalized Total Variation (NTV), a divergence based on the sole propensity
-score (Appendix \ref{apd:motivation_ntv}).
+Good overlap between treated and control population is crucial for causal
+inference (positivity assumption \ref{assumption:overlap}). We introduce the
+Normalized Total Variation (NTV), a divergence based on the propensity score
+summarizing the overlap between both populations (Appendix
+\ref{apd:motivation_ntv}).
 
@@
 \paragraph{Motivation of the Normalized Total Variation}
+
 %\idea{A simplier measure, NTV seems to capture what we want}
-Computing overlap when working only on samples of the observed distribution,
-outside of simulation, requires a sophisticated estimator of discrepancy
-between distributions, as two data points never have the same exact set of
-features. Maximum Mean Discrepancy \cite{gretton2012kernel} is typically
-used in the context of causal inference
-\cite{shalit_estimating_2017,johansson2022generalization}. However it
-needs a kernel, typically Gaussian, to extrapolate across neighboring
-observations. We prefer avoiding the need to specify such a kernel, as it must
-be adapted to the data which is tricky with categorical or non-Gaussian
-features, a common situation for medical data.
+
+Overlap is often assessed by comparing visually population distributions as in
+Figure \ref{fig:toy_example} or computing standardized difference on each
+feature \cite{austin_introduction_2011,austin_moving_2015}. While these methods
+are useful to decide if positivity holds, they do not yield a single measure.
+Rather, we compute the divergence between the population covariate distributions
+$\mathbb P(X|A=0)$ and $\mathbb P(X|A=1)$
+\cite{damour_overlap_2020,johansson2022generalization}. Computing overlap when
+working only on samples of the observed distribution, outside of simulation,
+requires a sophisticated estimator of discrepancy between distributions, as two
+data points never have the same exact set of features. Maximum Mean Discrepancy
+\cite{gretton2012kernel} is typically used in the context of causal inference
+\cite{shalit_estimating_2017,johansson2022generalization}. However it needs a
+kernel, typically Gaussian, to extrapolate across neighboring observations. We
+prefer avoiding the need to specify such a kernel, as it must be adapted to the
+data which is tricky with categorical or non-Gaussian features, a common
+situation for medical data.
 

### 2023-12-08 how_to_select_causal_models.tex
move prior work

@@
 \twocolumn
-\section{Background}\label{sec:intro}
+\section{Introduction}\label{sec:intro}
 
@@
 
-\paragraph{Objectives and structure of the paper}
-
-We study \textit{model selection procedures}
-in practical settings: \textit{finite samples} settings and without
-\textit{well-specification} assumption. Asymptotic causal-inference
-theory calls for complex risks, but a practical question is
-whether model-selection procedures, that rely on data split, can estimate
-these risks reliably enough. Indeed, they
-come with more quantities to estimate, which may
-bring additional variance, leading to worse model selection.
+\section{Objectives and structure of the paper}
+
+We study \textit{model selection procedures} in practical settings:
+\textit{finite samples} without the \textit{well-specification}
+assumptions which are often made in the literature (Appendix
+\ref{apd:prior_work}). Asymptotic causal-inference theory calls for complex
+risks, but a practical question is whether model-selection procedures, that rely
+on data split, can estimate these risks reliably enough. Indeed, they come with
+more quantities to estimate, which may bring additional variance, leading to
+worse model selection.
 
 We first illustrate the problem of causal model selection and briefly review
-prior art. Then we anchor causal model selection in the \emph{potential
-    outcome} framework and details the causal risks and model-selection procedure.
-We then rewrite the so-called $R\text{-risk}$, as a reweighted version of mean
-squared difference between the true and estimated individualized treatment
-effect. Finally, we conduct a thorough empirical study comparing the different
-metrics on a wide variety of data-generative processes based on simulations and
-real data.
+prior art. Then we anchor causal model selection in the \emph{potential outcome}
+framework and details the causal risks and model-selection procedure. We then
+rewrite the so-called $R\text{-risk}$, as a reweighted version of mean squared
+difference between the true and estimated individualized treatment effect.
+Finally, we conduct a thorough empirical study comparing the different metrics
+on a wide variety of data-generative processes based on simulations and real
+data, going beyond prior work
+\citep{schuler_comparison_2018,alaa_validating_2019} (reviewed in Appendix
+\ref{apd:prior_work}).
 
@@
 
-\subsection{Prior work: model selection for outcome modeling (g-computation)}%
-
-A natural way to select a predictive model for causal inference would be
-an error measure between a causal quantity such as the CATE and models' estimate. But such error is
-not a ``feasible'' risk: it cannot be computed solely from observed data
-and requires oracle knowledge.
-
-% XXX: I think that I should make the following paragraph shorter
-
-\paragraph{Simulation studies of causal model selection}
-
-Using eight simulations setups from \cite{powers_methods_2018}, where
-the oracle CATE is known, \citet{schuler_comparison_2018} compare four
-causal risks, concluding that for CATE estimation the best
-model-selection risk is the so-called $R\text{-risk}$
-\cite{nie_quasioracle_2017} --def.\,\ref{def:r_risk}, below. Their
-empirical results are clear for randomized treatment allocation but less
-convincing for observational settings where both simple Mean Squared
-Error --MSE, $\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
-reweighted MSE --$\mu\text{-risk}_{IPW}$ def.\,\ref{def:mu_ipw_risk}--
-appear to perform better than $R\text{-risk}$ on half of the simulations.
-Another work \cite{alaa_validating_2019} studied empirically both MSE and
-reweighted MSE risks on the semi-synthetic ACIC 2016 datasets
-\cite{dorie_automated_2019}, but did not include the $R\text{-risk}$. We complete these
-prior empirical work by studying a wider variety of data generative
-processes and varying the influence of overlap, an important parameter of
-the data generation process which makes a given causal metric appropriate
-\cite{damour_overlap_2020}. We also study how to best adapt
-cross-validation procedures to causal metrics which themselves come with
-models to estimate.
-
-\paragraph{Theoretical studies of causal model selection}
-
-Several theoretical works have proposed causal model selection procedures
-that are \emph{consistent}: select the best model in a family given
-asymptotically large data. These work rely on introducing a
-CATE estimator in the testing procedure: matching
-\citep{rolling_model_2014}, an IPW estimate
-\citep{gutierrez_causal_2016}, a doubly robust estimator
-\citep{saito_counterfactual_2020}, or debiasing the error with influence
-functions \cite{alaa_validating_2019}. However, for theoretical
-guarantees to hold, the test-set correction needs to converge to the
-oracle: it needs to be flexible enough --well-posed-- and asymptotic
-data. From a practical perspective, meeting such requirements
-implies having a good CATE estimate, thus having solved
-the original problem of causal model selection.
-
-\paragraph{Statistical guarantees on causal estimation procedures}
-
-Much work in causal inference has focused on procedures that
-guarantee asymptotically consistent estimators, such as Targeted
-Machine Learning
-Estimation (TMLE) \cite{laan_targeted_2011,schuler_targeted_2017} or
-Double Machine Learning \cite{chernozhukov_double_2018}. Here also, theories require asymptotic regimes and
-models to be \textit{well-specified}.
-
-By contrast, \citet{johansson2022generalization} studies causal estimation
-without assuming that estimators are well specified. They derive an upper bound
-on the oracle error to the CATE ($\tau\text{-risk}$) that involves the error on
-the outcome and the similarity of the distributions of treated and control
-patients. However, they use this upper bound for model optimization,
-and do not give insights on model selection. In addition, for hyperparameter
-selection, they rely on a plugin estimate of the $\tau\text{-risk}$ built with
-counterfactual nearest neighbors, which has been shown ineffective
-\cite{schuler_comparison_2018}.
-
-%\section{Formal setting of causal inference and model selection}\label{sec:framework}
+% A natural way to select a predictive model for causal inference would be
+% an error measure between a causal quantity such as the CATE and models' estimate. But such error is
+% not a ``feasible'' risk: it cannot be computed solely from observed data
+% and requires oracle knowledge.
+
 \section{Methods}\label{sec:framework}
@@
 Drawing inspiration from the TMLE literature that uses combination of flexible
-machine learning methods \cite{schuler_targeted_2017}, we use a meta-learner as models
-for the nuisances $\check e$ (respectively $\check m$): a
-stacked estimator of ridge and boosting classifiers (respectively
-regressions). We select hyper-parameters with randomized search on a validation set
-$\mathcal{V}$ and keep them fix for model selection
-(\ref{apd:experiments:nuisances_hp} lists hyperparameters).
+machine learning methods \cite{schuler_targeted_2017}, we model the nuisances
+$\check e$ (respectivley $\check m$) with a meta-learner: a stacked estimator of ridge
+and boosting classifiers (respectively regressions) (hyperparameter selection in Appendix \ref{apd:experiments:nuisances_hp}).
 
@@
 Good overlap between treated and control population is crucial for causal
-inference (positivity assumption \ref{assumption:overlap}). We introduce the
+inference (Assumption \ref{assumption:overlap}). We introduce the
 Normalized Total Variation (NTV), a divergence based on the propensity score
@@
 \end{table}
+
+\section{Prior work : model selection for outcome modeling (g-computation)}\label{apd:prior_work}
+
+A natural way to select a predictive model for causal inference would be
+an error measure between a causal quantity such as the CATE and models' estimate. But such error is
+not a ``feasible'' risk: it cannot be computed solely from observed data
+and requires oracle knowledge.
+
+% XXX: I think that I should make the following paragraph shorter
+
+\paragraph{Simulation studies of causal model selection}
+
+Using eight simulations setups from \cite{powers_methods_2018}, where
+the oracle CATE is known, \citet{schuler_comparison_2018} compare four
+causal risks, concluding that for CATE estimation the best
+model-selection risk is the so-called $R\text{-risk}$
+\cite{nie_quasioracle_2017} --def.\,\ref{def:r_risk}, below. Their
+empirical results are clear for randomized treatment allocation but less
+convincing for observational settings where both simple Mean Squared
+Error --MSE, $\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
+reweighted MSE --$\mu\text{-risk}_{IPW}$ def.\,\ref{def:mu_ipw_risk}--
+appear to perform better than $R\text{-risk}$ on half of the simulations.
+Another work \cite{alaa_validating_2019} studied empirically both MSE and
+reweighted MSE risks on the semi-synthetic ACIC 2016 datasets
+\cite{dorie_automated_2019}, but did not include the $R\text{-risk}$. We complete these
+prior empirical work by studying a wider variety of data generative
+processes and varying the influence of overlap, an important parameter of
+the data generation process which makes a given causal metric appropriate
+\cite{damour_overlap_2020}. We also study how to best adapt
+cross-validation procedures to causal metrics which themselves come with
+models to estimate.
+
+\paragraph{Theoretical studies of causal model selection}
+
+Several theoretical works have proposed causal model selection procedures
+that are \emph{consistent}: select the best model in a family given
+asymptotically large data. These work rely on introducing a
+CATE estimator in the testing procedure: matching
+\citep{rolling_model_2014}, an IPW estimate
+\citep{gutierrez_causal_2016}, a doubly robust estimator
+\citep{saito_counterfactual_2020}, or debiasing the error with influence
+functions \cite{alaa_validating_2019}. However, for theoretical
+guarantees to hold, the test-set correction needs to converge to the
+oracle: it needs to be flexible enough --well-posed-- and asymptotic
+data. From a practical perspective, meeting such requirements
+implies having a good CATE estimate, thus having solved
+the original problem of causal model selection.
+
+\paragraph{Statistical guarantees on causal estimation procedures}
+
+Much work in causal inference has focused on procedures that
+guarantee asymptotically consistent estimators, such as Targeted
+Machine Learning
+Estimation (TMLE) \cite{laan_targeted_2011,schuler_targeted_2017} or
+Double Machine Learning \cite{chernozhukov_double_2018}. Here also, theories require asymptotic regimes and
+models to be \textit{well-specified}.
+
+By contrast, \citet{johansson2022generalization} studies causal estimation
+without assuming that estimators are well specified. They derive an upper bound
+on the oracle error to the CATE ($\tau\text{-risk}$) that involves the error on
+the outcome and the similarity of the distributions of treated and control
+patients. However, they use this upper bound for model optimization,
+and do not give insights on model selection. In addition, for hyperparameter
+selection, they rely on a plugin estimate of the $\tau\text{-risk}$ built with
+counterfactual nearest neighbors, which has been shown ineffective
+\cite{schuler_comparison_2018}.
 
@@
 The nuisances are estimated with a stacked regressor inspired by the Super
-Learner framework, \cite{laan_super_2007}). The hyper-parameters are
-optimized with a random search with following search grid detailed in Table
+Learner framework, \cite{laan_super_2007}). We select hyper-parameters
+with randomized search on a validation set $\mathcal{V}$ and keep them fix for
+model selection  The searc grid is detailed in Table
 \ref{apd:experiments:nuisances_hp_grid}. All implementations come from

### 2023-12-08 how_to_select_causal_models.tex
iter

@@
 on a wide variety of data-generative processes based on simulations and real
-data, going beyond prior work
-\citep{schuler_comparison_2018,alaa_validating_2019} (reviewed in Appendix
+data, going beyond prior work based on smaller simulation sets
+\citep{schuler_comparison_2018,alaa_validating_2019} (Appendix
 \ref{apd:prior_work}).

### 2023-12-13 how_to_select_causal_models.tex
condense: move datasplit result in appendix

@@
 
+Practically, we recommand a train/test split of 90 \%/10\%. Use most of the data
+to estimate both the nuisances and outcome models, then compute the risks on the
+remaining test set for model selection (emprical argument in Appendix
+\ref{apd:results:data_split}).
+
 \begin{figure}[!tb]
@@
 \end{figure}
-
-\paragraph{Use 90\% of the data to estimate outcome models, 10\% to
-    select them}
-
-The analyst faces a compromise: given a finite
-data sample, should she allocate more data to estimate the outcome model,
-thus improving the quality of the outcome model but leaving
-little data for model selection. Or, she could choose a bigger test set for
-model selection and effect estimation. For causal model selection, there
-is no established practice (as reviewed in \ref{apd:results:k_fold_choices}).
-
-We investigate such tradeoff varying the ratio between train and test
-data size. For this, we first split out 30\% of the data as a holdout set
-$\mathcal{V}$ on which we use the oracle response functions to derive
-silver-standard estimates of causal quantities. We then
-use the standard estimation procedure on the remaining 70\% of the data,
-splitting it into train $\mathcal{T}$ and test $\mathcal{S}$ of varying
-sizes. We finally measure the error between this estimate and the
-silver standard.
-
-We consider two different analytic goals: estimating a average
-treatment effect --a single number used for policy making-- and a
-CATE --a full model of the treatment effect as a function of covariates
-$X$. Given that the latter is a much more complex object than the former,
-the optimal train/test ratio might vary. To measure errors, we use for
-the ATE the relative absolute ATE bias between the ATE computed with the
-selected outcome model on the test set, and the true ATE as evaluated on
-the holdout set $\mathcal{V}$. For the CATE, we compare the
-$\tau\text{-risk}$
-of the best selected model applied on the holdout set $\mathcal{V}$. We explore this trade-off for the ACIC 2016 dataset and the R-risk.
-
-\begin{figure}[!t]
-    \begin{minipage}{.5\textwidth}
-        \centerline{\textbf{a) CATE estimation error}}
-        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16_twocols.pdf}
-    \end{minipage}
-    %\hspace{0.3\textwidth}
-    \begin{minipage}{.5\textwidth}
-        \centerline{\textbf{b) ATE estimation error}}
-        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16_twocols.pdf}
-    \end{minipage}%
-    \caption{\textbf{a) For CATE, a train/test ratio of 0.9/0.1 appears a good
-            trade-off.} b) For ATE, there is a small signal pointing also to
-        0.9/0.1 (K=10).
-        for ATE. Experiences on 10 replications of all 78 instances of the ACIC 2016
-        data.}\label{fig:train_test_ratio}
-\end{figure}
-
-Figure \ref{fig:train_test_ratio} shows that a train/test ratio of
-0.9/0.1 (K=10) or 0.8/0.2 (K=5) appears best to estimate CATE and
-ATE.
 
@@
 \paragraph{Selecting different seeds and parameters is crucial to draw
-    conclucions}\label{{apd:results:seed_effect}}
+    conclusions}\label{{apd:results:seed_effect}}
 
@@
 
-\section{Heterogeneity in practices for data split}\label{apd:results:k_fold_choices}
+\section{Data split choices}\label{apd:results:data_split}
+
+\subsection{Use 90\% of the data to estimate outcome models, 10\% to
+    select them}
+
+The analyst faces a compromise: given a finite
+data sample, should she allocate more data to estimate the outcome model,
+thus improving the quality of the outcome model but leaving
+little data for model selection. Or, she could choose a bigger test set for
+model selection and effect estimation. For causal model selection, there
+is no established practice (as reviewed in \ref{apd:results:k_fold_choices}).
+
+We investigate such tradeoff varying the ratio between train and test
+data size. For this, we first split out 30\% of the data as a holdout set
+$\mathcal{V}$ on which we use the oracle response functions to derive
+silver-standard estimates of causal quantities. We then
+use the standard estimation procedure on the remaining 70\% of the data,
+splitting it into train $\mathcal{T}$ and test $\mathcal{S}$ of varying
+sizes. We finally measure the error between this estimate and the
+silver standard.
+
+We consider two different analytic goals: estimating a average
+treatment effect --a single number used for policy making-- and a
+CATE --a full model of the treatment effect as a function of covariates
+$X$. Given that the latter is a much more complex object than the former,
+the optimal train/test ratio might vary. To measure errors, we use for
+the ATE the relative absolute ATE bias between the ATE computed with the
+selected outcome model on the test set, and the true ATE as evaluated on
+the holdout set $\mathcal{V}$. For the CATE, we compare the
+$\tau\text{-risk}$
+of the best selected model applied on the holdout set $\mathcal{V}$. We explore this trade-off for the ACIC 2016 dataset and the R-risk.
+
+\begin{figure}[!t]
+    \begin{minipage}{.5\textwidth}
+        \centerline{\textbf{a) CATE estimation error}}
+        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16_twocols.pdf}
+    \end{minipage}
+    %\hspace{0.3\textwidth}
+    \begin{minipage}{.5\textwidth}
+        \centerline{\textbf{b) ATE estimation error}}
+        \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16_twocols.pdf}
+    \end{minipage}%
+    \caption{\textbf{a) For CATE, a train/test ratio of 0.9/0.1 appears a good
+            trade-off.} b) For ATE, there is a small signal pointing also to
+        0.9/0.1 (K=10).
+        for ATE. Experiences on 10 replications of all 78 instances of the ACIC 2016
+        data.}\label{fig:train_test_ratio}
+\end{figure}
+
+Figure \ref{fig:train_test_ratio} shows that a train/test ratio of
+0.9/0.1 (K=10) or 0.8/0.2 (K=5) appears best to estimate CATE and
+ATE.
+
+\subsection{Heterogeneity in practices for data split}\label{apd:results:k_fold_choices}
 

### 2023-12-13 how_to_select_causal_models.tex
condense: rm duplicate in intro and rm a sentence in ccl

@@
 \newcommand{\indep}{\perp \!\!\! \perp}
+\newcommand{\eg}{\emph{e.g.}}
 \DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
@@
     \textbf{Objective:}
-    As predictive models --\emph{eg} from machine learning-- give likely
+    As predictive models --\eg from machine learning-- give likely
     outcomes, they may be used to reason on the effect of an intervention, a
@@
 effects is that of confounding: spurious associations between treatment
-allocation and baseline health, \emph{eg} only prescribing a drug to mild cases
+allocation and baseline health, \eg only prescribing a drug to mild cases
 \citep{hernan_causal_2020,vanderweele2019principles}.
@@
 
-% TODO:move below
-Standard practices to select models in predictive settings rely on
-the error on the outcome
-\cite{poldrack2020establishment,varoquaux2022evaluating}. However, as we
-will see, these practices may not pick the best models
-for causal inference, as they can be misled by inhomogeneities due to
-treatment allocation.
-%
-
-\section{Objectives and structure of the paper}
+\subsection{Objectives and structure of the paper}
 
@@
 on a wide variety of data-generative processes based on simulations and real
-data, going beyond prior work based on smaller simulation sets
+data, going beyond prior work based on specific simulation settings
 \citep{schuler_comparison_2018,alaa_validating_2019} (Appendix
@@
 
-\begin{figure}[b!]
+Using a predictor to reason on causal effects relies on contrasting the
+prediction of the outcome for a given individual with and without the treatment.
+%
+Given various predictors of the outcome, which one should we use?
+%
+Standard predictive modeling or machine-learning practice selects the predictor
+that minimizes the expected error on the outcome
+\citep{poldrack2020establishment,varoquaux2022evaluating}. However, this
+predictor may not be the best model to reason about causal effects of an
+intervention as Figure \ref{fig:toy_example} illustrates. Consider the
+probability $Y$ of an undesirable outcome (\eg death), a binary treatment $A \in
+    \{0, 1\}$, and a covariate $X \in \mathbb R$ summarizing the patient health
+status (\eg the Charlson index \cite{charlson_new_1987}). We simulate a
+treatment beneficial (decreases mortality) for patients with high Charlson
+scores (bad health status) but with little effect for patients in good condition
+(low Charlson scores).
+
+Figure \ref{fig:toy_example}a shows a random forest predictor with a
+counter-intuitive behavior: it predicts well on average the outcome (as measured
+by a regression $R^2$ score) but perform poorly to estimate causal quantities:
+the average treatment effect $\tau$ (as visible via the error $|\tau -
+    \hat{\tau}|$) or the conditional average treatment effect (the error
+$\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$, called CATE).
+%
+On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
+smaller $R^2$ score but better causal inference.%
+
+The problem is that causal estimation requires controlling
+an error on both treated and non-treated outcome for the same individual:
+the observed outcome, and the non-observed \emph{counterfactual} one.
+The linear model is misspecified --the outcome functions are not
+linear--, leading to poor $R^2$; but it interpolates better to regions
+where there are few untreated individuals --high Charlson score-- and
+thus gives better causal estimates. Conversely, the random forest puts
+weaker assumptions on the data, thus has higher $R^2$ score but is biased
+by the treated population in the poor-overlap region, leading
+to bad causal estimates.
+
+This toy example illustrates that the classic minimum Mean Square Error
+criterion is not suited to choosing a model among candidate
+estimators for causal inference.
+
+% A natural way to select a predictive model for causal inference would be
+
+\begin{figure}[h!]
     %\centering
@@
 
-Using a predictor to reason on causal effects relies on contrasting the
-prediction of the outcome for a given individual with and without the treatment.
-%
-Given various predictors of the outcome, which one should we use?
-%
-Standard predictive modeling or machine-learning practice selects the predictor
-that minimizes the expected error. However, this predictor may not be the best
-model to reason about causal effects of an intervention as Figure
-\ref{fig:toy_example} illustrates. Consider the probability $Y$ of an undesirable outcome
-(\emph{eg} death), a binary treatment $A \in \{0, 1\}$, and a covariate $X \in
-    \mathbb R$ summarizing the patient health status (eg. the Charlson index
-\cite{charlson_new_1987}). We simulate a treatment beneficial (decreases mortality)
-for patients with high Charlson scores (bad health status) but with little
-effect for patients in good condition (low Charlson scores).
-
-Figure \ref{fig:toy_example}a shows a random forest predictor with a
-counter-intuitive behavior: it predicts well on average the outcome (as measured
-by a regression $R^2$ score) but perform poorly to estimate causal quantities:
-the average treatment effect $\tau$ (as visible via the error $|\tau -
-    \hat{\tau}|$) or the conditional average treatment effect (the error
-$\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$, called CATE).
-%
-On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
-smaller $R^2$ score but better causal inference.%
-
-The problem is that causal estimation requires controlling
-an error on both treated and non-treated outcome for the same individual:
-the observed outcome, and the non-observed \emph{counterfactual} one.
-The linear model is misspecified --the outcome functions are not
-linear--, leading to poor $R^2$; but it interpolates better to regions
-where there are few untreated individuals --high Charlson score-- and
-thus gives better causal estimates. Conversely, the random forest puts
-weaker assumptions on the data, thus has higher $R^2$ score but is biased
-by the treated population in the poor-overlap region, leading
-to bad causal estimates.
-
-This toy example illustrates that the classic minimum Mean Square Error
-criterion is not suited to choosing a model among candidate
-estimators for causal inference.
-
-% A natural way to select a predictive model for causal inference would be
 % an error measure between a causal quantity such as the CATE and models' estimate. But such error is
@@
 \cite{naimi2023defining,imbens_causal_2015} enables statistical reasoning on
-causal treatment effects: Given an outcome $Y \in \mathbb R$ (eg. mortality risk
+causal treatment effects: Given an outcome $Y \in \mathbb R$ (\eg mortality risk
 or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
-    \{0, 1\}$ (eg.~a medical procedure), and baseline
+    \{0, 1\}$ (\eg a medical procedure), and baseline
 covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
@@
 
-\section{The R-risk, a reweighted oracle metric}\label{theory:r_risk_rewrite}
+\subsection{R-risk as reweighted oracle metric}\label{theory:r_risk_rewrite}
 
@@
 We generate the response functions using random bases extension, a common method
-in biostatistics, \emph{eg} functional regression with splines
+in biostatistics, \eg functional regression with splines
 \cite{howe_splines_2011, perperoglou_review_2019}. By allowing the function to
@@
 Practically, we recommand a train/test split of 90 \%/10\%. Use most of the data
-to estimate both the nuisances and outcome models, then compute the risks on the
-remaining test set for model selection (emprical argument in Appendix
+to estimate both the nuisances and candidates, then compute the risks on the
+remaining test set for model selection (empirical argument in Appendix
 \ref{apd:results:data_split}).
@@
 to more valid causal estimates. Estimating the $R\text{-risk}$ requires a more
-complex procedure than standard cross-validation used \emph{e.g.}~in machine
+complex procedure than standard cross-validation used \eg in machine
 learning: it involves fitting nuisance models necessary for model evaluation,
@@
 framework studied here can directly apply.
-In particular, the log odds ratio is estimated by the common
-cross-entropy loss (or log loss) as in logistic regression.
+% In particular, the log odds ratio is estimated by the common
+% cross-entropy loss (or log loss) as in logistic regression.
 
@@
 code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
-Using the $R\text{-risk}$ does make evaluation more
-complicated not only because the procedure is more involved, but also
-because each intervention requires a dedicated evaluation. However, such
-off-policy evaluation remains much less costly than the recommended good
-practice of impact evaluation testing the ability of a prediction model
-to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
-the model-selection procedure puts no constraints on the models used to
+% Using the $R\text{-risk}$ does make evaluation more
+% complicated not only because the procedure is more involved, but also
+% because each intervention requires a dedicated evaluation. However, such
+% off-policy evaluation remains much less costly than the recommended good
+% practice of impact evaluation testing the ability of a prediction model
+% to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
+This model-selection procedure puts no constraints on the models used to
 build predictive models: it opens the door to evaluating a wide range of

### 2023-12-15 how_to_select_causal_models.tex
move fig 1 in first pageé
iter

@@
     the door to a plethora of models, but also the Pandora box of model
-    selection: which of these models yield the most valid causal estimates?  In
-    this paper, we study model selection procedures in practical settings:
+    selection: which of these models yield the most valid causal estimates? 
+    We study such model selection procedures in practical settings:
     finite samples settings and without well-specification assumption.
@@
     nuisance adds noise to model selection? Drawing from an extensive empirical
-    study based on a simulation and three semi-simulated healthcare datasets
+    study based on a simulation and three healthcare datasets
     based on real covariates, we compare different metrics for causal model
@@
 
-\subsection{Illustration: the best predictor may not estimate best causal
-    effects}%
-
-Using a predictor to reason on causal effects relies on contrasting the
-prediction of the outcome for a given individual with and without the treatment.
-%
-Given various predictors of the outcome, which one should we use?
-%
-Standard predictive modeling or machine-learning practice selects the predictor
-that minimizes the expected error on the outcome
-\citep{poldrack2020establishment,varoquaux2022evaluating}. However, this
-predictor may not be the best model to reason about causal effects of an
-intervention as Figure \ref{fig:toy_example} illustrates. Consider the
-probability $Y$ of an undesirable outcome (\eg death), a binary treatment $A \in
-    \{0, 1\}$, and a covariate $X \in \mathbb R$ summarizing the patient health
-status (\eg the Charlson index \cite{charlson_new_1987}). We simulate a
-treatment beneficial (decreases mortality) for patients with high Charlson
-scores (bad health status) but with little effect for patients in good condition
-(low Charlson scores).
-
-Figure \ref{fig:toy_example}a shows a random forest predictor with a
-counter-intuitive behavior: it predicts well on average the outcome (as measured
-by a regression $R^2$ score) but perform poorly to estimate causal quantities:
-the average treatment effect $\tau$ (as visible via the error $|\tau -
-    \hat{\tau}|$) or the conditional average treatment effect (the error
-$\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$, called CATE).
-%
-On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
-smaller $R^2$ score but better causal inference.%
-
-The problem is that causal estimation requires controlling
-an error on both treated and non-treated outcome for the same individual:
-the observed outcome, and the non-observed \emph{counterfactual} one.
-The linear model is misspecified --the outcome functions are not
-linear--, leading to poor $R^2$; but it interpolates better to regions
-where there are few untreated individuals --high Charlson score-- and
-thus gives better causal estimates. Conversely, the random forest puts
-weaker assumptions on the data, thus has higher $R^2$ score but is biased
-by the treated population in the poor-overlap region, leading
-to bad causal estimates.
-
-This toy example illustrates that the classic minimum Mean Square Error
-criterion is not suited to choosing a model among candidate
-estimators for causal inference.
-
-% A natural way to select a predictive model for causal inference would be
-
 \begin{figure}[h!]
@@
 \end{figure}
+
+\subsection{Illustration: the best predictor may not estimate best causal
+    effects}%
+
+Using a predictor to reason on causal effects relies on contrasting the
+prediction of the outcome for a given individual with and without the treatment.
+%
+Given various predictors of the outcome, which one should we use?
+%
+Standard predictive modeling or machine-learning practice selects the predictor
+that minimizes the expected error on the outcome
+\citep{poldrack2020establishment,varoquaux2022evaluating}. However, this
+predictor may not be the best model to reason about causal effects of an
+intervention as Figure \ref{fig:toy_example} illustrates. Consider the
+probability $Y$ of an undesirable outcome (\eg death), a binary treatment $A \in
+    \{0, 1\}$, and a covariate $X \in \mathbb R$ summarizing the patient health
+status (\eg the Charlson index \cite{charlson_new_1987}). We simulate a
+treatment beneficial (decreases mortality) for patients with high Charlson
+scores (bad health status) but with little effect for patients in good condition
+(low Charlson scores).
+
+Figure \ref{fig:toy_example}a shows a random forest predictor with a
+counter-intuitive behavior: it predicts well on average the outcome (as measured
+by a regression $R^2$ score) but perform poorly to estimate causal quantities:
+the average treatment effect $\tau$ (as visible via the error $|\tau -
+    \hat{\tau}|$) or the conditional average treatment effect (the error
+$\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$, called CATE).
+%
+On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
+smaller $R^2$ score but better causal inference.%
+
+The problem is that causal estimation requires controlling
+an error on both treated and non-treated outcome for the same individual:
+the observed outcome, and the non-observed \emph{counterfactual} one.
+The linear model is misspecified --the outcome functions are not
+linear--, leading to poor $R^2$; but it interpolates better to regions
+where there are few untreated individuals --high Charlson score-- and
+thus gives better causal estimates. Conversely, the random forest puts
+weaker assumptions on the data, thus has higher $R^2$ score but is biased
+by the treated population in the poor-overlap region, leading
+to bad causal estimates.
+
+This toy example illustrates that the classic minimum Mean Square Error
+criterion is not suited to choosing a model among candidate
+estimators for causal inference.
+
+% A natural way to select a predictive model for causal inference would be
 

### 2023-12-15 how_to_select_causal_models.tex
iter

@@
     Elaborate risks build proxies of the causal error using ``nuisance''
-    re-weighting to compute it on the observed data. But does computing these
-    nuisance adds noise to model selection? Drawing from an extensive empirical
+    re-weighting to compute it on the observed data. We evaluate whether
+    empirically estimated nuisances, which are necessarily noisy,
+    add noise to model selection. Drawing from an extensive empirical
     study based on a simulation and three healthcare datasets
@@
     scores, leads to the best performances.\\
-    \textbf{Conclusions:} We outline a good causal model-selection procedure
-    using the so-called $R\text{-risk}$; using flexible estimators to compute
-    the nuisance models on the train set; and splitting out 10\% of the data to
-    compute risks.
+    \textbf{Conclusions:} We outline that selecting model for causal
+    reasoning requires different procedures than standard predictive
+    settings: using the so-called $R\text{-risk}$ and computing nuisance
+    models with flexible estimators on the train set.
 \end{abstract}

### 2023-12-15 how_to_select_causal_models.tex
iter

@@
     scores, leads to the best performances.\\
-    \textbf{Conclusions:} We outline that selecting model for causal
+    \textbf{Conclusions:} We outline that evaluating models for causal
     reasoning requires different procedures than standard predictive

### 2023-12-15 how_to_select_causal_models.tex
iter

--- 2023-12-15 how_to_select_causal_models.tex
@@
     scores, leads to the best performances.\\
-    \textbf{Conclusions:} We outline that selecting model for causal
+    \textbf{Conclusions:} We outline that evaluating models for causal
     reasoning requires different procedures than standard predictive
-    settings: using the so-called $R\text{-risk}$ and computing nuisance
-    models with flexible estimators on the train set.
+    settings: using the so-called $R\text{-risk}$ with nuisance
+    corrections computed with flexible estimators.
 \end{abstract}

### 2023-12-16 how_to_select_causal_models.tex
iter

@@
 \maketitle
-\begin{abstract}
+\begin{abstract}\noindent
     \textbf{Objective:}
-    As predictive models --\eg from machine learning-- give likely
-    outcomes, they may be used to reason on the effect of an intervention, a
-    causal-inference task. The increasing complexity of health data has opened
-    the door to a plethora of models, but also the Pandora box of model
-    selection: which of these models yield the most valid causal estimates? 
-    We study such model selection procedures in practical settings:
+    To evaluate which procedure selects the predictive model most
+    trustworthy to reason on the effect of an intervention and support
+    decision making.\\
+    %As predictive models --\eg from machine learning-- give likely
+    %outcomes, they may be used to, a
+    %causal-inference task. The increasing complexity of health data has opened
+    %the door to a plethora of models, but also the Pandora box of model
+    %selection: which of these models yield the most valid causal estimates? 
+    \textbf{Methods:}
+    We study such a variety of model selection procedures in practical settings:
     finite samples settings and without well-specification assumption.
-    \textbf{Methods:}
     Elaborate risks build proxies of the causal error using ``nuisance''
@@
 \cite{beam2018big,rajkomar2019machine}, with the promise of precision medicine.
-Predictive models are increaslingly used for image analysis
+Predictive models are increasingly used for image analysis
 \cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
-routinely-collected patient records
+patient records
 \cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical

### 2023-12-16 how_to_select_causal_models.tex
One column

@@
-\documentclass[10pt,twocolumn]{article}
-
-\usepackage[a4paper, left=15mm, right=15mm, top=20mm,]{geometry}
+% XXX: for submission
+%\documentclass[10pt,twocolumn]{article}
+\documentclass[11pt]{article}
+
+%\usepackage[a4paper, left=15mm, right=15mm, top=20mm,]{geometry}
+% XXX: for submission
+\usepackage[a4paper, left=25mm, right=25mm, top=20mm,]{geometry}
 % declared functions and additional packages
@@
 
-\twocolumn
+% XXX: for submission
+%\twocolumn
 \section{Introduction}\label{sec:intro}
@@
 \begin{figure}[h!]
-    %\centering
+    % XXX: for submission
+    \centering\begin{minipage}{.6\linewidth}
     {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
@@
     \includegraphics[width=1\linewidth]{images/toy_tlinear_model_small_R2_small_tau_risk.pdf}%
+    \end{minipage}
 
@@
 \begin{figure}[h!]
+    % XXX: for submission
+    \centering\begin{minipage}{.6\linewidth}
     \includegraphics[width=\linewidth]{images/estimation_procedure_causal_selection_procedure.png}
+    \end{minipage}
     \caption{Estimation procedure for causal model
@@
 \begin{figure}[!b]
-    \centering
+    % XXX: for submission
+    \centering\begin{minipage}{.75\linewidth}
     \includegraphics[width=\linewidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
-    %\hfill
+    \end{minipage}
     \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$ agreement with $\tau\text{-risk}$.
@@
 \begin{figure}[!h]
-    \centering
+    % XXX: for submission
+    \centering\begin{minipage}{.75\linewidth}
     \includegraphics[width=\linewidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
-    %\hfill
+    \end{minipage}
     \caption{\textbf{Model selection is harder for low population
@@
 \begin{figure}[!tb]
+    % XXX: for submission
+    \centering\begin{minipage}{.75\linewidth}
     \includegraphics[width=\linewidth]{images/_3_procedure_r_risk_only_3datasets_twocols.pdf}
+    \end{minipage}
 
@@
 \begin{figure}[!tb]
+    % XXX: for submission
+    \centering\begin{minipage}{.75\linewidth}
     \includegraphics[width=\linewidth]{images/_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
+    \end{minipage}
 

### 2023-12-16 how_to_select_causal_models.tex
cosmetics
"Double spacing"
(or actually, something that looks a bit like it, but isn't, because
double spacing is too ugly)
iter

@@
 
+\usepackage[%
+    font={footnotesize,sf},
+    labelfont=bf,
+    format=hang,    
+    format=plain,
+    margin=0pt,
+]{caption}
+
+% XXX: for submission
+\usepackage{setspace}
+
 \usepackage{comment}
@@
     \textbf{Objective:}
-    To evaluate which procedure selects the predictive model most
+    We investigate which procedure selects the predictive model most
     trustworthy to reason on the effect of an intervention and support
@@
     We study such a variety of model selection procedures in practical settings:
-    finite samples settings and without well-specification assumption.
-    Elaborate risks build proxies of the causal error using ``nuisance''
+    finite samples settings and without theoretical assumption of 
+    \emph{well-specified} models. Beyond standard cross-validation or
+    internal validation procedures, we also study elaborate causal risks.
+    These build proxies of the causal error using ``nuisance''
     re-weighting to compute it on the observed data. We evaluate whether
     empirically estimated nuisances, which are necessarily noisy,
-    add noise to model selection. Drawing from an extensive empirical
+    add noise to model selection. We compare different metrics for
+    causal model selection in an extensive empirical
     study based on a simulation and three healthcare datasets
-    based on real covariates, we compare different metrics for causal model
-    selection.\\
+    based on real covariates. \\
     \textbf{Results:} Among all metrics the classical mean squared error (ie.
@@
 %\twocolumn
+% XXX: for submission
+\setstretch{1.1}
 \section{Introduction}\label{sec:intro}

### 2023-12-16 how_to_select_causal_models.tex
intro

@@
     based on real covariates. \\
-    \textbf{Results:} Among all metrics the classical mean squared error (ie.
-    factual $\mu\mathrm{-risk}$) is worse and re-weighting it with propensity score
-    ($\mu\mathrm{-risk}_{IPW}$) does not bring much improvements. The
-    $R\text{-risk}$, which includes a model of mean outcome and propensity
-    scores, leads to the best performances.\\
-    \textbf{Conclusions:} We outline that evaluating models for causal
-    reasoning requires different procedures than standard predictive
-    settings: using the so-called $R\text{-risk}$ with nuisance
-    corrections computed with flexible estimators.
+    \textbf{Results:} Among all metrics, the mean squared error,
+    classically used to evaluate predictive modes,
+    is worse. Re-weighting it with propensity score
+    does not bring much improvements. The
+    $R\text{-risk}$, which uses as nuisances a model of
+    mean outcome and propensity scores, leads to the best performances.
+    Nuisance corrections are best estimated with flexible estimators such
+    as a super learner.
+    \\
+    \textbf{Conclusions:} When predictive models are used to 
+    reason on the effect of an intervention, they must be evaluated 
+    with different procedures than standard predictive
+    settings; using the $R\text{-risk}$ from causal inference.
 \end{abstract}
-\keywords{Model Selection, Treatment Effect, G-formula, Observational Study, Machine Learning, }
+\keywords{Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning, }
 \clearpage

### 2023-12-16 how_to_select_causal_models.tex
iter

@@
 
-Progress in machine learning brings predictive models to new health data
-\cite{beam2018big,rajkomar2019machine}, with the promise of precision medicine.
-Predictive models are increasingly used for image analysis
+Progress in machine learning promises to integrate new data for
+precision medicine \cite{beam2018big,rajkomar2019machine}:
+health predictions building on medical images
 \cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},

### 2023-12-16 how_to_select_causal_models.tex
iter

@@
 
-Predictive modeling bridges to causal-inference with the so-called \emph{outcome
-    models}, or G-computation, G-formula \cite{robins_role_1986}, Q-model
-\cite{snowden_implementation_2011}, conditional mean regression
-\cite{wendling_comparing_2018}. A central challenge of inference of treatment
-effects is that of confounding: spurious associations between treatment
+Indeed, concluding on the effect of a treatment is a difficult
+causal-inference task, easily compromised by confounding:
+spurious associations between treatment
 allocation and baseline health, \eg only prescribing a drug to mild cases
 \citep{hernan_causal_2020,vanderweele2019principles}.
+Predictive modeling bridges to causal inference theory under the name of
+\emph{outcome models} (or G-computation, 
+G-formula \cite{robins_role_1986}, Q-model
+\cite{snowden_implementation_2011}, conditional mean regression
+\cite{wendling_comparing_2018}). 
 % Controlled allocation of treatment, as in randomized controlled trials (RCTs),
@@
 % \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
-Causal inference has been central to epidemiology, typically with methods that
-model treatment assignment \cite{austin_moving_2015,casucci2018estimating,grose_use_2020}, based on
-propensity scores \cite{rosenbaum_central_1983}. But outcome modeling go
+Medical statistics and epidemiology have mostly used other
+causal-inference methods, modeling treatment assignment
+with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
+the benefit of going
 beyond average effects, estimating individualized or conditional average
@@
 Outcome-modeling methods, even when specifically designed for causal
-inference are numerous: Bayesian Additive Regression Trees
+inference, are numerous: Bayesian Additive Regression Trees
 \cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation

### 2023-12-16 how_to_select_causal_models.tex
iter

@@
 
-Given complex,
-potentially noisy, data, which model is to be most trusted to yield valid causal
-estimates? As no single learner performs best on all data sets, there is a
+Given complex health data, which model is to be most trusted to yield 
+valid causal estimates needed to motivate individual treatment decisions?
+As no single learner performs best on all data sets, there is a
 pressing need for clear guidelines to select outcome models for causal
@@
 
-We study \textit{model selection procedures} in practical settings:
-\textit{finite samples} without the \textit{well-specification}
-assumptions which are often made in the literature (Appendix
-\ref{apd:prior_work}). Asymptotic causal-inference theory calls for complex
+We study \textit{model selection procedures} in practical settings,
+without theoretical assumptions often made in statistical literature such
+as \textit{infinite} data or \textit{well-specified} models (Appendix
+\ref{apd:prior_work}). Asymptotic causal-inference theory recommends complex
 risks, but a practical question is whether model-selection procedures, that rely
-on data split, can estimate these risks reliably enough. Indeed, they come with
+on data split, can estimate these risks reliably enough. Indeed, these
+risks come with
 more quantities to estimate, which may bring additional variance, leading to
@@
 framework and details the causal risks and model-selection procedure. We then
-rewrite the so-called $R\text{-risk}$, as a reweighted version of mean squared
+rewrite the so-called $R\text{-risk}$ as a reweighted version of mean squared
 difference between the true and estimated individualized treatment effect.
 Finally, we conduct a thorough empirical study comparing the different metrics
-on a wide variety of data-generative processes based on simulations and real
-data, going beyond prior work based on specific simulation settings
+on diverse datasets, using a family of simulations and real health
+data, going beyond prior work limited to specific simulation settings
 \citep{schuler_comparison_2018,alaa_validating_2019} (Appendix

### 2023-12-16 how_to_select_causal_models.tex
iter

@@
     \caption[The best predictor may not estimate best causal
-        effects]{\textbf{Illustration}: a) a random-forest estimator
+        effects]{\textbf{Illustration}: a) a random-forest predictor
         with high performance for standard prediction (high $R^2$) but that
         yields poor causal estimates (large error between true effect $\tau$ and
-        estimated $\hat{\tau}$), b) a linear estimator with smaller
+        estimated $\hat{\tau}$), b) a linear predictor with smaller
         prediction performance leading to better causal estimation. \\[1ex]
-        Selecting the estimator with the smallest error to the individual
+        Selecting the predictor with the smallest error to the individual
         treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$

### 2023-12-16 how_to_select_causal_models.tex
iter

@@
 
-    \includegraphics[width=1\linewidth]{images/toy_random_forest_high_R2_high_tau_risk.pdf}%
+    \setlength{\unitlength}{.1\linewidth}
+    \begin{picture}(10,5.2)(0,0)
+    \put(0, 0){\includegraphics[width=1\linewidth]{images/toy_random_forest_high_R2_high_tau_risk.pdf}}%
+    \put(7.05,4.4){\vector(0,1){.47}}
+    \put(7.05,4.4){\vector(0,-1){.55}}
+    \put(7.1,4.5){\scriptsize\sffamily treatment effect}
+    \put(7.1,4.15){\small $\tau(x)$}
+    \put(5.2,3.6){\small $\hat{\tau}(x)$}
+    \put(5.15,3.6){\vector(0,1){.29}}
+    \put(5.15,3.6){\vector(0,-1){.15}}
+    \end{picture}
 
@@
 or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
-    \{0, 1\}$ (\eg a medical procedure), and baseline
+    \{0, 1\}$ (\eg~a medical procedure), and baseline
 covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
 distribution, $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However,
-we want to model the existence of potential observations (unobserved ie.
-counterfactual) that correspond to a different treatment. Thus we want
+we want to model the existence of potential observations (unobserved ie.~counterfactual) that correspond to a different treatment. Thus we want
 quantities on the counterfactual distribution $O^{*} = (Y(1), Y(0), X, A) \sim

### 2023-12-16 how_to_select_causal_models.tex
iter

@@
 
-Assumptions are necessary for causal estimands to be
-identifiability
-in observational settings \cite{rubin_causal_2005}. We assume the usual
-strong ignorability assumptions: \emph{1)}
+A given data needs to meet a few assumptions to enable identifying
+causal estimands \cite{rubin_causal_2005}. The usual
+\emph{strong ignorability} assumptions are  (details in
+\ref{apd:causal_assumptions}): \emph{1)}
 \emph{unconfoundedness} \mbox{$\{Y(0),
@@
 strictly positive probability to receive each treatment, \emph{3)}
-\emph{consistency}, and \emph{4)} \emph{generalization} (detailed in
-\ref{apd:causal_assumptions}).
+\emph{consistency}, and \emph{4)} \emph{generalization}.
 

### 2023-12-16 how_to_select_causal_models.tex
iter

@@
 \usepackage{multirow}
-\usepackage[sorting=nyt,style=numeric, natbib=true,doi=false,isbn=false,url=false,eprint=false, maxnames=999, maxcitenames=2]{biblatex} 
+\usepackage[sorting=nyt,style=numeric,natbib=true,doi=false,isbn=false,url=false,eprint=false, maxnames=999, maxcitenames=2]{biblatex} 
+% To cite with superscript
+\let\cite=\supercite
 \DeclareNameAlias{default}{last-first}
@@
 \end{equation}
-This estimator is an oracle \textbf{finite sum estimator} by opposition to the
+This estimator is an oracle \emph{finite sum} estimator by opposition to the
 population expression of $\tau$, $\mathbb{E}[\mu_{1}(x_i) - \mu_{0}(x_i)]
@@
 The \emph{R-decomposition}
-of the outcome model plays an important role,
-\cite{robinson_rootnconsistent_1988}:
+of the outcome model\cite{robinson_rootnconsistent_1988} plays an important role:
 introducing two quantities, the conditional mean outcome

### 2023-12-16 how_to_select_causal_models.tex
iter

@@
                 \midrule
-                $mse(\tau(X), \tau_f(X))=\tau\text{-risk}$                                                     & $\mathbb E_{X\sim
+                $\tau\text{-risk}=\text{MSE}(\tau(X), \tau_f(X))$                                                     & $\mathbb E_{X\sim
                             p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
@@
 \ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
-                $mse(Y, f(X)) = \mu\text{-risk}$                                                               & $\mathbb{E}_{(Y, X, A)
+                $\mu\text{-risk} = \text{MSE}(Y, f(X))$                                                               & $\mathbb{E}_{(Y, X, A)
                         \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
@@
 \cite{schuler_comparison_2018,alaa_validating_2019}, we use many
-generative processes, to reach more general conclusions (as discussed
-in \ref{apd:results:fig:seed_effect}).
+generative processes, which is needed to reach general conclusions (Appendix \ref{apd:results:seed_effect}).
 
@@
 kernels. Unlike polynomial, Gaussian kernels have decreasing influences in the
-input space. This avoids unrealistic divergences of the response surfaces at the
+input space. This avoids unrealistic divergences of the functions at the
 ends of the feature space. We generate 1\,000 datasets based on these functions,
@@
 
-We compare different candidate estimators that approximate imperfectly the
+We test model selection across different candidate estimators that
+approximate imperfectly the
 data-generating process. To build such estimators, we first use a RBF
@@
 
-We also use three benchmarks of the causal-inference literature, semi-simulated
+We also use three semi-simulated
 data adding a known synthetic causal effect to real --non synthetic-- healthcare
@@
 
-\subsection{Results: factors driving good model selection}\label{empirical_study:results}
+\section{Results: factors driving good model selection}\label{empirical_study:results}
 
@@
 \paragraph{Selecting different seeds and parameters is crucial to draw
-    conclusions}\label{{apd:results:seed_effect}}
+    conclusions}\label{apd:results:seed_effect}
 

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 data-generating process. To build such estimators, we first use a RBF
-expansion similar to the one used for data generation. We choose two random
-knots and apply a transformation of the raw data features with a Gaussian
+expansion similar to that used for data generation. We choose two random
+knots and transform the raw data features with a Gaussian
 kernel. This step is referred as the featurization. Then, we fit a linear
 regression on these transformed features. We consider two ways of combining
-these steps for outcome mode; we use common nomenclature
+these steps for outcome model; we use common nomenclature
 \cite{kunzel_metalearners_2019,shen2023rctrep} to refer to these different

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 
+% TODO commands
+\usepackage{snaptodo}
+\setlength{\marginparwidth}{2.3cm}
+\setlength{\marginparsep}{.1cm}
+\snaptodoset{block rise=2em}
+\snaptodoset{margin block/.style={font=\tiny}}
+
+\newcommand{\gv}[1]{\snaptodo[margin block/.append style=green!50!black]{\sloppy\textbf{GV}: #1}}
+
+% Uncomment below to erase all comments
+%\renewcommand{\snaptodo}[2][1]{\relax}
+
 % XXX: for submission
@@
 
-\setlength{\marginparwidth}{0.5cm}
+%\setlength{\marginparwidth}{0.5cm}
 %\setlength\textwidth{177.8mm}
@@
 with 6 different choices of the regularization parameter $\lambda \in [10^{-3},
-        10^{-2}, 10^{-1}, 1, 10^{1}, 10^{2}]$, coupled with a TLearner or a SftLearner.
+        10^{-2}, 10^{-1}, 1, 10^{1}, 10^{2}]$, coupled with a TLearner or
+a SftLearner\gv{Si je lis bien, on n'utilise pas le SLearner. On pourrait
+le retirer de la liste au dessus}.
 We sample 10 different random basis for learning and

### 2023-12-17 how_to_select_causal_models.tex
iter

--- 2023-12-17 how_to_select_causal_models.tex
@@
 with 6 different choices of the regularization parameter $\lambda \in [10^{-3},
-        10^{-2}, 10^{-1}, 1, 10^{1}, 10^{2}]$, coupled with a TLearner or a SftLearner.
+        10^{-2}, 10^{-1}, 1, 10^{1}, 10^{2}]$, coupled with a TLearner or
+a SftLearner.
 We sample 10 different random basis for learning and
@@
 $N=11\,984$ samples, and $D=50$ covariates for which we simulated 1,000
-different treatment allocations. Datasets details are given in Appendix
-\ref{apd:experiments:datasets}.
+different treatment allocations. Appendix
+\ref{apd:experiments:datasets} gives datasets details.
 
@@
 
-Drawing inspiration from the TMLE literature that uses combination of flexible
+Drawing from the TMLE literature that uses combination of flexible
 machine learning methods \cite{schuler_targeted_2017}, we model the nuisances

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 datasets, the $R\text{-risk}$ consistently decreases the
-risk compared to the $\mu\text{-risk}$:
-0.1\% compared to 1\% on ACIC2016,  1\% compared to 20\% on ACIC2018,
-and 0.05\% compared to 1\% on Twins.
+risk compared to the $\mu\text{-risk}$: from
+0.1\% to 1\% on ACIC2016,  1\% from to 20\% on ACIC2018,
+and 0.05\% from to 1\% on Twins.
 
@@
 \ref{fig:all_datasets_overlap_effect_r_risk}). The absolute Kendall's
-coefficient correlation with $\tau\text{-risk}$ drops from values around 0.9
+coefficient correlation with $\tau\text{-risk}$ drops from 0.9
 (excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
@@
 
-Practically, we recommand a train/test split of 90 \%/10\%. Use most of the data
-to estimate both the nuisances and candidates, then compute the risks on the
-remaining test set for model selection (empirical argument in Appendix
+Empirically, the best split is 90\,\%/10\,\%:
+using 90\,\% of the data
+to estimate both the nuisances and candidates, then computing the risks on the
+remaining test set for model selection (experiments in Appendix
 \ref{apd:results:data_split}).
@@
 
-For every risk, the oracle version recovers better the best estimator.
-However,
-stacked nuisances estimators (boosting and linear) lead to feasible
+Stacked nuisances estimators (boosting and linear) lead to feasible
 metrics with close performances to the oracles ones: the

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 %
-Predictive models are increasingly used to reason about causal effects, for
+Predictive models are increasingly used to reason about treatment effects, for
 instance in precision medicine to drive individualized decision. Our results
 highlight that they should be selected, validated, and tuned using different
-procedures and error measures than those classically used to assess prediction
-(estimating the so-called $\mu\text{-risk}$). Rather, selecting the best outcome
+procedures and error measures than those classically used to assess prediction. Rather, selecting the best outcome
 model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
@@
 complex procedure than standard cross-validation used \eg in machine
-learning: it involves fitting nuisance models necessary for model evaluation,
-though our results show that these can be learned on the same set of data as the
-outcome model evaluated. The nuisance models must be well estimated (Figure
+learning: it involves fitting nuisance models necessary for model
+evaluation.
+Our results show that these can be learned on the same set of data as the
+outcome models evaluated. The nuisance models must be well estimated (Figure
 \ref{fig:all_datasets_nuisances_comparison}). Our results show that using for
 nuisance models a flexible stacking-based family of estimator suffices for good
-model selection. For propensity score, we used the Brier score,
-which is minimized by the true individual probability. An easy mistake is to
+model selection. To select propensity score models, we used the Brier score,
+minimized by the true individual probability. An easy mistake is to
 use calibration errors popular in machine learning
@@
 
-\paragraph{Extension to binary outcomes}
-While we focused on continuous outcomes, in medicine, the target outcome
-is often a categorical variable such as mortality status or diagnosis. In
-this case, it may be interesting to focus on other estimands than the
-Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
-$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
-or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
-    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
-\cite{austin2017estimating}. While the odds ratio is natural for
-case-control studies \cite{rothman2008case}, other measures
-can reduce heterogeneity \cite{colnet2023risk}.
-In the log domain, the ratios are written as a difference, the
-framework studied here can directly apply.
-% In particular, the log odds ratio is estimated by the common
-% cross-entropy loss (or log loss) as in logistic regression.
+%\paragraph{Extension to binary outcomes}
+%While we focused on continuous outcomes, in medicine, the target outcome
+%is often a categorical variable such as mortality status or diagnosis. In
+%this case, it may be interesting to focus on other estimands than the
+%Average Treatment Effect $\mathbb{E}[Y(1)] -\mathbb{E}[Y(0)] $, for instance the relative risk
+%$\frac{\mathbb P(Y(1) = 1)}{\mathbb P(Y(0) = 1)}$
+%or the odd ratio, $\frac{\mathbb P(Y(1) = 1) / [1 - \mathbb P(Y(1)
+%    =1)]}{\mathbb P(Y(0) = 1) / [1 - \mathbb P(Y(0) = 1]}$ are often used
+%\cite{austin2017estimating}. While the odds ratio is natural for
+%case-control studies \cite{rothman2008case}, other measures
+%can reduce heterogeneity \cite{colnet2023risk}.
+%In the log domain, the ratios are written as a difference, the
+%framework studied here can directly apply.
+%% In particular, the log odds ratio is estimated by the common
+%% cross-entropy loss (or log loss) as in logistic regression.
 

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 
-Progress in machine learning promises to integrate new data for
-precision medicine \cite{beam2018big,rajkomar2019machine}:
-health predictions building on medical images
+Prediction models have long been used in medicine, as with risk
+score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
+While these have historically been simple models on simple data, this is
+changing with progress in machine learning and richer sources of medical data
+\cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
 \cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
@@
 \cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
-notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}. Data may
-be difficult to control and model, but the accuracy of the prediction can be
-verified on left-out data
+notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
+Complex data is difficult to control and model, but these models are
+validated by verifying the accuracy of the prediction on left-out data
 \cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 guide decisions: will an individual benefit from an intervention such as surgery
-\cite{fontana2019can}? Contrasting predictions with and without the treatment
-gives an answer, but statistical validity requires causal inference
+\cite{fontana2019can}? An estimate of the effect of the treatment can be
+obtained by contrasting model predictions with and without the treatment,
+but statistical validity requires causal inference
 \cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.
@@
 Indeed, concluding on the effect of a treatment is a difficult
-causal-inference task, easily compromised by confounding:
+causal-inference task, as it can be easily compromised by confounding:
 spurious associations between treatment

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 \cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
-Given a model predicting a health outcome, precision medicine would like it to
-guide decisions: will an individual benefit from an intervention such as surgery
+Crucial to the clinical adoption of a model predicting a health outcome
+is that it ``can support decisions about patient care''
+\cite{wyatt1995commentary}. Precision medicine is about 
+guiding decisions: \emph{eg} will an individual benefit from an intervention such as surgery
 \cite{fontana2019can}? An estimate of the effect of the treatment can be

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 allocation and baseline health, \eg only prescribing a drug to mild cases
-\citep{hernan_causal_2020,vanderweele2019principles}.
+\cite{hernan_causal_2020,vanderweele2019principles}.
 Predictive modeling bridges to causal inference theory under the name of
@@
 data, going beyond prior work limited to specific simulation settings
-\citep{schuler_comparison_2018,alaa_validating_2019} (Appendix
+\cite{schuler_comparison_2018,alaa_validating_2019} (Appendix
 \ref{apd:prior_work}).
@@
 that minimizes the expected error on the outcome
-\citep{poldrack2020establishment,varoquaux2022evaluating}. However, this
+\cite{poldrack2020establishment,varoquaux2022evaluating}. However, this
 predictor may not be the best model to reason about causal effects of an
@@
 practice-- leads to 10\% excess risk and using $\tau\text{-risk}_{IPW}$
---as in some causal-inference methods \citep{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
+--as in some causal-inference methods \cite{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
 (Figure
@@
 CATE estimator in the testing procedure: matching
-\citep{rolling_model_2014}, an IPW estimate
-\citep{gutierrez_causal_2016}, a doubly robust estimator
-\citep{saito_counterfactual_2020}, or debiasing the error with influence
+\cite{rolling_model_2014}, an IPW estimate
+\cite{gutierrez_causal_2016}, a doubly robust estimator
+\cite{saito_counterfactual_2020}, or debiasing the error with influence
 functions \cite{alaa_validating_2019}. However, for theoretical

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 spurious associations between treatment
-allocation and baseline health, \eg only prescribing a drug to mild cases
+allocation and baseline health, \eg~only prescribing a drug to mild cases
 \cite{hernan_causal_2020,vanderweele2019principles}.
@@
 
-Given complex health data, which model is to be most trusted to yield 
+Given complex health data, which predictive model is to be most trusted to yield 
 valid causal estimates needed to motivate individual treatment decisions?
@@
 
-\subsection{Objectives and structure of the paper}
+\paragraph{Objectives and structure of the paper}
 

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 While these have historically been simple models on simple data, this is
-changing with progress in machine learning and richer sources of medical data
+changing with progress in machine learning and richer medical data
 \cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
@@
 valid causal estimates needed to motivate individual treatment decisions?
-As no single learner performs best on all data sets, there is a
+As no single machine-learning method performs best across all data sets, there is a
 pressing need for clear guidelines to select outcome models for causal

### 2023-12-17 how_to_select_causal_models.tex
iter

@@
 
-Prediction models have flourished because their predictions can be easily
-demonstrated and validated on left-out data. But they require more
-careful validation for decision making, using a metric accounting for the
-putative intervention, the $R\text{-risk}$. Even when treated and
-untreated population differ little, as in RCTs, the $R\text{-risk}$
-brings a sizeable benefit. To facilitate better model selection, we provide Python
+Increasingly complex prediction models integrating richer medical data
+have flourished because their predictions can be easily
+demonstrated and validated on left-out data. But using them to underpin a
+decision on whether to treat or not requires more careful validation,
+using a metric accounting for the
+putative intervention, the $R\text{-risk}$. The $R\text{-risk}$
+brings a sizeable benefit to select the most adequate model, even when
+model development is based on treated and
+untreated population with little differences, as in RCTs. To facilitate better model selection, we provide Python
 code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.

### 2024-01-10 how_to_select_causal_models.tex
iter
add graphical abstract
first proposition of reviewers
fix jamia demand
fix for submission

@@
 \usepackage{multirow}
+\usepackage{appendix}
 \usepackage[sorting=nyt,style=numeric,natbib=true,doi=false,isbn=false,url=false,eprint=false, maxnames=999, maxcitenames=2]{biblatex} 
@@
 \usepackage[dvipsnames]{xcolor}
+\usepackage{setspace} % double spaced required by journal
+\doublespacing
 
@@
 
-\author[1,2,*]{Matthieu Doutreligne}
-\author[1]{Gaël Varoquaux}
+\author[1,2,*]{Matthieu Doutreligne, PhD}
+\author[1]{Gaël Varoquaux, PhD}
 
@@
 %\twocolumn
-% XXX: for submission
-\setstretch{1.1}
+% XXX: for submission (double spaced required)
+%\setstretch{1.1}
+\doublespacing
 \section{Introduction}\label{sec:intro}
@@
 
-We first illustrate the problem of causal model selection and briefly review
-prior art. Then we anchor causal model selection in the \emph{potential outcome}
+We first illustrate the problem of causal model selection. Then we anchor causal model selection in the \emph{potential outcome}
 framework and details the causal risks and model-selection procedure. We then
@@
 
-\section{Discussion}\label{sec:discussion}
+\section{Discussion and conclusion}\label{sec:discussion}\label{sec:conclusion}
 
@@
 
-\section{Conclusion}\label{sec:conclusion}
 \paragraph{More $R\text{-risk}$ to select models driving decisions}
@@
 
-\appendix
-%\setcounter{secnumdepth}{0}
+\renewcommand\thefigure{S\arabic{figure}}
+\renewcommand\thetable{S\arabic{table}}
+\setcounter{figure}{0}
+\setcounter{table}{0}
+
+\begin{appendices}
+
+    \part*{Supplementary materials:\\
+        How to select predictive models for decision-making or causal inference? }
+
     \setcounter{secnumdepth}{3}
+
+    %\part{Supplementary materials}
 
@@
 
-\begin{figure}
+    \begin{figure}[!b]
         \centering
@@
     % \end{figure}
+\end{appendices}
 

### 2024-01-10 make_submission.sh

@@ -1,0 +1,5 @@
+#! /bin/bash
+rm -f how_to_select_causal_models.zip
+zip -r how_to_select_causal_models.zip images/ biblio.bib *.bbl how_to_select_causal_models.tex 
+pdftk how_to_select_causal_models.pdf cat 1-24 output main.pdf
+pdftk how_to_select_causal_models.pdf cat 25-49 output supplementary_materials.pdf

### 2024-06-02 _old_how_to_select_causal_models.tex
gigascience: new jounrnal compiling format

--- 2024-01-10 how_to_select_causal_models.tex
@@
 \documentclass[11pt]{article}
+%\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
 
@@
 %\usepackage{arxiv}
-\usepackage{multicol}
-\setlength{\columnsep}{0.5cm}
-\usepackage{wrapfig}
-%\usepackage{wraptable}
-\usepackage{authblk}% for author affiliations
-\usepackage[dvipsnames]{xcolor}
-\usepackage{setspace} % double spaced required by journal
-\doublespacing
-
-\usepackage[%
-    font={footnotesize,sf},
-    labelfont=bf,
-    format=hang,    
-    format=plain,
-    margin=0pt,
-]{caption}
-
-% TODO commands
-\usepackage{snaptodo}
-\setlength{\marginparwidth}{2.3cm}
-\setlength{\marginparsep}{.1cm}
-\snaptodoset{block rise=2em}
-\snaptodoset{margin block/.style={font=\tiny}}
-
-\newcommand{\gv}[1]{\snaptodo[margin block/.append style=green!50!black]{\sloppy\textbf{GV}: #1}}
-
-% Uncomment below to erase all comments
-%\renewcommand{\snaptodo}[2][1]{\relax}
-
-% XXX: for submission
-\usepackage{setspace}
-
-\usepackage{comment}
-\setcounter{secnumdepth}{0}
-
-%\setlength{\marginparwidth}{0.5cm}
-%\setlength\textwidth{177.8mm}
-
-% maths commands
-\newcommand{\indep}{\perp \!\!\! \perp}
-\newcommand{\eg}{\emph{e.g.}}
-\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
-\usepackage{mathtools}
-\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}
-
-\newtheorem{definition}{Definition}
-\newtheorem{condition}{Condition}
-\newtheorem{property}{Property}
-\newtheorem{remark}{Remark}
-\newtheorem{notation}{Notation}
-\newtheorem{proposition*}{Proposition}
-\newtheorem{proposition}{Proposition}
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{assumption}{Assumption}
-\newtheorem{proof}{Proof}
-%
-\def\keywordname{{\bfseries \emph{Keywords}}}%
-\def\keywords#1{\par\addvspace\medskipamount{\rightskip=0pt plus1cm
-\def\and{\ifhmode\unskip\nobreak\fi\ $\cdot$
-}\noindent\keywordname\enspace\ignorespaces#1\par}}
-
-\title{How to select predictive models for decision-making or causal inference?}
-\addbibresource{biblio.bib}
-
-\author[1,2,*]{Matthieu Doutreligne, PhD}
-\author[1]{Gaël Varoquaux, PhD}
-
-\affil[1]{Inria, Soda, Saclay, France}
-\affil[2]{Mission Data, Haute Autorité de Santé, Saint-Denis, France}
-\affil[*]{Corresponding author: matt.dout@gmail.com}
-
-%\raggedbottom
-
-\begin{document}
-
-%\corres{*\email{matthieu.doutreligne@inria.fr}}
-
-\onecolumn
-\maketitle
-\begin{abstract}\noindent
-    \textbf{Objective:}
-    We investigate which procedure selects the predictive model most
-    trustworthy to reason on the effect of an intervention and support
-    decision making.\\
-    %As predictive models --\eg from machine learning-- give likely
-    %outcomes, they may be used to, a
-    %causal-inference task. The increasing complexity of health data has opened
-    %the door to a plethora of models, but also the Pandora box of model
-    %selection: which of these models yield the most valid causal estimates? 
-    \textbf{Methods:}
-    We study such a variety of model selection procedures in practical settings:
-    finite samples settings and without theoretical assumption of
-    \emph{well-specified} models. Beyond standard cross-validation or
-    internal validation procedures, we also study elaborate causal risks.
-    These build proxies of the causal error using ``nuisance''
-    re-weighting to compute it on the observed data. We evaluate whether
-    empirically estimated nuisances, which are necessarily noisy,
-    add noise to model selection. We compare different metrics for
-    causal model selection in an extensive empirical
-    study based on a simulation and three healthcare datasets
-    based on real covariates. \\
-    \textbf{Results:} Among all metrics, the mean squared error,
-    classically used to evaluate predictive modes,
-    is worse. Re-weighting it with propensity score
-    does not bring much improvements. The
-    $R\text{-risk}$, which uses as nuisances a model of
-    mean outcome and propensity scores, leads to the best performances.
-    Nuisance corrections are best estimated with flexible estimators such
-    as a super learner.
-    \\
-    \textbf{Conclusions:} When predictive models are used to
-    reason on the effect of an intervention, they must be evaluated
-    with different procedures than standard predictive
-    settings; using the $R\text{-risk}$ from causal inference.
-\end{abstract}
-\keywords{Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning, }
-\clearpage
-
-% XXX: for submission
-%\twocolumn
-% XXX: for submission (double spaced required)
-%\setstretch{1.1}
-\doublespacing
-\section{Introduction}\label{sec:intro}
-
-\subsection{Extending prediction to prescription needs causality}
-
-Prediction models have long been used in medicine, as with risk
-score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
-While these have historically been simple models on simple data, this is
-changing with progress in machine learning and richer medical data
-\cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
-\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
-patient records
-\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
-notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
-Complex data is difficult to control and model, but these models are
-validated by verifying the accuracy of the prediction on left-out data
-\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
-Crucial to the clinical adoption of a model predicting a health outcome
-is that it ``can support decisions about patient care''
-\cite{wyatt1995commentary}. Precision medicine is about
-guiding decisions: \emph{eg} will an individual benefit from an intervention such as surgery
-\cite{fontana2019can}? An estimate of the effect of the treatment can be
-obtained by contrasting model predictions with and without the treatment,
-but statistical validity requires causal inference
-\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.
-
-% - Causal inference is difficult
-% - Beyond the dream of assessing effectiveness and safety in real-word
-%   practice \cite{black1996we} include off-label drug usage 
-%   \cite{radley2006off} (which may lead to drug repurposing)
-% - Plug in that outcome modeling is complementary to RCTs: observational
-%   data gives weaker evidence than trials, but on both data outcome
-%   modeling opens the door to individual.
-% - Epidemiology has focused on propensity-score methods
-% - Principle causal inference with outcome models is possible
-% - It brings the promises of \emph{individualized treatment effects} --a
-%   goal related to capturing heterogeneity and thus CATE--
-% - It also performs well on causal-inference competition, although they
-%   focused on the ATE
-
-% The plethora of approaches approaches give different causal-inference
-% estimates => need for guideline on causal model selection.
-
-Indeed, concluding on the effect of a treatment is a difficult
-causal-inference task, as it can be easily compromised by confounding:
-spurious associations between treatment
-allocation and baseline health, \eg~only prescribing a drug to mild cases
-\cite{hernan_causal_2020,vanderweele2019principles}.
-Predictive modeling bridges to causal inference theory under the name of
-\emph{outcome models} (or G-computation,
-G-formula \cite{robins_role_1986}, Q-model
-\cite{snowden_implementation_2011}, conditional mean regression
-\cite{wendling_comparing_2018}).
-% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
-% alleviate this concern. Yet most machine-learning models are trained on
-% \emph{observational} data, close to real-word practice
-% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
-Medical statistics and epidemiology have mostly used other
-causal-inference methods, modeling treatment assignment
-with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
-the benefit of going
-beyond average effects, estimating individualized or conditional average
-treatment effects (CATE), central to precision medicine.
-%
-For this purpose, such methods are also invaluable on randomized trials
-\cite{su2018random,lamont2018identification,hoogland2021tutorial}.
-
-% Recent empirical results \cite{wendling_comparing_2018,dorie_automated_2019}
-% show benefits of outcome modeling to estimate average treatment effects.
-
-%\idea{There is a variety of models for g-estimation}
-% There is an explosion of outcome modeling or machine learning methods, either 
-% developed for
-% For instance, many deep-learning methods have been developed for medical
-% image analysis \cite{shen2017deep,monshi2020deep}.
-Outcome-modeling methods, even when specifically designed for causal
-inference, are numerous: Bayesian Additive Regression Trees
-\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
-\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
-\cite{powers_methods_2018}, causal multivariate adaptive regression
-splines \cite{powers_methods_2018}, random forests
-\cite{wager_estimation_2018, athey_generalized_2019},
-Meta-learners \cite{kunzel_metalearners_2019}, R-learners
-\cite{nie_quasioracle_2017}, Doubly robust estimation
-\cite{chernozhukov_double_2018}...
-%\idea{The variety of estimators calls for model selection procedures}
-The wide variety of methods raises the problem
-of selecting between different estimators based on the data at hand. %TODO: might be removed since repeated two sentences below
-%
-Indeed, estimates of treatment effects can vary markedly across different
-predictive models \cite{fang2019applying,dorie_automated_2019} (illustration in
-Appendix \ref{apd:toy_example:acic_2016_ate_variability}).
-
-Given complex health data, which predictive model is to be most trusted to yield
-valid causal estimates needed to motivate individual treatment decisions?
-As no single machine-learning method performs best across all data sets, there is a
-pressing need for clear guidelines to select outcome models for causal
-inference.
-
-\paragraph{Objectives and structure of the paper}
-
-We study \textit{model selection procedures} in practical settings,
-without theoretical assumptions often made in statistical literature such
-as \textit{infinite} data or \textit{well-specified} models (Appendix
-\ref{apd:prior_work}). Asymptotic causal-inference theory recommends complex
-risks, but a practical question is whether model-selection procedures, that rely
-on data split, can estimate these risks reliably enough. Indeed, these
-risks come with
-more quantities to estimate, which may bring additional variance, leading to
-worse model selection.
-
-We first illustrate the problem of causal model selection. Then we anchor causal model selection in the \emph{potential outcome}
-framework and details the causal risks and model-selection procedure. We then
-rewrite the so-called $R\text{-risk}$ as a reweighted version of mean squared
-difference between the true and estimated individualized treatment effect.
-Finally, we conduct a thorough empirical study comparing the different metrics
-on diverse datasets, using a family of simulations and real health
-data, going beyond prior work limited to specific simulation settings
-\cite{schuler_comparison_2018,alaa_validating_2019} (Appendix
-\ref{apd:prior_work}).
-
-% not the place for results
-% Results outline how to best select outcome models for causal
-% inference with an adapted
-% cross-validation to estimate the so-called $R\text{-risk}$.
-% This risk compensates for systematic
-% differences between treated and non-treated individuals using
-% two \emph{nuisance} models,
-% themselves estimated from data and thus imperfect; yet these
-% imperfections do not undermine the $R\text{-risk}$.
-
-\begin{figure}[h!]
-    % XXX: for submission
-    \centering\begin{minipage}{.6\linewidth}
-        {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
-                causal inference}}
-
-        \setlength{\unitlength}{.1\linewidth}
-        \begin{picture}(10,5.2)(0,0)
-            \put(0, 0){\includegraphics[width=1\linewidth]{images/toy_random_forest_high_R2_high_tau_risk.pdf}}%
-            \put(7.05,4.4){\vector(0,1){.47}}
-            \put(7.05,4.4){\vector(0,-1){.55}}
-            \put(7.1,4.5){\scriptsize\sffamily treatment effect}
-            \put(7.1,4.15){\small $\tau(x)$}
-            \put(5.2,3.6){\small $\hat{\tau}(x)$}
-            \put(5.15,3.6){\vector(0,1){.29}}
-            \put(5.15,3.6){\vector(0,-1){.15}}
-        \end{picture}
-
-        {\sffamily\footnotesize\scalebox{1.03}{b) Linear model, worse average
-            prediction but better causal inference}}
-
-        \hfill%
-        \includegraphics[width=1\linewidth]{images/toy_tlinear_model_small_R2_small_tau_risk.pdf}%
-    \end{minipage}
-
-    \caption[The best predictor may not estimate best causal
-        effects]{\textbf{Illustration}: a) a random-forest predictor
-        with high performance for standard prediction (high $R^2$) but that
-        yields poor causal estimates (large error between true effect $\tau$ and
-        estimated $\hat{\tau}$), b) a linear predictor with smaller
-        prediction performance leading to better causal estimation. \\[1ex]
-        Selecting the predictor with the smallest error to the individual
-        treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$
-        --the $\tau\text{-risk}$, eq.\,\ref{eq:tau_risk} -- would lead to
-        the best causal estimates; however computing this error is not
-        feasible: it requires access to unknown quantities:
-        $\tau(x)$. \\[1ex]
-        While the random forest fits the data better than the linear model, it
-        gives worse causal inference because its error is inhomogeneous between
-        treated and untreated. The $R^2$ score does not capture this
-        inhomogeneity.
-        \label{fig:toy_example}
-    }
-\end{figure}
-
-\subsection{Illustration: the best predictor may not estimate best causal
-    effects}%
-
-Using a predictor to reason on causal effects relies on contrasting the
-prediction of the outcome for a given individual with and without the treatment.
-%
-Given various predictors of the outcome, which one should we use?
-%
-Standard predictive modeling or machine-learning practice selects the predictor
-that minimizes the expected error on the outcome
-\cite{poldrack2020establishment,varoquaux2022evaluating}. However, this
-predictor may not be the best model to reason about causal effects of an
-intervention as Figure \ref{fig:toy_example} illustrates. Consider the
-probability $Y$ of an undesirable outcome (\eg death), a binary treatment $A \in
-    \{0, 1\}$, and a covariate $X \in \mathbb R$ summarizing the patient health
-status (\eg the Charlson index \cite{charlson_new_1987}). We simulate a
-treatment beneficial (decreases mortality) for patients with high Charlson
-scores (bad health status) but with little effect for patients in good condition
-(low Charlson scores).
-
-Figure \ref{fig:toy_example}a shows a random forest predictor with a
-counter-intuitive behavior: it predicts well on average the outcome (as measured
-by a regression $R^2$ score) but perform poorly to estimate causal quantities:
-the average treatment effect $\tau$ (as visible via the error $|\tau -
-    \hat{\tau}|$) or the conditional average treatment effect (the error
-$\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$, called CATE).
-%
-On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
-smaller $R^2$ score but better causal inference.%
-
-The problem is that causal estimation requires controlling
-an error on both treated and non-treated outcome for the same individual:
-the observed outcome, and the non-observed \emph{counterfactual} one.
-The linear model is misspecified --the outcome functions are not
-linear--, leading to poor $R^2$; but it interpolates better to regions
-where there are few untreated individuals --high Charlson score-- and
-thus gives better causal estimates. Conversely, the random forest puts
-weaker assumptions on the data, thus has higher $R^2$ score but is biased
-by the treated population in the poor-overlap region, leading
-to bad causal estimates.
-
-This toy example illustrates that the classic minimum Mean Square Error
-criterion is not suited to choosing a model among candidate
-estimators for causal inference.
-
-% A natural way to select a predictive model for causal inference would be
-
-% an error measure between a causal quantity such as the CATE and models' estimate. But such error is
-% not a ``feasible'' risk: it cannot be computed solely from observed data
-% and requires oracle knowledge.
-
 \section{Methods}\label{sec:framework}
@@
 
+\setlength{\columnsep}{0.5cm}
+\usepackage{wrapfig}
+%\usepackage{wraptable}
+\usepackage{authblk}% for author affiliations
+\usepackage[dvipsnames]{xcolor}
+\usepackage{setspace} % double spaced required by journal
+\doublespacing
+
+\usepackage[%
+    font={footnotesize,sf},
+    labelfont=bf,
+    format=hang,    
+    format=plain,
+    margin=0pt,
+]{caption}
+
+% TODO commands
+\usepackage{snaptodo}
+\setlength{\marginparwidth}{2.3cm}
+\setlength{\marginparsep}{.1cm}
+\snaptodoset{block rise=2em}
+\snaptodoset{margin block/.style={font=\tiny}}
+
+\newcommand{\gv}[1]{\snaptodo[margin block/.append style=green!50!black]{\sloppy\textbf{GV}: #1}}
+
+% Uncomment below to erase all comments
+%\renewcommand{\snaptodo}[2][1]{\relax}
+
+% XXX: for submission
+\usepackage{setspace}
+
+\usepackage{comment}
+\setcounter{secnumdepth}{0}
+
+%\setlength{\marginparwidth}{0.5cm}
+%\setlength\textwidth{177.8mm}
+
+% maths commands
+\newcommand{\indep}{\perp \!\!\! \perp}
+\newcommand{\eg}{\emph{e.g.}}
+\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
+\usepackage{mathtools}
+\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}
+
+\newtheorem{definition}{Definition}
+\newtheorem{condition}{Condition}
+\newtheorem{property}{Property}
+\newtheorem{remark}{Remark}
+\newtheorem{notation}{Notation}
+\newtheorem{proposition*}{Proposition}
+\newtheorem{proposition}{Proposition}
+\newtheorem{lemma}{Lemma}
+\newtheorem{theorem}{Theorem}
+\newtheorem{assumption}{Assumption}
+\newtheorem{proof}{Proof}
+%
+\def\keywordname{{\bfseries \emph{Keywords}}}%
+\def\keywords#1{\par\addvspace\medskipamount{\rightskip=0pt plus1cm
+\def\and{\ifhmode\unskip\nobreak\fi\ $\cdot$
+}\noindent\keywordname\enspace\ignorespaces#1\par}}
+
+\title{How to select predictive models for decision-making or causal inference?}
+\addbibresource{biblio.bib}
+
+\author[1,2,*]{Matthieu Doutreligne, PhD}
+\author[1]{Gaël Varoquaux, PhD}
+
+\affil[1]{Inria, Soda, Saclay, France}
+\affil[2]{Mission Data, Haute Autorité de Santé, Saint-Denis, France}
+\affil[*]{Corresponding author: matt.dout@gmail.com}
+
+%\raggedbottom
+
+\begin{document}
+
+%\corres{*\email{matthieu.doutreligne@inria.fr}}
+
+\onecolumn
+\maketitle
+\begin{abstract}\noindent
+    \textbf{Background:}
+    We investigate which procedure selects the predictive model most
+    trustworthy to reason on the effect of an intervention and support
+    decision making.\\
+    %As predictive models --\eg from machine learning-- give likely
+    %outcomes, they may be used to, a
+    %causal-inference task. The increasing complexity of health data has opened
+    %the door to a plethora of models, but also the Pandora box of model
+    %selection: which of these models yield the most valid causal estimates? 
+    %\textbf{Methods:}
+    We study such a variety of model selection procedures in practical settings:
+    finite samples settings and without theoretical assumption of
+    \emph{well-specified} models. Beyond standard cross-validation or
+    internal validation procedures, we also study elaborate causal risks.
+    These build proxies of the causal error using ``nuisance''
+    re-weighting to compute it on the observed data. We evaluate whether
+    empirically estimated nuisances, which are necessarily noisy,
+    add noise to model selection. We compare different metrics for
+    causal model selection in an extensive empirical
+    study based on a simulation and three healthcare datasets
+    based on real covariates. \\
+    \textbf{Results:} Among all metrics, the mean squared error,
+    classically used to evaluate predictive modes,
+    is worse. Re-weighting it with propensity score
+    does not bring much improvements. The
+    $R\text{-risk}$, which uses as nuisances a model of
+    mean outcome and propensity scores, leads to the best performances.
+    Nuisance corrections are best estimated with flexible estimators such
+    as a super learner.
+    \\
+    \textbf{Conclusions:} When predictive models are used to
+    reason on the effect of an intervention, they must be evaluated
+    with different procedures than standard predictive
+    settings; using the $R\text{-risk}$ from causal inference.
+\end{abstract}
+\keywords{Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning, }
+\clearpage
+
+% XXX: for submission
+%\twocolumn
+% XXX: for submission (double spaced required)
+%\setstretch{1.1}
+\doublespacing
+\section{Introduction}\label{sec:intro}
+
+\subsection{Extending prediction to prescription needs causality}
+
+Prediction models have long been used in medicine, as with risk
+score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
+While these have historically been simple models on simple data, this is
+changing with progress in machine learning and richer medical data
+\cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
+\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
+patient records
+\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
+notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
+Complex data is difficult to control and model, but these models are
+validated by verifying the accuracy of the prediction on left-out data
+\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
+Crucial to the clinical adoption of a model predicting a health outcome
+is that it ``can support decisions about patient care''
+\cite{wyatt1995commentary}. Precision medicine is about
+guiding decisions: \emph{eg} will an individual benefit from an intervention such as surgery
+\cite{fontana2019can}? An estimate of the effect of the treatment can be
+obtained by contrasting model predictions with and without the treatment,
+but statistical validity requires causal inference
+\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.
+
+% - Causal inference is difficult
+% - Beyond the dream of assessing effectiveness and safety in real-word
+%   practice \cite{black1996we} include off-label drug usage 
+%   \cite{radley2006off} (which may lead to drug repurposing)
+% - Plug in that outcome modeling is complementary to RCTs: observational
+%   data gives weaker evidence than trials, but on both data outcome
+%   modeling opens the door to individual.
+% - Epidemiology has focused on propensity-score methods
+% - Principle causal inference with outcome models is possible
+% - It brings the promises of \emph{individualized treatment effects} --a
+%   goal related to capturing heterogeneity and thus CATE--
+% - It also performs well on causal-inference competition, although they
+%   focused on the ATE
+
+% The plethora of approaches approaches give different causal-inference
+% estimates => need for guideline on causal model selection.
+
+Indeed, concluding on the effect of a treatment is a difficult
+causal-inference task, as it can be easily compromised by confounding:
+spurious associations between treatment
+allocation and baseline health, \eg~only prescribing a drug to mild cases
+\cite{hernan_causal_2020,vanderweele2019principles}.
+Predictive modeling bridges to causal inference theory under the name of
+\emph{outcome models} (or G-computation,
+G-formula \cite{robins_role_1986}, Q-model
+\cite{snowden_implementation_2011}, conditional mean regression
+\cite{wendling_comparing_2018}).
+% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
+% alleviate this concern. Yet most machine-learning models are trained on
+% \emph{observational} data, close to real-word practice
+% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
+Medical statistics and epidemiology have mostly used other
+causal-inference methods, modeling treatment assignment
+with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
+the benefit of going
+beyond average effects, estimating individualized or conditional average
+treatment effects (CATE), central to precision medicine.
+%
+For this purpose, such methods are also invaluable on randomized trials
+\cite{su2018random,lamont2018identification,hoogland2021tutorial}.
+
+% Recent empirical results \cite{wendling_comparing_2018,dorie_automated_2019}
+% show benefits of outcome modeling to estimate average treatment effects.
+
+%\idea{There is a variety of models for g-estimation}
+% There is an explosion of outcome modeling or machine learning methods, either 
+% developed for
+% For instance, many deep-learning methods have been developed for medical
+% image analysis \cite{shen2017deep,monshi2020deep}.
+Outcome-modeling methods, even when specifically designed for causal
+inference, are numerous: Bayesian Additive Regression Trees
+\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
+\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
+\cite{powers_methods_2018}, causal multivariate adaptive regression
+splines \cite{powers_methods_2018}, random forests
+\cite{wager_estimation_2018, athey_generalized_2019},
+Meta-learners \cite{kunzel_metalearners_2019}, R-learners
+\cite{nie_quasioracle_2017}, Doubly robust estimation
+\cite{chernozhukov_double_2018}...
+%\idea{The variety of estimators calls for model selection procedures}
+The wide variety of methods raises the problem
+of selecting between different estimators based on the data at hand. %TODO: might be removed since repeated two sentences below
+%
+Indeed, estimates of treatment effects can vary markedly across different
+predictive models \cite{fang2019applying,dorie_automated_2019} (illustration in
+Appendix \ref{apd:toy_example:acic_2016_ate_variability}).
+
+Given complex health data, which predictive model is to be most trusted to yield
+valid causal estimates needed to motivate individual treatment decisions?
+As no single machine-learning method performs best across all data sets, there is a
+pressing need for clear guidelines to select outcome models for causal
+inference.
+
+\paragraph{Objectives and structure of the paper}
+
+We study \textit{model selection procedures} in practical settings,
+without theoretical assumptions often made in statistical literature such
+as \textit{infinite} data or \textit{well-specified} models (Appendix
+\ref{apd:prior_work}). Asymptotic causal-inference theory recommends complex
+risks, but a practical question is whether model-selection procedures, that rely
+on data split, can estimate these risks reliably enough. Indeed, these
+risks come with
+more quantities to estimate, which may bring additional variance, leading to
+worse model selection.
+
+We first illustrate the problem of causal model selection. Then we anchor causal model selection in the \emph{potential outcome}
+framework and details the causal risks and model-selection procedure. We then
+rewrite the so-called $R\text{-risk}$ as a reweighted version of mean squared
+difference between the true and estimated individualized treatment effect.
+Finally, we conduct a thorough empirical study comparing the different metrics
+on diverse datasets, using a family of simulations and real health
+data, going beyond prior work limited to specific simulation settings
+\cite{schuler_comparison_2018,alaa_validating_2019} (Appendix
+\ref{apd:prior_work}).
+
+% not the place for results
+% Results outline how to best select outcome models for causal
+% inference with an adapted
+% cross-validation to estimate the so-called $R\text{-risk}$.
+% This risk compensates for systematic
+% differences between treated and non-treated individuals using
+% two \emph{nuisance} models,
+% themselves estimated from data and thus imperfect; yet these
+% imperfections do not undermine the $R\text{-risk}$.
+
+\begin{figure}[h!]
+    % XXX: for submission
+    \centering\begin{minipage}{.6\linewidth}
+        {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
+                causal inference}}
+
+        \setlength{\unitlength}{.1\linewidth}
+        \begin{picture}(10,5.2)(0,0)
+            \put(0, 0){\includegraphics[width=1\linewidth]{images/toy_random_forest_high_R2_high_tau_risk.pdf}}%
+            \put(7.05,4.4){\vector(0,1){.47}}
+            \put(7.05,4.4){\vector(0,-1){.55}}
+            \put(7.1,4.5){\scriptsize\sffamily treatment effect}
+            \put(7.1,4.15){\small $\tau(x)$}
+            \put(5.2,3.6){\small $\hat{\tau}(x)$}
+            \put(5.15,3.6){\vector(0,1){.29}}
+            \put(5.15,3.6){\vector(0,-1){.15}}
+        \end{picture}
+
+        {\sffamily\footnotesize\scalebox{1.03}{b) Linear model, worse average
+            prediction but better causal inference}}
+
+        \hfill%
+        \includegraphics[width=1\linewidth]{images/toy_tlinear_model_small_R2_small_tau_risk.pdf}%
+    \end{minipage}
+
+    \caption[The best predictor may not estimate best causal
+        effects]{\textbf{Illustration}: a) a random-forest predictor
+        with high performance for standard prediction (high $R^2$) but that
+        yields poor causal estimates (large error between true effect $\tau$ and
+        estimated $\hat{\tau}$), b) a linear predictor with smaller
+        prediction performance leading to better causal estimation. \\[1ex]
+        Selecting the predictor with the smallest error to the individual
+        treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$
+        --the $\tau\text{-risk}$, eq.\,\ref{eq:tau_risk} -- would lead to
+        the best causal estimates; however computing this error is not
+        feasible: it requires access to unknown quantities:
+        $\tau(x)$. \\[1ex]
+        While the random forest fits the data better than the linear model, it
+        gives worse causal inference because its error is inhomogeneous between
+        treated and untreated. The $R^2$ score does not capture this
+        inhomogeneity.
+        \label{fig:toy_example}
+    }
+\end{figure}
+
+\subsection{Illustration: the best predictor may not estimate best causal
+    effects}%
+
+Using a predictor to reason on causal effects relies on contrasting the
+prediction of the outcome for a given individual with and without the treatment.
+%
+Given various predictors of the outcome, which one should we use?
+%
+Standard predictive modeling or machine-learning practice selects the predictor
+that minimizes the expected error on the outcome
+\cite{poldrack2020establishment,varoquaux2022evaluating}. However, this
+predictor may not be the best model to reason about causal effects of an
+intervention as Figure \ref{fig:toy_example} illustrates. Consider the
+probability $Y$ of an undesirable outcome (\eg death), a binary treatment $A \in
+    \{0, 1\}$, and a covariate $X \in \mathbb R$ summarizing the patient health
+status (\eg the Charlson index \cite{charlson_new_1987}). We simulate a
+treatment beneficial (decreases mortality) for patients with high Charlson
+scores (bad health status) but with little effect for patients in good condition
+(low Charlson scores).
+
+Figure \ref{fig:toy_example}a shows a random forest predictor with a
+counter-intuitive behavior: it predicts well on average the outcome (as measured
+by a regression $R^2$ score) but perform poorly to estimate causal quantities:
+the average treatment effect $\tau$ (as visible via the error $|\tau -
+    \hat{\tau}|$) or the conditional average treatment effect (the error
+$\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$, called CATE).
+%
+On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
+smaller $R^2$ score but better causal inference.%
+
+The problem is that causal estimation requires controlling
+an error on both treated and non-treated outcome for the same individual:
+the observed outcome, and the non-observed \emph{counterfactual} one.
+The linear model is misspecified --the outcome functions are not
+linear--, leading to poor $R^2$; but it interpolates better to regions
+where there are few untreated individuals --high Charlson score-- and
+thus gives better causal estimates. Conversely, the random forest puts
+weaker assumptions on the data, thus has higher $R^2$ score but is biased
+by the treated population in the poor-overlap region, leading
+to bad causal estimates.
+
+This toy example illustrates that the classic minimum Mean Square Error
+criterion is not suited to choosing a model among candidate
+estimators for causal inference.
+
+% A natural way to select a predictive model for causal inference would be
+
+% an error measure between a causal quantity such as the CATE and models' estimate. But such error is
+% not a ``feasible'' risk: it cannot be computed solely from observed data
+% and requires oracle knowledge.
+
+\section{Methods}\label{sec:framework}
+\subsection{Neyman-Rubin Potential Outcomes framework}%
+\label{sec:neyman_rubin}%
+
+\paragraph{Settings}
+
+The Neyman-Rubin Potential Outcomes framework
+\cite{naimi2023defining,imbens_causal_2015} enables statistical reasoning on
+causal treatment effects: Given an outcome $Y \in \mathbb R$ (\eg mortality risk
+or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
+    \{0, 1\}$ (\eg~a medical procedure), and baseline
+covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
+distribution, $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However,
+we want to model the existence of potential observations (unobserved ie.~counterfactual) that correspond to a different treatment. Thus we want
+quantities on the counterfactual distribution $O^{*} = (Y(1), Y(0), X, A) \sim
+    \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.
+
+Popular quantities of interest (estimands) are:
+at the population level, the
+Average Treatment Effect
+\begin{flalign*}
+    \text{ATE} &  &
+    \tau \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal D^*}[Y(1) - Y(0)];
+               &  &
+\end{flalign*}
+at the individual level, to model heterogeneity, the Conditional Average Treatment Effect
+\begin{flalign*}
+    \text{CATE} &  &
+    \tau (x) \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal{D}^\star}[Y(1) - Y(0) | X=x].
+                &  &
+\end{flalign*}
+
+\paragraph{Causal assumptions}
+
+A given data needs to meet a few assumptions to enable identifying
+causal estimands \cite{rubin_causal_2005}. The usual
+\emph{strong ignorability} assumptions are  (details in
+\ref{apd:causal_assumptions}): \emph{1)}
+\emph{unconfoundedness} \mbox{$\{Y(0),
+        Y(1) \} \indep A | X$}, \emph{~2)} \emph{strong overlap} ie. every patient has a
+strictly positive probability to receive each treatment, \emph{3)}
+\emph{consistency}, and \emph{4)} \emph{generalization}.
+
+\paragraph{Estimating treatment effects with outcome models -- g-computation \cite{robins_new_1986}}\label{subsec:estimators}
+
+Should we know the two expected outcomes for a given $X$,
+we could compute the
+difference between them, which gives the causal effect of the treatment.
+%
+These two expected outcomes can be estimated from observed data:
+the consistency \ref{assumption:consistency} and unconfoundedness
+\ref{assumption:ignorability} assumptions imply the following equality:
+\begin{equation}\label{eq:mu_identification}
+    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
+\end{equation}
+On the left, the expectation is taken on the counterfactual unobserved
+distribution. On the right, the expectation is taken on the factual observed
+distribution conditionally on the treatment. For the rest of the
+paper, the expectations will always be taken on the factual observed
+distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
+g-computation estimators \cite{snowden_implementation_2011}):
+\begin{eqnarray}
+    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x]
+    \notag
+    \\
+    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
+    \label{eq:tau_population}
+\end{eqnarray}
+This equation builds on two quantities: the conditional expectancy
+of the outcome given the covariates and either
+treatment or no no treatment, called \emph{response function}:
+\begin{flalign*}
+    \text{Response function}
+     &  &
+    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
+     &  &
+\end{flalign*}
+
+Given a sample of data and the oracle response functions $\mu_0, \mu_1$, the
+finite sum version of \autoref{eq:tau_population} leads to an
+estimator of the ATE written:
+\begin{equation}
+    \hat \tau = \frac{1}{n} \biggl(\sum_{i=1}^n \mu_{1}(x_i) - \mu_{0}(x_i) \biggr)
+    \label{eq:ate_estimate}
+\end{equation}
+This estimator is an oracle \emph{finite sum} estimator by opposition to the
+population expression of $\tau$, $\mathbb{E}[\mu_{1}(x_i) - \mu_{0}(x_i)]
+$,
+which involves an expectation taken on the full
+distribution $\mathcal D$, which is observable but requires infinite data. For
+each estimator $\ell$ taking an expectation over $\mathcal D$, we use the symbol
+$\hat \ell$ to note its finite sum version.
+
+Similarly to the ATE, at the individual level, the CATE:
+\begin{equation}
+    \tau(x) = \mu_{1}(x) - \mu_{0}(x)
+    \label{eq:cate_estimate}
+\end{equation}
+
+\paragraph{Robinson decomposition}
+The \emph{R-decomposition}
+of the outcome model\cite{robinson_rootnconsistent_1988} plays an important role:
+introducing two quantities, the conditional mean outcome
+and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
+\begin{flalign}
+    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
+    \mathcal{D}} [Y|X=x]            &                    &
+    \label{def:m}
+    \\
+    \text{Propensity score}         &                    &
+    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
+    \label{def:propensity_score}
+\end{flalign}
+the outcome can be written
+\begin{flalign}\label{eq:r_decomposition}
+    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
+    \tau(x) + \varepsilon(x; a)\notag &  &
+    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
+\end{flalign}
+$m$ and $e$ are often called
+\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown.
+
+%As noted by \cite{johansson2022generalization}, the machine learning
+%community often referred to the CATE by ITE, the Individual Treatment Effect.
+%From a purely causal point of view, the ITE is uniquely defined for each
+%individual and might not be accessible: $ITE(x_i) = Y_i(1) -  Y_i(0)$. On the
+%contrary, the CATE can always be derived by taking conditional expectancies. It
+%is the expected effect of the treatment in the region of the covariate space
+%around X. %Too cultivate
+
+\begin{table*}[bt!]
+    \makebox[\textwidth]{
+        \begin{threeparttable}[b]
+            \caption{Review of causal risks
+                ---
+                The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
+                \citet{schuler_comparison_2018}.
+                \label{tab:evaluation_metrics}}
+            \centering
+            \begin{tabular}{llr}
+                \toprule
+                Risk                                                                                           & Equation
+                                                                                                               & Reference                                                                                                                                           \\
+                \midrule
+                $\tau\text{-risk}=\text{MSE}(\tau(X), \tau_f(X))$                                              & $\mathbb E_{X\sim
+                            p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
+                                                                                                               & Eq.
+\ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
+                $\mu\text{-risk} = \text{MSE}(Y, f(X))$                                                        & $\mathbb{E}_{(Y, X, A)
+                        \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
+                                                                                                               & Def.
+\ref{def:mu_risk} \cite{schuler_comparison_2018}                                                                                               \\
+                $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
+                        \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
+                (Y-f(X ; A))^2 \right]$                                                                        & Def.
+                \ref{def:mu_ipw_risk} \cite{vanderlaan_unified_2003}                                                                                                                                                                                                 \\
+                $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
+                Def. \ref{def:tau_ipw_risk} \cite{wager_estimation_2018}
+                \\
+                $U\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A) \sim \mathcal D}  \big[
+                \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
+                Def. \ref{def:u_risk} \cite{nie_quasioracle_2017}
+                \\
+                $R\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A)
+                        \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
+                -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
+                Def. \ref{def:r_risk} \cite{nie_quasioracle_2017}
+                \\
+                \bottomrule
+            \end{tabular}
+        \end{threeparttable}
+    }
+\end{table*}
+
+\subsection{Model-selection risks, oracle and feasible}\label{sec:problem:model_selection}
+
+\paragraph{Causal model selection}\label{sec:problem:causal_selection}
+
+We formalize model selection for causal estimation. Thanks to the g-formula
+identification (\autoref{eq:mu_identification}), a given outcome model $f: \mathcal X
+    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
+domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
+\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
+%
+Let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
+a family of such estimators. Our goal is to select the best candidate in this
+family for the observed dataset $O$ using a risk
+$\ell$:
+\begin{equation}
+    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
+    \label{eq:causal_model_selection}
+\end{equation}
+
+We now detail possible risks $\ell$, risks useful for causal
+model selection, and how to compute them.
+
+\paragraph{The $\tau\text{-risk}$: an oracle error risk}\label{paragraph:oracle_metrics}
+%
+As we would like to target the CATE, the following
+evaluation risk is natural (also called PEHE \cite{schulam_reliable_2017, hill_bayesian_2011}):
+\begin{flalign}\label{eq:tau_risk}
+    \tau\text{-risk}(f) \myeq \mathbb E_{X\sim p(X)}[(\tau(X) - \hat \tau_f(X))^2]
+\end{flalign}
+
+Given observed data from $p(X)$, the expectation is computed with a
+finite sum, as in eq.\,\ref{eq:ate_estimate}, to give an estimated
+value $\widehat{\tau\text{-risk}}(f)$.
+%
+However this risk is not feasible as the oracles $\tau(x)$ are
+not accessible with the observed data $(Y, X, A) \sim \mathcal D$.
+
+\paragraph{Feasible error risks}\label{paragraph:feasible_metrics} Table
+\ref{tab:evaluation_metrics} lists \emph{feasible} risks (Detailed in Appendix
+\ref{def:feasible_risks}), based on the prediction error of the outcome model
+and \emph{observable} quantities. These observable, called nuisances are $e$
+--propensity score, eq \ref{def:propensity_score}-- and $m$ --conditional mean
+outcome, eq \ref{def:m}. We give the definitions as \textit{semi-oracles},
+function of the true unknown nuisances, but later instantiate them with
+estimated nuisances, noted $\big(\check e, \check m \big)$. Semi-oracles risks
+are superscripted with the $^{\star}$ symbol.
+
+\subsection{Estimation and model selection procedure}\label{problem:estimation_procedure}
+
+Causal model selection (eq
+\ref{eq:causal_model_selection}) may involve estimating various quantities
+from the observed data: the outcome model $f$, its induced risk as
+introduce in the previous section, and possibly nuisances required by the
+risk.
+Given a dataset with $N$ samples, we split out a train and a test sets
+$(\mathcal{T}, \mathcal{S})$. We
+fit each candidate estimator $f \in \mathcal{F}$ on $\mathcal{T}$. We also fit
+the nuisance models $(\check e, \check m)$ on the train set
+$\mathcal{T}$, setting hyperparameters by a nested
+cross-validation before fitting the nuisance estimators with these parameters
+on the full train set. Causal quantities are then computed by applying the fitted  candidates
+estimators $f \in \mathcal{F}$ on the test set $\mathcal{S}$. Finally, we
+compute the model-selection metrics for
+each candidate model on the test set. This procedure is described in Algorithm
+\ref{problem:estimation_procedure:algo} and Figure
+\ref{problem:estimation_procedure:figure}.
+
+% As extreme inverse propensity weights induce high variance, clipping can be
+% useful for numerical stability
+% \cite{swaminathan_counterfactual_2015, ionides_truncated_2008}.
+
+\begin{algorithm}[!htbp]
+    \caption{Model selection procedure}\label{problem:estimation_procedure:algo} {%
+        Given train and test sets $(\mathcal{T}, \mathcal{S}) \sim \mathcal{D}$,
+        a candidate estimators $f$, a causal
+        metrics $\ell$:
+        \begin{enumerate}
+            \item Prefit: Learn estimators for unknown nuisance quantities $(\check e,\,\check m)$ on the training set $\mathcal{T}$
+            \item Fit: learn $\hat f(\cdot, a)$ on
+                  $\mathcal T$
+            \item Model selection: $\forall{x} \in \mathcal{S}$ predict
+                  $\big(\hat f(x, 1), \hat f(x, 0)\big)$ and evaluate the estimator storing the metric value: $\ell(f, \mathcal S)$ -- possibly
+                  function of $\check e$ and $\check m$
+                  %\item Metric evaluation: return the oracle evaluation metrics
+                  %evaluated on $\mathcal{S}$: $\big(\widehat{\tau\text{-risk}}(\hat
+                  %f^*_{\ell}); \widehat{\ell}_{ATE}(\hat f^*_{\ell}) \big)$
+        \end{enumerate}
+
+    }
+\end{algorithm}
+
+\begin{figure}[h!]
+    % XXX: for submission
+    \centering\begin{minipage}{.6\linewidth}
+        \includegraphics[width=\linewidth]{images/estimation_procedure_causal_selection_procedure.png}
+    \end{minipage}
+    \caption{Estimation procedure for causal model
+        selection.}\label{problem:estimation_procedure:figure}
+\end{figure}
+
+\subsection{R-risk as reweighted oracle metric}\label{theory:r_risk_rewrite}
+
+The $R\text{-risk}$ can be rewritten as a rebalanced $\tau \text{-risk}$.
+
+% We now one feasible risks, $\mu \text{-risk}_{IPW}$ and the
+% $R\text{-risk}$ to the oracle $\tau\text{-risk}$. Both results make
+% explicit the role of overlap for the performances of causal risks.
+
+This rewriting involves reweighted residuals: for each potential outcome, $a \in
+    \{0; 1\}$, the variance conditionally on $x$ is \cite{shalit_estimating_2017}:
+\begin{equation*}\label{eq:residuals}
+    \sigma_{y}^{2}(x ; a) \overset{\text{def}}{=}
+    \int_{y}\left(y-\mu_{a}(x)\right)^{2} p(y \mid x=x ; A=a) \, d y
+\end{equation*}
+Integrating over the population, we get the Bayes squared error:
+$\sigma^2_{B}(a) = \int_{\mathcal X} \sigma_y^2(x;a) p(x)dx$
+and its propensity weighted version:
+$\tilde{\sigma}^2_{B}(a) = \int_{\mathcal X}\sigma_y^2(x;a)\,  p(x;
+    a)\,dx$. In case of a purely deterministic link between the
+covariates, the treatment, and the outcome, these residual terms are null.
+
+%\md{Lower bound: I think that there is no lower bound of the tau-risk with the
+%mu-risk. That means that we can have a mu risk of 0 wheras the taurisk is
+%non-null. Todo for myself: develop an example with overfitted 1NN on the
+%observed data.}
+
+\begin{proposition}[$R \text{-risk}$ as reweighted
+        $\tau\text{-risk}$]\label{theory:prop:r_risk_rewrite} Given an outcome model
+    $f$, its $R\text{-risk}$ appears as weighted version of its $\tau\text{-risk}$
+    (Proof in \ref{apd:proofs:r_risk_rewrite}):
+    \begin{align}
+        R\text{-risk}^*(f) & = \int_{x} e(x)\big(1-e(x)\big)\big(\tau(x)-\tau_ {f}(x)\big)^{2} p(x) d x \nonumber \\
+                           & \quad\; + \tilde{\sigma}_B^2(1) + \tilde{\sigma}_B^{2}(0)
+    \end{align}
+\end{proposition}
+
+The $R \text{-risk}$ targets the oracle at the cost of an overlap re-weighting
+and the addition of the reweighted Bayes residuals, which are independent of
+$f$. In good overlap regions the weights $e(x) \big(1-e(x) \big)$ are close to
+$\frac{1}{4}$, hence the $R \text{-risk}$ is close to the desired gold-standard
+$\tau \text{-risk}$. For randomized control trials, this weight is constant
+making the $R\text{-risk}$ particularly suited for exploring
+heterogeneity (Appendix \ref{apd:theory:special_cases})
+
 \section{Empirical Study}\label{sec:empirical_study}

### 2024-06-02 how_to_select.tex
gigascience: new jounrnal compiling format

--- 2024-06-02 _old_how_to_select_causal_models.tex
@@
+%%
+%% Copyright 2022 OXFORD UNIVERSITY PRESS
+%%
+%% This file is part of the 'oup-authoring-template Bundle'.
+%% ---------------------------------------------
+%%
+%% It may be distributed under the conditions of the LaTeX Project Public
+%% License, either version 1.2 of this license or (at your option) any
+%% later version.  The latest version of this license is in
+%%    http://www.latex-project.org/lppl.txt
+%% and version 1.2 or later is part of all distributions of LaTeX
+%% version 1999/12/01 or later.
+%%
+%% The list of all files belonging to the 'oup-authoring-template Bundle' is
+%% given in the file `manifest.txt'.
+%%
+%% Template article for OXFORD UNIVERSITY PRESS's document class `oup-authoring-template'
+%% with bibliographic references
+%%
+
+%%%CONTEMPORARY%%%
+\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
+%\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
+%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
+%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}
+
+%%%MODERN%%%
+%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
+%\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
+%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
+%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}
+
+%%%TRADITIONAL%%%
+%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
+%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
+%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
+%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}
+
+%\onecolumn % for one column layouts
+
+%\usepackage{showframe}
+
+\graphicspath{{images/}}
+
+% line numbers
+%\usepackage[mathlines, switch]{lineno}
+%\usepackage[right]{lineno}
+
+\theoremstyle{thmstyleone}%
+\newtheorem{theorem}{Theorem}%  meant for continuous numbers
+%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
+%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
+\newtheorem{proposition}[theorem]{Proposition}%
+%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
+\theoremstyle{thmstyletwo}%
+\newtheorem{example}{Example}%
+\newtheorem{remark}{Remark}%
+\theoremstyle{thmstylethree}%
+\newtheorem{definition}{Definition}
+
+% added commands and packages
+\newtheorem{assumption}{Assumption}
+\newtheorem{proposition*}{Proposition}
+\newcommand{\indep}{\perp \!\!\! \perp}
+\newcommand{\eg}{\emph{e.g.}}
+\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
+\usepackage{mathtools}
+\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}
+\usepackage{booktabs} % For formal tables (eg. bottomrules command)
+\usepackage{threeparttable} % For table 
+\usepackage{caption} % for subfigures
+\usepackage{subcaption}
+\usepackage{placeins} % for FloatBarrier command to prevent a figure to pass some point.
+
+\bibliographystyle{unsrtnat}
+% add external bibliography
+%\usepackage[sorting=nyt,style=numeric,natbib=true,doi=false,isbn=false,url=false,eprint=false, maxnames=999, maxcitenames=2]{biblatex} 
+%\let\cite=\supercite
+%\DeclareNameAlias{default}{last-first}
+%\addbibresource{biblio.bib}
+
+\begin{document}
+
+\journaltitle{GigaScience}
+\DOI{DOI HERE}
+\copyrightyear{2024}
+\pubyear{2024}
+\access{Advance Access Publication Date: Day Month Year}
+\appnotes{Paper}
+
+\firstpage{1}
+
+%\subtitle{Subject Section}
+
+\title[How to select predictive models for decision-making or causal inference?]{How to select predictive models for decision-making or causal inference?}
+
+\author[1, 2 $\ast$]{Matthieu Doutreligne}
+\author[1]{Gaël Varoquaux}
+
+\authormark{Author Name et al.}
+
+\address[1]{\orgdiv{Soda}, \orgname{Inria, Saclay}, \orgaddress{\country{France}}}
+\address[2]{\orgdiv{Mission Data}, \orgname{Haute Autorité de Santé}, \orgaddress{\country{France}}}
+
+\corresp[$\ast$]{Corresponding author. \href{email:matt.dout@gmail.com}{matt.dout@gmail.com}}
+
+\received{Date}{0}{Year}
+\revised{Date}{0}{Year}
+\accepted{Date}{0}{Year}
+
+%\editor{Associate Editor: Name}
+
+%\abstract{
+%\textbf{Motivation:} .\\
+%\textbf{Results:} .\\
+%\textbf{Availability:} .\\
+%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
+%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
+%online.}
+
+\abstract{\textbf{Background:}
+    We investigate which procedure selects the predictive model most
+    trustworthy to reason on the effect of an intervention and support
+    decision making.\\
+    %As predictive models --\eg from machine learning-- give likely
+    %outcomes, they may be used to, a
+    %causal-inference task. The increasing complexity of health data has opened
+    %the door to a plethora of models, but also the Pandora box of model
+    %selection: which of these models yield the most valid causal estimates? 
+    %\textbf{Methods:}
+    We study such a variety of model selection procedures in practical settings:
+    finite samples settings and without theoretical assumption of
+    \emph{well-specified} models. Beyond standard cross-validation or
+    internal validation procedures, we also study elaborate causal risks.
+    These build proxies of the causal error using ``nuisance''
+    re-weighting to compute it on the observed data. We evaluate whether
+    empirically estimated nuisances, which are necessarily noisy,
+    add noise to model selection. We compare different metrics for
+    causal model selection in an extensive empirical
+    study based on a simulation and three healthcare datasets
+    based on real covariates. \\
+    \textbf{Results:} Among all metrics, the mean squared error,
+    classically used to evaluate predictive modes,
+    is worse. Re-weighting it with propensity score
+    does not bring much improvements. The
+    $R\text{-risk}$, which uses as nuisances a model of
+    mean outcome and propensity scores, leads to the best performances.
+    Nuisance corrections are best estimated with flexible estimators such
+    as a super learner.
+    \\
+    \textbf{Conclusions:} When predictive models are used to
+    reason on the effect of an intervention, they must be evaluated
+    with different procedures than standard predictive
+    settings; using the $R\text{-risk}$ from causal inference.}
+\keywords{Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning, }
+
+% \boxedtext{
+% \begin{itemize}
+% \item Key boxed text here.
+% \item Key boxed text here.
+% \item Key boxed text here.
+% \end{itemize}}
+
+\maketitle
+
+\section{Introduction}
+
+\subsection{Extending prediction to prescription needs causality}
+
+Prediction models have long been used in medicine, as with risk
+score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
+While these have historically been simple models on simple data, this is
+changing with progress in machine learning and richer medical data
+\cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
+\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
+patient records
+\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
+notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
+Complex data is difficult to control and model, but these models are
+validated by verifying the accuracy of the prediction on left-out data
+\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
+Crucial to the clinical adoption of a model predicting a health outcome
+is that it ``can support decisions about patient care''
+\cite{wyatt1995commentary}. Precision medicine is about
+guiding decisions: \emph{eg} will an individual benefit from an intervention such as surgery
+\cite{fontana2019can}? An estimate of the effect of the treatment can be
+obtained by contrasting model predictions with and without the treatment,
+but statistical validity requires causal inference
+\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.
+
+% - Causal inference is difficult
+% - Beyond the dream of assessing effectiveness and safety in real-word
+%   practice \cite{black1996we} include off-label drug usage 
+%   \cite{radley2006off} (which may lead to drug repurposing)
+% - Plug in that outcome modeling is complementary to RCTs: observational
+%   data gives weaker evidence than trials, but on both data outcome
+%   modeling opens the door to individual.
+% - Epidemiology has focused on propensity-score methods
+% - Principle causal inference with outcome models is possible
+% - It brings the promises of \emph{individualized treatment effects} --a
+%   goal related to capturing heterogeneity and thus CATE--
+% - It also performs well on causal-inference competition, although they
+%   focused on the ATE
+
+% The plethora of approaches approaches give different causal-inference
+% estimates => need for guideline on causal model selection.
+
+Indeed, concluding on the effect of a treatment is a difficult
+causal-inference task, as it can be easily compromised by confounding:
+spurious associations between treatment
+allocation and baseline health, \eg~only prescribing a drug to mild cases
+\cite{hernan_causal_2020,vanderweele2019principles}.
+Predictive modeling bridges to causal inference theory under the name of
+\emph{outcome models} (or G-computation,
+G-formula \cite{robins_role_1986}, Q-model
+\cite{snowden_implementation_2011}, conditional mean regression
+\cite{wendling_comparing_2018}).
+% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
+% alleviate this concern. Yet most machine-learning models are trained on
+% \emph{observational} data, close to real-word practice
+% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
+Medical statistics and epidemiology have mostly used other
+causal-inference methods, modeling treatment assignment
+with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
+the benefit of going
+beyond average effects, estimating individualized or conditional average
+treatment effects (CATE), central to precision medicine.
+%
+For this purpose, such methods are also invaluable on randomized trials
+\cite{su2018random,lamont2018identification,hoogland2021tutorial}.
+
+% Recent empirical results \cite{wendling_comparing_2018,dorie_automated_2019}
+% show benefits of outcome modeling to estimate average treatment effects.
+
+%\idea{There is a variety of models for g-estimation}
+% There is an explosion of outcome modeling or machine learning methods, either 
+% developed for
+% For instance, many deep-learning methods have been developed for medical
+% image analysis \cite{shen2017deep,monshi2020deep}.
+Outcome-modeling methods, even when specifically designed for causal
+inference, are numerous: Bayesian Additive Regression Trees
+\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
+\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
+\cite{powers_methods_2018}, causal multivariate adaptive regression
+splines \cite{powers_methods_2018}, random forests
+\cite{wager_estimation_2018, athey_generalized_2019},
+Meta-learners \cite{kunzel_metalearners_2019}, R-learners
+\cite{nie_quasioracle_2017}, Doubly robust estimation
+\cite{chernozhukov_double_2018}...
+%\idea{The variety of estimators calls for model selection procedures}
+The wide variety of methods raises the problem
+of selecting between different estimators based on the data at hand. %TODO: might be removed since repeated two sentences below
+%
+Indeed, estimates of treatment effects can vary markedly across different
+predictive models \cite{fang2019applying,dorie_automated_2019} (illustration in
+Appendix \ref{apd:toy_example:acic_2016_ate_variability}).
+
+Given complex health data, which predictive model is to be most trusted to yield
+valid causal estimates needed to motivate individual treatment decisions?
+As no single machine-learning method performs best across all data sets, there is a
+pressing need for clear guidelines to select outcome models for causal
+inference.
+
+\paragraph{Objectives and structure of the paper}
+
+We study \textit{model selection procedures} in practical settings,
+without theoretical assumptions often made in statistical literature such
+as \textit{infinite} data or \textit{well-specified} models (Appendix
+\ref{apd:prior_work}). Asymptotic causal-inference theory recommends complex
+risks, but a practical question is whether model-selection procedures, that rely
+on data split, can estimate these risks reliably enough. Indeed, these
+risks come with
+more quantities to estimate, which may bring additional variance, leading to
+worse model selection.
+
+We first illustrate the problem of causal model selection. Then we anchor causal model selection in the \emph{potential outcome}
+framework and details the causal risks and model-selection procedure. We then
+rewrite the so-called $R\text{-risk}$ as a reweighted version of mean squared
+difference between the true and estimated individualized treatment effect.
+Finally, we conduct a thorough empirical study comparing the different metrics
+on diverse datasets, using a family of simulations and real health
+data, going beyond prior work limited to specific simulation settings
+\cite{schuler_comparison_2018,alaa_validating_2019} (Appendix
+\ref{apd:prior_work}).
+
+% not the place for results
+% Results outline how to best select outcome models for causal
+% inference with an adapted
+% cross-validation to estimate the so-called $R\text{-risk}$.
+% This risk compensates for systematic
+% differences between treated and non-treated individuals using
+% two \emph{nuisance} models,
+% themselves estimated from data and thus imperfect; yet these
+% imperfections do not undermine the $R\text{-risk}$.
+
+\begin{figure}[h!]
     % XXX: for submission
-%\documentclass[10pt,twocolumn]{article}
-\documentclass[11pt]{article}
-%\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
-
-%\usepackage[a4paper, left=15mm, right=15mm, top=20mm,]{geometry}
-% XXX: for submission
-\usepackage[a4paper, left=25mm, right=25mm, top=20mm,]{geometry}
-% declared functions and additional packages
-\usepackage{caption}
-\usepackage{subcaption}
-\usepackage{threeparttable}
-\usepackage{times}
-\usepackage{hyperref}
-\def\mycitecolor{green!50!black}
-\def\mylinkcolor{red!60!black}
-\hypersetup{ colorlinks=true, linkcolor=\mylinkcolor, citecolor=\mycitecolor }
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage{booktabs}
-\usepackage{amsmath}
-\usepackage{amssymb} %mathbb packages
-\usepackage{siunitx}
-\usepackage{algorithm, algorithmicx, algpseudocode}
-\usepackage{multirow}
-\usepackage{appendix}
-\usepackage[sorting=nyt,style=numeric,natbib=true,doi=false,isbn=false,url=false,eprint=false, maxnames=999, maxcitenames=2]{biblatex} 
-% To cite with superscript
-\let\cite=\supercite
-\DeclareNameAlias{default}{last-first}
-\usepackage{placeins}
-% associated arxiv class
-%\usepackage{arxiv}
+    \centering\begin{minipage}{.6\linewidth}
+        {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
+                causal inference}}
+
+        \setlength{\unitlength}{.1\linewidth}
+        \begin{picture}(10,5.2)(0,0)
+            \put(0, 0){\includegraphics[width=1\linewidth]{toy_random_forest_high_R2_high_tau_risk.pdf}}%
+            \put(7.05,4.4){\vector(0,1){.47}}
+            \put(7.05,4.4){\vector(0,-1){.55}}
+            \put(7.1,4.5){\scriptsize\sffamily treatment effect}
+            \put(7.1,4.15){\small $\tau(x)$}
+            \put(5.2,3.6){\small $\hat{\tau}(x)$}
+            \put(5.15,3.6){\vector(0,1){.29}}
+            \put(5.15,3.6){\vector(0,-1){.15}}
+        \end{picture}
+
+        {\sffamily\footnotesize\scalebox{1.03}{b) Linear model, worse average
+            prediction but better causal inference}}
+
+        \hfill%
+        \includegraphics[width=1\linewidth]{toy_tlinear_model_small_R2_small_tau_risk.pdf}%
+    \end{minipage}
+
+    \caption[The best predictor may not estimate best causal
+        effects]{\textbf{Illustration}: a) a random-forest predictor
+        with high performance for standard prediction (high $R^2$) but that
+        yields poor causal estimates (large error between true effect $\tau$ and
+        estimated $\hat{\tau}$), b) a linear predictor with smaller
+        prediction performance leading to better causal estimation. \\[1ex]
+        Selecting the predictor with the smallest error to the individual
+        treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$
+        --the $\tau\text{-risk}$, eq.\,\ref{eq:tau_risk} -- would lead to
+        the best causal estimates; however computing this error is not
+        feasible: it requires access to unknown quantities:
+        $\tau(x)$. \\[1ex]
+        While the random forest fits the data better than the linear model, it
+        gives worse causal inference because its error is inhomogeneous between
+        treated and untreated. The $R^2$ score does not capture this
+        inhomogeneity.
+        \label{fig:toy_example}
+    }
+\end{figure}
+
+\subsection{Illustration: the best predictor may not estimate best causal
+    effects}%
+
+Using a predictor to reason on causal effects relies on contrasting the
+prediction of the outcome for a given individual with and without the treatment.
+%
+Given various predictors of the outcome, which one should we use?
+%
+Standard predictive modeling or machine-learning practice selects the predictor
+that minimizes the expected error on the outcome
+\cite{poldrack2020establishment,varoquaux2022evaluating}. However, this
+predictor may not be the best model to reason about causal effects of an
+intervention as Figure \ref{fig:toy_example} illustrates. Consider the
+probability $Y$ of an undesirable outcome (\eg death), a binary treatment $A \in
+    \{0, 1\}$, and a covariate $X \in \mathbb R$ summarizing the patient health
+status (\eg the Charlson index \cite{charlson_new_1987}). We simulate a
+treatment beneficial (decreases mortality) for patients with high Charlson
+scores (bad health status) but with little effect for patients in good condition
+(low Charlson scores).
+
+Figure \ref{fig:toy_example}a shows a random forest predictor with a
+counter-intuitive behavior: it predicts well on average the outcome (as measured
+by a regression $R^2$ score) but perform poorly to estimate causal quantities:
+the average treatment effect $\tau$ (as visible via the error $|\tau -
+    \hat{\tau}|$) or the conditional average treatment effect (the error
+$\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$, called CATE).
+%
+On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
+smaller $R^2$ score but better causal inference.%
+
+The problem is that causal estimation requires controlling
+an error on both treated and non-treated outcome for the same individual:
+the observed outcome, and the non-observed \emph{counterfactual} one.
+The linear model is misspecified --the outcome functions are not
+linear--, leading to poor $R^2$; but it interpolates better to regions
+where there are few untreated individuals --high Charlson score-- and
+thus gives better causal estimates. Conversely, the random forest puts
+weaker assumptions on the data, thus has higher $R^2$ score but is biased
+by the treated population in the poor-overlap region, leading
+to bad causal estimates.
+
+This toy example illustrates that the classic minimum Mean Square Error
+criterion is not suited to choosing a model among candidate
+estimators for causal inference.
+
+% A natural way to select a predictive model for causal inference would be
+
+% an error measure between a causal quantity such as the CATE and models' estimate. But such error is
+% not a ``feasible'' risk: it cannot be computed solely from observed data
+% and requires oracle knowledge.
+
 \section{Methods}\label{sec:framework}
@@
                 The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
-                \citet{schuler_comparison_2018}.
+                \cite{schuler_comparison_2018}.
                 \label{tab:evaluation_metrics}}
@@
 \begin{figure}[h!]
-    % XXX: for submission
+
     \centering\begin{minipage}{.6\linewidth}
-        \includegraphics[width=\linewidth]{images/estimation_procedure_causal_selection_procedure.png}
-    \end{minipage}
-    \caption{Estimation procedure for causal model
-        selection.}\label{problem:estimation_procedure:figure}
-\end{figure}
-
-\subsection{R-risk as reweighted oracle metric}\label{theory:r_risk_rewrite}
-
-The $R\text{-risk}$ can be rewritten as a rebalanced $\tau \text{-risk}$.
-
-% We now one feasible risks, $\mu \text{-risk}_{IPW}$ and the
-% $R\text{-risk}$ to the oracle $\tau\text{-risk}$. Both results make
-% explicit the role of overlap for the performances of causal risks.
-
-This rewriting involves reweighted residuals: for each potential outcome, $a \in
-    \{0; 1\}$, the variance conditionally on $x$ is \cite{shalit_estimating_2017}:
-\begin{equation*}\label{eq:residuals}
-    \sigma_{y}^{2}(x ; a) \overset{\text{def}}{=}
-    \int_{y}\left(y-\mu_{a}(x)\right)^{2} p(y \mid x=x ; A=a) \, d y
-\end{equation*}
-Integrating over the population, we get the Bayes squared error:
-$\sigma^2_{B}(a) = \int_{\mathcal X} \sigma_y^2(x;a) p(x)dx$
-and its propensity weighted version:
-$\tilde{\sigma}^2_{B}(a) = \int_{\mathcal X}\sigma_y^2(x;a)\,  p(x;
-    a)\,dx$. In case of a purely deterministic link between the
-covariates, the treatment, and the outcome, these residual terms are null.
-
-%\md{Lower bound: I think that there is no lower bound of the tau-risk with the
-%mu-risk. That means that we can have a mu risk of 0 wheras the taurisk is
-%non-null. Todo for myself: develop an example with overfitted 1NN on the
-%observed data.}
-
-\begin{proposition}[$R \text{-risk}$ as reweighted
-        $\tau\text{-risk}$]\label{theory:prop:r_risk_rewrite} Given an outcome model
-    $f$, its $R\text{-risk}$ appears as weighted version of its $\tau\text{-risk}$
-    (Proof in \ref{apd:proofs:r_risk_rewrite}):
-    \begin{align}
-        R\text{-risk}^*(f) & = \int_{x} e(x)\big(1-e(x)\big)\big(\tau(x)-\tau_ {f}(x)\big)^{2} p(x) d x \nonumber \\
-                           & \quad\; + \tilde{\sigma}_B^2(1) + \tilde{\sigma}_B^{2}(0)
-    \end{align}
-\end{proposition}
-
-The $R \text{-risk}$ targets the oracle at the cost of an overlap re-weighting
-and the addition of the reweighted Bayes residuals, which are independent of
-$f$. In good overlap regions the weights $e(x) \big(1-e(x) \big)$ are close to
-$\frac{1}{4}$, hence the $R \text{-risk}$ is close to the desired gold-standard
-$\tau \text{-risk}$. For randomized control trials, this weight is constant
-making the $R\text{-risk}$ particularly suited for exploring
-heterogeneity (Appendix \ref{apd:theory:special_cases})
-
-\setlength{\columnsep}{0.5cm}
-\usepackage{wrapfig}
-%\usepackage{wraptable}
-\usepackage{authblk}% for author affiliations
-\usepackage[dvipsnames]{xcolor}
-\usepackage{setspace} % double spaced required by journal
-\doublespacing
-
-\usepackage[%
-    font={footnotesize,sf},
-    labelfont=bf,
-    format=hang,    
-    format=plain,
-    margin=0pt,
-]{caption}
-
-% TODO commands
-\usepackage{snaptodo}
-\setlength{\marginparwidth}{2.3cm}
-\setlength{\marginparsep}{.1cm}
-\snaptodoset{block rise=2em}
-\snaptodoset{margin block/.style={font=\tiny}}
-
-\newcommand{\gv}[1]{\snaptodo[margin block/.append style=green!50!black]{\sloppy\textbf{GV}: #1}}
-
-% Uncomment below to erase all comments
-%\renewcommand{\snaptodo}[2][1]{\relax}
-
-% XXX: for submission
-\usepackage{setspace}
-
-\usepackage{comment}
-\setcounter{secnumdepth}{0}
-
-%\setlength{\marginparwidth}{0.5cm}
-%\setlength\textwidth{177.8mm}
-
-% maths commands
-\newcommand{\indep}{\perp \!\!\! \perp}
-\newcommand{\eg}{\emph{e.g.}}
-\DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
-\usepackage{mathtools}
-\newcommand\myeq{\stackrel{\mathclap{\text{def}}}{=}}
-
-\newtheorem{definition}{Definition}
-\newtheorem{condition}{Condition}
-\newtheorem{property}{Property}
-\newtheorem{remark}{Remark}
-\newtheorem{notation}{Notation}
-\newtheorem{proposition*}{Proposition}
-\newtheorem{proposition}{Proposition}
-\newtheorem{lemma}{Lemma}
-\newtheorem{theorem}{Theorem}
-\newtheorem{assumption}{Assumption}
-\newtheorem{proof}{Proof}
-%
-\def\keywordname{{\bfseries \emph{Keywords}}}%
-\def\keywords#1{\par\addvspace\medskipamount{\rightskip=0pt plus1cm
-\def\and{\ifhmode\unskip\nobreak\fi\ $\cdot$
-}\noindent\keywordname\enspace\ignorespaces#1\par}}
-
-\title{How to select predictive models for decision-making or causal inference?}
-\addbibresource{biblio.bib}
-
-\author[1,2,*]{Matthieu Doutreligne, PhD}
-\author[1]{Gaël Varoquaux, PhD}
-
-\affil[1]{Inria, Soda, Saclay, France}
-\affil[2]{Mission Data, Haute Autorité de Santé, Saint-Denis, France}
-\affil[*]{Corresponding author: matt.dout@gmail.com}
-
-%\raggedbottom
-
-\begin{document}
-
-%\corres{*\email{matthieu.doutreligne@inria.fr}}
-
-\onecolumn
-\maketitle
-\begin{abstract}\noindent
-    \textbf{Background:}
-    We investigate which procedure selects the predictive model most
-    trustworthy to reason on the effect of an intervention and support
-    decision making.\\
-    %As predictive models --\eg from machine learning-- give likely
-    %outcomes, they may be used to, a
-    %causal-inference task. The increasing complexity of health data has opened
-    %the door to a plethora of models, but also the Pandora box of model
-    %selection: which of these models yield the most valid causal estimates? 
-    %\textbf{Methods:}
-    We study such a variety of model selection procedures in practical settings:
-    finite samples settings and without theoretical assumption of
-    \emph{well-specified} models. Beyond standard cross-validation or
-    internal validation procedures, we also study elaborate causal risks.
-    These build proxies of the causal error using ``nuisance''
-    re-weighting to compute it on the observed data. We evaluate whether
-    empirically estimated nuisances, which are necessarily noisy,
-    add noise to model selection. We compare different metrics for
-    causal model selection in an extensive empirical
-    study based on a simulation and three healthcare datasets
-    based on real covariates. \\
-    \textbf{Results:} Among all metrics, the mean squared error,
-    classically used to evaluate predictive modes,
-    is worse. Re-weighting it with propensity score
-    does not bring much improvements. The
-    $R\text{-risk}$, which uses as nuisances a model of
-    mean outcome and propensity scores, leads to the best performances.
-    Nuisance corrections are best estimated with flexible estimators such
-    as a super learner.
-    \\
-    \textbf{Conclusions:} When predictive models are used to
-    reason on the effect of an intervention, they must be evaluated
-    with different procedures than standard predictive
-    settings; using the $R\text{-risk}$ from causal inference.
-\end{abstract}
-\keywords{Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning, }
-\clearpage
-
-% XXX: for submission
-%\twocolumn
-% XXX: for submission (double spaced required)
-%\setstretch{1.1}
-\doublespacing
-\section{Introduction}\label{sec:intro}
-
-\subsection{Extending prediction to prescription needs causality}
-
-Prediction models have long been used in medicine, as with risk
-score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
-While these have historically been simple models on simple data, this is
-changing with progress in machine learning and richer medical data
-\cite{beam2018big,rajkomar2019machine}. Health predictions can now integrate medical images
-\cite{khojaste2022deep,zhang2019radiological,yala2019deep,shen2019deep,nassif2022breast},
-patient records
-\cite{mooney2018bigdata,desai2020comparison,simon2018predicting} or clinical
-notes \cite{horng2017creating,wang2020prediction,spasic2020clinical}.
-Complex data is difficult to control and model, but these models are
-validated by verifying the accuracy of the prediction on left-out data
-\cite{altman2009prognosis,poldrack2020establishment,varoquaux2022evaluating}.
-Crucial to the clinical adoption of a model predicting a health outcome
-is that it ``can support decisions about patient care''
-\cite{wyatt1995commentary}. Precision medicine is about
-guiding decisions: \emph{eg} will an individual benefit from an intervention such as surgery
-\cite{fontana2019can}? An estimate of the effect of the treatment can be
-obtained by contrasting model predictions with and without the treatment,
-but statistical validity requires causal inference
-\cite{snowden_implementation_2011,sperrin2019explicit,blakely2020reflection}.
-
-% - Causal inference is difficult
-% - Beyond the dream of assessing effectiveness and safety in real-word
-%   practice \cite{black1996we} include off-label drug usage 
-%   \cite{radley2006off} (which may lead to drug repurposing)
-% - Plug in that outcome modeling is complementary to RCTs: observational
-%   data gives weaker evidence than trials, but on both data outcome
-%   modeling opens the door to individual.
-% - Epidemiology has focused on propensity-score methods
-% - Principle causal inference with outcome models is possible
-% - It brings the promises of \emph{individualized treatment effects} --a
-%   goal related to capturing heterogeneity and thus CATE--
-% - It also performs well on causal-inference competition, although they
-%   focused on the ATE
-
-% The plethora of approaches approaches give different causal-inference
-% estimates => need for guideline on causal model selection.
-
-Indeed, concluding on the effect of a treatment is a difficult
-causal-inference task, as it can be easily compromised by confounding:
-spurious associations between treatment
-allocation and baseline health, \eg~only prescribing a drug to mild cases
-\cite{hernan_causal_2020,vanderweele2019principles}.
-Predictive modeling bridges to causal inference theory under the name of
-\emph{outcome models} (or G-computation,
-G-formula \cite{robins_role_1986}, Q-model
-\cite{snowden_implementation_2011}, conditional mean regression
-\cite{wendling_comparing_2018}).
-% Controlled allocation of treatment, as in randomized controlled trials (RCTs),
-% alleviate this concern. Yet most machine-learning models are trained on
-% \emph{observational} data, close to real-word practice
-% \cite{black1996we,hernan_methods_2021} but challenging for causal inference.
-Medical statistics and epidemiology have mostly used other
-causal-inference methods, modeling treatment assignment
-with propensity scores \cite{rosenbaum_central_1983,austin_moving_2015,casucci2018estimating,grose_use_2020}. Outcome modeling brings
-the benefit of going
-beyond average effects, estimating individualized or conditional average
-treatment effects (CATE), central to precision medicine.
-%
-For this purpose, such methods are also invaluable on randomized trials
-\cite{su2018random,lamont2018identification,hoogland2021tutorial}.
-
-% Recent empirical results \cite{wendling_comparing_2018,dorie_automated_2019}
-% show benefits of outcome modeling to estimate average treatment effects.
-
-%\idea{There is a variety of models for g-estimation}
-% There is an explosion of outcome modeling or machine learning methods, either 
-% developed for
-% For instance, many deep-learning methods have been developed for medical
-% image analysis \cite{shen2017deep,monshi2020deep}.
-Outcome-modeling methods, even when specifically designed for causal
-inference, are numerous: Bayesian Additive Regression Trees
-\cite{hill_bayesian_2011}, Targeted Maximum Likelihood Estimation
-\cite{laan_targeted_2011,schuler_targeted_2017}, causal boosting
-\cite{powers_methods_2018}, causal multivariate adaptive regression
-splines \cite{powers_methods_2018}, random forests
-\cite{wager_estimation_2018, athey_generalized_2019},
-Meta-learners \cite{kunzel_metalearners_2019}, R-learners
-\cite{nie_quasioracle_2017}, Doubly robust estimation
-\cite{chernozhukov_double_2018}...
-%\idea{The variety of estimators calls for model selection procedures}
-The wide variety of methods raises the problem
-of selecting between different estimators based on the data at hand. %TODO: might be removed since repeated two sentences below
-%
-Indeed, estimates of treatment effects can vary markedly across different
-predictive models \cite{fang2019applying,dorie_automated_2019} (illustration in
-Appendix \ref{apd:toy_example:acic_2016_ate_variability}).
-
-Given complex health data, which predictive model is to be most trusted to yield
-valid causal estimates needed to motivate individual treatment decisions?
-As no single machine-learning method performs best across all data sets, there is a
-pressing need for clear guidelines to select outcome models for causal
-inference.
-
-\paragraph{Objectives and structure of the paper}
-
-We study \textit{model selection procedures} in practical settings,
-without theoretical assumptions often made in statistical literature such
-as \textit{infinite} data or \textit{well-specified} models (Appendix
-\ref{apd:prior_work}). Asymptotic causal-inference theory recommends complex
-risks, but a practical question is whether model-selection procedures, that rely
-on data split, can estimate these risks reliably enough. Indeed, these
-risks come with
-more quantities to estimate, which may bring additional variance, leading to
-worse model selection.
-
-We first illustrate the problem of causal model selection. Then we anchor causal model selection in the \emph{potential outcome}
-framework and details the causal risks and model-selection procedure. We then
-rewrite the so-called $R\text{-risk}$ as a reweighted version of mean squared
-difference between the true and estimated individualized treatment effect.
-Finally, we conduct a thorough empirical study comparing the different metrics
-on diverse datasets, using a family of simulations and real health
-data, going beyond prior work limited to specific simulation settings
-\cite{schuler_comparison_2018,alaa_validating_2019} (Appendix
-\ref{apd:prior_work}).
-
-% not the place for results
-% Results outline how to best select outcome models for causal
-% inference with an adapted
-% cross-validation to estimate the so-called $R\text{-risk}$.
-% This risk compensates for systematic
-% differences between treated and non-treated individuals using
-% two \emph{nuisance} models,
-% themselves estimated from data and thus imperfect; yet these
-% imperfections do not undermine the $R\text{-risk}$.
-
-\begin{figure}[h!]
-    % XXX: for submission
-    \centering\begin{minipage}{.6\linewidth}
-        {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
-                causal inference}}
-
-        \setlength{\unitlength}{.1\linewidth}
-        \begin{picture}(10,5.2)(0,0)
-            \put(0, 0){\includegraphics[width=1\linewidth]{images/toy_random_forest_high_R2_high_tau_risk.pdf}}%
-            \put(7.05,4.4){\vector(0,1){.47}}
-            \put(7.05,4.4){\vector(0,-1){.55}}
-            \put(7.1,4.5){\scriptsize\sffamily treatment effect}
-            \put(7.1,4.15){\small $\tau(x)$}
-            \put(5.2,3.6){\small $\hat{\tau}(x)$}
-            \put(5.15,3.6){\vector(0,1){.29}}
-            \put(5.15,3.6){\vector(0,-1){.15}}
-        \end{picture}
-
-        {\sffamily\footnotesize\scalebox{1.03}{b) Linear model, worse average
-            prediction but better causal inference}}
-
-        \hfill%
-        \includegraphics[width=1\linewidth]{images/toy_tlinear_model_small_R2_small_tau_risk.pdf}%
-    \end{minipage}
-
-    \caption[The best predictor may not estimate best causal
-        effects]{\textbf{Illustration}: a) a random-forest predictor
-        with high performance for standard prediction (high $R^2$) but that
-        yields poor causal estimates (large error between true effect $\tau$ and
-        estimated $\hat{\tau}$), b) a linear predictor with smaller
-        prediction performance leading to better causal estimation. \\[1ex]
-        Selecting the predictor with the smallest error to the individual
-        treatment effect $\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$
-        --the $\tau\text{-risk}$, eq.\,\ref{eq:tau_risk} -- would lead to
-        the best causal estimates; however computing this error is not
-        feasible: it requires access to unknown quantities:
-        $\tau(x)$. \\[1ex]
-        While the random forest fits the data better than the linear model, it
-        gives worse causal inference because its error is inhomogeneous between
-        treated and untreated. The $R^2$ score does not capture this
-        inhomogeneity.
-        \label{fig:toy_example}
-    }
-\end{figure}
-
-\subsection{Illustration: the best predictor may not estimate best causal
-    effects}%
-
-Using a predictor to reason on causal effects relies on contrasting the
-prediction of the outcome for a given individual with and without the treatment.
-%
-Given various predictors of the outcome, which one should we use?
-%
-Standard predictive modeling or machine-learning practice selects the predictor
-that minimizes the expected error on the outcome
-\cite{poldrack2020establishment,varoquaux2022evaluating}. However, this
-predictor may not be the best model to reason about causal effects of an
-intervention as Figure \ref{fig:toy_example} illustrates. Consider the
-probability $Y$ of an undesirable outcome (\eg death), a binary treatment $A \in
-    \{0, 1\}$, and a covariate $X \in \mathbb R$ summarizing the patient health
-status (\eg the Charlson index \cite{charlson_new_1987}). We simulate a
-treatment beneficial (decreases mortality) for patients with high Charlson
-scores (bad health status) but with little effect for patients in good condition
-(low Charlson scores).
-
-Figure \ref{fig:toy_example}a shows a random forest predictor with a
-counter-intuitive behavior: it predicts well on average the outcome (as measured
-by a regression $R^2$ score) but perform poorly to estimate causal quantities:
-the average treatment effect $\tau$ (as visible via the error $|\tau -
-    \hat{\tau}|$) or the conditional average treatment effect (the error
-$\mathbb{E}[(\tau(x) - \hat{\tau}(x))^2]$, called CATE).
-%
-On the contrary, Figure \ref{fig:toy_example}b shows a linear model with
-smaller $R^2$ score but better causal inference.%
-
-The problem is that causal estimation requires controlling
-an error on both treated and non-treated outcome for the same individual:
-the observed outcome, and the non-observed \emph{counterfactual} one.
-The linear model is misspecified --the outcome functions are not
-linear--, leading to poor $R^2$; but it interpolates better to regions
-where there are few untreated individuals --high Charlson score-- and
-thus gives better causal estimates. Conversely, the random forest puts
-weaker assumptions on the data, thus has higher $R^2$ score but is biased
-by the treated population in the poor-overlap region, leading
-to bad causal estimates.
-
-This toy example illustrates that the classic minimum Mean Square Error
-criterion is not suited to choosing a model among candidate
-estimators for causal inference.
-
-% A natural way to select a predictive model for causal inference would be
-
-% an error measure between a causal quantity such as the CATE and models' estimate. But such error is
-% not a ``feasible'' risk: it cannot be computed solely from observed data
-% and requires oracle knowledge.
-
-\section{Methods}\label{sec:framework}
-\subsection{Neyman-Rubin Potential Outcomes framework}%
-\label{sec:neyman_rubin}%
-
-\paragraph{Settings}
-
-The Neyman-Rubin Potential Outcomes framework
-\cite{naimi2023defining,imbens_causal_2015} enables statistical reasoning on
-causal treatment effects: Given an outcome $Y \in \mathbb R$ (\eg mortality risk
-or hospitalization length), function of a binary treatment $A \in \mathcal{A} =
-    \{0, 1\}$ (\eg~a medical procedure), and baseline
-covariates $X \in \mathcal{X} \subset \mathbb{R}^d$, we observe the factual
-distribution, $O = (Y(A), X, A) \sim \mathcal D = \mathbb P(y, x, a)$. However,
-we want to model the existence of potential observations (unobserved ie.~counterfactual) that correspond to a different treatment. Thus we want
-quantities on the counterfactual distribution $O^{*} = (Y(1), Y(0), X, A) \sim
-    \mathcal D^{*} = \mathbb P(y(1), y(0), x, a)$.
-
-Popular quantities of interest (estimands) are:
-at the population level, the
-Average Treatment Effect
-\begin{flalign*}
-    \text{ATE} &  &
-    \tau \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal D^*}[Y(1) - Y(0)];
-               &  &
-\end{flalign*}
-at the individual level, to model heterogeneity, the Conditional Average Treatment Effect
-\begin{flalign*}
-    \text{CATE} &  &
-    \tau (x) \myeq \; \mathbb{E}_{Y(1),Y(0) \sim \mathcal{D}^\star}[Y(1) - Y(0) | X=x].
-                &  &
-\end{flalign*}
-
-\paragraph{Causal assumptions}
-
-A given data needs to meet a few assumptions to enable identifying
-causal estimands \cite{rubin_causal_2005}. The usual
-\emph{strong ignorability} assumptions are  (details in
-\ref{apd:causal_assumptions}): \emph{1)}
-\emph{unconfoundedness} \mbox{$\{Y(0),
-        Y(1) \} \indep A | X$}, \emph{~2)} \emph{strong overlap} ie. every patient has a
-strictly positive probability to receive each treatment, \emph{3)}
-\emph{consistency}, and \emph{4)} \emph{generalization}.
-
-\paragraph{Estimating treatment effects with outcome models -- g-computation \cite{robins_new_1986}}\label{subsec:estimators}
-
-Should we know the two expected outcomes for a given $X$,
-we could compute the
-difference between them, which gives the causal effect of the treatment.
-%
-These two expected outcomes can be estimated from observed data:
-the consistency \ref{assumption:consistency} and unconfoundedness
-\ref{assumption:ignorability} assumptions imply the following equality:
-\begin{equation}\label{eq:mu_identification}
-    \mathbb E_{Y(a) \sim \mathcal{D^{\star}}} [Y(a)|X=x] = \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
-\end{equation}
-On the left, the expectation is taken on the counterfactual unobserved
-distribution. On the right, the expectation is taken on the factual observed
-distribution conditionally on the treatment. For the rest of the
-paper, the expectations will always be taken on the factual observed
-distribution $\mathcal{D}$. This identification leads to outcome based estimators (ie.
-g-computation estimators \cite{snowden_implementation_2011}):
-\begin{eqnarray}
-    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x]
-    \notag
-    \\
-    = &\mathbb E_{Y \sim \mathcal{D}}[Y|A=1] - \mathbb E_{Y \sim \mathcal{D}}[Y| A=0]
-    \label{eq:tau_population}
-\end{eqnarray}
-This equation builds on two quantities: the conditional expectancy
-of the outcome given the covariates and either
-treatment or no no treatment, called \emph{response function}:
-\begin{flalign*}
-    \text{Response function}
-     &  &
-    \mu_{a}(x) \myeq \; \mathbb E_{Y \sim \mathcal{D}} [Y|X=x, A=a]
-     &  &
-\end{flalign*}
-
-Given a sample of data and the oracle response functions $\mu_0, \mu_1$, the
-finite sum version of \autoref{eq:tau_population} leads to an
-estimator of the ATE written:
-\begin{equation}
-    \hat \tau = \frac{1}{n} \biggl(\sum_{i=1}^n \mu_{1}(x_i) - \mu_{0}(x_i) \biggr)
-    \label{eq:ate_estimate}
-\end{equation}
-This estimator is an oracle \emph{finite sum} estimator by opposition to the
-population expression of $\tau$, $\mathbb{E}[\mu_{1}(x_i) - \mu_{0}(x_i)]
-$,
-which involves an expectation taken on the full
-distribution $\mathcal D$, which is observable but requires infinite data. For
-each estimator $\ell$ taking an expectation over $\mathcal D$, we use the symbol
-$\hat \ell$ to note its finite sum version.
-
-Similarly to the ATE, at the individual level, the CATE:
-\begin{equation}
-    \tau(x) = \mu_{1}(x) - \mu_{0}(x)
-    \label{eq:cate_estimate}
-\end{equation}
-
-\paragraph{Robinson decomposition}
-The \emph{R-decomposition}
-of the outcome model\cite{robinson_rootnconsistent_1988} plays an important role:
-introducing two quantities, the conditional mean outcome
-and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
-\begin{flalign}
-    \text{Conditional mean outcome} &                    & m(x) \myeq \; & \mathbb E_{Y \sim
-    \mathcal{D}} [Y|X=x]            &                    &
-    \label{def:m}
-    \\
-    \text{Propensity score}         &                    &
-    e(x) \myeq \;                   & \mathbb P[A=1|X=x]
-    \label{def:propensity_score}
-\end{flalign}
-the outcome can be written
-\begin{flalign}\label{eq:r_decomposition}
-    \text{R-decomposition}            &  & y(a) = m(x) + \big( a - e(x) \big)
-    \tau(x) + \varepsilon(x; a)\notag &  &
-    \\\ &&\text{with}\quad \mathbb E[\varepsilon(X; A)|X, A] = 0&&
-\end{flalign}
-$m$ and $e$ are often called
-\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown.
-
-%As noted by \cite{johansson2022generalization}, the machine learning
-%community often referred to the CATE by ITE, the Individual Treatment Effect.
-%From a purely causal point of view, the ITE is uniquely defined for each
-%individual and might not be accessible: $ITE(x_i) = Y_i(1) -  Y_i(0)$. On the
-%contrary, the CATE can always be derived by taking conditional expectancies. It
-%is the expected effect of the treatment in the region of the covariate space
-%around X. %Too cultivate
-
-\begin{table*}[bt!]
-    \makebox[\textwidth]{
-        \begin{threeparttable}[b]
-            \caption{Review of causal risks
-                ---
-                The $R\text{-risk}^*$ is called $\tau \text{-risk}_R$ in
-                \citet{schuler_comparison_2018}.
-                \label{tab:evaluation_metrics}}
-            \centering
-            \begin{tabular}{llr}
-                \toprule
-                Risk                                                                                           & Equation
-                                                                                                               & Reference                                                                                                                                           \\
-                \midrule
-                $\tau\text{-risk}=\text{MSE}(\tau(X), \tau_f(X))$                                              & $\mathbb E_{X\sim
-                            p(X)}[(\tau(X) - \hat \tau_f(X))^2] $
-                                                                                                               & Eq.
-\ref{eq:tau_risk} \cite{hill_bayesian_2011}                                                                                                     \\
-                $\mu\text{-risk} = \text{MSE}(Y, f(X))$                                                        & $\mathbb{E}_{(Y, X, A)
-                        \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
-                                                                                                               & Def.
-\ref{def:mu_risk} \cite{schuler_comparison_2018}                                                                                               \\
-                $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
-                        \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
-                (Y-f(X ; A))^2 \right]$                                                                        & Def.
-                \ref{def:mu_ipw_risk} \cite{vanderlaan_unified_2003}                                                                                                                                                                                                 \\
-                $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
-                Def. \ref{def:tau_ipw_risk} \cite{wager_estimation_2018}
-                \\
-                $U\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A) \sim \mathcal D}  \big[
-                \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
-                Def. \ref{def:u_risk} \cite{nie_quasioracle_2017}
-                \\
-                $R\text{-risk}^*$                                                                              & $\mathbb{E}_{(Y, X, A)
-                        \sim \mathcal D} \big[\big(\left(Y-m\left(X\right)\right)
-                -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
-                Def. \ref{def:r_risk} \cite{nie_quasioracle_2017}
-                \\
-                \bottomrule
-            \end{tabular}
-        \end{threeparttable}
-    }
-\end{table*}
-
-\subsection{Model-selection risks, oracle and feasible}\label{sec:problem:model_selection}
-
-\paragraph{Causal model selection}\label{sec:problem:causal_selection}
-
-We formalize model selection for causal estimation. Thanks to the g-formula
-identification (\autoref{eq:mu_identification}), a given outcome model $f: \mathcal X
-    \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
-domain knowledge-- induces feasible estimates of the ATE and CATE (eqs
-\ref{eq:ate_estimate} and \ref{eq:cate_estimate}), $\hat \tau_{f}$ and $\hat \tau_{f}(x)$.
-%
-Let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
-a family of such estimators. Our goal is to select the best candidate in this
-family for the observed dataset $O$ using a risk
-$\ell$:
-\begin{equation}
-    f^*_{\ell} = \argmin_{f \in \mathcal{F}} \ell(f, O)
-    \label{eq:causal_model_selection}
-\end{equation}
-
-We now detail possible risks $\ell$, risks useful for causal
-model selection, and how to compute them.
-
-\paragraph{The $\tau\text{-risk}$: an oracle error risk}\label{paragraph:oracle_metrics}
-%
-As we would like to target the CATE, the following
-evaluation risk is natural (also called PEHE \cite{schulam_reliable_2017, hill_bayesian_2011}):
-\begin{flalign}\label{eq:tau_risk}
-    \tau\text{-risk}(f) \myeq \mathbb E_{X\sim p(X)}[(\tau(X) - \hat \tau_f(X))^2]
-\end{flalign}
-
-Given observed data from $p(X)$, the expectation is computed with a
-finite sum, as in eq.\,\ref{eq:ate_estimate}, to give an estimated
-value $\widehat{\tau\text{-risk}}(f)$.
-%
-However this risk is not feasible as the oracles $\tau(x)$ are
-not accessible with the observed data $(Y, X, A) \sim \mathcal D$.
-
-\paragraph{Feasible error risks}\label{paragraph:feasible_metrics} Table
-\ref{tab:evaluation_metrics} lists \emph{feasible} risks (Detailed in Appendix
-\ref{def:feasible_risks}), based on the prediction error of the outcome model
-and \emph{observable} quantities. These observable, called nuisances are $e$
---propensity score, eq \ref{def:propensity_score}-- and $m$ --conditional mean
-outcome, eq \ref{def:m}. We give the definitions as \textit{semi-oracles},
-function of the true unknown nuisances, but later instantiate them with
-estimated nuisances, noted $\big(\check e, \check m \big)$. Semi-oracles risks
-are superscripted with the $^{\star}$ symbol.
-
-\subsection{Estimation and model selection procedure}\label{problem:estimation_procedure}
-
-Causal model selection (eq
-\ref{eq:causal_model_selection}) may involve estimating various quantities
-from the observed data: the outcome model $f$, its induced risk as
-introduce in the previous section, and possibly nuisances required by the
-risk.
-Given a dataset with $N$ samples, we split out a train and a test sets
-$(\mathcal{T}, \mathcal{S})$. We
-fit each candidate estimator $f \in \mathcal{F}$ on $\mathcal{T}$. We also fit
-the nuisance models $(\check e, \check m)$ on the train set
-$\mathcal{T}$, setting hyperparameters by a nested
-cross-validation before fitting the nuisance estimators with these parameters
-on the full train set. Causal quantities are then computed by applying the fitted  candidates
-estimators $f \in \mathcal{F}$ on the test set $\mathcal{S}$. Finally, we
-compute the model-selection metrics for
-each candidate model on the test set. This procedure is described in Algorithm
-\ref{problem:estimation_procedure:algo} and Figure
-\ref{problem:estimation_procedure:figure}.
-
-% As extreme inverse propensity weights induce high variance, clipping can be
-% useful for numerical stability
-% \cite{swaminathan_counterfactual_2015, ionides_truncated_2008}.
-
-\begin{algorithm}[!htbp]
-    \caption{Model selection procedure}\label{problem:estimation_procedure:algo} {%
-        Given train and test sets $(\mathcal{T}, \mathcal{S}) \sim \mathcal{D}$,
-        a candidate estimators $f$, a causal
-        metrics $\ell$:
-        \begin{enumerate}
-            \item Prefit: Learn estimators for unknown nuisance quantities $(\check e,\,\check m)$ on the training set $\mathcal{T}$
-            \item Fit: learn $\hat f(\cdot, a)$ on
-                  $\mathcal T$
-            \item Model selection: $\forall{x} \in \mathcal{S}$ predict
-                  $\big(\hat f(x, 1), \hat f(x, 0)\big)$ and evaluate the estimator storing the metric value: $\ell(f, \mathcal S)$ -- possibly
-                  function of $\check e$ and $\check m$
-                  %\item Metric evaluation: return the oracle evaluation metrics
-                  %evaluated on $\mathcal{S}$: $\big(\widehat{\tau\text{-risk}}(\hat
-                  %f^*_{\ell}); \widehat{\ell}_{ATE}(\hat f^*_{\ell}) \big)$
-        \end{enumerate}
-
-    }
-\end{algorithm}
-
-\begin{figure}[h!]
-    % XXX: for submission
-    \centering\begin{minipage}{.6\linewidth}
-        \includegraphics[width=\linewidth]{images/estimation_procedure_causal_selection_procedure.png}
+        \includegraphics[width=\linewidth]{estimation_procedure_causal_selection_procedure.png}
     \end{minipage}
@@
     \centering\begin{minipage}{.75\linewidth}
-        \includegraphics[width=\linewidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
+        \includegraphics[width=\linewidth]{_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
     \end{minipage}
@@
     \centering\begin{minipage}{.75\linewidth}
-        \includegraphics[width=\linewidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
+        \includegraphics[width=\linewidth]{_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
     \end{minipage}
@@
     \centering\begin{minipage}{.75\linewidth}
-        \includegraphics[width=\linewidth]{images/_3_procedure_r_risk_only_3datasets_twocols.pdf}
+        \includegraphics[width=\linewidth]{_3_procedure_r_risk_only_3datasets_twocols.pdf}
     \end{minipage}
@@
     \centering\begin{minipage}{.75\linewidth}
-        \includegraphics[width=\linewidth]{images/_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
+        \includegraphics[width=\linewidth]{_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
     \end{minipage}
@@
 
-\section*{Acknowledgments}
-
-We acknowledge fruitful discussions with Bénédicte Colnet.
-
-\subsection*{Financial disclosure}
-
-None reported.
-
-\subsection*{Conflict of interest}
-
-The authors declare no potential conflict of interests.
-
-%\if@Mybibrefstyle%
-\printbibliography
-%\fi if using njdnatbib.sty \bibliography{bibliography}%
-
 \clearpage
@@
 
+%%%%%%%%%%%%%%
+
 \begin{appendices}
 
-    \part*{Supplementary materials:\\
-        How to select predictive models for decision-making or causal inference? }
+    %\part*{Supplementary materials:\\How to select predictive models for decision-making or causal inference? }
 
@@
         \centering
-        \includegraphics[width=0.95\linewidth]{images/2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
+        \includegraphics[width=0.95\linewidth]{2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
         \caption{\textbf{Different outcome models lead to different
@@
     Using eight simulations setups from \cite{powers_methods_2018}, where
-    the oracle CATE is known, \citet{schuler_comparison_2018} compare four
+    the oracle CATE is known, \cite{schuler_comparison_2018} compare four
     causal risks, concluding that for CATE estimation the best
@@
 
-    By contrast, \citet{johansson2022generalization} studies causal estimation
-    without assuming that estimators are well specified. They derive an upper bound
+    By contrast, without assuming that estimators are well specified, there exists an upper bound
     on the oracle error to the CATE ($\tau\text{-risk}$) that involves the error on
     the outcome and the similarity of the distributions of treated and control
-    patients. However, they use this upper bound for model optimization,
+    patients \cite{johansson2022generalization}. However, they use this upper bound for model optimization,
     and do not give insights on model selection. In addition, for hyperparameter
@@
     between $\tau\text{-risk}$ and $\mu\text{-risk}_{IPW}$ coming from heterogeneous
-    errors between populations as shown experimentally in \citet{schuler_comparison_2018} and our
+    errors between populations as shown experimentally in \cite{schuler_comparison_2018} and our
     results below.
@@
             \caption{\textbf{Uncalibrated classifiers}}
-            \includegraphics[width=\linewidth]{images/overlap_measure_diff_oracle_ntv_to_n_tv_calibration=False_vs_oracle_n_tv.pdf}
+            \includegraphics[width=\linewidth]{overlap_measure_diff_oracle_ntv_to_n_tv_calibration=False_vs_oracle_n_tv.pdf}
         \end{subfigure}
@@
                     classifiers}}
-            \includegraphics[width=\linewidth]{images/overlap_measure_diff_oracle_ntv_to_n_tv_calibration=True_vs_oracle_n_tv.pdf}
+            \includegraphics[width=\linewidth]{overlap_measure_diff_oracle_ntv_to_n_tv_calibration=True_vs_oracle_n_tv.pdf}
         \end{subfigure}
@@
             \cite{dorie_automated_2019}}\label{apd:overlap:penalized_overlap}
-        \includegraphics[width=0.7\linewidth]{images/overlap_measure_acic_2016_recovery_overlap_setup.pdf}
+        \includegraphics[width=0.7\linewidth]{overlap_measure_acic_2016_recovery_overlap_setup.pdf}
     \end{figure}
@@
             datasets)}\label{apd:overlap:caussim:mmd_vs_ntv}
-        \includegraphics[width=0.7\linewidth]{images/overlap_measure_caussim_transformed_mmd_vs_ntv.pdf}
+        \includegraphics[width=0.7\linewidth]{overlap_measure_caussim_transformed_mmd_vs_ntv.pdf}
     \end{figure}
@@
             \caption{\textbf{Caussim}}
-            \includegraphics[width=\textwidth]{images/overlap_measure_caussim_max_ipw_vs_ntv.pdf}
+            \includegraphics[width=\textwidth]{overlap_measure_caussim_max_ipw_vs_ntv.pdf}
             \label{apd:caussim:max_ipw_vs_ntv}
@@
             \caption{\textbf{ACIC 2016}}
-            \includegraphics[width=\textwidth]{images/overlap_measure_acic_2016_max_ipw_vs_ntv.pdf}
+            \includegraphics[width=\textwidth]{overlap_measure_acic_2016_max_ipw_vs_ntv.pdf}
             \label{apd:acic_2016:ntv_vs_max_ipw}
@@
             \caption{\textbf{ACIC 2018}}
-            \includegraphics[width=\textwidth]{images/overlap_measure_acic_2018_max_ipw_vs_ntv.pdf}
+            \includegraphics[width=\textwidth]{overlap_measure_acic_2018_max_ipw_vs_ntv.pdf}
             \label{apd:acic_2018:ntv_vs_max_ipw}
@@
             \caption{\textbf{TWINS}}
-            \includegraphics[width=\textwidth]{images/overlap_measure_twins_max_ipw_vs_ntv.pdf}
+            \includegraphics[width=\textwidth]{overlap_measure_twins_max_ipw_vs_ntv.pdf}
             \label{apd:twins:ntv_vs_max_ipw}
@@
         \centering
-        \includegraphics[width=\linewidth]{images/caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
+        \includegraphics[width=\linewidth]{caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
 
@@
 
-    \begin{description}
+    \begin{itemize}
         \item[ACIC 2018] \cite{dorie_automated_2019}:  The initial intervention was a child’s birth
@@
             set by the original outcomes.
-    \end{description}
+    \end{itemize}
 
@@
         \centering
-        \includegraphics[width=\linewidth]{images/_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset.pdf}
+        \includegraphics[width=\linewidth]{_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset.pdf}
         %\hfill
@@
             \caption{\textbf{Caussim}}
-            \includegraphics[width=\textwidth]{images/kendalls_tau_caussim__nuisance_non_linear__candidates_ridge__overlap_01-247.pdf}
+            \includegraphics[width=\textwidth]{kendalls_tau_caussim__nuisance_non_linear__candidates_ridge__overlap_01-247.pdf}
             \label{fig:ranking_agreement_w_tau_risk_caussim}
@@
             \caption{\textbf{ACIC 2016}}
-            \includegraphics[width=\textwidth]{images/kendalls_tau_acic_2016__nuisance_non_linear__candidates_hist_gradient_boosting__dgp_1-77__rs_1-10.pdf}
+            \includegraphics[width=\textwidth]{kendalls_tau_acic_2016__nuisance_non_linear__candidates_hist_gradient_boosting__dgp_1-77__rs_1-10.pdf}
             \label{fig:ranking_agreement_tau_risk_acic_2016}
@@
             \centering
-            \includegraphics[width=1.5\textwidth]{images/legend_metrics.pdf}
+            \includegraphics[width=1.5\textwidth]{legend_metrics.pdf}
         \end{subfigure}
@@
             \caption{\textbf{ACIC 2018}}
-            \includegraphics[width=\textwidth]{images/kendalls_tau_acic_2018__nuisance_stacking__candidates_hist_gradient_boosting__tset_50.pdf}
+            \includegraphics[width=\textwidth]{kendalls_tau_acic_2018__nuisance_stacking__candidates_hist_gradient_boosting__tset_50.pdf}
             \label{fig:ranking_agreement_w_tau_risk_acic_2018}
@@
             \caption{\textbf{TWINS}}
-            \includegraphics[width=\textwidth]{images/kendalls_tau_twins__nuisance_stacking__candidates_hist_gradient_boosting__tset_50__overlap_11-296__rs_0-9_noise10.pdf}
+            \includegraphics[width=\textwidth]{kendalls_tau_twins__nuisance_stacking__candidates_hist_gradient_boosting__tset_50__overlap_11-296__rs_0-9_noise10.pdf}
             \label{fig:ranking_agreement_tau_risk_twins}
@@
         %   \centering
-        %   \includegraphics[width=0.5\textwidth]{images/legend_metrics_horizontal.pdf}
+        %   \includegraphics[width=0.5\textwidth]{legend_metrics_horizontal.pdf}
         % \end{subfigure}
@@
             \caption{\textbf{Caussim}}
-            \includegraphics[width=\textwidth]{images/normalized_bias_tau_risk_to_best_method_caussim__nuisance_non_linear__candidates_ridge__overlap_01-247.pdf}
+            \includegraphics[width=\textwidth]{normalized_bias_tau_risk_to_best_method_caussim__nuisance_non_linear__candidates_ridge__overlap_01-247.pdf}
             \label{fig:normalized_bias_tau_risk_to_best_method_caussim}
@@
             \caption{\textbf{ACIC 2016}}
-            \includegraphics[width=\textwidth]{images/normalized_bias_tau_risk_to_best_method_acic_2016__nuisance_non_linear__candidates_hist_gradient_boosting__dgp_1-77__rs_1-10.pdf}
+            \includegraphics[width=\textwidth]{normalized_bias_tau_risk_to_best_method_acic_2016__nuisance_non_linear__candidates_hist_gradient_boosting__dgp_1-77__rs_1-10.pdf}
             \label{fig:normalized_bias_tau_risk_to_best_method_acic_2016}
@@
             \caption{\textbf{ACIC 2018}}
-            \includegraphics[width=\textwidth]{images/normalized_bias_tau_risk_to_best_method_acic_2018__nuisance_stacking__candidates_hist_gradient_boosting__tset_50.pdf}
+            \includegraphics[width=\textwidth]{normalized_bias_tau_risk_to_best_method_acic_2018__nuisance_stacking__candidates_hist_gradient_boosting__tset_50.pdf}
             \label{fig:normalized_bias_tau_risk_to_best_method_acic_2018}
@@
             \caption{\textbf{TWINS}}
-            \includegraphics[width=\textwidth]{images/normalized_bias_tau_risk_to_best_method_twins__nuisance_stacking__candidates_hist_gradient_boosting__tset_50__overlap_11-296__rs_0-9_noise10.pdf}
+            \includegraphics[width=\textwidth]{normalized_bias_tau_risk_to_best_method_twins__nuisance_stacking__candidates_hist_gradient_boosting__tset_50__overlap_11-296__rs_0-9_noise10.pdf}
             \label{fig:normalized_bias_tau_risk_to_best_method_twins}
@@
             \caption{\textbf{Caussim}}
-            \includegraphics[width=1.15\textwidth]{images/_3_procedure_Caussim_N5000__ref_metric_oracle_r_risk_training_procedure.pdf}
+            \includegraphics[width=1.15\textwidth]{_3_procedure_Caussim_N5000__ref_metric_oracle_r_risk_training_procedure.pdf}
             \label{fig:experiments:procedures_comparison:caussim}
@@
             \caption{\textbf{ACIC 2016}}
-            \includegraphics[width=1\textwidth]{images/_3_procedure_ACIC2016_N4802__ref_metric_oracle_r_risk_training_procedure.pdf}
+            \includegraphics[width=1\textwidth]{_3_procedure_ACIC2016_N4802__ref_metric_oracle_r_risk_training_procedure.pdf}
             \label{fig:experiments:procedures_comparison:acic_2016}
@@
             \caption{\textbf{Twins}}
-            \includegraphics[width=1\textwidth]{images/_3_procedure_Twins_N11984__ref_metric_oracle_r_risk_training_procedure.pdf}
+            \includegraphics[width=1\textwidth]{_3_procedure_Twins_N11984__ref_metric_oracle_r_risk_training_procedure.pdf}
             \label{fig:experiments:procedures_comparison:twins}
@@
         \centering
-        \includegraphics[width=\textwidth]{images/_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset.pdf}
+        \includegraphics[width=\textwidth]{_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset.pdf}
         \hfill
@@
             \caption{\textbf{Twins}}
-            \includegraphics[width=1\textwidth]{images/_4_nuisance_models_Twins_N11984__ref_metric_oracle_r_risk_nuisance_models.pdf}
+            \includegraphics[width=1\textwidth]{_4_nuisance_models_Twins_N11984__ref_metric_oracle_r_risk_nuisance_models.pdf}
             \label{fig:experiments:nuisance_comparison:twins}
@@
             \caption{\textbf{Twins downsampled}}
-            \includegraphics[width=1\textwidth]{images/_4_nuisance_models_Twinsdownsampled_N4794__ref_metric_oracle_r_risk_nuisance_models.pdf}
+            \includegraphics[width=1\textwidth]{_4_nuisance_models_Twinsdownsampled_N4794__ref_metric_oracle_r_risk_nuisance_models.pdf}
             \label{fig:experiments:nuisance_comparison:twins_ds}
@@
             \caption{\textbf{Caussim}}
-            \includegraphics[width=1.15\textwidth]{images/_4_nuisance_models_Caussim_N5000__ref_metric_oracle_r_risk_nuisance_models.pdf}
+            \includegraphics[width=1.15\textwidth]{_4_nuisance_models_Caussim_N5000__ref_metric_oracle_r_risk_nuisance_models.pdf}
             \label{fig:experiments:nuisance_comparison:caussim}
@@
             \caption{\textbf{ACIC 2016}}
-            \includegraphics[width=1\textwidth]{images/_4_nuisance_models_ACIC2016_N4802__ref_metric_oracle_r_risk_nuisance_models.pdf}
+            \includegraphics[width=1\textwidth]{_4_nuisance_models_ACIC2016_N4802__ref_metric_oracle_r_risk_nuisance_models.pdf}
             \label{fig:experiments:nuisance_comparison:acic_2016}
@@
         \centering
-        \includegraphics[width=\textwidth]{images/_4_nuisance_models_r_risk_only_twins_nuisances.pdf}
+        \includegraphics[width=\textwidth]{_4_nuisance_models_r_risk_only_twins_nuisances.pdf}
         \hfill
@@
             process.}\label {apd:results:fig:seed_effect}
-        \includegraphics[width=\linewidth]{images/caussim_seed_effect.pdf}
+        \includegraphics[width=\linewidth]{caussim_seed_effect.pdf}
     \end{figure}
@@
             \centerline{\textbf{a) CATE estimation error}}
-            \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16_twocols.pdf}
+            \includegraphics[width=\linewidth]{_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16_twocols.pdf}
         \end{minipage}
@@
             \centerline{\textbf{b) ATE estimation error}}
-            \includegraphics[width=\linewidth]{images/_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16_twocols.pdf}
+            \includegraphics[width=\linewidth]{_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16_twocols.pdf}
         \end{minipage}%
@@
     \cite{causalevaluations} or K=3 in econML \cite{econml}.
-    Likewise, for ATE estimation, \citet{chernozhukov_double_2018}
+    Likewise, for ATE estimation, \cite{chernozhukov_double_2018}
     introduce doubly-robust machine learning,
@@
     subsequent works use doubly robust ML with varying choices
-    of K: \citet{loiseau_external_2022} use K=3, \citet{gao_assessment_2021} use
+    of K: \cite{loiseau_external_2022} use K=3, \cite{gao_assessment_2021} use
     K=2. In the econML implementation, K is set to 3 \cite{econml}.
-    \citet{naimi2021challenges} evaluate various machine-learning approaches
+    \cite{naimi2021challenges} evaluate various machine-learning approaches
     --including R-learners-- using K=5 and 10, drawing inspiration from the
@@
     Causal model selection has been much less discussed. The only study that
-    we are aware of, \citet{schuler_comparison_2018}, use a different data
+    we are aware of, \cite{schuler_comparison_2018}, use a different data
     split: a 2-folds train/test procedure,
@@
 
-    % \paragraph{Simulations: naive reweighting of the $R \text{-risk}$}
-
-    % Applying a naive reweighting, $w(x, a)=\frac{1}{e(x)\big(1-e(x) \big )}$ to the
-    % $R \text{-risk}$ to recover the $\tau \text{-risk}$ in the first part of
-    % \ref{theory:prop:r_risk_rewrite} makes the residuals explode in case of noise as
-    % shown in Figure \ref{apd:simu_noised}.
-
-    % \begin{figure}[htbp]
-    %   \centering
-    %   \caption{Caussim simulations (500 repetitions):
-    %   $R\text{-risk}_{IPW^*_{naive}}$ (in green) is exploding}\label
-    %   {apd:simu_noised}
-    %   \includegraphics[width=0.7\linewidth]{images/noised_0.001_simu_best_normalized_abs_bias_ate_vs_mmd.pdf}
-    % \end{figure}
 \end{appendices}
 
+\section{Competing interests}
+No competing interest is declared.
+
+\section{Author contributions statement}
+
+M.D. conceived and conducted the experiments, M.D. and G.V. analysed the results. M.D. and G.V. wrote and reviewed the manuscript.
+
+\section{Acknowledgments}
+
+We acknowledge fruitful discussions with Bénédicte Colnet.
+
+\bibliography{biblio}
+%\printbibliography
+
 \end{document}

### 2024-06-02 how_to_select.tex
format figure for new format

@@
 \usepackage{placeins} % for FloatBarrier command to prevent a figure to pass some point.
+\usepackage{titlesec} % for title formatting
+\titleformat*{\subsection}{\color{jnlclr}\bfseries} % force colored subsection font for paragraphs
+\titleformat*{\paragraph}{\bfseries} % force bold font for paragraphs
 
@@
     %selection: which of these models yield the most valid causal estimates? 
-    %\textbf{Methods:}
+    \textbf{Methods:}
     We study such a variety of model selection procedures in practical settings:
@@
     % XXX: for submission
-    \centering\begin{minipage}{.6\linewidth}
-        {\sffamily\footnotesize\scalebox{1.03}{a) Random forest, good average prediction but bad
+    \centering\begin{minipage}{.95\linewidth}
+        {\sffamily\footnotesize\scalebox{0.95}{a) Random forest, good average prediction but bad
                 causal inference}}
@@
 
-        {\sffamily\footnotesize\scalebox{1.03}{b) Linear model, worse average
+        {\sffamily\footnotesize\scalebox{0.95}{b) Linear model, worse average
             prediction but better causal inference}}
@@
 
-    \centering\begin{minipage}{.6\linewidth}
+    \centering\begin{minipage}{.95\linewidth}
         \includegraphics[width=\linewidth]{estimation_procedure_causal_selection_procedure.png}
@@
     % XXX: for submission
-    \centering\begin{minipage}{.75\linewidth}
+    \centering\begin{minipage}{\linewidth}
         \includegraphics[width=\linewidth]{_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
@@
     % XXX: for submission
-    \centering\begin{minipage}{.75\linewidth}
+    \centering\begin{minipage}{\linewidth}
         \includegraphics[width=\linewidth]{_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
@@
     % XXX: for submission
-    \centering\begin{minipage}{.75\linewidth}
+    \centering\begin{minipage}{\linewidth}
         \includegraphics[width=\linewidth]{_3_procedure_r_risk_only_3datasets_twocols.pdf}
@@
     % XXX: for submission
-    \centering\begin{minipage}{.75\linewidth}
+    \centering\begin{minipage}{\linewidth}
         \includegraphics[width=\linewidth]{_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
@@
         \centering
-        \includegraphics[width=\linewidth]{caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
+        \includegraphics[width=0.8\linewidth]{caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
 

### 2024-06-02 how_to_select.tex
iterr gigascience format

@@
 %%%CONTEMPORARY%%%
-\documentclass[unnumsec,webpdf,contemporary,large]{oup-authoring-template}%
-%\documentclass[unnumsec,webpdf,contemporary,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
-%\documentclass[unnumsec,webpdf,contemporary,medium]{oup-authoring-template}
-%\documentclass[unnumsec,webpdf,contemporary,small]{oup-authoring-template}
-
-%%%MODERN%%%
-%\documentclass[unnumsec,webpdf,modern,large]{oup-authoring-template}
-%\documentclass[unnumsec,webpdf,modern,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
-%\documentclass[unnumsec,webpdf,modern,medium]{oup-authoring-template}
-%\documentclass[unnumsec,webpdf,modern,small]{oup-authoring-template}
-
-%%%TRADITIONAL%%%
-%\documentclass[unnumsec,webpdf,traditional,large]{oup-authoring-template}
-%\documentclass[unnumsec,webpdf,traditional,large,namedate]{oup-authoring-template}% uncomment this line for author year citations and comment the above
-%\documentclass[unnumsec,namedate,webpdf,traditional,medium]{oup-authoring-template}
-%\documentclass[namedate,webpdf,traditional,small]{oup-authoring-template}
-
-%\onecolumn % for one column layouts
-
-%\usepackage{showframe}
+\documentclass[a4paper,num-refs]{oup-contemporary}%
+
+%%% Journal toggle; only specific options recognised.
+%%% (Only "gigascience" and "general" are implemented now. Support for other journals is planned.)
+\journal{gigascience}
+
+\usepackage{graphicx}
+\usepackage{siunitx}
+
+%%% Flushend: You can add this package to automatically balance the final page, but if things go awry (e.g.
+section contents appearing out-of-order or entire blocks or paragraphs are coloured), remove it!
+% \usepackage{flushend}
 
@@
 
-\theoremstyle{thmstyleone}%
+% added commands and packages
+\newtheorem{proof}{Proof}
 \newtheorem{theorem}{Theorem}%  meant for continuous numbers
-%%\newtheorem{theorem}{Theorem}[section]% meant for sectionwise numbers
-%% optional argument [theorem] produces theorem numbering sequence instead of independent numbers for Proposition
 \newtheorem{proposition}[theorem]{Proposition}%
-%%\newtheorem{proposition}{Proposition}% to get separate numbers for theorem and proposition etc.
-\theoremstyle{thmstyletwo}%
 \newtheorem{example}{Example}%
 \newtheorem{remark}{Remark}%
-\theoremstyle{thmstylethree}%
 \newtheorem{definition}{Definition}
-
-% added commands and packages
 \newtheorem{assumption}{Assumption}
@@
 \newcommand{\eg}{\emph{e.g.}}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{algorithm, algorithmicx, algpseudocode}
 \DeclareMathOperator*{\argmin}{argmin} \def\mycitecolor{green!50!black}
@@
 \usepackage{booktabs} % For formal tables (eg. bottomrules command)
+\usepackage{multirow} % For multirow in tables
 \usepackage{threeparttable} % For table 
@@
 \usepackage{titlesec} % for title formatting
-\titleformat*{\subsection}{\color{jnlclr}\bfseries} % force colored subsection font for paragraphs
-\titleformat*{\paragraph}{\bfseries} % force bold font for paragraphs
+%\titleformat*{\subsection}{\color{jnlclr}\bfseries} % force colored subsection font for paragraphs
+%\titleformat*{\paragraph}{\bfseries} % force bold font for paragraphs
 
@@
 
+\title{How to select predictive models for decision-making or causal inference?}
+
+%%% Use the \authfn to add symbols for additional footnotes, if any. 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
+\author[1,\authfn{1}]{Matthieu Doutreligne}
+\author[2]{Gaël Varoquaux}
+
+\affil[1]{Soda, Inria Saclay, France}
+\affil[2]{Mission Data, Haute Autorité de Santé, France}
+
+%%% Author Notes
+\authnote{\authfn{1}matt.dout@gmail.com; matthieu.doutreligne@inria.fr}
+
+%%% Paper category
+\papercat{Paper}
+
+%%% "Short" author for running page header
+\runningauthor{Doutreligne et al.}
+
+%%% Should only be set by an editor
+\jvolume{00}
+\jnumber{0}
+\jyear{2024}
+
 \begin{document}
 
-\journaltitle{GigaScience}
-\DOI{DOI HERE}
-\copyrightyear{2024}
-\pubyear{2024}
-\access{Advance Access Publication Date: Day Month Year}
-\appnotes{Paper}
-
-\firstpage{1}
-
-%\subtitle{Subject Section}
-
-\title[How to select predictive models for decision-making or causal inference?]{How to select predictive models for decision-making or causal inference?}
-
-\author[1, 2 $\ast$]{Matthieu Doutreligne}
-\author[1]{Gaël Varoquaux}
-
-\authormark{Author Name et al.}
-
-\address[1]{\orgdiv{Soda}, \orgname{Inria, Saclay}, \orgaddress{\country{France}}}
-\address[2]{\orgdiv{Mission Data}, \orgname{Haute Autorité de Santé}, \orgaddress{\country{France}}}
-
-\corresp[$\ast$]{Corresponding author. \href{email:matt.dout@gmail.com}{matt.dout@gmail.com}}
-
-\received{Date}{0}{Year}
-\revised{Date}{0}{Year}
-\accepted{Date}{0}{Year}
-
-%\editor{Associate Editor: Name}
-
-%\abstract{
-%\textbf{Motivation:} .\\
-%\textbf{Results:} .\\
-%\textbf{Availability:} .\\
-%\textbf{Contact:} \href{name@email.com}{name@email.com}\\
-%\textbf{Supplementary information:} Supplementary data are available at \textit{Journal Name}
-%online.}
-
-\abstract{\textbf{Background:}
+\begin{frontmatter}
+    \maketitle
+    \begin{abstract}
+        \textbf{Background:}
         We investigate which procedure selects the predictive model most
@@
         with different procedures than standard predictive
-    settings; using the $R\text{-risk}$ from causal inference.}
-\keywords{Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning, }
-
-% \boxedtext{
+        settings; using the $R\text{-risk}$ from causal inference.
+    \end{abstract}
+
+    \begin{keywords}
+        Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning
+    \end{keywords}
+
+\end{frontmatter}
+
+%%% Key points will be printed at top of second page
+% \begin{keypoints*}
 %     \begin{itemize}
-% \item Key boxed text here.
-% \item Key boxed text here.
-% \item Key boxed text here.
-% \end{itemize}}
-
-\maketitle
+%         \item This is the first point
+%         \item This is the second point
+%         \item One last point.
+%     \end{itemize}
+% \end{keypoints*}
 
@@
 (excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
-(\ref{apd:experiments:additional_results}).
+(\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}).
 
@@
         Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
-        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents results for all
+        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. Supplementary materials presents results for all
         metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
@@
 
+%\renewcommand\thefigure{S\arabic{figure}}
+%\renewcommand\thetable{S\arabic{table}}
+%\setcounter{figure}{0}
+%\setcounter{table}{0}
+
+%%%%%%%%%%%%%%
+
+\section{Availability of source code and requirements}
+
+Lists the following:
+\begin{itemize}
+    \item Project name: Caussim
+    \item Project home page: \url{https://github.com/soda-inria/caussim}
+    \item Operating system(s): Platform independent
+    \item Programming language: Python
+    \item License: BSD 3-Clause License
+\end{itemize}
+
 \clearpage
@@
 \onecolumn
-
-\renewcommand\thefigure{S\arabic{figure}}
-\renewcommand\thetable{S\arabic{table}}
-\setcounter{figure}{0}
-\setcounter{table}{0}
-
-%%%%%%%%%%%%%%
-
-\begin{appendices}
+\setcounter{secnumdepth}{2}
+
+\appendix
+\section{Additional files}
 
 %\part*{Supplementary materials:\\How to select predictive models for decision-making or causal inference? }
-
-    \setcounter{secnumdepth}{3}
+%\setcounter{secnumdepth}{1}
+%\renewcommand{\thesubsection}{\Alph{subsection}}
+%\setcounter{secnumdepth}{3}
 
@@
 
-    \section{Variability of ATE estimation on ACIC
+\subsection{Variability of ATE estimation on ACIC
     2016}\label{apd:toy_example:acic_2016_ate_variability}
@@
 
-    \section{Prior work : model selection for outcome modeling (g-computation)}\label{apd:prior_work}
+\subsection{Prior work : model selection for outcome modeling (g-computation)}\label{apd:prior_work}
 
@@
 
-    \section{Causal assumptions}\label{apd:causal_assumptions}
+\subsection{Causal assumptions}\label{apd:causal_assumptions}
 
@@
 
-    \section{Definitions of feasible risks}\label{def:feasible_risks}
+\subsection{Definitions of feasible risks}\label{def:feasible_risks}
 
@@
 
-    \section{Proofs: Links between feasible and oracle risks}\label{apd:proofs}
+\subsection{Proofs: Links between feasible and oracle risks}\label{apd:proofs}
 
@@
 
-    \subsection{Reformulation of the $R\text{-risk}$ as reweighted
+\subsubsection{Reformulation of the $R\text{-risk}$ as reweighted
     $\tau\text{-risk}$}\label{apd:proofs:r_risk_rewrite}
@@
 
-    \subsection{Interesting special cases}\label{apd:theory:special_cases}
+\subsubsection{Interesting special cases}\label{apd:theory:special_cases}
 
@@
 
-    \section{Measuring overlap}\label{apd:motivation_ntv}
+\subsection{Measuring overlap}\label{apd:motivation_ntv}
 
@@
 
-    \section{Experiments}
-
-    \subsection{Details on the data generation process}
+\subsection{Experiments}
+
+\subsubsection{Details on the data generation process}
 \label{apd:experiments:generation}
@@
 
-    \subsection{ Details on the semi-simulated datasets}\label{apd:experiments:datasets}
+\subsubsection{ Details on the semi-simulated datasets}\label{apd:experiments:datasets}
 
@@
 
-    \subsection{Model selection procedures}
+\subsubsection{Model selection procedures}
 
@@
 
-    \subsection{Additional Results}\label{apd:experiments:additional_results}
-
-    \paragraph{Definition of the Kendall's tau, $\kappa$}
+\subsubsection{Additional Results}
+
+\paragraph{Definition of the Kendall's tau, $\kappa$}\label{apd:experiments:additional_results}
 
@@
 
-    \section{Data split choices}\label{apd:results:data_split}
-
-    \subsection{Use 90\% of the data to estimate outcome models, 10\% to
+\subsection{Data split choices}\label{apd:results:data_split}
+
+\subsubsection{Use 90\% of the data to estimate outcome models, 10\% to
     select them}
@@
 
-    \subsection{Heterogeneity in practices for data split}\label{apd:results:k_fold_choices}
+\subsubsection{Heterogeneity in practices for data split}\label{apd:results:k_fold_choices}
 
@@
 
-\end{appendices}
-
-\section{Competing interests}
+\section{Declaration}
+\subsection{Competing interests}
 No competing interest is declared.

### 2024-06-02 make_submission.sh

@@ -1,5 +1,5 @@
 #! /bin/bash
-rm -f how_to_select_causal_models.zip
-zip -r how_to_select_causal_models.zip images/ biblio.bib *.bbl how_to_select_causal_models.tex 
-pdftk how_to_select_causal_models.pdf cat 1-24 output main.pdf
-pdftk how_to_select_causal_models.pdf cat 25-49 output supplementary_materials.pdf
+rm -f how_to_select.zip
+zip -r how_to_select.zip images/ biblio.bib *.bbl how_to_select.tex how_to_select.pdf
+pdftk how_to_select.pdf cat 1-6 output main.pdf
+pdftk how_to_select.pdf cat 7-26 output supplementary_materials.pdf

### 2024-06-02 how_to_select.tex
rm zone_identifier figures

@@
     \centering
-    \includegraphics[width=0.8\linewidth]{caussim_example_rs_gaussian=8_rs_rotation=8_ntv=0.37_D=2_overlap=0.7_p_A=0.1.pdf}
+    \includegraphics[width=0.8\linewidth]{caussim_example.pdf}
 
@@
         \caption{\textbf{TWINS}}
-        \includegraphics[width=\textwidth]{normalized_bias_tau_risk_to_best_method_twins__nuisance_stacking__candidates_hist_gradient_boosting__tset_50__overlap_11-296__rs_0-9_noise10.pdf}
+        \includegraphics[width=\textwidth]{normalized_bias_tau_risk_to_best_method_twins__nuisance_stacking__candidates_hgb.pdf}
         \label{fig:normalized_bias_tau_risk_to_best_method_twins}

### 2024-06-14 how_to_select.tex
manuscript for gigascience: Rename images to match jounral <=60 chars

@@
 
-Prediction models have long been used in medicine, as with risk
+Prediction models have long been used in biomedical settings, as with risk
 score or prognostic models \cite{moons2009prognosis,steyerberg2019clinical}.
@@
 %
-For this purpose, such methods are also invaluable on randomized trials
+For this purpose, such methods are also invaluable for randomized trials
 \cite{su2018random,lamont2018identification,hoogland2021tutorial}.
@@
     \centering\begin{minipage}{\linewidth}
-        \includegraphics[width=\linewidth]{_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset_feasible_only.pdf}
+        \includegraphics[width=\linewidth]{_1_r_risk_domination_mrisks_by_ds_feasible.pdf}
     \end{minipage}
@@
     \centering\begin{minipage}{\linewidth}
-        \includegraphics[width=\linewidth]{_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset_r_risk_only_twocols.pdf}
+        \includegraphics[width=\linewidth]{_2_overlap_influence_by_bin_k_by_ds_r_risk_2cols.pdf}
     \end{minipage}
@@
     \centering\begin{minipage}{\linewidth}
-        \includegraphics[width=\linewidth]{_4_nuisance_models_r_risk_only_3datasets_twocols.pdf}
+        \includegraphics[width=\linewidth]{_4_nuisance_models_r_risk_3ds_2cols.pdf}
     \end{minipage}
@@
 
+\section{Declaration}
+\subsection{Competing interests}
+No competing interest is declared.
+
+\section{Author contributions statement}
+
+M.D. conceived and conducted the experiments, M.D. and G.V. analyzed the results. M.D. and G.V. wrote and reviewed the manuscript.
+
+\section{Acknowledgments}
+
+We acknowledge fruitful discussions with Bénédicte Colnet.
+
 \clearpage
@@
     \centering
-    \includegraphics[width=0.95\linewidth]{2023-03-08-11-10-28_acic_2016_ate_heterogeneity.parquet_abs_bias_ylog_scale=True.pdf}%
+    \includegraphics[width=0.95\linewidth]{acic_2016_ate_hte.pdf}%
     \caption{\textbf{Different outcome models lead to different
@@
         \caption{\textbf{Uncalibrated classifiers}}
-        \includegraphics[width=\linewidth]{overlap_measure_diff_oracle_ntv_to_n_tv_calibration=False_vs_oracle_n_tv.pdf}
+        \includegraphics[width=\linewidth]{overlap_measure_diff_oracle_ntv_to_pred_ntv.pdf}
     \end{subfigure}
@@
                 classifiers}}
-        \includegraphics[width=\linewidth]{overlap_measure_diff_oracle_ntv_to_n_tv_calibration=True_vs_oracle_n_tv.pdf}
+        \includegraphics[width=\linewidth]{overlap_measure_diff_oracle_ntv_to_ntv_wcalibration.pdf}
     \end{subfigure}
@@
     \centering
-    \includegraphics[width=\linewidth]{_1_r_risk_domination_r_risk_domination__ref_metric_mean_risks_by_Dataset.pdf}
+    \includegraphics[width=\linewidth]{_1_r_risk_domination_mrisks_by_ds.pdf}
     %\hfill
@@
         \caption{\textbf{Caussim}}
-        \includegraphics[width=\textwidth]{kendalls_tau_caussim__nuisance_non_linear__candidates_ridge__overlap_01-247.pdf}
+        \includegraphics[width=\textwidth]{ktau_caussim_nuisance_nlin_models_ridge_ov1-247.pdf}
         \label{fig:ranking_agreement_w_tau_risk_caussim}
@@
         \caption{\textbf{ACIC 2016}}
-        \includegraphics[width=\textwidth]{kendalls_tau_acic_2016__nuisance_non_linear__candidates_hist_gradient_boosting__dgp_1-77__rs_1-10.pdf}
+        \includegraphics[width=\textwidth]{ktau_acic2016_nuisance_nlin_models_hgb_dgp1-77_rs1-10.pdf}
         \label{fig:ranking_agreement_tau_risk_acic_2016}
@@
         \caption{\textbf{ACIC 2018}}
-        \includegraphics[width=\textwidth]{kendalls_tau_acic_2018__nuisance_stacking__candidates_hist_gradient_boosting__tset_50.pdf}
+        \includegraphics[width=\textwidth]{ktau_acic2018_nuisance_stacking_models_hgb_tset_50.pdf}
         \label{fig:ranking_agreement_w_tau_risk_acic_2018}
@@
         \caption{\textbf{TWINS}}
-        \includegraphics[width=\textwidth]{kendalls_tau_twins__nuisance_stacking__candidates_hist_gradient_boosting__tset_50__overlap_11-296__rs_0-9_noise10.pdf}
+        \includegraphics[width=\textwidth]{ktau_twins_nuisance_stacking__models_hgb.pdf}
         \label{fig:ranking_agreement_tau_risk_twins}
@@
         \caption{\textbf{Caussim}}
-        \includegraphics[width=\textwidth]{normalized_bias_tau_risk_to_best_method_caussim__nuisance_non_linear__candidates_ridge__overlap_01-247.pdf}
+        \includegraphics[width=\textwidth]{norm_bias_trisk_tobest_caussim_nuis_nlin_models_ridge.pdf}
         \label{fig:normalized_bias_tau_risk_to_best_method_caussim}
@@
         \caption{\textbf{ACIC 2016}}
-        \includegraphics[width=\textwidth]{normalized_bias_tau_risk_to_best_method_acic_2016__nuisance_non_linear__candidates_hist_gradient_boosting__dgp_1-77__rs_1-10.pdf}
+        \includegraphics[width=\textwidth]{norm_bias_trisk_tobest_acic2016_nuis_nlin_models_hgb.pdf}
         \label{fig:normalized_bias_tau_risk_to_best_method_acic_2016}
@@
         \caption{\textbf{ACIC 2018}}
-        \includegraphics[width=\textwidth]{normalized_bias_tau_risk_to_best_method_acic_2018__nuisance_stacking__candidates_hist_gradient_boosting__tset_50.pdf}
+        \includegraphics[width=\textwidth]{norm_bias_trisk_tobest_acic2018_nuis_nlin_models_hgb.pdf}
         \label{fig:normalized_bias_tau_risk_to_best_method_acic_2018}
@@
         \caption{\textbf{TWINS}}
-        \includegraphics[width=\textwidth]{normalized_bias_tau_risk_to_best_method_twins__nuisance_stacking__candidates_hgb.pdf}
+        \includegraphics[width=\textwidth]{norm_bias_trisk_tobest_twins_nuis_stack_models_hgb.pdf}
         \label{fig:normalized_bias_tau_risk_to_best_method_twins}
@@
         \caption{\textbf{Caussim}}
-        \includegraphics[width=1.15\textwidth]{_3_procedure_Caussim_N5000__ref_metric_oracle_r_risk_training_procedure.pdf}
+        \includegraphics[width=1.15\textwidth]{_3_procedure_Caussim_N5000__ref_metric_oracle_r_risk.pdf}
         \label{fig:experiments:procedures_comparison:caussim}
@@
         \caption{\textbf{ACIC 2016}}
-        \includegraphics[width=1\textwidth]{_3_procedure_ACIC2016_N4802__ref_metric_oracle_r_risk_training_procedure.pdf}
+        \includegraphics[width=1\textwidth]{_3_procedure_ACIC2016_N4802__ref_oracle_r_risk.pdf}
         \label{fig:experiments:procedures_comparison:acic_2016}
@@
         \caption{\textbf{Twins}}
-        \includegraphics[width=1\textwidth]{_3_procedure_Twins_N11984__ref_metric_oracle_r_risk_training_procedure.pdf}
+        \includegraphics[width=1\textwidth]{_3_procedure_Twins_N11984__ref_oracle_r_risk.pdf}
         \label{fig:experiments:procedures_comparison:twins}
@@
     \centering
-    \includegraphics[width=\textwidth]{_2_overlap_influence_overlap_by_bin_comparaison_kendall_by_Dataset.pdf}
+    \includegraphics[width=\textwidth]{_2_overlap_influence_by_bin_k_by_ds.pdf}
     \hfill
@@
         \caption{\textbf{Twins}}
-        \includegraphics[width=1\textwidth]{_4_nuisance_models_Twins_N11984__ref_metric_oracle_r_risk_nuisance_models.pdf}
+        \includegraphics[width=1\textwidth]{_4_nuisance_models_Twins_N11984__ref_oracle_r_risk.pdf}
         \label{fig:experiments:nuisance_comparison:twins}
@@
         \caption{\textbf{Twins downsampled}}
-        \includegraphics[width=1\textwidth]{_4_nuisance_models_Twinsdownsampled_N4794__ref_metric_oracle_r_risk_nuisance_models.pdf}
+        \includegraphics[width=1\textwidth]{_4_nuisance_models_Twinssmall_N4794__ref_oracle_r_risk.pdf}
         \label{fig:experiments:nuisance_comparison:twins_ds}
@@
         \caption{\textbf{Caussim}}
-        \includegraphics[width=1.15\textwidth]{_4_nuisance_models_Caussim_N5000__ref_metric_oracle_r_risk_nuisance_models.pdf}
+        \includegraphics[width=1.15\textwidth]{_4_nuisance_models_Caussim_N5000__ref_oracle_r_risk.pdf}
         \label{fig:experiments:nuisance_comparison:caussim}
@@
         \caption{\textbf{ACIC 2016}}
-        \includegraphics[width=1\textwidth]{_4_nuisance_models_ACIC2016_N4802__ref_metric_oracle_r_risk_nuisance_models.pdf}
+        \includegraphics[width=1\textwidth]{_4_nuisance_models_ACIC2016_N4802__ref_oracle_r_risk.pdf}
         \label{fig:experiments:nuisance_comparison:acic_2016}
@@
         \centerline{\textbf{a) CATE estimation error}}
-        \includegraphics[width=\linewidth]{_5_train_size_evaluated_metric_r_risk__evaluation_validation_tau_risk__best_tau_norm__False_norm_False__acic16_twocols.pdf}
+        \includegraphics[width=\linewidth]{_5_train_size_r_risk__cate__acic16_2cols.pdf}
     \end{minipage}
@@
         \centerline{\textbf{b) ATE estimation error}}
-        \includegraphics[width=\linewidth]{_5_train_size_evaluated_metric_r_risk__evaluation_validation_test_abs_bias_ate__best_tau_norm__False_norm_True__acic16_twocols.pdf}
+        \includegraphics[width=\linewidth]{_5_train_size_r_risk__ate__acic16_2cols.pdf}
     \end{minipage}%
@@
 
-\section{Declaration}
-\subsection{Competing interests}
-No competing interest is declared.
-
-\section{Author contributions statement}
-
-M.D. conceived and conducted the experiments, M.D. and G.V. analysed the results. M.D. and G.V. wrote and reviewed the manuscript.
-
-\section{Acknowledgments}
-
-We acknowledge fruitful discussions with Bénédicte Colnet.
-
 \bibliography{biblio}

### 2024-06-14 images/replace_img_names.sh

@@ -1,0 +1,23 @@
+# Check if exactly three arguments are provided
+if [ "$#" -ne 2 ]; then
+    echo "Usage: $0 old_string new_string"
+    exit 1
+fi
+
+# Assign arguments to variables
+OLD_STRING=$1
+NEW_STRING=$2
+
+# Replace string in how_to_select.tex
+sed -i "s/$OLD_STRING/$NEW_STRING/g" ../how_to_select.tex
+
+# Check if the file exists in the images folder
+if [ -f "$OLD_STRING" ]; then
+    # Rename the file in the images folder
+    mv "$OLD_STRING" "$NEW_STRING"
+    echo "Renamed $OLD_STRING to $NEW_STRING"
+else
+    echo "File $OLD_STRING does not exist."
+fi
+
+echo "String replacement in how_to_select.tex complete."

### 2024-06-14 make_submission.sh

@@ -1,5 +1,0 @@
-#! /bin/bash
-rm -f how_to_select.zip
-zip -r how_to_select.zip images/ biblio.bib *.bbl how_to_select.tex how_to_select.pdf
-pdftk how_to_select.pdf cat 1-6 output main.pdf
-pdftk how_to_select.pdf cat 7-26 output supplementary_materials.pdf

### 2024-08-25 how_to_select_latex.tex
WIP: Reviewer1

--- 2024-06-14 how_to_select.tex
@@
 \begin{eqnarray}
-    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)|X=x]
+    \tau =& \mathbb E_{Y \sim \mathcal{D^{\star}}}[Y(1) - Y(0)]
     \notag
@@
 \end{eqnarray}
-This equation builds on two quantities: the conditional expectancy
-of the outcome given the covariates and either
-treatment or no no treatment, called \emph{response function}:
+This equation builds on the conditional expectation of the outcome given the treatment $\mathbb E_{\sim \mathcal{D}}[Y|A]$.
+Outcome based methods target this quantity conditionally on the covariates, called \emph{response function}:
 \begin{flalign*}
@@
 Given a sample of data and the oracle response functions $\mu_0, \mu_1$, the
-finite sum version of \autoref{eq:tau_population} leads to an
+finite sum version of \autoref{eq:tau_population} leads to an unbiased
 estimator of the ATE written:
@@
 $\hat \ell$ to note its finite sum version.
+In practice the response function can be estimated with one statistical model taking the treatment as a supplementary covariate $\hat \mu(x, a)$ (called S-learner) or by two separate model fitted on each population, treated or not $(\hat \mu_0(x), \hat \mu_1(x))$ (T-learner) \cite{kunzel_metalearners_2019}.
 
@@
 
+%TODO: not well located here. The reader does not know why we introduce this decomoositon her.
 \paragraph{Robinson decomposition}
 The \emph{R-decomposition}
-of the outcome model\cite{robinson_rootnconsistent_1988} plays an important role:
-introducing two quantities, the conditional mean outcome
-and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
+of the outcome model\cite{robinson_rootnconsistent_1988} links the counterfactual outcomes $y(a)$ to the CATE by introducing two quantities, the conditional mean outcome
+and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}).:
 \begin{flalign}
@@
 $m$ and $e$ are often called
-\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown.
+\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown and are estimated in practice thanks to statistical models.
 

### 2024-08-29 response_to_reviewer.md
WIP: response to reviewer

# Reviewer 1

I appreciate the approch and the level of detail that this paper offers on an issue that is, to say the least, a hot topic in bioinformatics and data science in general: the evaluation of the best 'error' model when choosing a causal model for predictive purposes. Indeed, there is much confusion on the subject and very often error models are applied without any real awareness of their potential and limitations.

That is why I think the paper can be of great educational value as well as research.

Of particular interest is the statement "the best predictor may not estimate best causal effects".
Central to the understanding of the paper is Figure 1 and the concept it seeks to explain. The figure clearly shows that if we rely on R^2, a predictor is preferred which on average is efficient but not on estimating causal quantities. The authors conceivably state that "The problem is that causal estimation requires controlling an error on both treated and non-treated outcome for the same individual.". Consequently, they report on estimator that attempt to consider this (in Section "Methods"). However, I cannot understand how this hint ("controlling error on both treated and non-treated sample (or patient?)") has been implemented by the procedure the authors describe in Equations (1), (2), and (3). If, I correctly understand", in Eq. (2) we see the g-computation estimators, a well-known concept introduced by [21] (and recently used also by https://www.nature.com/articles/s41598-020-65917-x ), that splits the two configurations "treated" and "non-treated".

Other reference on g-computation to consider in the study can be for example:

-       [Le Borgne et al. 2021](https://www.nature.com/articles/s41598-021-81110-0)
-       https://www.sciencedirect.com/science/article/pii/S0167947306002076
-       https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-023-01835-6
-       https://pubmed.ncbi.nlm.nih.gov/34861799/

I understand that Eq. (2) and in Eq. (3) consider both treated and non-treated sample (or patient?), but those sums can be also equal to 0, even if their single terms are different from 0. Moreover, it is not clear if, and how Eq. (3) is related to Eq. (2): is it a variant? Is Eq. (3) derived from Eq. (2) under special assumptions?

> Eq. (3) is the definition of the estimator of the ATE based on a model of the outcome. It is sometimes called g-estimation. It  is unbiased under under consistency (Assumption. 3) and unconfoundedness (Assumption 1).  To prove that, take the expectation of Eq. (3).  Then apply  Eq. (1) which is true under consistency (Assumption 3) and unconfoundedness (Assumption 1). Then by the law of total expectancy, the conditionnality on X disappear and we recover Eq. (2).

Furthermore, the Robinson decomposition (R-decomposition) should be better discussed and its role in the context of causal inference should be better commented.
R-decomposition introduces two quantities, the conditional mean outcome and the probability to be treated (known as propensity score). What is the role of the propensity score? How is it calculated? Is it calculated simply by a frequentist approach (number fo treated/total number of patients?

> The R-decomposition can be understood  as a writing trick that separate the mean outcome from the treatment effect. The mean outcome is the part of the counterfactual outcome that can be best estimated from the full population (treated or not). The treatment effect $(a-e(x)) \tau(x)$ should take into account the difference between the two population covariates which is captured by the propensity score.

> The R-decomposition inspires two different risks, this is why we are introducing it here. I developped a little bit on the role of this decomposition and why we are introducing it.

> The propensity score is a nuisance quantity that is necessary to estimate .

Finally, I note that the paper mentions inference methods specific to machine learning (random forest, etc.). In this regard, there is a great deal of study in the literature as to their actual usefulness in this field. I suggest the authors consider/review some of the following literature in the paper (those that the authors consider most appropriate to be included in their manuscript):

https://link.springer.com/book/10.1007/978-3-031-35051-1

https://www.causal-machine-learning.com/

https://www.frontiersin.org/journals/bioinformatics/articles/10.3389/fbinf.2021.746712/full

https://www.research-collection.ethz.ch/handle/20.500.11850/650703

https://aisel.aisnet.org/pacis2022/181/

or some other that can be easily found in literature.

# Reviewer 2

Overall assessment:

Let me begin with positive comments.  I think the paper is very important because it describes a sophisticated methodology for experimental examination of causal models.
Despite some attempts this is still lacking in literature.  I especially like the simulation scheme which uses different features for data generation and learning models.  Moreover thorough experimental evaluations of causal models are not very frequent making the paper important.  Code is available on Github which is a big plus.

I am however not sure about the overall conclusions, which are that R-risk is always the best.  Performance of risk functions may depend on several factors and the overall best performance might be a result of bias in available data.  The only factor taken into account by the authors is overlap.  However, only one type of lack of overlap is tested in experiments, based on shifted Gaussians.  This will likely lead to violation of the positivity assumption.  How about different scenarios, e.g.: perfect overlap but different treatment/control distributions?  Confounding need not necessarily lead to lack of overlap.  Another possible factor is the ratio of causal effect to background responses.  It is known that it may affect relative performance of HTE estimators.

Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.

Another question: do the differences between risks in experiments follow known theoretical results?  For example, it is well known (e.g. https://www.semanticscholar.org/reader/112aba9c62f98b9684693f4c20c30274bc9eb1dc or [Rudas ans jaroszewisz, 2017](https://link.springer.com/content/pdf/10.1007/s10618-018-0576-8.pdf)) that the tau risk performs poorly.

> The second paper is from the uplift regression literature. It focus on full randomization problems and a linear specification with gaussian predictors. It shows theoretically that a linear T-learner (called double model approach in their approach) has smaller asymptotical variance than a treatment ratio reweighted outcome S learner. However, they note that in non normal cases the  T-learner can be ill-defined or with high variance. They propose a corrected reweighted S-learner approach to reduce this  variance. They show empirically that it behaves better than the non corrected version on a simulation and two real world problems. 
> I am not sure to see the link with the tau-risk. The paper does not compare risks but corrected or uncorrected vs T-learners. In addition, they only consider full randomization and linear problems. In our case, we consider a more general setting with a variety of data generation procedures and we focus on non-randomized situations where the distribution of the treatment is conditioned by the predictors.

Another problem is that the paper is hard to follow. The authors rapidly switch between topics: risk, selection, nuisance parameter estimation so the reader quickly gets lost.  The sections in supplementary materials contain important information but not all are referenced in main text.  The paper gives an impression of several experiments put together in some order without guiding the reader through the maze.  I think organization should be improved before publication.

Other comments:
---------------

In the Methods section g-computation is described.  It seems indeed the first name under which the technique appeared in literature, but other names, most importantly S-learner are being used frequently nowadays.  Interestingly those names are mentioned in Empirical study on page 4, however, the close relationship (equivalence?) between S-learner and g-computation is not mentioned.

> G-computation is larger than S-learner. The first mention of S-learner that we are aware of is [Künzel, 2019](https://arxiv.org/pdf/1706.03461). In this paper, they distinguish T-learners from S-learners (or more elaborated ones). T-learners model $\mu(0)$ and $\mu(1)$ separately and train separately on each population. On the contrary S learner use only one model for both population and add the treatment as a covariate. A good example to understand the difference is to note that a linear T-learner is the same thing as a linear S-learner where we add an interaction term between the treatment and the covariates. Ultimately, both methods are G-computation in the sense that they rely on the identification of eq. (1) making the link between potential outcomes and observed quantities. To improve the manuscript, we added just after the eq. (3) that one can choose between T-learner and S-learner.

Equation 2: should the expectation be over X not Y?

> Thanks for the remark. We corrected a typo in the equation. There should not have any mention of X in the conditional expectation.

page 17: proposition number is missing

> Thanks. Fixed.

A.7: the data generation procedure needs better explanations.  Why not call (7) P(d|A) since that's how its used later?  On page 17 you say '[b1,bD]~P(x)' and later 'x~P(x)' which causes confusion.  Also, how are beta vectors generated?

page 4: TLearner uses different feature representation for each treatment arm.  Does this really happen in practice?  Usually all subgroups are described with the same set of features.

page 4/5: Family of candidate estimators: why are only linear models used in some dataset and forests for other?  The leaf sizes of forests are chosen from a rather small range, why not use some percentage of the size of the training set?

Language comments:
------------------

It seems that English might need improvement in some cases, for example:

Abstract: "We study such a variety of model selection procedures in practical settings": a "large variety" seems more appropriate.

On page 1, the sentence "Predictive modeling bridges to causal inference theory under the name of outcome models ..." contains "bridges under the name" which also sounds strange.

On page 2: In the sentence "Then we anchor causal model selection in the potential outcome framework and details the causal risks and model-selection procedure." I believe 'details' should be 'detail'.

Algorithm 1: 'a candidate estimators f' should be 'a candidate estimator f'

throughout: "an RBF expansion" is more commonly used than "a RBF expansion"

page 4: 'for both population' should be 'for both populations'

### 2024-08-29 how_to_select_latex.tex
WIP: response to reviewer

@@
 
-\paragraph{Estimating treatment effects with outcome models -- g-computation \cite{robins_new_1986}}\label{subsec:estimators}
-
-Should we know the two expected outcomes for a given $X$,
+\paragraph{Identifying treatment effects with outcome models -- g-computation \cite{robins_new_1986}}\label{subsec:estimators}
+
+Should we know the two potential outcomes for a given $X$,
 we could compute the
@@
 %
-These two expected outcomes can be estimated from observed data:
+These two potential outcomes can be estimated from observed data:
 the consistency \ref{assumption:consistency} and unconfoundedness
@@
           non-treated populations, an important parameter for causal inference
-          (related to $\eta$ in Proposition
-          \ref{theory:prop:mu_risk_ipw_bound}). Finally, we generate observed
+          (related to $\eta$ in Assumption
+          \ref{assumption:overlap}). Finally, we generate observed
           outcomes adding Gaussian noise.

### 2024-08-29 response_to_reviewer.md
wip: review iterate

@@
 To improve the manuscript, we added just after the eq.
-(3) that one can choose between T-learner and S-learner.
+(3) that one can choose to model the response function with either a T-learner or a S-learner.
 
@@
 
+> It is indeed true that all subgroups are described with the same set of features. However,
+
 page 4/5: Family of candidate estimators: why are only linear models used in some dataset and forests for other?
 The leaf sizes of forests are chosen from a rather small range, why not use some percentage of the size of the training set?
+
+> The choice of the candidate family of estimators was guided by the idea to compare estimators that are not too different from each other, making the model selection task more challenging.
+The important point is that we are not focusing on the best possible estimator but on the best possible model selection procedure.
+To identify the optimal procedure, it is more convincing to compare close candidate estimators than a wide variety of candidate estimators.
+In the later case, one estimator could dominate all other estimators for every selection procedure.
+This explains why we are choosing a narrow range of leaf sizes for the forests.
+
+> For the simulations we chose estimators that are out of the functional class of the  true data generation process but are quite similar to it.
+They are out of the funtional class because the representers are different from the one of the true data.
+They are close because they have the same functional form : a linear model on top of representers.
+We used the simulation to explore the wider range of data generation processes, focusing on a gradient of overlap.
+Because of the sheer number of generated dataset, we did not include computationnaly costly boosted trees in the candidate family.
+than the one used for the data generation process.
+An important point is that the .
+For the semi-simulated datasets, we chose .
+For a forest estimator, a T learner or a S learner are quite similar .
+The fact that the .
 

### 2024-08-29 response_to_reviewer.md
wip: review iterate

@@
 
+> For the simulations we chose estimators that are out of the functional class of the  true data generation process (a candidate estimator cannot model exactly the data generation process) but that are quite similar to it.
+They are out of the functional class because the representers are different from the one of the true data (they are two randomly sampled data points).
+They are close because they have the same functional form : a linear model on top of two data representers.
+We used the simulations to explore the wider range of data generation processes, focusing on a gradient of overlap.
+Because of the sheer number of generated dataset, we did not include computationally costly boosted trees in the candidate family.
+
 > The choice of the candidate family of estimators was guided by the idea to compare estimators that are not too different from each other, making the model selection task more challenging.
@@
 This explains why we are choosing a narrow range of leaf sizes for the forests.
-
-> For the simulations we chose estimators that are out of the functional class of the  true data generation process but are quite similar to it.
-They are out of the funtional class because the representers are different from the one of the true data.
-They are close because they have the same functional form : a linear model on top of representers.
-We used the simulation to explore the wider range of data generation processes, focusing on a gradient of overlap.
-Because of the sheer number of generated dataset, we did not include computationnaly costly boosted trees in the candidate family.
-than the one used for the data generation process.
-An important point is that the .
-For the semi-simulated datasets, we chose .
-For a forest estimator, a T learner or a S learner are quite similar .
-The fact that the .
 

### 2024-08-29 response_to_reviewer.md
wip: review iterate

@@
 
+> The differences between risks have been explored *theoretically* by XX for the lose performance of the U-risk. For the other risk, we are not aware of any theoretical results. 
+> The differences between risks have been highlighted *experimentally* in other papers such as [Schuler, et al., 2018](https://arxiv.org/pdf/1804.05146) for quadratic and stepwise data generation processes or [Curt et Van der Schaar, 2023](https://proceedings.mlr.press/v202/curth23b/curth23b.pdf) exploring more diverse outcome generation processes such as CATE complexity.
+They both underline good performances for the R risk.
+
 > The second paper is from the uplift regression literature.
@@
 
-> Thanks. Fixed.
+> Thanks for this remark. We fixed it.
 
@@
 
-> It is indeed true that all subgroups are described with the same set of features. However,
+> It is indeed true that in usual practice all subgroups are described with the same set of features.
+However, as soon as we are using a machine learning (necessary to model CATE) such as a random forest or boosted trees, the inner representations of the features inside the model is different for both population.
+This idea has been exploited to build representation aware neural net models for cate estimation by [Johansson et al., 2022](https://www.jmlr.org/papers/volume23/19-511/19-511.pdf).
+It is also the case for simpler models such as a linear model with treatment interaction and lasso regularization.
 
@@
 page 4: 'for both population' should be 'for both populations'
+
+> Thanks for the remarks. We corrected the typos and improved the language as suggested.

### 2024-08-29 how_to_select_latex.tex
wip: review iterate

@@
         \textbf{Methods:}
-        We study such a variety of model selection procedures in practical settings:
+        We study a large variety of model selection procedures in practical settings:
         finite samples settings and without theoretical assumption of
@@
 \cite{hernan_causal_2020,vanderweele2019principles}.
-Predictive modeling bridges to causal inference theory under the name of
+Predictive modeling is linked to causal inference theory by the concept of
 \emph{outcome models} (or G-computation,
@@
 We first illustrate the problem of causal model selection. Then we anchor causal model selection in the \emph{potential outcome}
-framework and details the causal risks and model-selection procedure. We then
+framework and detail the causal risks and model-selection procedure. We then
 rewrite the so-called $R\text{-risk}$ as a reweighted version of mean squared
@@
         Given train and test sets $(\mathcal{T}, \mathcal{S}) \sim \mathcal{D}$,
-        a candidate estimators $f$, a causal
+        a candidate estimator $f$, a causal
         metrics $\ell$:
@@
 approximate imperfectly the
-data-generating process. To build such estimators, we first use a RBF
+data-generating process. To build such estimators, we first use an RBF
 expansion similar to that used for data generation. We choose two random
@@
 \begin{itemize}
-    \item SLearner: A single learner for both population, taking the treatment as
+    \item SLearner: A single learner for both populations, taking the treatment as
           a supplementary covariate.

### 2024-08-29 response_to_reviewer.md
iter

@@
 > The propensity score is a nuisance quantity that is necessary to estimate . 
+**TODO: unfinished**
 
@@
 or some other that can be easily found in literature.
+
+> Thanks for the references.
+I am in favor of adding some of them to the manuscript such as [Chernozhukov et al., 2024](https://causalml-book.org/) or [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475) which clearly distangle the different subfields at the intersection of causal inference and machine learning.
+I have added them at the beginning of the manuscript before focusing on the model selection problem for cate.
 
@@
 
+**TODO**
+
 Another question: do the differences between risks in experiments follow known theoretical results?
@@
 
-> The differences between risks have been explored *theoretically* by XX for the lose performance of the U-risk. For the other risk, we are not aware of any theoretical results. 
-> The differences between risks have been highlighted *experimentally* in other papers such as [Schuler, et al., 2018](https://arxiv.org/pdf/1804.05146) for quadratic and stepwise data generation processes or [Curt et Van der Schaar, 2023](https://proceedings.mlr.press/v202/curth23b/curth23b.pdf) exploring more diverse outcome generation processes such as CATE complexity.
-They both underline good performances for the R risk.
+> The high variability of the U-learner has already been noted by [Nie and Wager, 2017](https://arxiv.org/pdf/1712.04912).
+This instability is naturally found in the U-risk.
+For the other risk, we are not aware of any theoretical results justifying differences between risks.
+> Empirically, the differences between risks have been highlighted in other papers such as [Schuler, et al., 2018](https://arxiv.org/pdf/1804.05146) for quadratic and stepwise data generation processes or [Curth et Van der Schaar, 2023](https://proceedings.mlr.press/v202/curth23b/curth23b.pdf) exploring more diverse outcome generation processes focusing on CATE sparsity (as a proxy of its complexity) based on a semi-simulated dataset for the covariate (ACIC2016).
+They both underline good performances for the R-risk (see Table.1 for Schuler and Fig.
+3 for Curth).
+A remark that we made in our manuscript is that the results of Schuler are convincing for randomization setups but not for observational (non-randomized) setups.
+Curth's results are more convincing for observational setups but did not explore the overlap assumption.
 
@@
 I think organization should be improved before publication.
+
+**TODO**
 
@@
 Also, how are beta vectors generated?
+
+**TODO**
 

### 2024-08-31 response_to_reviewer.md
answer to R1

@@
 (2).
+> G-estimation itself does not control error on both populations as highlighted in Figure 1.
+Specific inductive biases of ML models are able to favor such regularization(see for example [Johannson et al., 2022](https://www.jmlr.org/papers/volume23/19-511/19-511.pdf)).
+Selection procedures, more elaborated than RMSE are needed to favor such models.
+I added this last explanation in the method (par.
+Model-selection risks, oracle and feasible) trying to clarify the link between the hint on error control on both population, g-estimation and the selection procedure which is our main focus.
 
@@
 
-> The R-decomposition can be understood  as a writing trick that separate the mean outcome from the treatment effect.
-The mean outcome is the part of the counterfactual outcome that can be best estimated from the full population (treated or not).
-The treatment effect $(a-e(x)) \tau(x)$ should take into account the difference between the two population covariates which is captured by the propensity score.
-
-> The R-decomposition inspires two different risks, this is why we are introducing it here. I developped a little bit on the role of this decomposition and why we are introducing it.
-
-> The propensity score is a nuisance quantity that is necessary to estimate . 
-**TODO: unfinished**
+> G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function µa(x) [49 , chapter 15].
+Other choices of decomposition exist, such as the R-decomposition [57], the Double robust estimation, the X-learner...
+The R-decomposition is one way of identifying the causal quantity of interest which introduce two statistical quantity that have to be modelled: the mean outcome from the treatment effect and the propensity score.
+Any statistical estimator can be chosen to model these quantities.
+The frequentist approach is one (naive) such estimator.
+In practice, researchers often estimate the propensity score with a logistic regression, but other methods such as Boosted trees, Random Forests, or Neural Networks can be used.
+A good ps estimator should be well calibrated but this question is out of scope for this paper.
+We briefly mentioned in the discussion this research question on selecting a good estimator for the propensity score.
 
@@
 > Thanks for the references.
-I am in favor of adding some of them to the manuscript such as [Chernozhukov et al., 2024](https://causalml-book.org/) or [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475) which clearly distangle the different subfields at the intersection of causal inference and machine learning.
-I have added them at the beginning of the manuscript before focusing on the model selection problem for cate.
+I added two references to Causal ML to the manuscript such as [Chernozhukov et al., 2024](https://causalml-book.org/) or [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475).
+Both of them cover most of the subfields at the intersection of causal inference and machine learning.
+I have added them at the beginning of the manuscript before stating our focus on model selection for CATE.
 

### 2024-08-31 how_to_select_latex.tex
answer to R1

@@
     \begin{keywords}
-        Model Selection, Predictive model, Treatment Effect, G-formula, Machine Learning
+        Model Selection, Predictive model, Treatment Effect, G-computation, Machine Learning
     \end{keywords}
@@
 Predictive modeling is linked to causal inference theory by the concept of
-\emph{outcome models} (or G-computation,
-G-formula \cite{robins_role_1986}, Q-model
+\emph{outcome models} (or g-computation, g-estimation,
+g-formula \cite{robins_role_1986}, Q-model
 \cite{snowden_implementation_2011}, conditional mean regression
@@
 
-%\idea{There is a variety of models for g-estimation}
+%\idea{There is a variety of models for g-computation}
 % There is an explosion of outcome modeling or machine learning methods, either 
@@
 Indeed, estimates of treatment effects can vary markedly across different
-predictive models \cite{fang2019applying,dorie_automated_2019} (illustration in
+predictive models \cite{fang2019applying,dorie_automated_2019,leborgne2021g,ren2023comparing} (illustration in
 Appendix \ref{apd:toy_example:acic_2016_ate_variability}).
@@
 
-We study \textit{model selection procedures} in practical settings,
+The intersection between machine learning and causal inference is growing rapidly
+\cite{kaddour2022causal,chernozhukov2024applied}. We focus on \textit{model selection procedures} in practical settings,
 without theoretical assumptions often made in statistical literature such
@@
 
-This toy example illustrates that the classic minimum Mean Square Error
+This toy example illustrates that the classic minimum mean squared error (MSE)
 criterion is not suited to choosing a model among candidate
@@
 the consistency \ref{assumption:consistency} and unconfoundedness
-\ref{assumption:ignorability} assumptions imply the following equality:
+\ref{assumption:ignorability} assumptions imply the following equality, linking the target potential quantity to the observed data:
 \begin{equation}\label{eq:mu_identification}
@@
 
-%TODO: not well located here. The reader does not know why we introduce this decomoositon her.
 \paragraph{Robinson decomposition}
-The \emph{R-decomposition}
-of the outcome model\cite{robinson_rootnconsistent_1988} links the counterfactual outcomes $y(a)$ to the CATE by introducing two quantities, the conditional mean outcome
-and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}).:
+G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function $\mu_{a}(x)$ \cite[chapter 15]{chernozhukov2024applied}.
+Other choices of decomposition exist, such as the R-decomposition \cite{robinson_rootnconsistent_1988}.
+It introduces two new statistical estimates, the conditional mean outcome
+and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
 \begin{flalign}
@@
 
-We formalize model selection for causal estimation. Thanks to the g-formula
-identification (\autoref{eq:mu_identification}), a given outcome model $f: \mathcal X
+We formalize model selection for causal estimation. Thanks to the outcome model identification (\autoref{eq:mu_identification}), a given model $f: \mathcal X
     \times \mathcal A \rightarrow \mathcal{Y}$ --learned from data or built from
@@
 %
-Let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
+However, g-computation itself does not control error on both populations as highlighted in \ref{fig:toy_example}.
+Specific inductive biases of ML models are able to favor such regularization \cite{johansson2022generalization}.
+Thus, selection procedures, more elaborated than MSE are needed to favor such models.
+A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ on the observed data $O$ and a splitting strategy of the data.
+%
+Formally, let $\mathcal F=\{f: \mathcal X \times \mathcal A \rightarrow \mathcal{Y}\}$ be
 a family of such estimators. Our goal is to select the best candidate in this
@@
 empirical results are clear for randomized treatment allocation but less
-convincing for observational settings where both simple Mean Squared
-Error --MSE, $\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
+convincing for observational settings where both simple MSE --$\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
 reweighted MSE --$\mu\text{-risk}_{IPW}$ def.\,\ref{def:mu_ipw_risk}--

### 2024-08-31 response_to_reviewer.md
R 2: answer to all but the criticism on simulation too focused on overalpé

@@
 
-> The high variability of the U-learner has already been noted by [Nie and Wager, 2017](https://arxiv.org/pdf/1712.04912).
-This instability is naturally found in the U-risk.
-For the other risk, we are not aware of any theoretical results justifying differences between risks.
+> The high variability of the U-learner has already been noted by [Nie and Wager, 2021](https://arxiv.org/pdf/1712.04912).
+This advantage of the R-loss that does not have a propensity score to the denominator is also discussed in [Chernozhukov et al.
+2024, chapter 15.](https://causalml-book.org/assets/chapters/CausalML_chap_15.pdf).
+The convergence rate of the R-loss has been proven in [Foster and Syrgkanis, 2023](https://arxiv.org/pdf/1901.09036).
+We are  not aware of theoretical results comparing simple robust losses (such as mu-risk or mu-risk IPW) to orthogonal losses such as R-loss, U-loss or DR-Loss.
 > Empirically, the differences between risks have been highlighted in other papers such as [Schuler, et al., 2018](https://arxiv.org/pdf/1804.05146) for quadratic and stepwise data generation processes or [Curth et Van der Schaar, 2023](https://proceedings.mlr.press/v202/curth23b/curth23b.pdf) exploring more diverse outcome generation processes focusing on CATE sparsity (as a proxy of its complexity) based on a semi-simulated dataset for the covariate (ACIC2016).
@@
 
-**TODO**
+> We tried to better guide the reader through the method section and to clarify the motivation behind the different experiments. 
+> We checked the supplementary materials but we did not find any appendix that was not referenced at least once in the main text.
+Did you have a specific appendix in mind that would require a reference in the main text?
 
@@
 
-**TODO**
+> We assumed that you talked about the equation 17 rather than 7. Could you point where we use the P(d|A) notation? We did not find it in the manuscript. 
+> The notation b1, ..bD ~P(x) was meant to be understood as the beta vectors are drawn from a distribution P(x).
+The later mention of 'x~P(x)' was meant to be understood as the definition if the RBF function.
+We removed the second mention which causes confusion.
+> The beta vectors are drawn from two normal distributions with mean 0 and unit variances. We added this information in the manuscript.
 

### 2024-08-31 how_to_select_latex.tex
R 2: answer to all but the criticism on simulation too focused on overalpé

@@
 
+The simulation has been designed to control precisely the overlap parameter.
+The semi-simulated datasets have been included to explore more diverse and noisy covariate distributions than simulated covariates.
+These simulations cover a wide range of causal settings, such ratio of causal effect to background responses, and functional links between covariates, outcome and treatment.
+
 \subsection{Caussim: Extensive simulation settings}\label{subsec:simulations}
@@
 \cite{schuler_comparison_2018}.
+An interesting direction is taken in \cite{foster2023orthogonal} where the authors derive convergence rates for orthogonal losses such as the R-loss, or the DR-Loss without assuming well-specification of the model for target parameter.
 
@@
           with D=2 in our simulations. Then, we compute an approximation
-          of the full kernel of the data generation process $RBF(x,
-              \cdot) \;  with \; x \sim \mathbb P(x)$ with these
+          of the full kernel of the data generation process with these
           representers: $z(x) = [RBF_{\gamma}(x, b_d)]_{d=1..D} \cdot
@@
           \end{equation*}
+          Where $\beta_{\mu}$ and $\beta_{\tau}$ are sampled from two normal distributions with mean 0 and unit variances $\mathcal{N}(0, I_{D+1})$.
     \item Adding a Gaussian noise, $\varepsilon \sim \mathcal N(0, \sigma(x;a))$,

### 2024-09-01 how_to_select_latex.tex
iter R2

@@
 valid causal estimates needed to motivate individual treatment decisions?
-As no single machine-learning method performs best across all data sets, there is a
+As no single machine-learning method performs best across all datasets, there is a
 pressing need for clear guidelines to select outcome models for causal
@@
 
-Similarly to the ATE, at the individual level, the CATE:
+Similarly to the ATE, at the individual level Eq.\ref{eq:mu_identification} links the CATE to statistical quantities:
 \begin{equation}
@@
 $m$ and $e$ are often called
-\emph{nuisances} \cite{chernozhukov_double_2018}; they are unknown and are estimated in practice thanks to statistical models.
+\emph{nuisances} \cite{chernozhukov_double_2018}. They are unknown and are estimated in practice thanks to statistical models (eg. linear models or forest based estimators).
 
@@
 
-\subsection{Model-selection risks, oracle and feasible}\label{sec:problem:model_selection}
+\subsection{Model-selection, oracle and feasible risks}\label{sec:problem:model_selection}
 
@@
 %
-However, g-computation itself does not control error on both populations as highlighted in \ref{fig:toy_example}.
+However, g-computation itself does not control error on both populations as highlighted in Figure \ref{fig:toy_example}.
 Specific inductive biases of ML models are able to favor such regularization \cite{johansson2022generalization}.
-Thus, selection procedures, more elaborated than MSE are needed to favor such models.
-A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ on the observed data $O$ and a splitting strategy of the data.
+Thus, selection procedures, more elaborated than MSE are needed to favor the best model for each dataset.
+A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ with observed data $O$ and a splitting strategy of the data for estimating the different regressions (nuisances) involved in the risk.
 %
@@
 
-\subsection{Estimation and model selection procedure}\label{problem:estimation_procedure}
+\subsection{Model selection procedure}\label{problem:estimation_procedure}
 
@@
 
-The simulation has been designed to control precisely the overlap parameter.
-The semi-simulated datasets have been included to explore more diverse and noisy covariate distributions than simulated covariates.
-These simulations cover a wide range of causal settings, such ratio of causal effect to background responses, and functional links between covariates, outcome and treatment.
+The simulations have been designed to evaluate the effect of the overlap parameter.
+The semi-simulated datasets have been included to explore more diverse and noisy covariate distributions.
+These simulations cover a wide range of causal settings such as different ratio of causal effect to background responses, and functional links between covariates, outcome and treatment.
 
@@
 selection and, strictly speaking, push
-to split out a third separated data set --a ``nuisance set''-- to fit the
+to split out a third separated dataset --a ``nuisance set''-- to fit the
 nuisance models. The drawback is that it depletes the data available for

### 2024-09-11 how_to_select_latex.tex
apd:simus add generation algo

@@
 %\titleformat*{\paragraph}{\bfseries} % force bold font for paragraphs
+\usepackage{algpseudocode} % for algorithmic
 
@@
 control the amount of overlap between treated and control populations, a crucial
-assumption for causal inference. Figure \ref{fig:simulation_examples}
+assumption for causal inference. Algorithm \ref{apd:experiments:generation:algo} describes the generation process for one simulation. Figure \ref{fig:simulation_examples}
 illustrates 2D examples of the simulation.
@@
           we construct the potential outcomes:
-          $y(a) = \mu_0(x) + a\,\tau(x) + \varepsilon(x, a)$
+          $y(a) = (1-\omega)\mu_0(x) + a\,\omega\tau(x) + \varepsilon(x, a)$ with $\omega$ the treatment effect relative size with respect to the baseline response.
 \end{itemize}
@@
 parameters $\theta \in \left[ 0, 2.5 \right]$.
+
+\begin{algorithm}
+    \caption{Data simulation for the simulated dataset caussim}
+    \label{apd:experiments:generation:algo}
+    Let the parameters of a given simulation be:
+    \begin{itemize}
+        \item N the number of samples.
+        \item D the dimension of the basis expansion.
+        \item $p_A$ the proportion of treated individuals.
+        \item $\theta$ the overlap parameter.
+        \item $\mathbb P(x|A=a) = W \cdot \mathcal N \Big( \begin{bmatrix} (1-2a) \theta \\ 0\end{bmatrix} ; \begin{bmatrix} \sigma_0 & 0 \\ 0 & \sigma_1\end{bmatrix} \Big)$ the rotated gaussian for each population covariate.
+$W$ is a random rotation matrix and $\sigma_0^2=2;\sigma_1^2=5$ are scaling parameters.
+        \item $e(x)=\mathbb P(A=1|X=x)$ the oracle propensity score, obtained analytically from $\mathbb P(x|A=a)$.
+        \item $\mathbb P(X) = p_A \mathbb P(X|A=1) + (1- p_A) \mathbb P(X|A=0)$ the mixture of gaussian for the whole population covariates.
+        \item $b_1, .., b_D \sim \mathbb P(X)$, the basis sampled from the gaussian mixture.
+        \item $z(x) = [RBF_{\gamma}(x, b_d)]_{d=1..D} \cdot Z^T \in
+                  \mathbb{R}^D$ the Nystroem expansion. $RBF_{\gamma}$ is the
+              Gaussian kernel $K(x, y) = exp(-\gamma ||x-y||^2)$ and Z is the
+              normalization constant of the kernel basis, computed as the root
+              inverse of the basis kernel $Z=[K(b_i, b_j)]_{i, j \in
+                  {1..D}}^{-1/2}$.
+        \item $\beta_{\mu}$ and $\beta_{\tau}$, the linear coefficients on top of the basis expansion.
+They are sampled from two normal distributions with mean 0 and unit variances $\mathcal{N}(0, I_{D+1})$.
+        \item $\omega$ the treatment effect relative size with respect to the baseline response.
+        \item $\varepsilon(x, a) \sim \mathcal N(0, \sigma_y)$ an exogen gaussian noise with $\sigma_y$ the scale of the noise.
+    \end{itemize}
+
+    Generation process for one sample ($x$, $a$, e(x), $y(0)$, $y(1)$, $y$):
+    \begin{algorithmic}[1]
+        \State $x \sim \mathbb P(X)$ \Comment{Sample from rotated gaussian mixture}
+        \State  $z(x)$ \Comment{Compute the Nystroem expansion}
+        \State $\mu_0(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\mu}^T$ \Comment{Compute the non-treated response surface, ie. the baseline}
+        \State  $\tau(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\tau}^T$ \Comment{Compute the conditional treatment effect (CATE)}
+        \State $y(a) = (1-\omega) \mu_0(x) + a\,\omega\tau(x) + \varepsilon(x, a) + \varepsilon(x, a)$ \Comment{Compute the potential outcomes}
+        \State $a \sim \operatorname{Bernoulli}(e(x))$ \Comment{Compute the treatment status}
+        \State $y = a\,y(1) + (1-a)\,y(0)$ \Comment{Compute the observed outcome}
+    \end{algorithmic}
+
+\end{algorithm}
 

### 2024-09-30 response_to_reviewer.md
reviewer response : wip on the last point

@@
 
-**TODO**
+> We thank the reviewer for this remark and the suggestion to explore more diverse simulation scenarii.
+Exploring the sensitivity of our result on the R-risk appears to us particularly necessary.
+Thus, we ran a dedicated experiment focusing on the ratio of causal effect to background response.
+
+> We did not explore more sophisticated treatment/control distributions for our simulation.
+Building these new data generation processes would imply a big modification of our simulation.
+We rely on the easy inversion of the distribution functions of the gaussians (p(X|A=a)) to analytically compute the true propensity score.
+In addition, we think that the semi-simulated datasets already provide some interesting argument in favor of r-risk with more complex treatment/control covariate distributions.
+However, we mentionned in the discussion the need to explore more diverse data generation processes for simulations.
 

### 2024-09-30 how_to_select_latex.tex
reviewer response : wip on the last point

@@
         is worse. Re-weighting it with propensity score
-        does not bring much improvements. The
+        does not bring much improvements in most cases. On average, the
         $R\text{-risk}$, which uses as nuisances a model of
@@
 
+\paragraph{R-risk is robust to a wide range of effect ratio}
+
+We focused on the
+
+\begin{figure}[!tb]
+    % XXX: for submission
+    \centering\begin{minipage}{\linewidth}
+        \includegraphics[width=\linewidth]{_6_effect_ratio_by_bin_comparaison_kendall_by_Dataset.pdf}
+    \end{minipage}
+
+    \caption{\textbf{R-risk is robust to a wide range of effect ratio}: Kendall’s $\tau$ agreement with $\tau$-risk.
+Strong, medium and Weak Causal effect ratio are the tertiles of the absolute ratio causal effect to baseline response, ranging from 1 to 300.
+Appendix \ref{apd:effect_ratio} details this simulation.}\label{fig:effect_ratio_influence}
+\end{figure}
+
 \section{Discussion and conclusion}\label{sec:discussion}\label{sec:conclusion}
@@
 
+\subsection{Sensitivity of the results to different ratio of causal effect to baseline response}\label{apd:effect_ratio}
+
+TODO
+
 \bibliography{biblio}

### 2024-10-03 response_to_reviewer.md
answer to reader, finished v1

@@
 
-I appreciate the approch and the level of detail that this paper offers on an issue that is, to say the least, a hot topic in bioinformatics and data science in general: the evaluation of the best 'error' model when choosing a causal model for predictive purposes.
+I appreciate the approach and the level of detail that this paper offers on an issue that is, to say the least, a hot topic in bioinformatics and data science in general: the evaluation of the best 'error' model when choosing a causal model for predictive purposes.
 Indeed, there is much confusion on the subject and very often error models are applied without any real awareness of their potential and limitations.
@@
 
-> G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function µa(x) [49 , chapter 15].
-Other choices of decomposition exist, such as the R-decomposition [57], the Double robust estimation, the X-learner...
+> G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function µa(x)  [Chernozhukov et al., 2024, chapter 15](https://causalml-book.org/).
+Other choices of decomposition exist, such as the R-decomposition, the Double robust estimation, the X-learner...
 The R-decomposition is one way of identifying the causal quantity of interest which introduce two statistical quantity that have to be modelled: the mean outcome from the treatment effect and the propensity score.
@@
 A good ps estimator should be well calibrated but this question is out of scope for this paper.
-We briefly mentioned in the discussion this research question on selecting a good estimator for the propensity score.
+In our discussion, we briefly mentioned this interesting research question on what properties should have a good estimator for the propensity score.
 
@@
 > Thanks for the references.
-I added two references to Causal ML to the manuscript such as [Chernozhukov et al., 2024](https://causalml-book.org/) or [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475).
+I added two references for Causal ML into the manuscript: [Chernozhukov et al., 2024](https://causalml-book.org/) and [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475).
 Both of them cover most of the subfields at the intersection of causal inference and machine learning.
@@
 Thus, we ran a dedicated experiment focusing on the ratio of causal effect to background response.
+We measure the ratio by the absolute mean difference between the causal effect and the baseline response, measured empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$.
+We vary a dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this ratio to cover a wide range of values.
+In one experiment it ranges from 1 to 300.
+Figure 7 gives the main results of this experiment.
+We see that R-risk is still the best risk on average, but when the effect ratio is high (up to 100), the reweighted mu-risk and the the tau-risk_IPW are better than the R-risk.
+For low value of the ratio, however, the R-risk still outperforms all other metrics.
+We added these new results to the main paper and the.
+We also nuanced our conclusion on the R-risk, underlying that it is better on average but that in some situations, other risks might be more adapted.
+A typical such situation would be when the applied scientist is interested in estimating an effect with a big ratio for the part of the population where there is low overlap.
 
 > We did not explore more sophisticated treatment/control distributions for our simulation.
-Building these new data generation processes would imply a big modification of our simulation.
+Building these new data generation processes would imply a big modification to our setup.
 We rely on the easy inversion of the distribution functions of the gaussians (p(X|A=a)) to analytically compute the true propensity score.
-In addition, we think that the semi-simulated datasets already provide some interesting argument in favor of r-risk with more complex treatment/control covariate distributions.
-However, we mentionned in the discussion the need to explore more diverse data generation processes for simulations.
+In addition, we think that the semi-simulated datasets already provide some interesting argument in favor of r-risk with more more diverse treatment/control covariate distributions.
 
@@
 > The beta vectors are drawn from two normal distributions with mean 0 and unit variances. We added this information in the manuscript.
+> We added the Algorithm 2, hopping that it would help the reader to understand exactly how we generate the data for our simulation.
 

### 2024-10-03 how_to_select_latex.tex
answer to reader, finished v1

@@
 
-\paragraph{The $R\text{-risk}$ is the best metric}
+\paragraph{The $R\text{-risk}$ is the best metric on average}
 
@@
 using a metric accounting for the
-putative intervention, the $R\text{-risk}$. The $R\text{-risk}$
+putative intervention, the $R\text{-risk}$. On average, the $R\text{-risk}$
 brings a sizeable benefit to select the most adequate model, even when
@@
 models.
+
+\paragraph{Research question needing more work}
+
+On average, the R-risk outperforms other risks, but it is not optimal for every
+situations. If one possess some prior knowledge on the data generation process,
+another risk might be more adapted. This is typically the case for big causal
+effect ratio to baseline and low overlap where the $\tau\text{-risk}_{IPW}$
+should be better.
 

### 2024-10-03 how_to_select_latex.tex
apd: add effect ratio

@@
 
-We focused on the
+Beyond overlap, we study the effect of varying ratio of causal effect ratio to
+baseline on model selection. In our simulation, we vary the empirical mean
+absolute difference between the causal effect and the baseline, $\Delta_{\mu} =
+    \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)}
+    \big|$, covering a range of ratio from 1 to 300. Figure
+\ref{fig:effect_ratio_influence} shows that for high values of the ratio,
+$R-\text{risk}$ is outperformed by the $\mu\text{-risk}_{IPW}$ and the
+$\tau\text{-risk}_{IPW}$. However, on average, the $R-\text{risk}$ is still the
+better risk.
 
@@
 
-    \caption{\textbf{R-risk is robust to a wide range of effect ratio}: Kendall’s $\tau$ agreement with $\tau$-risk.
-Strong, medium and Weak Causal effect ratio are the tertiles of the absolute ratio causal effect to baseline response, ranging from 1 to 300.
-Appendix \ref{apd:effect_ratio} details this simulation.}\label{fig:effect_ratio_influence}
+    \caption{\textbf{R-risk is robust to a wide range of effect ratio}:
+        Kendall’s $\tau$ agreement with $\tau$-risk. Strong, medium and Weak Causal
+        effect ratio are the tertiles of the absolute ratio causal effect to
+        baseline response, ranging from 1 to 300. Appendix \ref{apd:effect_ratio}
+        details this simulation.}\label{fig:effect_ratio_influence}
 \end{figure}
@@
 
-\paragraph{Research question needing more work}
+\paragraph{Limitations}
 
@@
 
-TODO
+Going beyond overlap, we study the effect of another important parameter of the
+data generation process: the ratio of causal effect to the baseline response.
+
+For each dataset of our simulation (caussim), we measure it as the absolute mean
+difference between the causal effect and the baseline response, measured
+empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N
+    \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a
+dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this
+ratio to cover a wide range of values. For 90 dataset instances, it ranges from 1
+to 300 with the full distribution of values given in Table \ref{apd:effect_ratio_distribution}.
+
+\begin{table}[!ht]
+    \centering
+    \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l|}
+        \hline
+        ~             & count & mean    & std     & min     & 10\%    & 25\%    & 50\%    & 75\%    & 90\%    & max     \\ \hline
+        effect\_ratio & 89    & 99.1291 & 132.525 & 1.24843 & 2.82881 & 4.07361 & 30.1524 & 184.996 & 316.999 & 510.897 \\ \hline
+    \end{tabular}
+    \caption{Distribution of the causal effect ratio $\Delta_\mu$ for the experiment.}\label{apd:effect_ratio_distribution}
+\end{table}
 
 \bibliography{biblio}
-%\printbibliography
 

### 2024-10-06 response_to_reviewer.md
apd: add effect ratio
effect ratio: include result of expe
- diff with original publication
- answer to reviewer: complete with new results

@@
 We vary a dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this ratio to cover a wide range of values.
-In one experiment it ranges from 1 to 300.
+In 90 dataset instance, it ranges several orders of magnitude from 0.005 to 9675.
 Figure 7 gives the main results of this experiment.
-We see that R-risk is still the best risk on average, but when the effect ratio is high (up to 100), the reweighted mu-risk and the the tau-risk_IPW are better than the R-risk.
-For low value of the ratio, however, the R-risk still outperforms all other metrics.
-We added these new results to the main paper and the.
+We see that R-risk is still the best risk on average, but when the effect ratio is high (tertile between 9 and 9675), the reweighted mu-risk and the the tau-risk_IPW are better than the R-risk.
+However, for low value of the ratio,  the R-risk largely outperforms all other metrics.
+We added these new results to the main paper.
 We also nuanced our conclusion on the R-risk, underlying that it is better on average but that in some situations, other risks might be more adapted.
-A typical such situation would be when the applied scientist is interested in estimating an effect with a big ratio for the part of the population where there is low overlap.
+A typical such situation would be when the applied scientist is interested in estimating an effect with a big ratio and want to focus on the part of the population where there is low overlap.
 
@@
 > We tried to better guide the reader through the method section and to clarify the motivation behind the different experiments. 
-> We checked the supplementary materials but we did not find any appendix that was not referenced at least once in the main text.
+> We checked the supplementary materials but we did not find any appendix that was not referenced at least once in the main text (several are referred in figure captions).
 Did you have a specific appendix in mind that would require a reference in the main text?

### 2024-10-06 how_to_select_latex.tex
effect ratio: include result of expe
- diff with original publication
- answer to reviewer: complete with new results

--- 2024-10-03 how_to_select_latex.tex
@@
 \paragraph{Robinson decomposition}
-G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function $\mu_{a}(x)$ \cite[chapter 15]{chernozhukov2024applied}.
-Other choices of decomposition exist, such as the R-decomposition \cite{robinson_rootnconsistent_1988}.
-It introduces two new statistical estimates, the conditional mean outcome
-and the probability to be treated (known as propensity score \cite{rosenbaum_central_1983}):
+G-computation is a choice of decomposition of the CATE estimation into a
+sequence of regressions, where the only statistical estimate is the response
+function $\mu_{a}(x)$ \cite[chapter 15]{chernozhukov2024applied}. Other choices
+of decomposition exist, such as the R-decomposition
+\cite{robinson_rootnconsistent_1988}. The latter introduces two new statistical
+estimates, the conditional mean outcome and the probability of being treated (known
+as propensity score \cite{rosenbaum_central_1983}):
 \begin{flalign}
@@
 
-\paragraph{R-risk is robust to a wide range of effect ratio}
-
-We focused on the
+\paragraph{R-risk is robust to a wide range of effect ratio values}
+
+Beyond overlap, we study for caussim simulations, the effect on model selection
+of different causal effect ratio to baseline. We vary the empirical mean
+absolute difference between the causal effect and the baseline, $\Delta_{\mu} =
+    \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)}
+    \big|$, covering a ratio range from 0.05 to 1000 ($\text{median}=1.9$, full
+distribution in Appendix \ref{apd:effect_ratio}). Figure
+\ref{fig:effect_ratio_influence} shows that for high values of the ratio,
+$R-\text{risk}$ is outperformed by the $\mu\text{-risk}_{IPW}$ and the
+$\tau\text{-risk}_{IPW}$. However, on average, the $R-\text{risk}$ is still the
+better risk.
 
 \begin{figure}[!tb]
-    % XXX: for submission
-    \centering\begin{minipage}{\linewidth}
-        \includegraphics[width=\linewidth]{_6_effect_ratio_by_bin_comparaison_kendall_by_Dataset.pdf}
+    \centering
+    \begin{minipage}{\linewidth}
+        \includegraphics[width=0.9\linewidth]{_6_effect_ratio_by_bin_comparaison_kendall_by_Dataset.pdf}
     \end{minipage}
 
-    \caption{\textbf{R-risk is robust to a wide range of effect ratio}: Kendall’s $\tau$ agreement with $\tau$-risk.
-Strong, medium and Weak Causal effect ratio are the tertiles of the absolute ratio causal effect to baseline response, ranging from 1 to 300.
-Appendix \ref{apd:effect_ratio} details this simulation.}\label{fig:effect_ratio_influence}
+    \caption{\textbf{R-risk is robust to a wide range of effect ratio}:
+    Kendall's $\tau$ agreement with $\tau$-risk. Strong, medium and Weak Causal
+    effect ratio are the tertiles of the absolute ratio causal effect to
+    baseline response, $\Delta_{\mu}$: Low $[0.005;0.4[$, Medium $[0.44; 9[$,
+    High $[9;9675[$. Appendix \ref{apd:effect_ratio} details this
+    simulation.}\label{fig:effect_ratio_influence}
 \end{figure}
@@
 model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
-to more valid causal estimates. Estimating the $R\text{-risk}$ requires a more
+to more valid causal estimates on average. Estimating the $R\text{-risk}$ requires a more
 complex procedure than standard cross-validation used \eg in machine
@@
 build predictive models: it opens the door to evaluating a wide range of
-models, from gradient boosting to convolutional neutral, or language
+models, from gradient boosting to convolutional neural network, or language
 models.
 
-\paragraph{Research question needing more work}
+\paragraph{Limitations}
 
 On average, the R-risk outperforms other risks, but it is not optimal for every
-situations. If one possess some prior knowledge on the data generation process,
+situations. If one possesses some prior knowledge on the data generation process,
 another risk might be more adapted. This is typically the case for big causal
@@
 
-TODO
+Going beyond overlap, we study the effect of another important parameter of the
+data generation process: the ratio of causal effect to the baseline response.
+
+For each dataset of our simulation (caussim), we measure this ratio as the absolute mean
+difference between the causal effect and the baseline response, measured
+empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N
+    \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a
+dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this
+ratio to cover a wide range of values. For 179 dataset instances, it ranges
+several order of magnitudes with the full distribution of values given in
+Table \ref{apd:effect_ratio_distribution}.
+
+\begin{table}[!ht]
+    \centering
+    \resizebox{\textwidth}{!}{ % Adjusts the table to fit the text width
+        \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
+            \hline
+                                   & \textbf{count} & \textbf{mean} & \textbf{std} & \textbf{min} & \textbf{1\%} & \textbf{10\%} & \textbf{25\%} & \textbf{50\%} & \textbf{75\%} & \textbf{90\%} & \textbf{99\%} & \textbf{max} \\ \hline
+            \textbf{effect\_ratio} & 179            & 159.6         & 852.4        & 0.004674     & 0.004765     & 0.02483       & 0.1566        & 1.879         & 21.12         & 135.1         & 3394          & 9675         \\ \hline
+        \end{tabular}
+    }
+    \caption{Distribution of the causal effect ratio $\Delta_\mu$ for the experiment on causal effect ratio, results in Figure \ref{fig:effect_ratio_influence}.}\label{apd:effect_ratio_distribution}
+\end{table}
 
 \bibliography{biblio}
-%\printbibliography
 

### 2024-10-19 response_to_reviewer.md
iter

@@
+We thank the reviewers for their time, thoughts, and feedback on the
+manuscript. We have addresed the comments through additional work as well
+as clarifications in the revised manuscript.
+
 # Reviewer 1
 
-I appreciate the approach and the level of detail that this paper offers on an issue that is, to say the least, a hot topic in bioinformatics and data science in general: the evaluation of the best 'error' model when choosing a causal model for predictive purposes.
-Indeed, there is much confusion on the subject and very often error models are applied without any real awareness of their potential and limitations.
-
-That is why I think the paper can be of great educational value as well as research.
-
-Of particular interest is the statement "the best predictor may not estimate best causal effects".
-Central to the understanding of the paper is Figure 1 and the concept it seeks to explain.
-The figure clearly shows that if we rely on R^2, a predictor is preferred which on average is efficient but not on estimating causal quantities.
-The authors conceivably state that "The problem is that causal estimation requires controlling an error on both treated and non-treated outcome for the same individual.".
+*[...] The authors conceivably state that "The problem is that causal estimation requires controlling an error on both treated and non-treated outcome for the same individual.".
 Consequently, they report on estimator that attempt to consider this (in Section "Methods").
@@
 (2) we see the g-computation estimators, a well-known concept introduced by [21] (and recently used also by https://www.nature.com/articles/s41598-020-65917-x ), that splits the two configurations "treated" and "non-treated".
-
-Other reference on g-computation to consider in the study can be for example:
-
--       [Le Borgne et al. 2021](https://www.nature.com/articles/s41598-021-81110-0)
--       https://www.sciencedirect.com/science/article/pii/S0167947306002076
--       https://bmcmedresmethodol.biomedcentral.com/articles/10.1186/s12874-023-01835-6
--       https://pubmed.ncbi.nlm.nih.gov/34861799/
-
-I understand that Eq.
+[...] I understand that Eq.
 (2) and in Eq.
@@
 (3) derived from Eq.
-(2) under special assumptions?
+(2) under special assumptions?*
+
+> The reviewer is right that Eq. (2) is part of constructing the classic
+> g-computation estimator, as well as Eq. (1) and Eq. (3). We added a
+> sentence at the begining of the section to stress that it is a section
+> summarizing a classic framework.
 
@@
 (2).
+
 > G-estimation itself does not control error on both populations as highlighted in Figure 1.

### 2024-10-19 response_to_reviewer.md
iter

@@
 > Eq.
-(3) is the definition of the estimator of the ATE based on a model of the outcome.
-It is sometimes called g-estimation.
-It  is unbiased under under consistency (Assumption.
+(3) is an expression of the estimator of the ATE based on a model of the outcome, sometimes called g-estimation.
+It is unbiased under consistency (Assumption.
 3) and unconfoundedness (Assumption 1).
@@
 
-> G-estimation itself does not control error on both populations as highlighted in Figure 1.
-Specific inductive biases of ML models are able to favor such regularization(see for example [Johannson et al., 2022](https://www.jmlr.org/papers/volume23/19-511/19-511.pdf)).
-Selection procedures, more elaborated than RMSE are needed to favor such models.
-I added this last explanation in the method (par.
-Model-selection risks, oracle and feasible) trying to clarify the link between the hint on error control on both population, g-estimation and the selection procedure which is our main focus.
+> G-estimation itself does not control error on both populations.
+The formulas in Eq.
+(1-3) are all partly oracle formulas: they rely on conditional expectations but give no specific procedures on how to compute them.
+We added a clarification on this topic above Eq.
+(4) and above Eq.
+(8).
 

### 2024-10-19 how_to_select_latex.tex
iter

@@
 
+We first expose the classic construction of the outcome modeling (or
+G-computation) estimators of causal effect
+\cite{robins_new_1986,snowden_implementation_2011,hernan_causal_2020}.
+
 \paragraph{Settings}
@@
 the consistency \ref{assumption:consistency} and unconfoundedness
-\ref{assumption:ignorability} assumptions imply the following equality, linking the target potential quantity to the observed data:
+\ref{assumption:ignorability} assumptions imply the following
+equality, linking the target potential quantity to the observed data:
 \begin{equation}\label{eq:mu_identification}
@@
 $\hat \ell$ to note its finite sum version. 
-In practice the response function can be estimated with one statistical model taking the treatment as a supplementary covariate $\hat \mu(x, a)$ (called S-learner) or by two separate model fitted on each population, treated or not $(\hat \mu_0(x), \hat \mu_1(x))$ (T-learner) \cite{kunzel_metalearners_2019}.
+The formulas in Eq. (1-3) are all partly oracle formulas: they rely on
+conditional expectations, the response functions, but give no specific
+procedures on how to compute or select them. This last point is the topic
+of our work, describe in the next section.
+%In practice the response function can be estimated with one statistical model taking the treatment as a supplementary covariate $\hat \mu(x, a)$ (called S-learner) or by two separate model fitted on each population, treated or not $(\hat \mu_0(x), \hat \mu_1(x))$ (T-learner) \cite{kunzel_metalearners_2019}.
 
@@
 \paragraph{Robinson decomposition}
-G-computation is a choice of decomposition of the CATE estimation into a
-sequence of regressions, where the only statistical estimate is the response
-function $\mu_{a}(x)$ \cite[chapter 15]{chernozhukov2024applied}. Other choices
+G-computation is a choice of decomposition of the CATE estimation. Other choices
 of decomposition exist, such as the R-decomposition
@@
 %
-However, g-computation itself does not control error on both populations as highlighted in Figure \ref{fig:toy_example}.
-Specific inductive biases of ML models are able to favor such regularization \cite{johansson2022generalization}.
-Thus, selection procedures, more elaborated than MSE are needed to favor the best model for each dataset.
+However, the g-computation framework presented above is written in terms
+of ``perfect'' conditional expectations (oracles), it does not control an
+error, \emph{eg} on both populations as highlighted in Figure
+\ref{fig:toy_example}. Selection procedures are needed to favor the best model for each dataset.
+
 A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ with observed data $O$ and a splitting strategy of the data for estimating the different regressions (nuisances) involved in the risk.

### 2024-10-20 response_to_reviewer.md
iter

@@
 
-Furthermore, the Robinson decomposition (R-decomposition) should be better discussed and its role in the context of causal inference should be better commented.
+*Furthermore, the Robinson decomposition (R-decomposition) should be better discussed and its role in the context of causal inference should be better commented.
 R-decomposition introduces two quantities, the conditional mean outcome and the probability to be treated (known as propensity score).
@@
 How is it calculated?
-Is it calculated simply by a frequentist approach (number fo treated/total number of patients?
+Is it calculated simply by a frequentist approach (number of treated/total number of patients)?*
 
-> G-computation is one choice of decomposition of the CATE estimation into a sequence of regressions, where the only statistical estimate is the response function µa(x)  [Chernozhukov et al., 2024, chapter 15](https://causalml-book.org/).
-Other choices of decomposition exist, such as the R-decomposition, the Double robust estimation, the X-learner...
+> G-computation relies on one choice of decomposition of the CATE estimation (Eq.
+(2)), which leads to into a sequence of regressions, where the only statistical estimate is the response function µa(x)  [Chernozhukov et al., 2024, chapter 15](https://causalml-book.org/).
+Other decomposition of the outcome exist, as a decomposition is just a mathematical rewriting...
 The R-decomposition is one way of identifying the causal quantity of interest which introduce two statistical quantity that have to be modelled: the mean outcome from the treatment effect and the propensity score.

### 2024-10-20 how_to_select_latex.tex
iter

@@
 \ref{assumption:ignorability} assumptions imply the following
-equality, linking the target potential quantity to the observed data:
+equality, linking the target quantity to the observed data:
 \begin{equation}\label{eq:mu_identification}
@@
 error, \emph{eg} on both populations as highlighted in Figure
-\ref{fig:toy_example}. Selection procedures are needed to favor the best model for each dataset.
-
-A selection procedure is defined by a risk $\ell$ evaluating the quality of a model $f$ with observed data $O$ and a splitting strategy of the data for estimating the different regressions (nuisances) involved in the risk.
+\ref{fig:toy_example}. Selection procedures are needed to find the best
+conditional-expectation models.
+
+A selection procedure combines a risk $\ell$, evaluating the quality of a
+model $f$ with observed data $O$, and a splitting strategy of the data to
+estimate different regressions (nuisances) involved in the risk.
 %

### 2024-10-20 response_to_reviewer.md
iter

@@
 as clarifications in the revised manuscript.
+
+To clarify the origin of the Robinson decomposition, as suggested by
+reviewer 1, we have added an equation. As a result the equation numbers
+in the revised manuscript are all shifted by one.
 
@@
 > Eq.
-(3) is an expression of the estimator of the ATE based on a model of the outcome, sometimes called g-estimation.
+(3) (now 4) is an expression of the estimator of the ATE based on a model of the outcome, sometimes called g-estimation.
 It is unbiased under consistency (Assumption.
@@
 Then apply  Eq.
-(1) which is true under consistency (Assumption 3) and unconfoundedness (Assumption 1).
+(1) (now 2) which is true under consistency (Eq.
+1 and Assumption 1) and unconfoundedness (Assumption 3).
 Then by the law of total expectancy, the conditionnality on X disappear and we recover Eq.
-(2).
+(2) (now 3)
 
@@
 The formulas in Eq.
-(1-3) are all partly oracle formulas: they rely on conditional expectations but give no specific procedures on how to compute them.
+(1-3) (now 2-4) are all partly oracle formulas: they rely on conditional expectations but give no specific procedures on how to compute them.
 We added a clarification on this topic above Eq.
-(4) and above Eq.
-(8).
+(4) (now 5) and above Eq.
+(8) (now 9).
 

### 2024-10-20 how_to_select_latex.tex
iter

@@
 A given data needs to meet a few assumptions to enable identifying
-causal estimands \cite{rubin_causal_2005}. The usual
-\emph{strong ignorability} assumptions are  (details in
-\ref{apd:causal_assumptions}): \emph{1)}
+causal estimands \cite{rubin_causal_2005}. \emph{1)} an individual's
+outcome $Y$ is solely governed by the corresponding potential outcome:
+\begin{flalign}
+    \text{\emph{Consistency} assumption,}
+    &&
+    Y = A\, Y(1) + (1 - A)\, Y(0)
+    \label{eq:consistency}
+    &&
+\end{flalign}
+\emph{2)}
 \emph{unconfoundedness} \mbox{$\{Y(0),
-        Y(1) \} \indep A | X$}, \emph{~2)} \emph{strong overlap} ie. every patient has a
-strictly positive probability to receive each treatment, \emph{3)}
-\emph{consistency}, and \emph{4)} \emph{generalization}.
+        Y(1) \} \indep A | X$}, \emph{~3)} \emph{strong overlap} ie. every patient has a
+strictly positive probability to receive each treatment, 
+and \emph{4)} \emph{generalization} --no covariate shift.
+These classic assumptions, called
+\emph{strong ignorability}, are formally detailed in
+\ref{apd:causal_assumptions}.
 
@@
 $\hat \ell$ to note its finite sum version. 
-The formulas in Eq. (1-3) are all partly oracle formulas: they rely on
+The formulas in Eq. (\ref{eq:mu_identification}-\ref{eq:ate_estimate}) are all partly oracle formulas: they rely on
 conditional expectations, the response functions, but give no specific
@@
 \end{flalign}
-the outcome can be written
+with these, the outcome (Eq.\,\ref{eq:consistency}) can be written
 \begin{flalign}\label{eq:r_decomposition}
@@
 data \cite{rubin_causal_2005}:
-\begin{assumption}[Unconfoundedness]\label{assumption:ignorability}
-    \begin{equation*}\label{eq:ignorability}
-        \{Y(0), Y(1) \} \indep A | X
-    \end{equation*}
-    This condition --also called ignorability-- is equivalent to the conditional
-    independence on $e(X)$ \cite{rosenbaum_central_1983}: $\{Y(0), Y(1) \}
-        \indep  A | e(X)$.
-\end{assumption}
-
-\begin{assumption}[Overlap, also known as Positivity)]\label{assumption:overlap}
-    \begin{equation*}\label{eq:overlap}
-        \eta < e(x) < 1 - \eta \quad \forall x \in \mathcal X \text{ and some } \eta > 0
-    \end{equation*}
-    The treatment is not perfectly predictable. Or with different words, every
-    patient has a chance to be treated and not to be treated. For a given set of
-    covariates, we need examples of both to recover the ATE.
-\end{assumption}
-
-As noted by \cite{damour_overlap_2020}, the choice of covariates $X$ can
-be viewed as a trade-off between these two central assumptions. A bigger
-covariates set generally reinforces the ignorability assumption. In the
-contrary, overlap can be weakened by large $\mathcal{X}$ because of the
-potential inclusion of instruments: variables only linked to the treatment which
-could lead to arbitrarily small propensity scores.
-
-% remark: There is a major counter example to these colliders (variables which
-% are caused by both the outcome and the treatment),
-
 \begin{assumption}[Consistency]\label{assumption:consistency} The observed
@@
 \end{assumption}
+
+\begin{assumption}[Unconfoundedness]\label{assumption:ignorability}
+    \begin{equation*}\label{eq:ignorability}
+        \{Y(0), Y(1) \} \indep A | X
+    \end{equation*}
+    This condition --also called ignorability-- is equivalent to the conditional
+    independence on $e(X)$ \cite{rosenbaum_central_1983}: $\{Y(0), Y(1) \}
+        \indep  A | e(X)$.
+\end{assumption}
+
+\begin{assumption}[Overlap, also known as Positivity)]\label{assumption:overlap}
+    \begin{equation*}\label{eq:overlap}
+        \eta < e(x) < 1 - \eta \quad \forall x \in \mathcal X \text{ and some } \eta > 0
+    \end{equation*}
+    The treatment is not perfectly predictable. Or with different words, every
+    patient has a chance to be treated and not to be treated. For a given set of
+    covariates, we need examples of both to recover the ATE.
+\end{assumption}
+
+As noted by \cite{damour_overlap_2020}, the choice of covariates $X$ can
+be viewed as a trade-off between these two central assumptions. A bigger
+covariates set generally reinforces the ignorability assumption. In the
+contrary, overlap can be weakened by large $\mathcal{X}$ because of the
+potential inclusion of instruments: variables only linked to the treatment which
+could lead to arbitrarily small propensity scores.
+
+% remark: There is a major counter example to these colliders (variables which
+% are caused by both the outcome and the treatment),
 

### 2024-10-20 response_to_reviewer.md
iter

@@
 
-> G-computation relies on one choice of decomposition of the CATE estimation (Eq.
-(2)), which leads to into a sequence of regressions, where the only statistical estimate is the response function µa(x)  [Chernozhukov et al., 2024, chapter 15](https://causalml-book.org/).
-Other decomposition of the outcome exist, as a decomposition is just a mathematical rewriting...
-The R-decomposition is one way of identifying the causal quantity of interest which introduce two statistical quantity that have to be modelled: the mean outcome from the treatment effect and the propensity score.
-Any statistical estimator can be chosen to model these quantities.
-The frequentist approach is one (naive) such estimator.
-In practice, researchers often estimate the propensity score with a logistic regression, but other methods such as Boosted trees, Random Forests, or Neural Networks can be used.
-A good ps estimator should be well calibrated but this question is out of scope for this paper.
-In our discussion, we briefly mentioned this interesting research question on what properties should have a good estimator for the propensity score.
+> G-computation relies on one choice of decomposition of the CATE
+> estimation (Eq. (2)), which leads to into a sequence of regressions,
+> where the only statistical estimate is the response function µa(x)
+> [Chernozhukov et al., 2024, chapter 15](https://causalml-book.org/). We
+> have clarified that this decomposition is just a mathematical
+> rewriting. The mean outcome from the treatment effect and the
+> propensity score are two statistical quantity that enable this
+> rewritting . Any statistical estimator can be chosen to model these
+> quantities, in particular in the section were we introduce them, they
+> are written in term of expectations, not detailing an estimation
+> procedure, as this is discussed later. We have clarfied this below
+> formula 4 (formely 3) and in the begining of the section containing
+> formula 9 (formely 8). A frequentist approach counting rations of
+> patients is one such estimator, but it will work only for a small
+> number of discrete covariate. In practice, researchers often estimate
+> the propensity score with a logistic regression, but other methods such
+> as Boosted trees, Random Forests, or Neural Networks can be used. A
+> good ps estimator should be well calibrated but this question is out of
+> scope for this paper. In our discussion, we briefly mentioned this
+> interesting research question on what properties should have a good
+> estimator for the propensity score.
 
-Finally, I note that the paper mentions inference methods specific to machine learning (random forest, etc.).
+*Finally, I note that the paper mentions inference methods specific to machine learning (random forest, etc.).
 In this regard, there is a great deal of study in the literature as to their actual usefulness in this field.
-I suggest the authors consider/review some of the following literature in the paper (those that the authors consider most appropriate to be included in their manuscript):
+I suggest the authors consider/review some of the following literature in the paper (those that the authors consider most appropriate to be included in their manuscript):*
 
@@
 
-> Thanks for the references.
-I added two references for Causal ML into the manuscript: [Chernozhukov et al., 2024](https://causalml-book.org/) and [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475).
+> We thank the reviewer for the references.
+We have added two references for Causal ML into the manuscript: [Chernozhukov et al., 2024](https://causalml-book.org/) and [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475).
 Both of them cover most of the subfields at the intersection of causal inference and machine learning.

### 2024-10-20 response_to_reviewer.md
Finish address rev 1

@@
 > are written in term of expectations, not detailing an estimation
-> procedure, as this is discussed later. We have clarfied this below
+> procedure, as this is discussed later. We have clarified this below
 > formula 4 (formely 3) and in the begining of the section containing
-> formula 9 (formely 8). A frequentist approach counting rations of
+> formula 9 (formely 8). A frequentist approach counting ratios of
 > patients is one such estimator, but it will work only for a small
@@
 
-https://link.springer.com/book/10.1007/978-3-031-35051-1
+*https://link.springer.com/book/10.1007/978-3-031-35051-1*
 
-https://www.causal-machine-learning.com/
+*https://www.causal-machine-learning.com/*
 
-https://www.frontiersin.org/journals/bioinformatics/articles/10.3389/fbinf.2021.746712/full
+*https://www.frontiersin.org/journals/bioinformatics/articles/10.3389/fbinf.2021.746712/full*
 
-https://www.research-collection.ethz.ch/handle/20.500.11850/650703
+*https://www.research-collection.ethz.ch/handle/20.500.11850/650703*
 
-https://aisel.aisnet.org/pacis2022/181/
+*https://aisel.aisnet.org/pacis2022/181/*
 
-or some other that can be easily found in literature.
+*or some other that can be easily found in literature.*
 
-> We thank the reviewer for the references.
-We have added two references for Causal ML into the manuscript: [Chernozhukov et al., 2024](https://causalml-book.org/) and [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475).
-Both of them cover most of the subfields at the intersection of causal inference and machine learning.
-I have added them at the beginning of the manuscript before stating our focus on model selection for CATE.
+> We thank the reviewer for the references. We have added several
+> references for Causal ML into the manuscript, in particular [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475), corresponding to
+> https://www.causal-machine-learning.com/, and [Chernozhukov et al., 2024](https://causalml-book.org/) which both cover most of the
+> subfields at the intersection of causal inference and machine learning.
 

### 2024-10-20 how_to_select_latex.tex
Finish address rev 1

@@
 $m$ and $e$ are often called
-\emph{nuisances} \cite{chernozhukov_double_2018}. They are unknown and are estimated in practice thanks to statistical models (eg. linear models or forest based estimators).
+\emph{nuisances} \cite{chernozhukov_double_2018}. They are unknown and must be estimated from the data.
+
+\bigskip
+
+Both the in the ATE and CATE formula Eq.\,(\ref{eq:ate_estimate},
+\ref{eq:cate_estimate}), and the Robinson decomposition involve
+conditional expectations --the response functions $\mu_a(x)$ or the
+nuisances $m(x)$ and $e(x)$. In practice those are given by statistical
+models: linear models, random forests, etc
+\cite{kaddour2022causal,chernozhukov2024applied}.
 

### 2024-10-20 response_to_reviewer.md
iter

@@
 
-Overall assessment:
+*Overall assessment:*
 
-Let me begin with positive comments.  I think the paper is very important because it describes a sophisticated methodology for experimental examination of causal models.
+*Let me begin with positive comments.  I think the paper is very important because it describes a sophisticated methodology for experimental examination of causal models.
 Despite some attempts this is still lacking in literature.
@@
 Moreover thorough experimental evaluations of causal models are not very frequent making the paper important.
-Code is available on Github which is a big plus.
+Code is available on Github which is a big plus.*
 
-I am however not sure about the overall conclusions, which are that R-risk is always the best.
+*I am however not sure about the overall conclusions, which are that R-risk is always the best.
 Performance of risk functions may depend on several factors and the overall best performance might be a result of bias in available data.
@@
 Another possible factor is the ratio of causal effect to background responses.
-It is known that it may affect relative performance of HTE estimators.
+It is known that it may affect relative performance of HTE estimators.*
 
-Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.
+*Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.*
 
-> We thank the reviewer for this remark and the suggestion to explore more diverse simulation scenarii.
-Exploring the sensitivity of our result on the R-risk appears to us particularly necessary.
-Thus, we ran a dedicated experiment focusing on the ratio of causal effect to background response.
-We measure the ratio by the absolute mean difference between the causal effect and the baseline response, measured empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$.
-We vary a dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this ratio to cover a wide range of values.
-In 90 dataset instance, it ranges several orders of magnitude from 0.005 to 9675.
-Figure 7 gives the main results of this experiment.
-We see that R-risk is still the best risk on average, but when the effect ratio is high (tertile between 9 and 9675), the reweighted mu-risk and the the tau-risk_IPW are better than the R-risk.
-However, for low value of the ratio,  the R-risk largely outperforms all other metrics.
-We added these new results to the main paper.
-We also nuanced our conclusion on the R-risk, underlying that it is better on average but that in some situations, other risks might be more adapted.
-A typical such situation would be when the applied scientist is interested in estimating an effect with a big ratio and want to focus on the part of the population where there is low overlap.
+> We thank the reviewer for the thoughful comments and the call for
+> nuance. Our first response is that our thoughts did not come out right
+> and that it is not clear, indeed, that R-risk is always the best.
+> Rather, that it is a good default solution. We have rewritten parts of
+> the paper to convey this nuance better.
+
+> In addition, the reviewer is correct that factors beyond overlap
+> probably affect which risk is best. We have followed the suggestion to
+> explore more diverse simulation scenarii, as detailing the sensitivity
+> of our result on the R-risk makes a more nuanced and stronger analysis.
+> Thus, we ran a dedicated experiment focusing on the ratio of causal
+> effect to background response. We measure the ratio by the absolute
+> mean difference between the causal effect and the baseline response,
+> measured empirically on each dataset instance $\Delta_{\mu} =
+> \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) -
+> \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a dedicated parameter
+> ($\omega$ in Algorithm 2) in the simulation to force this ratio to
+> cover a wide range of values. In 90 dataset instance, it ranges several
+> orders of magnitude from 0.005 to 9675. Figure 7 gives the main results
+> of this experiment. We see that R-risk is still the best risk on
+> average, but when the effect ratio is high (tertile between 9 and
+> 9675), the reweighted mu-risk and the the tau-risk_IPW are better than
+> the R-risk. However, for low value of the ratio,  the R-risk largely
+> outperforms all other metrics. We added these new results to the main
+> paper. We also nuanced our conclusion on the R-risk, underlying that it
+> is better on average but that in some situations, other risks might be
+> more adapted. A typical such situation would be when the applied
+> scientist is interested in estimating an effect with a big ratio and
+> want to focus on the part of the population where there is low overlap.
 
@@
 
-Another question: do the differences between risks in experiments follow known theoretical results?
+*Another question: do the differences between risks in experiments follow known theoretical results?
 For example, it is well known (e.g.
-https://www.semanticscholar.org/reader/112aba9c62f98b9684693f4c20c30274bc9eb1dc or [Rudas ans jaroszewisz, 2017](https://link.springer.com/content/pdf/10.1007/s10618-018-0576-8.pdf)) that the tau risk performs poorly.
+https://www.semanticscholar.org/reader/112aba9c62f98b9684693f4c20c30274bc9eb1dc or [Rudas ans jaroszewisz, 2017](https://link.springer.com/content/pdf/10.1007/s10618-018-0576-8.pdf)) that the tau risk performs poorly.*
 
@@
 
-Another problem is that the paper is hard to follow.
+*Another problem is that the paper is hard to follow.
 The authors rapidly switch between topics: risk, selection, nuisance parameter estimation so the reader quickly gets lost.
@@
 The paper gives an impression of several experiments put together in some order without guiding the reader through the maze.
-I think organization should be improved before publication.
+I think organization should be improved before publication.*
 
@@
 
-In the Methods section g-computation is described.
+*In the Methods section g-computation is described.
 It seems indeed the first name under which the technique appeared in literature, but other names, most importantly S-learner are being used frequently nowadays.
-Interestingly those names are mentioned in Empirical study on page 4, however, the close relationship (equivalence?) between S-learner and g-computation is not mentioned.
+Interestingly those names are mentioned in Empirical study on page 4, however, the close relationship (equivalence?) between S-learner and g-computation is not mentioned.*
 
@@
 
-Equation 2: should the expectation be over X not Y?
+*Equation 2: should the expectation be over X not Y?*
 
@@
 
-page 17: proposition number is missing
+*page 17: proposition number is missing*
 
@@
 
-A.7: the data generation procedure needs better explanations.
+*A.7: the data generation procedure needs better explanations.
 Why not call (7) P(d|A) since that's how its used later?
 On page 17 you say '[b1,bD]~P(x)' and later 'x~P(x)' which causes confusion.
-Also, how are beta vectors generated?
+Also, how are beta vectors generated?*
 
@@
 
-page 4: TLearner uses different feature representation for each treatment arm.  Does this really happen in practice?  Usually all subgroups are described with the same set of features.
+*page 4: TLearner uses different feature representation for each treatment arm.  Does this really happen in practice?  Usually all subgroups are described with the same set of features.*
 
@@
 
-page 4/5: Family of candidate estimators: why are only linear models used in some dataset and forests for other?
-The leaf sizes of forests are chosen from a rather small range, why not use some percentage of the size of the training set?
+*page 4/5: Family of candidate estimators: why are only linear models used in some dataset and forests for other?
+The leaf sizes of forests are chosen from a rather small range, why not use some percentage of the size of the training set?*
 
@@
 
-It seems that English might need improvement in some cases, for example:
+*It seems that English might need improvement in some cases, for
+example:*
 
-Abstract: "We study such a variety of model selection procedures in practical settings": a "large variety" seems more appropriate.
+*Abstract: "We study such a variety of model selection procedures in practical settings": a "large variety" seems more appropriate.*
 
-On page 1, the sentence "Predictive modeling bridges to causal inference theory under the name of outcome models ..." contains "bridges under the name" which also sounds strange.
+*On page 1, the sentence "Predictive modeling bridges to causal inference theory under the name of outcome models ..." contains "bridges under the name" which also sounds strange.*
 
-On page 2: In the sentence "Then we anchor causal model selection in the potential outcome framework and details the causal risks and model-selection procedure." I believe 'details' should be 'detail'.
+*On page 2: In the sentence "Then we anchor causal model selection in the potential outcome framework and details the causal risks and model-selection procedure." I believe 'details' should be 'detail'.*
 
-Algorithm 1: 'a candidate estimators f' should be 'a candidate estimator f'
+*Algorithm 1: 'a candidate estimators f' should be 'a candidate estimator f'*
 
-throughout: "an RBF expansion" is more commonly used than "a RBF expansion"
+*throughout: "an RBF expansion" is more commonly used than "a RBF expansion"*
 
-page 4: 'for both population' should be 'for both populations'
+*page 4: 'for both population' should be 'for both populations'*
 

### 2024-10-20 how_to_select_latex.tex
iter

@@
 
-The simulations have been designed to evaluate the effect of the overlap parameter.
-The semi-simulated datasets have been included to explore more diverse and noisy covariate distributions.
-These simulations cover a wide range of causal settings such as different ratio of causal effect to background responses, and functional links between covariates, outcome and treatment.
+The simulations, designed to evaluate the effect of the overlap
+parameter, also explore more diverse and noisy covariate distributions.
+They cover a diversity of causal settings such as different ratio of causal effect to background responses, and functional links between covariates, outcome and treatment.
 
@@
 model development is based on treated and
-untreated population with little differences, as in RCTs. To facilitate better model selection, we provide Python
+untreated population with little differences, as in RCTs. 
+%
+Our conclusions are that without prior knowledge, the R-risk is a good
+default. However, there is much remaining variation, and the R-risk will
+not be optimal for every situation. We have identified one such specific
+situation: when the causal effect is large compared to the variation of
+the baseline effect.
+
+%If one possesses some prior knowledge on the data generation process,
+%another risk might be more adapted. This is typically the case for big causal
+%effect ratio to baseline and low overlap where the $\tau\text{-risk}_{IPW}$
+%should be better.
+
+To facilitate better model selection, we provide Python
 code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
@@
 models.
-
-\paragraph{Limitations}
-
-On average, the R-risk outperforms other risks, but it is not optimal for every
-situations. If one possesses some prior knowledge on the data generation process,
-another risk might be more adapted. This is typically the case for big causal
-effect ratio to baseline and low overlap where the $\tau\text{-risk}_{IPW}$
-should be better.
 
@@
     %\hfill
-    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$
+    \caption{\textbf{The $R$-risk is the best metric on average}: Relative Kendall's $\tau$
         agreement with $\tau\text{-risk}$. Strong and Weak overlap correspond to

### 2024-10-20 how_to_select_latex.tex
iter

@@
 situation: when the causal effect is large compared to the variation of
-the baseline effect.
+the baseline effect where the $\tau\text{-risk}_{IPW}$ should be better.
 

### 2024-10-20 how_to_select_latex.tex
iter

--- 2024-10-20 how_to_select_latex.tex
@@
 situation: when the causal effect is large compared to the variation of
-the baseline effect.
+the baseline effect, the $\tau\text{-risk}_{IPW}$ performs slightly
+better.
 

### 2024-10-20 response_to_reviewer.md
iter

@@
 > and that it is not clear, indeed, that R-risk is always the best.
-> Rather, that it is a good default solution. We have rewritten parts of
-> the paper to convey this nuance better.
+> Rather, that it is a good default solution. We have reformulate parts of
+> the paper to convey this nuance better and added the nuance in the
+> conclusion.
 
@@
 > outperforms all other metrics. We added these new results to the main
-> paper. We also nuanced our conclusion on the R-risk, underlying that it
-> is better on average but that in some situations, other risks might be
-> more adapted. A typical such situation would be when the applied
-> scientist is interested in estimating an effect with a big ratio and
-> want to focus on the part of the population where there is low overlap.
+> paper.
 

### 2024-10-20 response_to_reviewer.md
iter

@@
 > average, but when the effect ratio is high (tertile between 9 and
-> 9675), the reweighted mu-risk and the the tau-risk_IPW are better than
+> 9675), the reweighted mu-risk and the tau-risk_IPW are better than
 > the R-risk. However, for low value of the ratio,  the R-risk largely

### 2024-11-10 response_to_reviewer.md
update causal effect xp with the symmetric measure of causal effect.

@@
 > \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) -
-> \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a dedicated parameter
+> \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_{0}(x_i) + \mu_{1}(x_i)} \big|$. We vary a dedicated parameter
 > ($\omega$ in Algorithm 2) in the simulation to force this ratio to
 > cover a wide range of values. In 90 dataset instance, it ranges several
-> orders of magnitude from 0.005 to 9675. Figure 7 gives the main results
+> orders of magnitude from 0.04 to 206. Figure 7 gives the main results
 > of this experiment. We see that R-risk is still the best risk on
@@
 > paper.
+We also studied the effect of the variance of the effect measured as the empirical version of $\frac{var(\mu_1 - \mu_0)}{ var(\mu_1+\mu_0)}$.
+For this last experiment, we found similar results to the causal effect experiment.
 

### 2024-11-10 how_to_select_latex.tex
update causal effect xp with the symmetric measure of causal effect.

--- 2024-10-20 how_to_select_latex.tex
@@
     \text{\emph{Consistency} assumption,}
-    &&
+     &  &
     Y = A\, Y(1) + (1 - A)\, Y(0)
     \label{eq:consistency}
-    &&
+     &  &
 \end{flalign}
@@
 of different causal effect ratio to baseline. We vary the empirical mean
-absolute difference between the causal effect and the baseline, $\Delta_{\mu} =
-    \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)}
-    \big|$, covering a ratio range from 0.05 to 1000 ($\text{median}=1.9$, full
+absolute difference between the causal effect and the baseline, $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_0(x_j) + \mu_1(x_j)} \big |$, covering a ratio range from 0.04 to 206 ($\text{median}=9.1$, full
 distribution in Appendix \ref{apd:effect_ratio}). Figure
@@
     \begin{minipage}{\linewidth}
-        \includegraphics[width=0.9\linewidth]{_6_effect_ratio_by_bin_comparaison_kendall_by_Dataset.pdf}
+        \includegraphics[width=0.9\linewidth]{_6_effect_ratio_sym_by_bin_comparaison_kendall_by_Dataset.pdf}
     \end{minipage}
@@
     effect ratio are the tertiles of the absolute ratio causal effect to
-    baseline response, $\Delta_{\mu}$: Low $[0.005;0.4[$, Medium $[0.44; 9[$,
-    High $[9;9675[$. Appendix \ref{apd:effect_ratio} details this
+    baseline response, $\Delta_{\mu}$: Low $[0.04;2.86[$, Medium $[2.86; 16.65[$,
+    High $[16.65;206.53[$. Appendix \ref{apd:effect_ratio} details this
     simulation.}\label{fig:effect_ratio_influence}
@@
 situation: when the causal effect is large compared to the variation of
-the baseline effect.
+the baseline effect, the $\mu\text{-risk}_{IPW}$ performs slightly
+better.
 
@@
 data generation process: the ratio of causal effect to the baseline response.
+Here, we detail the experimental setup of the results presented in Figure \ref{fig:effect_ratio_influence}.
 
@@
 difference between the causal effect and the baseline response, measured
-empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N
-    \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a
+empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_0(x_j) + \mu_1(x_j)} \big |$.
+We vary a
 dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this
@@
                                    & \textbf{count} & \textbf{mean} & \textbf{std} & \textbf{min} & \textbf{1\%} & \textbf{10\%} & \textbf{25\%} & \textbf{50\%} & \textbf{75\%} & \textbf{90\%} & \textbf{99\%} & \textbf{max} \\ \hline
-            \textbf{effect\_ratio} & 179            & 159.6         & 852.4        & 0.004674     & 0.004765     & 0.02483       & 0.1566        & 1.879         & 21.12         & 135.1         & 3394          & 9675         \\ \hline
+            \textbf{effect\_ratio} & 179            & 16.95         & 24.92        & 0.043        & 0.054        & 0.32          & 1.34          & 9.12          & 22.55         & 39.97         & 101.12        & 206.53       \\ \hline
         \end{tabular}

### 2024-11-10 response_to_reviewer.md
iter submission of revised manuscript

@@
 > paper.
-We also studied the effect of the variance of the effect measured as the empirical version of $\frac{var(\mu_1 - \mu_0)}{ var(\mu_1+\mu_0)}$.
+We also studied another characteristic of the simulation: the variance of the causal effect measured as the empirical version of $\frac{var(\mu_1 - \mu_0)}{var(\mu_1+\mu_0)}$.
 For this last experiment, we found similar results to the causal effect experiment.
+We include it in the supp mat for completeness about our reflexion.
 

### 2024-11-10 how_to_select_latex.tex
iter submission of revised manuscript

@@
 of different causal effect ratio to baseline. We vary the empirical mean
-absolute difference between the causal effect and the baseline, $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_0(x_j) + \mu_1(x_j)} \big |$, covering a ratio range from 0.04 to 206 ($\text{median}=9.1$, full
-distribution in Appendix \ref{apd:effect_ratio}). Figure
+absolute difference between the causal effect and the baseline, $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_0(x_j) + \mu_1(x_j)} \big |$, covering a ratio range from 0.04 to 206 ($\text{median}=9.1$).
+Appendix \ref{apd:effect_ratio} details this setup as well as an alternive measure of effect ratio.
+Figure
 \ref{fig:effect_ratio_influence} shows that for high values of the ratio,
@@
 data generation process: the ratio of causal effect to the baseline response.
+
 Here, we detail the experimental setup of the results presented in Figure \ref{fig:effect_ratio_influence}.
@@
 
+We also studied another measure, focusing on the ratio of the variance of the causal effect to the variance of the mean shared response ($\mu_0 + \mu_1$) of both population: $\frac{\hat{var}(\mu_{1}(x_i) - \mu_{0}(x_i))}{\hat{var}(\mu_{1}(x_i) + \mu_{0}(x_i))}$, covering a range from 0 to 11.08 with the full
+distribution described in Table \ref{table:effect_variance}. Figure
+\ref{fig:effect_variance_influence} shows that for high values of the variance,
+$R-\text{risk}$ is outperformed by the $\mu\text{-risk}_{IPW}$ and the
+$\tau\text{-risk}_{IPW}$. However, on average, the $R-\text{risk}$ is still the
+better risk.
+
+\begin{figure}[!tb]
+    \centering
+    \begin{minipage}{\linewidth}
+        \includegraphics[width=0.9\linewidth]{_6_effect_variance_by_bin_comparaison_kendall_by_Dataset.pdf}
+    \end{minipage}
+
+    \caption{\textbf{R-risk is robust to a wide range of effect variance}:
+    Kendall's $\tau$ agreement with $\tau$-risk. Strong, medium and Weak Causal
+    effect variance are the tertiles of the ratio of causal effect variance to
+    the mean outcome variance: Low $[0;0.04[$, Medium $[0.04; 1.04[$,
+    High $[1.04;11.08[$.}\label{fig:effect_variance_influence}
+\end{figure}
+
+\begin{table}[!ht]
+    \centering
+    \resizebox{\textwidth}{!}{ % Adjusts the table to fit the text width
+        \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
+            \hline
+                                      & \textbf{count} & \textbf{mean} & \textbf{std} & \textbf{min} & \textbf{1\%} & \textbf{10\%} & \textbf{25\%} & \textbf{50\%} & \textbf{75\%} & \textbf{90\%} & \textbf{99\%} & \textbf{max} \\ \hline
+            \textbf{effect\_variance} & 179            & 1.013         & 1.781        & 1.680e-5     & 1.734e-5     & 4.875e-4      & 6.449e-3      & 0.789         & 1.160         & 1.807         & 9.641         & 11.079       \\
+            \hline
+        \end{tabular}
+    }
+    \caption{Statistical summary of the causal effect variance.}
+    \label{table:effect_variance}
+\end{table}
+
 \bibliography{biblio}

### 2025-01-16 how_to_select_latex.tex
publication: add data availability section

@@
 
-Lists the following:
+\subsection{Source code to replicate the experiments of the paper}
+
+The following repository is self-contained to allow anyone to replicate all experiments of the paper.
+It contains different codes to cover the full content of the paper.
+Thus it requires a bit of work to install and run the code.
+
 \begin{itemize}
@@
 \end{itemize}
+
+\subsection{Standalone source code for the selection procedure}
+
+Code for cross-validation to select the best predictive model to guide decision on potential intervention.
+The procedure implemented here is described in details in Algorithm \ref{problem:estimation_procedure:algo}.
+It serves as a simple plug-and-play code for a user that would want to use our procedure.
+
+\begin{itemize}
+    \item Project name: Causal Model Selection
+    \item Project home page: \url{https://github.com/soda-inria/causal_model_selection}
+    \item Operating system(s): Platform independent
+    \item Programming language: Python
+    \item License: BSD 2-Clause License
+\end{itemize}
+
+\section{Data availability}
+
+\subsection{Semi-simulated datasets for the experiments}
+
+The semi-simulated and simulated datasets used for the experiments are available at the following urls.
+
+Detailed explanations on how to generate the datasets are available in the readme of the data section of our github: \url{https://github.com/soda-inria/caussim/tree/main/data}.
+
+\begin{itemize}
+    \item ACIC 2016: dataset provided through R package aciccomp2016: \url{https://github.com/vdorie/aciccomp}
+
+    \item ACIC 2018: scaling subset from link to Synapse : \url{https://www.synapse.org/Synapse:syn11738767}
+
+    \item TWINS: dataset from \url{https://github.com/AMLab-Amsterdam/CEVAE}, the version that we are using is on the following zenodo repository: \url{https://zenodo.org/records/14674618}.
+
+    \item Caussim Dataset: generated dataset - instructions from \url{https://github.com/soda-inria/caussim}
+\end{itemize}
+\subsection{Experiments result data}
+
+The result data generated by the experiments are available at a dedicated zenodo repository \url{https://zenodo.org/records/13765465}.
+We provide the data for the simulations and the semi-simulated datasets to allow an easy replication of the main graphic (Fig.
+\ref{fig:relative_kendalls_all_datasets}) in the result section (detailed steps on our github: \url{https://github.com/soda-inria/caussim/tree/main?tab=readme-ov-file#reports}).
 
@@
 \begin{equation}
-    \mu\text{-risk}_{IPW}(\mu) = \sigma_B^{2}(0) + \sigma_B^{2}(1)  \quad \text{from Lemma \ref{apd:proofs:mu_risk_ipw_link_mu}}
+    \mu\text{-risk}_{IPW}(\mu) = \sigma_B^{2}(0) + \sigma_B^{2}(1)
     %\notag

### 2025-02-07 how_to_select_latex.tex
prepare main_submission latex and pdf

@@
 %%% Use the \authfn to add symbols for additional footnotes, if any. 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
-\author[1,\authfn{1}]{Matthieu Doutreligne}
-\author[2]{Gaël Varoquaux}
+\author[1,\authfn{1}]{Matthieu Doutreligne (ORCID: 0000-0001-8072-9966)}
+\author[2]{Gaël Varoquaux (ORCID: 0000-0003-1076-5122)}
 
@@
 many different simulated data generation
-processes and three semi-simulated datasets \footnote{Scripts for the simulations and the selection procedure are available at
-    \url{https://github.com/soda-inria/caussim}.
-}.
+processes and three semi-simulated datasets.We provide scripts for the simulations \cite{doutreligne2025caussim} and the selection procedure \cite{doutreligne2025causal}.
 
@@
 boosting trees for both the response surfaces and the treatment
-\footnote{Scikit-learn regressor,
-    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingRegressor.html}{HistGradientBoostingRegressor},
-    and classifier,
-    \href{https://scikit-learn.org/stable/modules/generated/sklearn.ensemble.HistGradientBoostingClassifier.html}{HistGradientBoostingClassifier}.}
+(scikit-learn HistGradientBoostingRegressor HistGradientBoostingClassifier  \cite{pedregosa_scikitlearn_2011})
 with S-learner, learning rate
@@
 Figure \ref{fig:relative_kendalls_all_datasets} shows the agreement between the
-ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the different
-feasible causal metrics. We measure this agreement with relative\footnote{To
+ideal ranking of outcome models given the oracle $\tau\text{-risk}$ and the
+different feasible causal metrics. We measure this agreement with relative
+Kendall tau $\kappa$ (eq. \ref{eq:kendall_tau}) \cite{kendall_new_1938}. To
 remove the variance across datasets (some datasets lead to easier model
@@
     \kappa(\ell,\tau\mathrm{{-risk}})= \kappa(\ell,\tau\mathrm{{-risk}}) -
-        mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$} Kendall tau $\kappa$
-(eq. \ref{eq:kendall_tau}) \cite{kendall_new_1938}. Given the importance of
+    mean_{\ell}\big(\kappa(\ell,\tau\mathrm{{-risk}})\big)$. Given the importance of
 overlap in how well metrics approximate the oracle $\tau\text{-risk}$, we
@@
 To facilitate better model selection, we provide Python
-code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
+code \cite{doutreligne2025causal}.
 % Using the $R\text{-risk}$ does make evaluation more
@@
     \item Project name: Caussim
-    \item Project home page: \url{https://github.com/soda-inria/caussim}
+    \item Project home page: \cite{doutreligne2025caussim}
     \item Operating system(s): Platform independent
@@
     \item Project name: Causal Model Selection
-    \item Project home page: \url{https://github.com/soda-inria/causal_model_selection}
+    \item Project home page: \cite{doutreligne2025causal}
     \item Operating system(s): Platform independent
@@
 
-Detailed explanations on how to generate the datasets are available in the readme of the data section of our github: \url{https://github.com/soda-inria/caussim/tree/main/data}.
+Detailed explanations on how to generate the datasets are available in the readme of the data section of our github \cite{doutreligne2025caussim}.
 
 \begin{itemize}
-    \item ACIC 2016: dataset provided through R package aciccomp2016: \url{https://github.com/vdorie/aciccomp}
-
-    \item ACIC 2018: scaling subset from link to Synapse : \url{https://www.synapse.org/Synapse:syn11738767}
-
-    \item TWINS: dataset from \url{https://github.com/AMLab-Amsterdam/CEVAE}, the version that we are using is on the following zenodo repository: \url{https://zenodo.org/records/14674618}.
-
-    \item Caussim Dataset: generated dataset - instructions from \url{https://github.com/soda-inria/caussim}
+    \item ACIC 2016: dataset provided through R package aciccomp2016 \cite{dorie_automated_2019}
+
+    \item ACIC 2018: scaling subset from the Synapse repository : \cite{shimoni2018ibm}
+
+    \item TWINS: dataset from \cite{louizos_causal_2017}, our version is on a zenodo repository \cite{doutreligne2025twins}.
+
+    \item Caussim Dataset: generated dataset with the source code \cite{doutreligne2025caussim}.
 \end{itemize}
@@
 
-The result data generated by the experiments are available at a dedicated zenodo repository \url{https://zenodo.org/records/13765465}.
+The result data generated by the experiments are available at a dedicated zenodo repository \cite{doutreligne2024how}.
 We provide the data for the simulations and the semi-simulated datasets to allow an easy replication of the main graphic (Fig.
-\ref{fig:relative_kendalls_all_datasets}) in the result section (detailed steps on our github: \url{https://github.com/soda-inria/caussim/tree/main?tab=readme-ov-file#reports}).
+\ref{fig:relative_kendalls_all_datasets}) in the result section.
+
+\bibliography{biblio}
 
@@
 
-\subsubsection{Additional Results}
+\subsubsection{Additional Results}\label{apd:experiments:additional_results}
 
@@
 
-\bibliography{biblio}
-
 \end{document}

### 2025-02-07 main_submission.tex
prepare main_submission latex and pdf

--- 2025-02-07 how_to_select_latex.tex
@@
 These two potential outcomes can be estimated from observed data:
-the consistency \ref{assumption:consistency} and unconfoundedness
-\ref{assumption:ignorability} assumptions imply the following
+the consistency and unconfoundedness
+\ref{apd:causal_assumptions} assumptions imply the following
 equality, linking the target quantity to the observed data:
@@
                         \sim \mathcal D}\left[(Y-f(X ; A))^2 \right]$
-                                                                                                               & Def.
-\ref{def:mu_risk} \cite{schuler_comparison_2018}                                                                                               \\
+                                                                                                               & \cite{schuler_comparison_2018}                                                                                                                      \\
                 $\mu\text{-risk}_{IPW}^*$                                                                      & $\mathbb{E}_{(Y, X, A)
                         \sim \mathcal D}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big)
-                (Y-f(X ; A))^2 \right]$                                                                        & Def.
-                \ref{def:mu_ipw_risk} \cite{vanderlaan_unified_2003}                                                                                                                                                                                                 \\
+                (Y-f(X ; A))^2 \right]$                                                                        & \cite{vanderlaan_unified_2003}                                                                                                                      \\
                 $\tau\text{-risk}^{\star}_{IPW}$                                                               & $\mathbb{E}_{(Y, X, A) \sim \mathcal D} \left[ \Big(Y \left( \frac{A}{e(X)} - \frac{1-A}{1-e(X)}\right)-\hat \tau_f\left(X\right)\Big)^2 \right]$ &
-                Def. \ref{def:tau_ipw_risk} \cite{wager_estimation_2018}
+                \cite{wager_estimation_2018}
                 \\
@@
                 \big( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -  \hat \tau_f\left(X\right)\big)^{2} \big]$ &
-                Def. \ref{def:u_risk} \cite{nie_quasioracle_2017}
+                \cite{nie_quasioracle_2017}
                 \\
@@
                 -\left(A-e\left(X\right)\right) \hat \tau_f\left(X\right)\big)^{2} \big]$                      &
-                Def. \ref{def:r_risk} \cite{nie_quasioracle_2017}
+                \cite{nie_quasioracle_2017}
                 \\
@@
     $f$, its $R\text{-risk}$ appears as weighted version of its $\tau\text{-risk}$
-    (Proof in \ref{apd:proofs:r_risk_rewrite}):
+    (Proof in Appendix \ref{apd:proofs}):
     \begin{align}
@@
 making the $R\text{-risk}$ particularly suited for exploring
-heterogeneity (Appendix \ref{apd:theory:special_cases})
+heterogeneity (Appendix \ref{apd:proofs})
 
@@
 \cite{schuler_comparison_2018,alaa_validating_2019}, we use many
-generative processes, which is needed to reach general conclusions (Appendix \ref{apd:results:seed_effect}).
+generative processes, which is needed to reach general conclusions (Appendix \ref{apd:experiments:additional_results}).
 
@@
 ends of the feature space. We generate 1\,000 datasets based on these functions,
-with random overlap parameters. Example shown in Figure
-\ref{fig:simulation_examples} and details in \ref{apd:experiments:generation}.
+with random overlap parameters. Example shown in Appendix
+\ref{apd:experiments:additional_results}, Fig  13. and details in Appendix \ref{apd:experiments:additional_results}.
 
@@
 different treatment allocations. Appendix
-\ref{apd:experiments:datasets} gives datasets details.
+\ref{apd:experiments:additional_results} gives datasets details.
 
@@
 $\check e$ (respectivley $\check m$) with a meta-learner: a stacked estimator of ridge
-and boosting classifiers (respectively regressions) (hyperparameter selection in Appendix \ref{apd:experiments:nuisances_hp}).
+and boosting classifiers (respectively regressions) (hyperparameter selection in Appendix \ref{apd:experiments:additional_results}).
 
@@
 Good overlap between treated and control population is crucial for causal
-inference (Assumption \ref{assumption:overlap}). We introduce the
+inference (Assumption \ref{apd:causal_assumptions}). We introduce the
 Normalized Total Variation (NTV), a divergence based on the propensity score
@@
         Strong and Weak overlap correspond to the first and last tertiles of the overlap distribution measured with
-        Normalized Total Variation eq. \ref{eq:ntv}. \ref{apd:experiments:additional_results} presents the same results
-        by adding semi-oracle risks in Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics}, measured with
-        absolute Kendall's in Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} and with $\tau\mathrm{-risk}$
-        gains in Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}. Table
-        \ref{apd:table:relative_kendalls_all_datasets} gives median and
+        Normalized Total Variation (Appendix \ref{apd:motivation_ntv}). Appendix \ref{apd:experiments:additional_results} presents the same results
+        by adding semi-oracle risks in Figure 14, measured with
+        absolute Kendall's in Figure 15 and with $\tau\mathrm{-risk}$
+        gains in Figure 16. Table
+        4 gives median and
         IQR of the relative Kendall.}\label{fig:relative_kendalls_all_datasets}
@@
 different feasible causal metrics. We measure this agreement with relative
-Kendall tau $\kappa$ (eq. \ref{eq:kendall_tau}) \cite{kendall_new_1938}. To
+Kendall tau $\kappa$ (see Appendix \ref{apd:experiments:additional_results}, eq. 20) \cite{kendall_new_1938}. To
 remove the variance across datasets (some datasets lead to easier model
@@
 --as in some causal-inference methods \cite{athey2016recursive,gutierrez_causal_2016}--leads to 100\% excess risk
-(Figure
-\ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}). Across
+(Appendix \ref{apd:experiments:additional_results}, Figure 16). Across
 datasets, the $R\text{-risk}$ consistently decreases the
@@
 (excellent agreement with oracle selection) to 0.6 on both Caussim and ACIC 2018
-(\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}).
+(Appendix \ref{apd:experiments:additional_results}, Figure 15).
 
@@
         Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
-        are the tertiles of the overlap measured with NTV eq. \ref{eq:ntv}. Supplementary materials presents results for all
-        metrics in Figure \ref{apd:fig:all_datasets_overlap_effect} in absolute
+        are the tertiles of the overlap measured with NTV (Appendix \ref{apd:motivation_ntv}, eq. 17). Appendix \ref{apd:experiments:additional_results} presents results for all
+        metrics in Figure 18 in absolute
         Kendall's and continuous overlap values in Figure
-            {\ref{apd:fig:all_datasets_tau_risk_ranking_agreement}}.}\label{fig:all_datasets_overlap_effect_r_risk}
+            {15}.}\label{fig:all_datasets_overlap_effect_r_risk}
 \end{figure}
@@
             nuisances/candidate set} and
-        the \textcolor{RedOrange}{separated nuisances set} procedures. Figure
-        \ref{apd:fig:procedures_comparison_all_metrics} details results for all metrics.}\label{fig:procedures_comparison}
+        the \textcolor{RedOrange}{separated nuisances set} procedures. Appendix \ref{apd:experiments:additional_results} Figure
+        17 details results for all metrics.}\label{fig:procedures_comparison}
 \end{figure}
@@
 stacked and linear estimations of the nuisances performs equivalently.
-Detailed analysis (Figure \ref{apd:fig:nuisances_comparison_twins})
+Detailed analysis (Appendix \ref{apd:experiments:additional_results} Figure 20)
 confirms that using adaptive models --as built by
@@
         Results are shown only for the
-        R-risk; Figure \ref{apd:fig:nuisances_comparison}
+        R-risk; \ref{apd:experiments:additional_results} Figure 19
         details every metrics. For Twins, where the true propensity
@@
 procedures and error measures than those classically used to assess prediction. Rather, selecting the best outcome
-model according to the $R\text{-risk}$ (eq.\, Definition \ref{def:r_risk}) leads
+model according to the $R\text{-risk}$ (Appendix \ref{def:feasible_risks} eq.\, Definition 5) leads
 to more valid causal estimates on average. Estimating the $R\text{-risk}$ requires a more
@@
 \appendix
-\section{Additional files}
-
-%\part*{Supplementary materials:\\How to select predictive models for decision-making or causal inference? }
-%\setcounter{secnumdepth}{1}
-%\renewcommand{\thesubsection}{\Alph{subsection}}
-%\setcounter{secnumdepth}{3}
-
-%\part{Supplementary materials}
-
-%The following supporting information is available as part of the online article:
+\section{List of additional files}
 
@@
 
-Figure \ref{fig:acic_2016_ate_heterogeneity} shows ATE estimations for six
-different models used in g-computation estimators on the 76 configurations of
-the ACIC 2016 dataset. Outcome models are fitted on half of the data and
-inference is done on the other half --ie. train/test with a split ratio of 0.5.
-For each configuration, and each model, this train test split was repeated ten
-times, yielding non parametric variance estimates
-\cite{bouthillier_accounting_2021}. Figure \ref{fig:acic_2016_ate_heterogeneity}
-shows large variations obtained across different outcome estimators on
-semi-synthetic datasets \cite{dorie_automated_2019}. Flexible models such as
-random forests are doing well in most settings except when treated and untreated
-populations differ noticeably, in which case a linear model (ridge) is to be
-preferred. However random forests with different hyper-parameters (max depth= 2)
-yield poor estimates. A simple rule of thumb such as preferring flexible models
-does not work in general; model selection is needed.
-
-%\idea{It is crucial to have tools for model selection.}
-
-\begin{figure}[!b]
-    \centering
-    \includegraphics[width=0.95\linewidth]{acic_2016_ate_hte.pdf}%
-    \caption{\textbf{Different outcome models lead to different
-            estimation errors on the Average Treatment Effects},
-        on 77 classic simulations with known true causal effect
-        \cite{dorie_automated_2019}. The different models are ridge regression
-        and random forests with different hyper-parameters
-        (details
-        \ref{apd:toy_example:acic_2016_ate_variability}). The different configurations are
-        plotted as a function of increasing difference between treated and
-        untreated population --see
-        \autoref{subsec:measuring_overlap}.
-        There is no systematic best performer; data-driven model
-        selection is important.
-        \label{fig:acic_2016_ate_heterogeneity}%
-    }
-\end{figure}
-
-Outcome models are implemented with
-\href{https://scikit-learn.org/stable/}{scikit-learn}
-\cite{pedregosa_scikitlearn_2011} and the following hyper-parameters:
-
-\begin{table}[h!]
-    \centering
-    \begin{tabular}{llll}
-        \toprule
-        Outcome Model                                  & Hyper-parameters grid
-        \\
-        \midrule
-        Random Forests                                 & Max depth: [2,
-        10]                                                                    \\
-
-        Ridge regression without treatment interaction & Ridge regularization:
-        [0.1]                                                                  \\
-
-        Ridge regression with treatment interaction    & Ridge regularization:
-        [0.1]                                                                  \\
-        \bottomrule
-    \end{tabular}
-    \caption{Hyper-parameters grid used for ACIC 2016 ATE variability}
-    \label{apd:toy_example:acic_2016_ate_variability:table}
-\end{table}
-
 \subsection{Prior work : model selection for outcome modeling (g-computation)}\label{apd:prior_work}
 
-A natural way to select a predictive model for causal inference would be
-an error measure between a causal quantity such as the CATE and models' estimate. But such error is
-not a ``feasible'' risk: it cannot be computed solely from observed data
-and requires oracle knowledge.
-
-% XXX: I think that I should make the following paragraph shorter
-
-\paragraph{Simulation studies of causal model selection}
-
-Using eight simulations setups from \cite{powers_methods_2018}, where
-the oracle CATE is known, \cite{schuler_comparison_2018} compare four
-causal risks, concluding that for CATE estimation the best
-model-selection risk is the so-called $R\text{-risk}$
-\cite{nie_quasioracle_2017} --def.\,\ref{def:r_risk}, below. Their
-empirical results are clear for randomized treatment allocation but less
-convincing for observational settings where both simple MSE --$\mu\text{-risk}(f)$ def.\,\ref{def:mu_risk}-- and
-reweighted MSE --$\mu\text{-risk}_{IPW}$ def.\,\ref{def:mu_ipw_risk}--
-appear to perform better than $R\text{-risk}$ on half of the simulations.
-Another work \cite{alaa_validating_2019} studied empirically both MSE and
-reweighted MSE risks on the semi-synthetic ACIC 2016 datasets
-\cite{dorie_automated_2019}, but did not include the $R\text{-risk}$. We complete these
-prior empirical work by studying a wider variety of data generative
-processes and varying the influence of overlap, an important parameter of
-the data generation process which makes a given causal metric appropriate
-\cite{damour_overlap_2020}. We also study how to best adapt
-cross-validation procedures to causal metrics which themselves come with
-models to estimate.
-
-\paragraph{Theoretical studies of causal model selection}
-
-Several theoretical works have proposed causal model selection procedures
-that are \emph{consistent}: select the best model in a family given
-asymptotically large data. These work rely on introducing a
-CATE estimator in the testing procedure: matching
-\cite{rolling_model_2014}, an IPW estimate
-\cite{gutierrez_causal_2016}, a doubly robust estimator
-\cite{saito_counterfactual_2020}, or debiasing the error with influence
-functions \cite{alaa_validating_2019}. However, for theoretical
-guarantees to hold, the test-set correction needs to converge to the
-oracle: it needs to be flexible enough --well-posed-- and asymptotic
-data. From a practical perspective, meeting such requirements
-implies having a good CATE estimate, thus having solved
-the original problem of causal model selection.
-
-\paragraph{Statistical guarantees on causal estimation procedures}
-
-Much work in causal inference has focused on procedures that
-guarantee asymptotically consistent estimators, such as Targeted
-Machine Learning
-Estimation (TMLE) \cite{laan_targeted_2011,schuler_targeted_2017} or
-Double Machine Learning \cite{chernozhukov_double_2018}. Here also, theories require asymptotic regimes and
-models to be \textit{well-specified}.
-
-By contrast, without assuming that estimators are well specified, there exists an upper bound
-on the oracle error to the CATE ($\tau\text{-risk}$) that involves the error on
-the outcome and the similarity of the distributions of treated and control
-patients \cite{johansson2022generalization}. However, they use this upper bound for model optimization,
-and do not give insights on model selection. In addition, for hyperparameter
-selection, they rely on a plugin estimate of the $\tau\text{-risk}$ built with
-counterfactual nearest neighbors, which has been shown ineffective
-\cite{schuler_comparison_2018}.
-An interesting direction is taken in \cite{foster2023orthogonal} where the authors derive convergence rates for orthogonal losses such as the R-loss, or the DR-Loss without assuming well-specification of the model for target parameter.
-
 \subsection{Causal assumptions}\label{apd:causal_assumptions}
 
-We assume the following four assumptions, referred as strong ignorability and
-necessary to assure identifiability of the causal estimands with observational
-data \cite{rubin_causal_2005}:
-\begin{assumption}[Consistency]\label{assumption:consistency} The observed
-    outcome is the potential outcome of the assigned treatment:
-    \begin{equation*}\label{eq:consistancy}
-        Y = A \, Y(1) + (1-A) \, Y(0)
-    \end{equation*}
-    Here, we assume that the intervention $A$ has been well defined. This
-    assumption focuses on the design of the experiment. It clearly states the link
-    between the observed outcome and the potential outcomes through the
-    intervention \cite{hernan_causal_2020}.
-\end{assumption}
-
-\begin{assumption}[Unconfoundedness]\label{assumption:ignorability}
-    \begin{equation*}\label{eq:ignorability}
-        \{Y(0), Y(1) \} \indep A | X
-    \end{equation*}
-    This condition --also called ignorability-- is equivalent to the conditional
-    independence on $e(X)$ \cite{rosenbaum_central_1983}: $\{Y(0), Y(1) \}
-        \indep  A | e(X)$.
-\end{assumption}
-
-\begin{assumption}[Overlap, also known as Positivity)]\label{assumption:overlap}
-    \begin{equation*}\label{eq:overlap}
-        \eta < e(x) < 1 - \eta \quad \forall x \in \mathcal X \text{ and some } \eta > 0
-    \end{equation*}
-    The treatment is not perfectly predictable. Or with different words, every
-    patient has a chance to be treated and not to be treated. For a given set of
-    covariates, we need examples of both to recover the ATE.
-\end{assumption}
-
-As noted by \cite{damour_overlap_2020}, the choice of covariates $X$ can
-be viewed as a trade-off between these two central assumptions. A bigger
-covariates set generally reinforces the ignorability assumption. In the
-contrary, overlap can be weakened by large $\mathcal{X}$ because of the
-potential inclusion of instruments: variables only linked to the treatment which
-could lead to arbitrarily small propensity scores.
-
-% remark: There is a major counter example to these colliders (variables which
-% are caused by both the outcome and the treatment),
-
-\begin{assumption}[Generalization]\label{assumption:generalization} The training
-    data on which we build the estimator and the test data on which we make the
-    estimation are drawn from the same distribution $\mathcal D^*$, also known as
-    the ``no covariate shift'' assumption \cite{jesson_identifying_2020}.
-\end{assumption}
-
 \subsection{Definitions of feasible risks}\label{def:feasible_risks}
 
-\begin{definition}[Factual $\mu\text{-risk}$]\label{def:mu_risk}
-    \cite{shalit_estimating_2017} This is the usual Mean Squared Error on
-    the target y. It is what is typically meant by ``generalization error'' in
-    supervised learning:
-    \begin{equation*}\label{eq:mu_risk}
-        \mu\text{-risk}(f)=\mathbb{E}\left[(Y-f(X ; A))^2 \right]
-    \end{equation*}
-\end{definition}
-
-\begin{definition}[$\mu\text{-risk}_{IPW}^{\star}$]\label{def:mu_ipw_risk}
-    \cite{vanderlaan_unified_2003} Let the inverse propensity weighting
-    function $w(x, a) = \frac{a}{e(x)} + \frac{1 - a}{1 - e(x)}$, we define the
-    semi-oracle Inverse Propensity Weighting risk,
-    \begin{equation*}\label{eq:mu_ipw_risk}
-        \mu\text{-risk}_{IPW}^{\star}(f) = \mathbb{E}\left[ \Big( \frac{A}{e(X)} + \frac{1-A}{1-e(X)} \Big) (Y-f(X ; A))^2 \right]
-    \end{equation*}
-\end{definition}
-
-\smallskip
-
-\begin{definition}[$\tau\text{-risk}^{\star}_{IPW}$]\label{def:tau_ipw_risk}
-    \cite{wager_estimation_2018}
-    The CATE $\tau(x)$ can be estimated with a
-    regression against inverse propensity weighted outcomes \cite{
-        athey2016recursive,gutierrez_causal_2016,wager_estimation_2018},
-    the $\tau\text{-risk}_{IPW}$.
-    \begin{equation*}
-        \tau\text{-risk}^{\star}_{IPW}(f) =\mathbb{E}
-        \left[ \Big(Y \frac{A - e(X)}{e(X)
-                (1-e(X))}-\tau_f\left(X\right)\Big)^2 \right]
-        %  =\mathbb{E}_{(Y, X, A)
-        %            \sim \mathcal D} \left[ \left(Y \left( \frac{A}{e(X)} -
-        %            \frac{1-A}{1-e(X)}\right)-\tau_f\left(X\right)\right)^2 \right]
-    \end{equation*}
-\end{definition}
-
-\begin{definition}[$U\text{-risk}^{\star}$]\label{def:u_risk}
-    \cite{kunzel_metalearners_2019,nie_quasioracle_2017} Based on
-    the Robinson decomposition --eq. \ref{eq:r_decomposition}, the U-learner
-    uses the $A-e(X)$ term
-    in the denominator. The derived risk is:
-    \begin{equation*}
-        U\text{-risk}^{\star}(f) =\mathbb{E}
-        \left[
-            \left( \frac{Y-m\left(X\right)}{A-e\left(X\right)} -
-            \tau_f\left(X\right)\right)^{2} \right]
-    \end{equation*}
-    Note that extreme propensity weights in the
-    denominator term might inflate errors in the numerator due to imperfect
-    estimation of the mean outcome $m$.
-\end{definition}
-
-\begin{definition}[$R\text{-risk}^{\star}$]\label{def:r_risk}
-    \cite{nie_quasioracle_2017,schuler_comparison_2018}
-    The $R\text{-risk}$ also uses two nuisance $m$ and $e$:
-    \begin{equation*}
-        R\text{-risk}^{\star}(f) =\mathbb{E} \big[
-            \big(\left(Y-m\left(X\right)\right) -\left(A-e\left(X\right)\right) \tau_f\left(X\right)\big)^{2} \big]
-    \end{equation*}
-\end{definition}
-
-It is also based on the Robinson decomposition --eq. \ref{eq:r_decomposition}. %It performs well in various
-%simulations, estimating the nuisances
-%$(\check e, \check m)$ with
-%lasso, boosting or kernel ridge regressions
-%\cite{nie_quasioracle_2017}.
-
 \subsection{Proofs: Links between feasible and oracle risks}\label{apd:proofs}
 
-% \subsection{Upper bound of $\tau\text{-risk}$ with
-%     $\mu\text{-risk}_{IPW}$}\label{apd:proofs:mu_risk_ipw_bound}
-
-% For the bound with the $\mu\text{-risk}_{IPW}$, we will decompose the CATE risk
-% on each factual population risks:
-
-% \begin{definition}[Population Factual $\mu\text{-risk}$]\label{mu_risk_a}
-%     \cite{shalit_estimating_2017}
-%     \begin{equation*}
-%         \mu\text{-risk}_{a}(f)= \int_{\mathcal Y \times \mathcal X} (y-f(x ; A=a))^{2}  p(y ; x=x \mid A=a) \; dy dx
-%     \end{equation*}
-% \end{definition}
-
-% Applying Bayes rule, we can decompose the $\mu\text{-risk}$ on each
-% intervention:
-% \begin{equation*}
-%     \mu\text{-risk}(f)
-%     =p_{A} \,\mu\text{-risk}_{1}(f)+\left(1-p_{A}\right) \,\mu\text{-risk}_{0}(f)
-%     \text{with } p_A=\mathbb P(A=1)
-% \end{equation*}
-
-% These definitions allows to state a intermediary result on each population:
-% \begin{lemma}[Mean-variance decomposition]\label{apd:proofs:mu_risk_ipw_link_mu}
-%     We need a reweighted version of the classical mean-variance decomposition.
-
-%     For an outcome model $f: x \times A \rightarrow \mathcal X$. Let the inverse
-%     propensity weighting function $w(a ; x)=a e(x)^{-1}+(1-a)(1-e(x))^{-1}$.
-%     \begin{align*}
-%          & \int_{\mathcal X}(\mu_{1}(x)-f(x ; 1))^{2} p(x) dx  = p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  -\sigma^{2}_{Bayes}(1)
-%     \end{align*}
-%     And
-%     \begin{align*}
-%          & \int_{\mathcal X}(\mu_{0}(x)-f(x; 0))^{2} p(x) dx   = (1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)  -\sigma^{2}_{Bayes}(0)
-%     \end{align*}
-
-%     \begin{proof}
-%         \begin{align*}
-%              & p_{A} \mu\text{-risk}_{IPW, 1}(w, f)  = \int_{\mathcal X \times \mathcal Y} \frac{1}{e(x)}(y-f(x ; 1))^{2} p(y \mid x ; A=1) p(x ; A=1) d y d x                                                      \\
-%              & = \int_{\mathcal X \times \mathcal Y} (y-f(x ; 1))^{2} p(y \mid x ; A=1) \frac{p(x ; A=1)}{p(x ; A=1)}p(x)dy dx                                                                                      \\
-%              & = \int_{\mathcal X \times \mathcal Y} \big[(y-\mu_1(x))^{2}+\left(\mu_{1}(x)-f(x ; 1)\right)^{2} + 2\left(y-\mu_{1}(x)\right)\left(\mu_{1}(x)-f(x, 1)\right) \big] p(y \mid x ; A=1) p(x) d y d x    \\
-%              & =\int_{\mathcal X} \big [ \int_{\mathcal Y} (y-\mu_1(x))^{2} p(y \mid x ; A=1) dy\big ] p(x)dx + \int_{\mathcal X \times \mathcal Y} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x)p(y \mid x ; A=1)dx dy \\
-%              & + \qquad 2 \int_{\mathcal X} \big [ \int_{\mathcal Y} \left(y-\mu_{1}(x)\right) p(y \mid x ; A=1) dy \big ] \left(\mu_{1}(x)-f(x, 1)\right)p(x)dx                                                    \\
-%              & =\int_{\mathcal X} \sigma_{y}^{2}(x, 1) p(x) d x +\int_{\mathcal X} \left(\mu_{1}(x)-f(x ; 1)\right)^{2} p(x) d x+0
-%         \end{align*}
-%     \end{proof}
-% \end{lemma}
-
-% \begin{proposition*}[Upper bound with mu-IPW]\label{apd:proofs:prop:upper_bound}
-%     Let f be a given outcome model, let the weighting function $w$ be the Inverse
-%     Propensity Weight $w(x; a) = \frac{a}{e(x)} + \frac{1-a}{1-e(x)}$. Then, under
-%     overlap (assumption \ref{assumption:overlap}),
-%     \begin{align*}
-%         \tau\text{-risk}(f) \leq & \; 2 \, \mu\text{-risk}_{IPW}(w, f)  \; - 2 \, (\sigma^2_{Bayes}(1) +  \sigma^2_{Bayes}(0))
-%     \end{align*}
-
-%     \begin{proof}
-%         \begin{align*}
-%              & \tau\text{-risk}(f) =\int_{\mathcal X}(\mu_{1}(x)-\mu_{0}(x)-(f(x ; 1)-f(x ; 0))^{2} p(x) d x
-%         \end{align*}
-%         By the triangle inequality $(u+v)^2 \leq 2(u^2 + v^2)$:
-%         \begin{align*}
-%              & \tau\text{-risk}(f) \leq
-%             2 \int_{\mathcal X}\big[\left(\mu_{1}(x)-f(x ; 1)\right)^{2}+\left(\mu_{0}(x)-f(x ; 0)\right)^{2}\big] p(x) d x \\
-%         \end{align*}
-%         Applying Lemma \ref{apd:proofs:mu_risk_ipw_link_mu},
-%         \begin{align*}
-%              & \tau\text{-risk}(f) \leq 2\big[p_A \mu\text{-risk}_{IPW, 1}(w, f) +(1-p_A) \mu\text{-risk}_{IPW, 0}(w, f)(w, f)\big] -2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1)) \\
-%              & = 2 \mu\text{-risk}_{IPW}(w, f)-2(\sigma^{2}_{Bayes}(0) + \sigma^{2}_{Bayes}(1))
-%         \end{align*}
-%     \end{proof}
-% \end{proposition*}
-
-\subsubsection{Reformulation of the $R\text{-risk}$ as reweighted
-    $\tau\text{-risk}$}\label{apd:proofs:r_risk_rewrite}
-
-\begin{proposition*}[$R\text{-risk}$ as reweighted $\tau
-            \text{-risk}$]\label{apd:proofs:prop:r_risk_rewrite}
-
-    \begin{proof}
-
-        We consider the R-decomposition: \cite{robinson_rootnconsistent_1988},
-        \begin{equation}\label{apd:eq:r_decomposition}
-            y(a) = m(x) + \big( a - e(x) \big) \tau(x) + \varepsilon(x; a)
-        \end{equation}
-        Where $\mathbb E[\varepsilon(X; A)|X, A] = 0$ We can use it as plug in the
-        $R\text{-risk}$ formula:
-
-        \begin{align*}
-             & R\text {-risk}(f) =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}[(y-m(x))-\big(a-e(x)\big) \tau_f(x)]^{2} p(y ; x ; a) d y d x d a                                     \\
-             & =\int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}} \left[\big(a-e(x)\big)\tau(x)+\varepsilon(x ; a)-\big(a-e(x)\big) \tau_f(x)\right]^{2} p(y ; x ; a) d y d x da          \\
-             & =\int_{\mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)^{2}\big(\tau(x)- \tau_f(x)\big)^{2} p(x ; a) d x d a                                                                       \\
-             & + 2  \int_{\mathcal{Y} \times \mathcal{X} \times \mathcal{A}}\big(a-e(x)\big)\big(\tau(x)-\tau_f(x)\big)  \int_{\mathcal{Y}} \varepsilon(x ; a) p(y \mid x ; a) d y p(x ; a)dx da \\
-             & +\int_{\mathcal{X} \times \mathcal{A}} \int_{\mathcal{Y}} \varepsilon^{2}(x ; a) p(y \mid x ; a) d y p(x ; a) d x d a
-        \end{align*}
-
-        The first term can be decomposed on control and treated populations to force
-        $e(x)$ to appear:
-        \begin{align*}
-             & \int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}\left[e(x)^{2}p(x;0) + \big(1-e(x)\big)^{2} p(x;1)\right] d x                    \\
-             & =\int_{\mathcal{X}}\big(\tau(x)-\tau_f(x)\big)^{2}  \left[e(x)^{2}\big(1-e(x)\big)p(x) + \big(1-e(x)\big)^{2}e(x) p(x)\right] d x \\
-             & =\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x)[1-e(x)+e(x)] p(x) d x                                                     \\ &=\int_{\mathcal{X}}(\tau(x)-\tau_f(x))^{2}(1-e(x)) e(x) p(x) d x.
-        \end{align*}
-
-        The second term is null since, $\mathbb E[\varepsilon(x, a) |X, A]=0$.
-
-        The third term corresponds to the modulated residuals \ref{eq:residuals} :
-        $\tilde{\sigma}_B^2(0) + \tilde{\sigma}_B^2(1)$
-
-    \end{proof}
-\end{proposition*}
-
-\subsubsection{Interesting special cases}\label{apd:theory:special_cases}
-
-\paragraph{Randomization special case}\label{remark:rct} If the treatment is
-randomized as in RCTs, $p(A=1 \mid X=x) = p(A=1)=p_A$, thus
-$\mu\text{-risk}_{IPW}$ takes a simpler form:
-\begin{equation*}
-    \mu\text{-risk}_{IPW} = \mathbb{E}_{(Y, X, A) \sim \mathcal D}\left[ \Big( \frac{A}{p_A} + \frac{1-A}{1-p_A} \Big) (Y-f(X ; A))^2 \right]
-\end{equation*}
-However, we still can have large differences
-between $\tau\text{-risk}$ and $\mu\text{-risk}_{IPW}$ coming from heterogeneous
-errors between populations as shown experimentally in \cite{schuler_comparison_2018} and our
-results below.
-
-Concerning the $R\text{-risk}$, replacing $e(x)$ by its randomized value $p_A$
-in Proposition \ref{theory:prop:r_risk_rewrite} yields the oracle
-$\tau\text{-risk}$ up to multiplicative and additive constants:
-\begin{equation*}
-    R\text{-risk} = p_A \, (1-p_A) \, \tau\text{-risk} \;+\; (1 - p_A) \,\sigma_B^2(0) \;+\; p_A \sigma_B^2(1)
-\end{equation*}
-Thus, selecting estimators with $R\text{-risk}^*$ in
-randomized setting controls the $\tau\text{-risk}$. This explains
-the strong performances of $R\text{-risk}$ in randomized setups
-\cite{schuler_comparison_2018} and is a strong argument to use it
-to estimate heterogeneity in RCTs.
-
-\paragraph{Oracle Bayes predictor}\label{remark:bayes_oracle} If we
-have access to the oracle Bayes predictor for the outcome ie.~$f(x,
-    a)=\mu(x, a)$, then all risks are equivalent up to the residual variance:
-\begin{equation}
-    \tau\text{-risk}(\mu) = \mathbb E_{X\sim p(X)}[(\tau(X) - \tau_{\mu}(X))^2] = 0
-\end{equation}
-\begin{align}
-    \mu\text{-risk}(\mu) & = \mathbb E_{(Y, X, A) \sim p(Y;X;A)}[\big( Y - \mu_A(X)\big)^2] \\
-                         & = \int_{\mathcal X, \mathcal A}
-    \,\varepsilon(x,a)^2 p(a \mid x) \,p(x) \,dx\,da  \leq \sigma_B^{2}(0) + \sigma_B^{2}(1) \nonumber
-\end{align}
-\begin{equation}
-    \mu\text{-risk}_{IPW}(\mu) = \sigma_B^{2}(0) + \sigma_B^{2}(1)
-    %\notag
-\end{equation}
-\begin{multline}
-    R\text{-risk}(\mu) = \tilde{\sigma}_B^{2}(0) + \tilde{\sigma}_B^{2}(1)
-    \leq \sigma_B^{2}(0) + \sigma_B^{2}(1)  \\  \text{from Proposition \ref{theory:prop:r_risk_rewrite}}%\notag
-\end{multline}
-
-Thus, differences between causal risks only matter in finite sample regimes.
-Universally consistent learners converge to the Bayes risk in asymptotic
-regimes, making all model selection risks equivalent. In practice however,
-choices must be made in non-asymptotic regimes.
-
 \subsection{Measuring overlap}\label{apd:motivation_ntv}
 
-\paragraph{Motivation of the Normalized Total Variation}
-
-%\idea{A simplier measure, NTV seems to capture what we want}
-
-Overlap is often assessed by comparing visually population distributions as in
-Figure \ref{fig:toy_example} or computing standardized difference on each
-feature \cite{austin_introduction_2011,austin_moving_2015}. While these methods
-are useful to decide if positivity holds, they do not yield a single measure.
-Rather, we compute the divergence between the population covariate distributions
-$\mathbb P(X|A=0)$ and $\mathbb P(X|A=1)$
-\cite{damour_overlap_2020,johansson2022generalization}. Computing overlap when
-working only on samples of the observed distribution, outside of simulation,
-requires a sophisticated estimator of discrepancy between distributions, as two
-data points never have the same exact set of features. Maximum Mean Discrepancy
-\cite{gretton2012kernel} is typically used in the context of causal inference
-\cite{shalit_estimating_2017,johansson2022generalization}. However it needs a
-kernel, typically Gaussian, to extrapolate across neighboring observations. We
-prefer avoiding the need to specify such a kernel, as it must be adapted to the
-data which is tricky with categorical or non-Gaussian features, a common
-situation for medical data.
-
-For simulated and some semi-simulated data, we have access to the probability of
-treatment for each data point, which sample both densities in the same data
-point. Thus, we can directly use distribution discrepancy measures and rely on
-the Normalized Total Variation (NTV) distance to measure the overlap between the
-treated and control propensities. This is the empirical measure of the total
-variation distance \cite{sriperumbudur_integral_2009} between the distributions,
-$TV(\mathbb{P}(X|A=1), \mathbb{P}(X|A=0))$. As we have both distribution sampled
-on the same points, we can rewrite it a sole function of the propensity score, a
-low dimensional score more tractable than the full distribution $\mathbb
-    P(X|A)$:
-
-\begin{equation}\label{eq:ntv}
-    \widehat{NTV}(e, 1-e) = \frac{1}{2N} \sum_{i =1}^{N} \big |\frac{e(x_i)}{p_A}-\frac{1-e(x_i)}{1-{p_A}} \big |
-\end{equation}
-
-Formally, we can rewrite NTV as the Total Variation distance
-between the two population distributions. For a population $O = (Y(A), X, A)
-    \sim \mathcal{D}$:
-
-\begin{align*}
-    NTV(O) & = \frac{1}{2N} \sum_{i =1}^{N} \big | \frac{e(x_i)}{p_A}-\frac{1-e(x_i)}{1-{p_A}} \big|         \\
-           & = \frac{1}{2N} \sum_{i =1}^{N} \big|\frac{P(A=1|X=x_i)}{p_A}-\frac{P(A=0|X=x_i)}{1-{p_A}} \big|
-\end{align*}
-
-Thus NTV approximates the following quantity in expectation over the data
-distribution $\mathcal{D}$:
-
-\begin{align*}
-    NTV(\mathcal{D}) & = \int_{\mathcal{X}} \big | \frac{p(A=1|X=x)}{p_A}-\frac{p(A=0|X=x)}{1-{p_A}} \big | p(x)dx \\
-                     & = \int_{\mathcal{X}} \big | \frac{p(A=1, X=x)}{p_A}-\frac{p(A=0, X=x)}{1-{p_A}} \big |dx    \\
-                     & = \int_{\mathcal{X}} \big | p(X=x|A=1)- p(X=x|A=0) \big |dx
-\end{align*}
-
-For countable sets, this expression corresponds to the Total Variation distance
-between treated and control populations covariate distributions : $TV(p_0(x),
-    p_1(x))$.
-
-\paragraph{Measuring overlap without the oracle propensity scores:} For ACIC
-2018, or for non-simulated data, the true propensity scores are not known. To
-measure overlap, we rely on flexible estimations of the Normalized Total
-Variation, using gradient boosting trees to approximate the propensity score.
-Empirical arguments for this plug-in approach is given in Figure
-\ref{apd:overlap:ntv_approximation}.
-
-\paragraph{Empirical arguments}
-
-We show empirically that NTV is an appropriate measure of overlap by :
-\begin{itemize}
-    \item Comparing the NTV distance with the MMD for Caussim which is gaussian
-          distributed in Figure \ref{apd:overlap:caussim:mmd_vs_ntv},
-    \item Verifying that setups with penalized overlap from ACIC 2016 have a
-          higher total variation distance than unpenalized setups in Figure
-          \ref{apd:overlap:penalized_overlap}.
-    \item Verifying that the Inverse Propensity Weights extrema (the inverse of
-          the $\nu$ overlap constant appearing in the overlap Assumption
-          \ref{assumption:overlap}) positevely correlates with NTV for Caussim,
-          ACIC 2016 and Twins in Figure \ref{apd:ntv_vs_max_ipw}. Even if the same
-          value of the maximum IPW could lead to different values of NTV, we
-          expect both measures to be correlated : the higher the extrem propensity
-          weights, the higher the NTV.
-\end{itemize}
-
-\paragraph{Estimating NTV in practice}
-
-Finally, we verify that approximating the NTV distance with a learned plug-in
-estimates of e(x) is reasonnable. We used either a logistic regression or a
-gradient boosting classifier to learn the propensity models for the three
-datasets where we have access to the ground truth propensity scores: Caussim,
-Twins and ACIC 2016. We respectively sampled 1000, 1000 and 770 instances of
-these datasets with different seeds and overlap settings. We first run a
-hyperparameter search with cross-validation on the train set, then select the
-best estimator. We refit on the train set this estimator with or without
-calibration by cross validation and finally estimate the normalized TV with the
-obtained model. This training procedure reflects the one described in Algorithm
-\ref{problem:estimation_procedure:algo} where nuisance models are fitted only on
-the train set.
-
-The hyper parameters are : learning rate $ \in [1e-3, 1e-2, 1e-1, 1]$, minimum
-samples leaf $\in [2, 10, 50, 100, 200]$ for boosting and L2 regularization $\in
-    [1e-3, 1e-2, 1e-1, 1]$ for logistic regression.
-
-Results in Figure \ref{apd:overlap:ntv_approximation} comparing bias to the true
-normalized Total Variation of each dataset instances versus growing true NTV
-indicate that calibration of the propensity model is crucial to recover a good
-approximation of the NTV.
-
-\begin{figure}
-    \begin{subfigure}[b]{\textwidth}
-        \centering
-        \caption{\textbf{Uncalibrated classifiers}}
-        \includegraphics[width=\linewidth]{overlap_measure_diff_oracle_ntv_to_pred_ntv.pdf}
-    \end{subfigure}
-    \begin{subfigure}[b]{\textwidth}
-        \centering
-        \caption{\textbf{Calibrated
-                classifiers}}
-        \includegraphics[width=\linewidth]{overlap_measure_diff_oracle_ntv_to_ntv_wcalibration.pdf}
-    \end{subfigure}
-    \caption{a) Without calibration, estimation of NTV is not trivial even for
-        boosting models. b) Calibrated classifiers are able to recover the true
-        Normalized Total Variation for all datasets where it is
-        available.}\label{apd:overlap:ntv_approximation}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \caption{NTV recovers well the overlap settings described in the ACIC paper
-        \cite{dorie_automated_2019}}\label{apd:overlap:penalized_overlap}
-    \includegraphics[width=0.7\linewidth]{overlap_measure_acic_2016_recovery_overlap_setup.pdf}
-\end{figure}
-
-\begin{figure}[htbp]
-    \centering
-    \caption{Good correlation between overlap measured as normalized Total
-        Variation and Maximum Mean Discrepancy (200 sampled Caussim
-        datasets)}\label{apd:overlap:caussim:mmd_vs_ntv}
-    \includegraphics[width=0.7\linewidth]{overlap_measure_caussim_transformed_mmd_vs_ntv.pdf}
-\end{figure}
-
-\begin{figure}[ht]
-    \centering
-    \begin{subfigure}[b]{0.47\textwidth}
-        \centering
-        \caption{\textbf{Caussim}}
-        \includegraphics[width=\textwidth]{overlap_measure_caussim_max_ipw_vs_ntv.pdf}
-        \label{apd:caussim:max_ipw_vs_ntv}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.47\textwidth}
-        \centering
-        \caption{\textbf{ACIC 2016}}
-        \includegraphics[width=\textwidth]{overlap_measure_acic_2016_max_ipw_vs_ntv.pdf}
-        \label{apd:acic_2016:ntv_vs_max_ipw}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.47\textwidth}
-        \centering
-        \caption{\textbf{ACIC 2018}}
-        \includegraphics[width=\textwidth]{overlap_measure_acic_2018_max_ipw_vs_ntv.pdf}
-        \label{apd:acic_2018:ntv_vs_max_ipw}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.49\textwidth}
-        \centering
-        \caption{\textbf{TWINS}}
-        \includegraphics[width=\textwidth]{overlap_measure_twins_max_ipw_vs_ntv.pdf}
-        \label{apd:twins:ntv_vs_max_ipw}
-    \end{subfigure}
-    \caption{Maximal value of Inverse Propensity Weights increases exponentially with the overlap as measure by Normalized Total Variation.}
-    \label{apd:ntv_vs_max_ipw}
-\end{figure}
-
-\subsection{Experiments}
-
-\subsubsection{Details on the data generation process}
-\label{apd:experiments:generation}
-
-We use Gaussian-distributed covariates and random basis expansion based on
-Radial Basis Function kernels. A random basis of RBF kernel enables modeling
-non-linear and complex relationships between covariates in a similar way to the
-well known spline expansion. The estimators of the response function are learned
-with a linear model on another random basis (which can be seen as a stochastic
-approximation of the full data kernel \cite{rahimi_random_2008}). We carefully
-control the amount of overlap between treated and control populations, a crucial
-assumption for causal inference. Algorithm \ref{apd:experiments:generation:algo} describes the generation process for one simulation. Figure \ref{fig:simulation_examples}
-illustrates 2D examples of the simulation.
-
-\begin{itemize}
-    \item The raw features for both populations are drawn from a mixture of
-          Gaussians:
-          $\mathbb P(X) = p_A \mathbb P(X|A=1) + (1- p_A) \mathbb P(X|A=0)$
-          where $\mathbb P(x|A=a)$ is a rotated Gaussian:
-          \begin{equation}
-              \mathbb P(x|A=a) = W \cdot \mathcal N \Big( \begin{bmatrix} (1-2a) \theta \\ 0\end{bmatrix} ; \begin{bmatrix} \sigma_0 & 0 \\ 0 & \sigma_1\end{bmatrix} \Big)
-          \end{equation}
-          with $\theta$ a parameter controlling overlap (bigger yields poorer
-          overlap), $W$ a random rotation matrix and $\sigma_0^2=2;\sigma_1^2=5$.
-
-          This generation process allows to analytically compute the oracle
-          propensity scores $e(x)$, to simply control for overlap with the
-          parameter $\theta$, the distance between the two Gaussian main axes and
-          to  visualize response surfaces.
-
-    \item A basis expansion of the raw features increases the problem dimension.
-          Using Radial Basis Function (RBF) Nystroem transformation \footnote{We
-              use the
-              \href{https://scikit-learn.org/stable/modules/generated/sklearn.kernel_approximation.Nystroem.html}{Sklearn
-                  implementation, \cite{pedregosa_scikitlearn_2011}}}, we expand the raw
-          features into a transformed space. The basis expansion samples
-          randomly a small number of representers in the raw data. Then,  it
-          computes an approximation of the full N-dimensional kernel with these
-          basis components, yielding the transformed features $z(x)$. The number
-          of basis functions --\emph{ie. knots}--, controls the complexity of
-          the ground-truth response surfaces and treatment. We first use this
-          process to draw the non-treated response surface $\mu_0$ and the
-          causal effect $\tau$. We then draw the observations from a mixture two
-          Gaussians, for the treated and non treated. We vary the separation
-          between the two Gaussians to control the overlap between treated and
-          non-treated populations, an important parameter for causal inference
-          (related to $\eta$ in Assumption
-          \ref{assumption:overlap}). Finally, we generate observed
-          outcomes adding Gaussian noise.
-
-          More formally, we generate the basis following the original data
-          distribution, $\left [ b_1 .. b_D \right ] \sim \mathbb P(x)$,
-          with D=2 in our simulations. Then, we compute an approximation
-          of the full kernel of the data generation process with these
-          representers: $z(x) = [RBF_{\gamma}(x, b_d)]_{d=1..D} \cdot
-              Z^T \in \mathbb{R}^D$ with $RBF_{\gamma}$ being the Gaussian
-          kernel $K(x, y) = exp(-\gamma ||x-y||^2)$ and Z the
-          normalization constant of the kernel basis, computed as the
-          root inverse of the basis kernel $Z=[K(b_i, b_j)]_{i, j \in
-              {1..D}}^{-1/2}$
-
-    \item Functions $\mu_0$, $\tau$ are distinct linear functions of the
-          transformed features:
-          \begin{equation*}
-              \mu_0(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\mu}^T
-          \end{equation*}
-          \begin{equation*}
-              \tau(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\tau}^T
-          \end{equation*}
-          Where $\beta_{\mu}$ and $\beta_{\tau}$ are sampled from two normal distributions with mean 0 and unit variances $\mathcal{N}(0, I_{D+1})$.
-    \item Adding a Gaussian noise, $\varepsilon \sim \mathcal N(0, \sigma(x;a))$,
-          we construct the potential outcomes:
-          $y(a) = (1-\omega)\mu_0(x) + a\,\omega\tau(x) + \varepsilon(x, a)$ with $\omega$ the treatment effect relative size with respect to the baseline response.
-\end{itemize}
-We generated 1000 instances of this dataset with uniformly random overlap
-parameters $\theta \in \left[ 0, 2.5 \right]$.
-
-\begin{algorithm}
-    \caption{Data simulation for the simulated dataset caussim}
-    \label{apd:experiments:generation:algo}
-    Let the parameters of a given simulation be:
-    \begin{itemize}
-        \item N the number of samples.
-        \item D the dimension of the basis expansion.
-        \item $p_A$ the proportion of treated individuals.
-        \item $\theta$ the overlap parameter.
-        \item $\mathbb P(x|A=a) = W \cdot \mathcal N \Big( \begin{bmatrix} (1-2a) \theta \\ 0\end{bmatrix} ; \begin{bmatrix} \sigma_0 & 0 \\ 0 & \sigma_1\end{bmatrix} \Big)$ the rotated gaussian for each population covariate.
-$W$ is a random rotation matrix and $\sigma_0^2=2;\sigma_1^2=5$ are scaling parameters.
-        \item $e(x)=\mathbb P(A=1|X=x)$ the oracle propensity score, obtained analytically from $\mathbb P(x|A=a)$.
-        \item $\mathbb P(X) = p_A \mathbb P(X|A=1) + (1- p_A) \mathbb P(X|A=0)$ the mixture of gaussian for the whole population covariates.
-        \item $b_1, .., b_D \sim \mathbb P(X)$, the basis sampled from the gaussian mixture.
-        \item $z(x) = [RBF_{\gamma}(x, b_d)]_{d=1..D} \cdot Z^T \in
-                  \mathbb{R}^D$ the Nystroem expansion. $RBF_{\gamma}$ is the
-              Gaussian kernel $K(x, y) = exp(-\gamma ||x-y||^2)$ and Z is the
-              normalization constant of the kernel basis, computed as the root
-              inverse of the basis kernel $Z=[K(b_i, b_j)]_{i, j \in
-                  {1..D}}^{-1/2}$.
-        \item $\beta_{\mu}$ and $\beta_{\tau}$, the linear coefficients on top of the basis expansion.
-They are sampled from two normal distributions with mean 0 and unit variances $\mathcal{N}(0, I_{D+1})$.
-        \item $\omega$ the treatment effect relative size with respect to the baseline response.
-        \item $\varepsilon(x, a) \sim \mathcal N(0, \sigma_y)$ an exogen gaussian noise with $\sigma_y$ the scale of the noise.
-    \end{itemize}
-
-    Generation process for one sample ($x$, $a$, e(x), $y(0)$, $y(1)$, $y$):
-    \begin{algorithmic}[1]
-        \State $x \sim \mathbb P(X)$ \Comment{Sample from rotated gaussian mixture}
-        \State  $z(x)$ \Comment{Compute the Nystroem expansion}
-        \State $\mu_0(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\mu}^T$ \Comment{Compute the non-treated response surface, ie. the baseline}
-        \State  $\tau(x) = \begin{bmatrix} z(x); 1 \end{bmatrix} \cdot \beta_{\tau}^T$ \Comment{Compute the conditional treatment effect (CATE)}
-        \State $y(a) = (1-\omega) \mu_0(x) + a\,\omega\tau(x) + \varepsilon(x, a) + \varepsilon(x, a)$ \Comment{Compute the potential outcomes}
-        \State $a \sim \operatorname{Bernoulli}(e(x))$ \Comment{Compute the treatment status}
-        \State $y = a\,y(1) + (1-a)\,y(0)$ \Comment{Compute the observed outcome}
-    \end{algorithmic}
-
-\end{algorithm}
-
-\begin{figure}[!t]
-    \centering
-    \includegraphics[width=0.8\linewidth]{caussim_example.pdf}
-
-    \caption{Example of the simulation setup in the input space with two
-        knots --\emph{ie.} basis functions. The top panel
-        shows the observations in feature space, while the bottom panel displays the
-        two response surfaces on a 1D cut along the black lines drawn on
-        the top panel.}
-    \label{fig:simulation_examples}
-\end{figure}
-
-\subsubsection{ Details on the semi-simulated datasets}\label{apd:experiments:datasets}
-
-\begin{itemize}
-    \item[ACIC 2018] \cite{dorie_automated_2019}:  The initial intervention was a child’s birth
-          weight $(A = 1 \text{ if weight} < 2.5 kg)$, and outcome was the child’s
-          IQ after a follow-up period. The study contained $N=4\,802$ data
-          points with $D=55$ features (5 binary, 27 count data, and 23
-          continuous). They simulated 77 different setups varying parameters
-          for treatment and response models, overlap, and interactions between treatment and
-          covariates \footnote{Original R code available at
-              \url{https://github.com/vdorie/aciccomp/tree/master/2016}
-              to generate 77 simulations settings.}. We used 10 different seeds
-          for each setup, totaling 770 dataset instances.
-
-    \item[ACIC 2018] \cite{shimoni_benchmarking_2018}: Starting from data from
-          the Linked Births and Infant Deaths Database (LBIDD)
-          \cite{macdorman_infant_1998} with $D=177$ covariates, treatment and
-          outcome models are simulated with complex models to reflect different
-          scenarii. The data do not provide the true propensity scores, so we
-          evaluate only feasible metrics, which do not require this nuisance
-          parameter. We used all 432 datasets\footnote{Using the scaling part of
-              the data, from
-              \href{https://github.com/IBM-HRL-MLHLS/IBM-Causal-Inference-Benchmarking-Framework}{github.com/IBM-HRL-MLHLS/IBM-Causal-Inference-Benchmarking-Framework}}
-          of size $N=5\,000$.
-
-    \item[Twins] \cite{louizos_causal_2017}: It is an augmentation of
-          real data on twin births and mortality rates
-          \cite{almond_costs_2005}. There are $N=11\,984$ samples (pairs of twins),
-          and $D=50$ covariates\footnote{We obtained the dataset from
-              \href{https://github.com/AMLab-Amsterdam/CEVAE/tree/master/datasets/TWINS}{https://github.com/AMLab-Amsterdam/CEVAE/tree/master/datasets/TWINS}}, The outcome is the mortality and the treatment is the
-          weight of the heavier twin at birth. This is a "true" counterfactual dataset
-          \cite{curth_really_2021} in the sense that we have
-          both potential outcomes with each twin. They simulate the treatment with a
-          sigmoid model based on GESTAT10 (number of gestation weeks before birth) and x
-          the 45 other covariates:
-          \begin{align}
-              \mathbf{t}_{i} & \mid \mathbf{x}_{i}, \mathbf{z}_{i} \sim
-              \operatorname{Bern}\left(\sigma\left(w_{o}^{\top}
-              \mathbf{x}+w_{h}(\mathbf{z} / 10-0.1)\right)\right)       \\ & \text{with} \;
-              w_{o} \sim \mathcal{N}(0,0.1 \cdot I),\; w_{h} \sim \mathcal{N}(5,0.1) \nonumber
-          \end{align}
-          We add a non-constant slope in the
-          sigmoid to control the overlap between
-          treated and control populations.
-          We sampled uniformly 1\,000 different overlap parameters between 0 and
-          2.5, totaling 1\,000 dataset instances. Unlike the previous datasets,
-          only the overlap varies for these instances. The response surfaces are
-          set by the original outcomes.
-\end{itemize}
-
-\subsubsection{Model selection procedures}
-
-\paragraph{Nuisances estimation}\label{apd:experiments:nuisances_hp}
-
-The nuisances are estimated with a stacked regressor inspired by the Super
-Learner framework, \cite{laan_super_2007}). We select hyper-parameters
-with randomized search on a validation set $\mathcal{V}$ and keep them fix for
-model selection  The searc grid is detailed in Table
-\ref{apd:experiments:nuisances_hp_grid}. All implementations come from
-\href{https://scikit-learn.org/stable/}{scikit-learn}
-\cite{pedregosa_scikitlearn_2011}.  As extreme inverse propensity weights induce
-high variance, we use clipping \cite{swaminathan_counterfactual_2015,
-    ionides_truncated_2008} to bound $min(\check e, 1-\check e)$ away from 0 with a
-fixed $\eta=10^{-10}$, ensuring strict overlap for numerical stability.
-
-\begin{table}[h!]
-    \begin{tabular}{llll}
-        \toprule
-        Model                & Estimator
-                             & Hyper-parameters grid                                         \\
-        \midrule
-        Outcome, m           & StackedRegressor
-                             & ridge regularization: [0.0001, 0.001, 0.01, 0.1, 1, 10, 100]  \\
-        \multirow[c]{3}{*}{} & (HistGradientBoostingRegressor, ridge)
-                             & HistGradientBoostingRegressor  learning rate: [0.01, 0.1, 1]  \\
-                             &
-                             & HistGradientBoostingRegressor  max leaf nodes: [10,
-        20, 30, 50]                                                                          \\
-        \midrule
-        Treatment, e         & StackedClassifier
-                             & LogisticRegression  C: [0.0001, 0.001, 0.01, 0.1, 1, 10, 100] \\
-        \multirow[c]{3}{*}{} & (HistgradientBoostingClassifier, LogisticRegression)
-                             & HistGradientBoostingClassifier  learning rate: [0.01, 0.1, 1] \\
-                             &
-                             & HistGradientBoostingClassifier  max leaf nodes: [10,
-        20, 30, 50]                                                                          \\
-        \bottomrule
-    \end{tabular}
-    \caption{Hyper-parameters grid used for nuisance models}
-    \label{apd:experiments:nuisances_hp_grid}
-\end{table}
-
-\subsubsection{Additional Results}\label{apd:experiments:additional_results}
-
-\paragraph{Definition of the Kendall's tau, $\kappa$}\label{apd:experiments:additional_results}
-
-The Kendall's tau  is a widely used statistics to measure the rank correlation
-between two set of observations. It measures the number of concordant pairs
-minus the discordant pairs normalized by the total number of pairs. It takes values in the
-$[-1, 1]$ range.
-\begin{equation}\label{eq:kendall_tau}
-    \kappa=\frac{\text { (number of concordant pairs })-\text { (number of discordant pairs) }}{\text { (number of pairs) }}
-\end{equation}
-
-\paragraph{Values of relative $\kappa(\ell,\tau\mathrm{{-risk}})$ compared to
-    the mean over all metrics Kendall's as shown in the boxplots of Figure \ref{fig:relative_kendalls_all_datasets}}
-
-\begin{table}
-    \centering
-    \resizebox{0.7\textwidth}{!}{
-        \input{images/_1_r_risk_domination_boxplot_df.tex}
-    }
-    \caption{Values of relative $\kappa(\ell,\tau\mathrm{{-risk}})$ compared to
-        the mean over all metrics Kendall's as shown in the boxplots of Figure
-        \ref{fig:relative_kendalls_all_datasets}}\label{apd:table:relative_kendalls_all_datasets}
-\end{table}
-
-\paragraph{Figure \ref{apd:fig:relative_kendalls_all_datasets_all_metrics} -
-    Results measured in relative Kendall's for feasible and semi-oracle risks}
-
-Because of extreme propensity scores in the denominator and bayes error residuals in the numerator, the semi-oracle
-$U$-risk has poor performances at bad overlap. Estimating these propensity scores in the is feasible $U$-risk reduces
-the variance since clipping is performed.
-
-\begin{figure}[!b]
-    \centering
-    \includegraphics[width=\linewidth]{_1_r_risk_domination_mrisks_by_ds.pdf}
-    %\hfill
-    \caption{\textbf{The $R$-risk is the best metric on average}: Relative Kendall's $\tau$
-        agreement with $\tau\text{-risk}$. Strong and Weak overlap correspond to
-        the first and last tertiles of the overlap distribution measured with
-        Normalized Total Variation eq. \ref{eq:ntv}.
-    }\label{apd:fig:relative_kendalls_all_datasets_all_metrics}
-\end{figure}
-
-\paragraph{Figure \ref{apd:fig:all_datasets_tau_risk_ranking_agreement} -
-    Results measured in absolute Kendall's}
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{0.44\textwidth}
-        \centering
-        \caption{\textbf{Caussim}}
-        \includegraphics[width=\textwidth]{ktau_caussim_nuisance_nlin_models_ridge_ov1-247.pdf}
-        \label{fig:ranking_agreement_w_tau_risk_caussim}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.44\textwidth}
-        \centering
-        \caption{\textbf{ACIC 2016}}
-        \includegraphics[width=\textwidth]{ktau_acic2016_nuisance_nlin_models_hgb_dgp1-77_rs1-10.pdf}
-        \label{fig:ranking_agreement_tau_risk_acic_2016}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.10\textwidth}
-        \centering
-        \includegraphics[width=1.5\textwidth]{legend_metrics.pdf}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.44\textwidth}
-        \centering
-        \caption{\textbf{ACIC 2018}}
-        \includegraphics[width=\textwidth]{ktau_acic2018_nuisance_stacking_models_hgb_tset_50.pdf}
-        \label{fig:ranking_agreement_w_tau_risk_acic_2018}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.44\textwidth}
-        \centering
-        \caption{\textbf{TWINS}}
-        \includegraphics[width=\textwidth]{ktau_twins_nuisance_stacking__models_hgb.pdf}
-        \label{fig:ranking_agreement_tau_risk_twins}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.10\textwidth}
-        ~
-    \end{subfigure}
-    % \begin{subfigure}[b]{\textwidth}
-    %   \centering
-    %   \includegraphics[width=0.5\textwidth]{legend_metrics_horizontal.pdf}
-    % \end{subfigure}
-    \caption{Agreement with $\tau\text{-risk}$ ranking of methods function
-        of overlap violation. The lines represent medians, estimated with a
-        lowess. The transparent
-        bands denote the 5\% and 95\% confidence intervals.}\label{apd:fig:all_datasets_tau_risk_ranking_agreement}
-\end{figure}
-
-\paragraph{Figure \ref{apd:all_datasets_normalized_bias_tau_risk_to_best_method}
-    - Results measured as distance to the oracle tau-risk}
-
-To see practical gain in term of $\tau\text{-risk}$, we plot the results as the
-normalized distance between the estimator selected by the oracle
-$\tau\text{-risk}$ and the estimator selected by each causal metric.
-
-Then, $\widehat{R\text{-risk}}^*$ is more efficient than all other metrics. The
-gain are substantial for every datasets.
-
-\begin{figure}
-    \centering
-    \begin{subfigure}[b]{0.44\textwidth}
-        \centering
-        \caption{\textbf{Caussim}}
-        \includegraphics[width=\textwidth]{norm_bias_trisk_tobest_caussim_nuis_nlin_models_ridge.pdf}
-        \label{fig:normalized_bias_tau_risk_to_best_method_caussim}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.44\textwidth}
-        \centering
-        \caption{\textbf{ACIC 2016}}
-        \includegraphics[width=\textwidth]{norm_bias_trisk_tobest_acic2016_nuis_nlin_models_hgb.pdf}
-        \label{fig:normalized_bias_tau_risk_to_best_method_acic_2016}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.10\textwidth}
-        \centering
-        \includegraphics[width=1.5\textwidth]{legend_metrics.pdf}
-    \end{subfigure}
-    \begin{subfigure}[b]{0.44\textwidth}
-        \centering
-        \caption{\textbf{ACIC 2018}}
-        \includegraphics[width=\textwidth]{norm_bias_trisk_tobest_acic2018_nuis_nlin_models_hgb.pdf}
-        \label{fig:normalized_bias_tau_risk_to_best_method_acic_2018}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.44\textwidth}
-        \centering
-        \caption{\textbf{TWINS}}
-        \includegraphics[width=\textwidth]{norm_bias_trisk_tobest_twins_nuis_stack_models_hgb.pdf}
-        \label{fig:normalized_bias_tau_risk_to_best_method_twins}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.10\textwidth}
-        ~
-    \end{subfigure}
-    \caption{Metric performances by normalized tau-risk distance to the best
-        method selected with $\tau\text{-risk}$. All nuisances are learned with the
-        same estimator stacking gradient boosting and ridge regression. Doted and
-        plain lines corresponds to 60\% lowess quantile estimates. This choice of
-        quantile allows to see better the oracle metrics lines for which outliers with a value
-        of 0 distord the curves.}
-    \label
-    {apd:all_datasets_normalized_bias_tau_risk_to_best_method}
-\end{figure}
-
-\paragraph{Figure \ref{apd:fig:procedures_comparison_all_metrics} - Stacked models for the nuisances is more efficient}
-For each metrics the benefit of
-using a stacked model of linear and boosting estimators for nuisances compared
-to a linear model. The evaluation measure is Kendall's tau relative to the
-oracle $R\text{-risk}^{\star}$ to have a stable reference between exepriments.
-Thus, we do not include in this analysis the ACIC 2018 dataset since
-$R\text{-risk}^{\star}$ is not available due to the lack of the true propensity
-score.
-
-\begin{figure}
-    \begin{subfigure}[b]{0.9\textwidth}
-        %\centering
-        \caption{\textbf{Caussim}}
-        \includegraphics[width=1.15\textwidth]{_3_procedure_Caussim_N5000__ref_metric_oracle_r_risk.pdf}
-        \label{fig:experiments:procedures_comparison:caussim}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.9\textwidth}
-        \centering
-        \caption{\textbf{ACIC 2016}}
-        \includegraphics[width=1\textwidth]{_3_procedure_ACIC2016_N4802__ref_oracle_r_risk.pdf}
-        \label{fig:experiments:procedures_comparison:acic_2016}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.9\textwidth}
-        \centering
-        \caption{\textbf{Twins}}
-        \includegraphics[width=1\textwidth]{_3_procedure_Twins_N11984__ref_oracle_r_risk.pdf}
-        \label{fig:experiments:procedures_comparison:twins}
-    \end{subfigure}
-    \hfill
-    \caption{Results are similar between the \textcolor{MidnightBlue}{Shared
-            nuisances/candidate set} and
-        the \textcolor{RedOrange}{Separated nuisances set} procedure. The
-        experience has not been run on the full metrics for Caussim due to computation costs.}\label
-    {apd:fig:procedures_comparison_all_metrics}
-\end{figure}
-
-\paragraph{Figure \ref{apd:fig:all_datasets_overlap_effect} Low population
-    overlap hinders model selection for all metrics}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{_2_overlap_influence_by_bin_k_by_ds.pdf}
-    \hfill
-    \caption{\textbf{Low population overlap hinders causal model selection for all
-            metrics}:
-        Kendall's $\tau$ agreement with $\tau\text{-risk}$. Strong, medium and Weak overlap
-        correspond
-        to the tertiles of the overlap distribution measured with Normalized Total
-        Varation eq. \ref{eq:ntv}.}\label{apd:fig:all_datasets_overlap_effect}
-\end{figure}
-
-\paragraph{Figure \ref{apd:fig:nuisances_comparison} - Stacked models for the nuisances is more efficient}
-For each metrics the benefit of
-using a stacked model of linear and boosting estimators for nuisances compared
-to a linear model. The evaluation measure is Kendall's tau relative to the
-oracle $R\text{-risk}^{\star}$ to have a stable reference between exepriments.
-Thus, we do not include in this analysis the ACIC 2018 dataset since
-$R\text{-risk}^{\star}$ is not available due to the lack of the true propensity
-score.
-
-\begin{figure}
-    \begin{subfigure}[b]{0.49\textwidth}
-        \centering
-        \caption{\textbf{Twins}}
-        \includegraphics[width=1\textwidth]{_4_nuisance_models_Twins_N11984__ref_oracle_r_risk.pdf}
-        \label{fig:experiments:nuisance_comparison:twins}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.49\textwidth}
-        \centering
-        \caption{\textbf{Twins downsampled}}
-        \includegraphics[width=1\textwidth]{_4_nuisance_models_Twinssmall_N4794__ref_oracle_r_risk.pdf}
-        \label{fig:experiments:nuisance_comparison:twins_ds}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.49\textwidth}
-        %\centering
-        \caption{\textbf{Caussim}}
-        \includegraphics[width=1.15\textwidth]{_4_nuisance_models_Caussim_N5000__ref_oracle_r_risk.pdf}
-        \label{fig:experiments:nuisance_comparison:caussim}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{0.49\textwidth}
-        \centering
-        \caption{\textbf{ACIC 2016}}
-        \includegraphics[width=1\textwidth]{_4_nuisance_models_ACIC2016_N4802__ref_oracle_r_risk.pdf}
-        \label{fig:experiments:nuisance_comparison:acic_2016}
-    \end{subfigure}
-    \hfill
-    \caption{Learning the nuisances with \textcolor{DarkOrchid}{stacked models} (linear and
-        gradient boosting) is important for successful model selection with R-risk.
-        For Twins dataset, there is no improvement for \textcolor{DarkOrchid}{stacked models} compared to
-        \textcolor{ForestGreen}{linear models} because of the linearity of the propensity model.}\label
-    {apd:fig:nuisances_comparison}
-\end{figure}
-
-\paragraph{Figure \ref{apd:fig:nuisances_comparison_twins} - Flexible models are performant in recovering nuisances even
-    in linear setups}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\textwidth]{_4_nuisance_models_r_risk_only_twins_nuisances.pdf}
-    \hfill
-    \caption{\textbf{Flexible models are performant in recovering nuisances
-            in the downsampled Twins dataset.} The propensity score is linear in this
-        setup, making it particularly challenging for flexible models compared to
-        linear methods.}\label{apd:fig:nuisances_comparison_twins}
-\end{figure}
-
-\paragraph{Selecting different seeds and parameters is crucial to draw
-    conclusions}\label{apd:results:seed_effect}
-
-One strength of our study is the various number of different simulated and
-semi-simulated datasets. We are convinced that the usual practice of using only
-a small number of generation processes does not allow to draw statistically
-significant conclusions.
-
-Figure \ref{apd:results:fig:seed_effect} illustrate the dependence of the
-results on the generation process for caussim simulations. We highlighted the
-different trajectories induced by three different seeds for data generation and
-three different treatment ratio instead of 1000 different seeds. The result
-curves are relatively stable from one setup to another for $R{-risk}$, but vary
-strongly for $\mu\text{-risk}$ and $\mu\text{-risk}_{IPW}$.
-
-\begin{figure}
-    \centering
-    \caption{Kendall correlation coefficients for each causal metric. Each (color,
-        shape) pair indicates a different (treatment ratio, seed) of the generation
-        process.}\label {apd:results:fig:seed_effect}
-    \includegraphics[width=\linewidth]{caussim_seed_effect.pdf}
-\end{figure}
-
-\FloatBarrier
+\subsection{Experiments}\label{apd:experiments:additional_results}
 
@@
 
-\subsubsection{Use 90\% of the data to estimate outcome models, 10\% to
-    select them}
-
-The analyst faces a compromise: given a finite
-data sample, should she allocate more data to estimate the outcome model,
-thus improving the quality of the outcome model but leaving
-little data for model selection. Or, she could choose a bigger test set for
-model selection and effect estimation. For causal model selection, there
-is no established practice (as reviewed in \ref{apd:results:k_fold_choices}).
-
-We investigate such tradeoff varying the ratio between train and test
-data size. For this, we first split out 30\% of the data as a holdout set
-$\mathcal{V}$ on which we use the oracle response functions to derive
-silver-standard estimates of causal quantities. We then
-use the standard estimation procedure on the remaining 70\% of the data,
-splitting it into train $\mathcal{T}$ and test $\mathcal{S}$ of varying
-sizes. We finally measure the error between this estimate and the
-silver standard.
-
-We consider two different analytic goals: estimating a average
-treatment effect --a single number used for policy making-- and a
-CATE --a full model of the treatment effect as a function of covariates
-$X$. Given that the latter is a much more complex object than the former,
-the optimal train/test ratio might vary. To measure errors, we use for
-the ATE the relative absolute ATE bias between the ATE computed with the
-selected outcome model on the test set, and the true ATE as evaluated on
-the holdout set $\mathcal{V}$. For the CATE, we compare the
-$\tau\text{-risk}$
-of the best selected model applied on the holdout set $\mathcal{V}$. We explore this trade-off for the ACIC 2016 dataset and the R-risk.
-
-\begin{figure}[!t]
-    \begin{minipage}{.5\textwidth}
-        \centerline{\textbf{a) CATE estimation error}}
-        \includegraphics[width=\linewidth]{_5_train_size_r_risk__cate__acic16_2cols.pdf}
-    \end{minipage}
-    %\hspace{0.3\textwidth}
-    \begin{minipage}{.5\textwidth}
-        \centerline{\textbf{b) ATE estimation error}}
-        \includegraphics[width=\linewidth]{_5_train_size_r_risk__ate__acic16_2cols.pdf}
-    \end{minipage}%
-    \caption{\textbf{a) For CATE, a train/test ratio of 0.9/0.1 appears a good
-            trade-off.} b) For ATE, there is a small signal pointing also to
-        0.9/0.1 (K=10).
-        for ATE. Experiences on 10 replications of all 78 instances of the ACIC 2016
-        data.}\label{fig:train_test_ratio}
-\end{figure}
-
-Figure \ref{fig:train_test_ratio} shows that a train/test ratio of
-0.9/0.1 (K=10) or 0.8/0.2 (K=5) appears best to estimate CATE and
-ATE.
-
-\subsubsection{Heterogeneity in practices for data split}\label{apd:results:k_fold_choices}
-
-Splitting the data is common when using machine learning for causal
-inference, but practices vary widely in terms of the fraction of data to
-allocate to train models, outcomes and nuisances, and to evaluate them.
-
-Before even model selection, data splitting is often required for
-estimation of the treatment effect, ATE or CATE, for instance to compute
-the nuisances required to optimize the outcome model (as the
-$R\text{-risk}$, Definition \ref{def:r_risk}).
-%
-The most frequent choice is use 80\% of the data to fit the models,
-and 20\% to evaluate them.
-%
-For instance, for CATE estimation, the R-learner has been introduced using K-folds with K = 5
-and K = 10: 80\% of the data (4 folds) to train the nuisances and the remaining
-fold to minimize the corresponding R-loss \cite{nie_quasioracle_2017}.
-Yet, it has been implemented with K=5 in causallib
-\cite{causalevaluations} or K=3 in econML \cite{econml}.
-Likewise, for ATE estimation, \cite{chernozhukov_double_2018}
-introduce doubly-robust machine learning,
-recommending K=5 based on an empirical comparison K=2. However,
-subsequent works use doubly robust ML with varying choices
-of K: \cite{loiseau_external_2022} use K=3, \cite{gao_assessment_2021} use
-K=2. In the econML implementation, K is set to 3 \cite{econml}.
-\cite{naimi2021challenges} evaluate various machine-learning approaches
---including R-learners-- using K=5 and 10, drawing inspiration from the
-TMLE literature which sets
-K=5 in the TMLE package \cite{tmle_package_2012}.
-
-Causal model selection has been much less discussed. The only study that
-we are aware of, \cite{schuler_comparison_2018}, use a different data
-split: a 2-folds train/test procedure,
-training the nuisances on the first half of the data, and using the
-second half to estimate the $R\text{-risk}$ and select the best treatment
-effect model.
-
 \subsection{Sensitivity of the results to different ratio of causal effect to baseline response}\label{apd:effect_ratio}
 
-Going beyond overlap, we study the effect of another important parameter of the
-data generation process: the ratio of causal effect to the baseline response.
-
-Here, we detail the experimental setup of the results presented in Figure \ref{fig:effect_ratio_influence}.
-
-For each dataset of our simulation (caussim), we measure this ratio as the absolute mean
-difference between the causal effect and the baseline response, measured
-empirically on each dataset instance $\Delta_{\mu} = \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) - \mu_{0}(x_i)}{\mu_{0}(x_i) + \mu_{1}(x_i) - \frac{1}{N} \sum_{j=1}^N \mu_0(x_j) + \mu_1(x_j)} \big |$.
-We vary a
-dedicated parameter ($\omega$ in Algorithm 2) in the simulation to force this
-ratio to cover a wide range of values. For 179 dataset instances, it ranges
-several order of magnitudes with the full distribution of values given in
-Table \ref{apd:effect_ratio_distribution}.
-
-\begin{table}[!ht]
-    \centering
-    \resizebox{\textwidth}{!}{ % Adjusts the table to fit the text width
-        \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
-            \hline
-                                   & \textbf{count} & \textbf{mean} & \textbf{std} & \textbf{min} & \textbf{1\%} & \textbf{10\%} & \textbf{25\%} & \textbf{50\%} & \textbf{75\%} & \textbf{90\%} & \textbf{99\%} & \textbf{max} \\ \hline
-            \textbf{effect\_ratio} & 179            & 16.95         & 24.92        & 0.043        & 0.054        & 0.32          & 1.34          & 9.12          & 22.55         & 39.97         & 101.12        & 206.53       \\ \hline
-        \end{tabular}
-    }
-    \caption{Distribution of the causal effect ratio $\Delta_\mu$ for the experiment on causal effect ratio, results in Figure \ref{fig:effect_ratio_influence}.}\label{apd:effect_ratio_distribution}
-\end{table}
-
-We also studied another measure, focusing on the ratio of the variance of the causal effect to the variance of the mean shared response ($\mu_0 + \mu_1$) of both population: $\frac{\hat{var}(\mu_{1}(x_i) - \mu_{0}(x_i))}{\hat{var}(\mu_{1}(x_i) + \mu_{0}(x_i))}$, covering a range from 0 to 11.08 with the full
-distribution described in Table \ref{table:effect_variance}. Figure
-\ref{fig:effect_variance_influence} shows that for high values of the variance,
-$R-\text{risk}$ is outperformed by the $\mu\text{-risk}_{IPW}$ and the
-$\tau\text{-risk}_{IPW}$. However, on average, the $R-\text{risk}$ is still the
-better risk.
-
-\begin{figure}[!tb]
-    \centering
-    \begin{minipage}{\linewidth}
-        \includegraphics[width=0.9\linewidth]{_6_effect_variance_by_bin_comparaison_kendall_by_Dataset.pdf}
-    \end{minipage}
-
-    \caption{\textbf{R-risk is robust to a wide range of effect variance}:
-    Kendall's $\tau$ agreement with $\tau$-risk. Strong, medium and Weak Causal
-    effect variance are the tertiles of the ratio of causal effect variance to
-    the mean outcome variance: Low $[0;0.04[$, Medium $[0.04; 1.04[$,
-    High $[1.04;11.08[$.}\label{fig:effect_variance_influence}
-\end{figure}
-
-\begin{table}[!ht]
-    \centering
-    \resizebox{\textwidth}{!}{ % Adjusts the table to fit the text width
-        \begin{tabular}{|l|r|r|r|r|r|r|r|r|r|r|r|r|}
-            \hline
-                                      & \textbf{count} & \textbf{mean} & \textbf{std} & \textbf{min} & \textbf{1\%} & \textbf{10\%} & \textbf{25\%} & \textbf{50\%} & \textbf{75\%} & \textbf{90\%} & \textbf{99\%} & \textbf{max} \\ \hline
-            \textbf{effect\_variance} & 179            & 1.013         & 1.781        & 1.680e-5     & 1.734e-5     & 4.875e-4      & 6.449e-3      & 0.789         & 1.160         & 1.807         & 9.641         & 11.079       \\
-            \hline
-        \end{tabular}
-    }
-    \caption{Statistical summary of the causal effect variance.}
-    \label{table:effect_variance}
-\end{table}
-
 \end{document}
